### 2017-12-23 cuthesis_phd.tex
using thesis template for now

\documentstyle[cuthesis]{report}

\author{Valerie Hayot-Sasson}
\title{Thesis Title}
\titleOfPhDAuthor{Ms.}          % or Ms., Mrs., Miss, etc. (only for PhD's)
\PhD                            % Masters by default
\degree{Computer Science}
% \dept{Computer Science}       % default is Comp.Sci.
\cosupervisor                   % if you also have a co-supervisor

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{abstract}
  Text of abstract.  
\end{abstract}

\begin{acknowledgments}
  Text of acknowledgments.
\end{acknowledgments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Body of Thesis goes here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\addcontentsline{toc}{chapter}{Bibliography}
%\bibliography{abbr,chalin,common,larch,tn}  %place your .bib files here
%\bibliographystyle{alpha}                   %the bibliography style to use

\end{document}

### 2017-12-25 cuthesis_phd.tex
very primitive outline
summary of [1]

@@
-\documentstyle[cuthesis]{report}
-
+\documentclass{report}
+\usepackage{cuthesis}
+\usepackage{hyperref}
+\usepackage{xcolor}
 \author{Valerie Hayot-Sasson}
@@
 
+\newcommand{\note}[1]{\textcolor{green}{\textit{note}: #1}}
+
 \begin{document}
-
 \begin{abstract}
@@
 
+\chapter{Parallelization frameworks}
+\section{Databases}
+\section{Pipelining engines}
+\chapter{MapReduce}
+Summary of ~\cite{mapred}
+
+Typically simple computations need to be performed on increasingly growing large datasets.
+Processing of such large datasets require parallelization.
+As a result, these computations are complicated by details of parallelization, fault tolerance, data distribution and load balancing.
+The MapReduce library addresses this issue by allowing the expression of simple computations while abstracting the other details.
+The map and reduce paradigm was selected for this purpose as it was already commonly used in functional programming languages, such as Lisp, and it was observed that many computations could be expressed as such.
+
+The MapReduce library operates by first dividing the data into manageable chunks (16 to 64MB).
+It then creates multiple copies of the program and distributes them across nodes.
+There exists a single master node; the remainder are worker nodes.
+The master node is responsible for delegating map and reduce tasks to the idle workers.
+There are two data structures found in the master.
+The first stores the state (\textit{idle}, \textit{in-progress} or \textit{completed}) of each \textit{map} and \textit{reduce} task, and the second stores the identity of the non-idle worker nodes.
+In contrast, the worker nodes are responsible for the execution of \textit{map} or \textit{reduce} tasks.
+A worker node assigned a \textit{map} task will parse out the key-value pairs from the input file and pass it to the \textit{map} function.
+The intermediate keys produced by the \textit{map} function are buffered in memory and periodically written to disk.
+The locations of these intermediate files are sent to the master who is responsible for forwarding these locations to the reducer.
+A worker node assigned a \textit{reduce} task uses remote procedure calls to obtain the intermediate local files created by the map tasks and sorts the keys such that identical keys are grouped together.
+Each unique key and its list of values are then submitted to the the \textit{reduce} task and the output of the \textit{reduce} task is stored in a final output file.
+Once all \textit{reduce} tasks have completed, the master resumes the program and the output is returned to the user.
+
+In order to ensure fault tolerance, the master regularly pings the worker nodes.
+Should a worker node not respond after a certain delay, the master labels the worker as failed.
+The \textit{map} tasks that were completed by the worker are all returned to their original state (\textit{idle}) as \textit{map} task data is stored locally.
+Both the completed and in-progress worker's \textit{map} tasks are subsequently reassigned to other workers.
+Only the failed worker's \textit{in-progress} \textit{reduce} tasks need to be reassigned.
+
+The master node undergoes periodic checkpoints, such that the master node can be reinitialized from its last checkpoint in case of failure. Master node failure is, however, unlikely.
+
+As network bandwidth is a scarce resource, the master attempts to schedule \textit{map} tasks closest to where the data is located.
+it will first attempt to schedule a map task on the host that contains the data; should that option not be available, the master the will attempts to schedule the \textit{map} task on a host nearest to the data (e.g.
+same network switch)
+
+Ideally, the number of \textit{map} and \textit{reduce} tasks should be much larger than the number of available worker nodes.
+Having it as such improves both dynamic load balancing and recovery time.
+The number of \textit{map} tasks used in practice typically corresponds to the size of the input chunks, whereas the number of \textit{reduce} tasks are a small multiple of the number of worker nodes used.
+
+When nearing completion of the MapReduce program, it is possible for  ``stragglers" (\textit{in-progress} tasks running on nodes with below average performance) to delay completion.
+As a measure to counteract this, the master node assigns the same task to idle workers.
+The task is then marked as completed as soon as either the primary or the backups complete the task.
+This strategy has been found to significantly improve performance when used with very large datasets.
+\note{44\% speedup with their sort program}.
+
+A few extensions have also been added to the library in order to improve user experience.
+An important extension is the introduction of a \textit{combiner}, which enables the commutative and associative reduce function to be applied to the same keys on the hosts where the \textit{map} task was executed.
+Like the \textit{map} task, the \textit{combiner} produces intermediate output which is sent to the \textit{reducer}.
+This alleviates the amount of key-value pairs that must be transferred over the network from the map host to the reduce host, and as a result, improves performance.
+Other extensions include the ability to create user-defined partitioning functions, a guaranteed increasing key sort order, the ability for the user to specify input types, an option to skip records that continuously result in execution failure, the ability to produce auxiliary output files in \textit{map} or \textit{reduce} task, live status updates stream to HTTP server,  the ability to run all code on a single local machine, and a \textit{Counter} class.
+
+\section{Hadoop}
+\cite{hadoop}
+
+\section{Apache Spark}
+\cite{spark}
+\chapter{MapReduce applications in neuroscience}
+\cite{thunder}
+
+\chapter{Pipelining engines used in neuroscience}
+\cite{nipype}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \addcontentsline{toc}{chapter}{Bibliography}
+\bibliography{bibliography}
+\bibliographystyle{ieeetr}
 %\bibliography{abbr,chalin,common,larch,tn}  %place your .bib files here

### 2017-12-25 cuthesis_phd.tex
fixed formatting

@@
 
-Typically simple computations need to be performed on increasingly growing large datasets.
-Processing of such large datasets require parallelization.
-As a result, these computations are complicated by details of parallelization, fault tolerance, data distribution and load balancing.
-The MapReduce library addresses this issue by allowing the expression of simple computations while abstracting the other details.
-The map and reduce paradigm was selected for this purpose as it was already commonly used in functional programming languages, such as Lisp, and it was observed that many computations could be expressed as such.
+Typically simple computations need to be performed on increasingly growing large datasets. Processing of such large datasets require
+parallelization. As a result, these computations are complicated by details of parallelization, fault tolerance, data distribution and load
+balancing. The MapReduce library addresses this issue by allowing the expression of simple computations while abstracting the other
+details. The map and reduce paradigm was selected for this purpose as it was already commonly used in functional programming languages,
+ such as Lisp, and it was observed that many computations could be expressed as such.
 
-The MapReduce library operates by first dividing the data into manageable chunks (16 to 64MB).
-It then creates multiple copies of the program and distributes them across nodes.
-There exists a single master node; the remainder are worker nodes.
-The master node is responsible for delegating map and reduce tasks to the idle workers.
-There are two data structures found in the master.
-The first stores the state (\textit{idle}, \textit{in-progress} or \textit{completed}) of each \textit{map} and \textit{reduce} task, and the second stores the identity of the non-idle worker nodes.
-In contrast, the worker nodes are responsible for the execution of \textit{map} or \textit{reduce} tasks.
-A worker node assigned a \textit{map} task will parse out the key-value pairs from the input file and pass it to the \textit{map} function.
-The intermediate keys produced by the \textit{map} function are buffered in memory and periodically written to disk.
-The locations of these intermediate files are sent to the master who is responsible for forwarding these locations to the reducer.
-A worker node assigned a \textit{reduce} task uses remote procedure calls to obtain the intermediate local files created by the map tasks and sorts the keys such that identical keys are grouped together.
-Each unique key and its list of values are then submitted to the the \textit{reduce} task and the output of the \textit{reduce} task is stored in a final output file.
-Once all \textit{reduce} tasks have completed, the master resumes the program and the output is returned to the user.
+The MapReduce library operates by first dividing the data into manageable chunks (16 to 64MB). It then creates multiple copies of the
+program and distributes them across nodes. There exists a single master node; the remainder are worker nodes. The master node is
+responsible for delegating map and reduce tasks to the idle workers. There are two data structures found in the master. The first stores the
+state (\textit{idle}, \textit{in-progress} or \textit{completed}) of each \textit{map} and \textit{reduce} task, and the second stores the identity of
+the non-idle worker nodes. In contrast, the worker nodes are responsible for the execution of \textit{map} or \textit{reduce} tasks. A worker
+node assigned a \textit{map} task will parse out the key-value pairs from the input file and pass it to the \textit{map} function. The
+intermediate keys produced by the \textit{map} function are buffered in memory and periodically written to disk. The locations of these
+intermediate files are sent to the master who is responsible for forwarding these locations to the reducer. A worker node assigned a 
+\textit{reduce} task uses remote procedure calls to obtain the intermediate local files created by the map tasks and sorts the keys such that 
+identical keys are grouped together. Each unique key and its list of values are then submitted to the the \textit{reduce} task and the output of 
+the \textit{reduce} task is stored in a final output file. Once all \textit{reduce} tasks have completed, the master resumes the program and the 
+output is returned to the user.
 
-In order to ensure fault tolerance, the master regularly pings the worker nodes.
-Should a worker node not respond after a certain delay, the master labels the worker as failed.
-The \textit{map} tasks that were completed by the worker are all returned to their original state (\textit{idle}) as \textit{map} task data is stored locally.
-Both the completed and in-progress worker's \textit{map} tasks are subsequently reassigned to other workers.
-Only the failed worker's \textit{in-progress} \textit{reduce} tasks need to be reassigned.
+In order to ensure fault tolerance, the master regularly pings the worker nodes. Should a worker node not respond after a certain delay, the 
+master labels the worker as failed. The \textit{map} tasks that were completed by the worker are all returned to their original state 
+(\textit{idle}) as \textit{map} task data is stored locally. Both the completed and in-progress worker's \textit{map} tasks are subsequently 
+reassigned to other workers. Only the failed worker's \textit{in-progress} \textit{reduce} tasks need to be reassigned.
 
-The master node undergoes periodic checkpoints, such that the master node can be reinitialized from its last checkpoint in case of failure. Master node failure is, however, unlikely.
+The master node undergoes periodic checkpoints, such that the master node can be reinitialized from its last checkpoint in case of failure. 
+Master node failure is, however, unlikely.
 
-As network bandwidth is a scarce resource, the master attempts to schedule \textit{map} tasks closest to where the data is located.
-it will first attempt to schedule a map task on the host that contains the data; should that option not be available, the master the will attempts to schedule the \textit{map} task on a host nearest to the data (e.g.
-same network switch)
+As network bandwidth is a scarce resource, the master attempts to schedule \textit{map} tasks closest to where the data is located. it will first 
+attempt to schedule a map task on the host that contains the data; should that option not be available, the master the will attempts to 
+schedule the \textit{map} task on a host nearest to the data (e.g. same network switch)
 
-Ideally, the number of \textit{map} and \textit{reduce} tasks should be much larger than the number of available worker nodes.
-Having it as such improves both dynamic load balancing and recovery time.
-The number of \textit{map} tasks used in practice typically corresponds to the size of the input chunks, whereas the number of \textit{reduce} tasks are a small multiple of the number of worker nodes used.
+Ideally, the number of \textit{map} and \textit{reduce} tasks should be much larger than the number of available worker nodes. Having it as 
+such improves both dynamic load balancing and recovery time. The number of \textit{map} tasks used in practice typically corresponds to the 
+size of the input chunks, whereas the number of \textit{reduce} tasks are a small multiple of the number of worker nodes used.
 
-When nearing completion of the MapReduce program, it is possible for  ``stragglers" (\textit{in-progress} tasks running on nodes with below average performance) to delay completion.
-As a measure to counteract this, the master node assigns the same task to idle workers.
-The task is then marked as completed as soon as either the primary or the backups complete the task.
-This strategy has been found to significantly improve performance when used with very large datasets.
-\note{44\% speedup with their sort program}.
+When nearing completion of the MapReduce program, it is possible for  ``stragglers" (\textit{in-progress} tasks running on nodes with below 
+average performance) to delay completion. As a measure to counteract this, the master node assigns the same task to idle workers. The 
+task is then marked as completed as soon as either the primary or the backups complete the task. This strategy has been found to 
+significantly improve performance when used with very large datasets. \note{44\% speedup with their sort program}.
 
-A few extensions have also been added to the library in order to improve user experience.
-An important extension is the introduction of a \textit{combiner}, which enables the commutative and associative reduce function to be applied to the same keys on the hosts where the \textit{map} task was executed.
-Like the \textit{map} task, the \textit{combiner} produces intermediate output which is sent to the \textit{reducer}.
-This alleviates the amount of key-value pairs that must be transferred over the network from the map host to the reduce host, and as a result, improves performance.
-Other extensions include the ability to create user-defined partitioning functions, a guaranteed increasing key sort order, the ability for the user to specify input types, an option to skip records that continuously result in execution failure, the ability to produce auxiliary output files in \textit{map} or \textit{reduce} task, live status updates stream to HTTP server,  the ability to run all code on a single local machine, and a \textit{Counter} class.
+A few extensions have also been added to the library in order to improve user experience. An important extension is the introduction of a 
+\textit{combiner}, which enables the commutative and associative reduce function to be applied to the same keys on the hosts where the 
+\textit{map} task was executed. Like the \textit{map} task, the \textit{combiner} produces intermediate output which is sent to the 
+\textit{reducer}. This alleviates the amount of key-value pairs that must be transferred over the network from the map host to the reduce host, 
+and as a result, improves performance. Other extensions include the ability to create user-defined partitioning functions, a guaranteed 
+increasing key sort order, the ability for the user to specify input types, an option to skip records that continuously result in execution failure, 
+the ability to produce auxiliary output files in \textit{map} or \textit{reduce} task, live status updates stream to HTTP server,  the ability to run 
+all code on a single local machine, and a \textit{Counter} class.
 

### 2018-01-05 cuthesis_phd.tex
Added some 'notes' on the HDFS paper.. to be fixed

--- 2017-12-25 cuthesis_phd.tex
@@
 \title{Thesis Title}
-\titleOfPhDAuthor{Ms.}          % or Ms., Mrs., Miss, etc. (only for PhD's)
+\titleOfPhDAuthor{}          % or Ms., Mrs., Miss, etc. (only for PhD's)
 \PhD                            % Masters by default
@@
 
+\chapter{Introduction}
+intro here
 \chapter{Parallelization frameworks}
@@
 Typically simple computations need to be performed on increasingly growing large datasets. 
-Processing of such large datasets require parallelization.
-As a result, these computations are complicated by details of parallelization, fault tolerance, data distribution and load balancing.
-The MapReduce library addresses this issue by allowing the expression of simple computations while abstracting the other details.
-The map and reduce paradigm was selected for this purpose as it was already commonly used in functional programming languages, such as Lisp, and it was observed that many computations could be expressed as such.
+Processing of such large datasets require parallelization. As a result, these computations are 
+complicated by details of parallelization, fault tolerance, data distribution and load balancing. The 
+MapReduce library addresses this issue by allowing the expression of simple computations while 
+abstracting the other details. The map and reduce paradigm was selected for this purpose as it 
+was already commonly used in functional programming languages, such as Lisp, and it was 
+observed that many computations could be expressed as such.
 
-The MapReduce library operates by first dividing the data into manageable chunks (16 to 64MB).
-It then creates multiple copies of the program and distributes them across nodes.
-There exists a single master node; the remainder are worker nodes.
-The master node is responsible for delegating map and reduce tasks to the idle workers.
-There are two data structures found in the master.
-The first stores the state (\textit{idle}, \textit{in-progress} or \textit{completed}) of each \textit{map} and \textit{reduce} task, and the second stores the identity of the non-idle worker nodes.
-In contrast, the worker nodes are responsible for the execution of \textit{map} or \textit{reduce} tasks.
-A worker node assigned a \textit{map} task will parse out the key-value pairs from the input file and pass it to the \textit{map} function.
-The intermediate keys produced by the \textit{map} function are buffered in memory and periodically written to disk.
-The locations of these intermediate files are sent to the master who is responsible for forwarding these locations to the reducer.
-A worker node assigned a \textit{reduce} task uses remote procedure calls to obtain the intermediate local files created by the map tasks and sorts the keys such that identical keys are grouped together.
-Each unique key and its list of values are then submitted to the the \textit{reduce} task and the output of the \textit{reduce} task is stored in a final output file.
-Once all \textit{reduce} tasks have completed, the master resumes the program and the output is returned to the user.
+The MapReduce library operates by first dividing the data into manageable chunks (16 to 64MB). It 
+then creates multiple copies of the program and distributes them across nodes. There exists a 
+single master node; the remainder are worker nodes. The master node is responsible for 
+delegating map and reduce tasks to the idle workers. There are two data structures found in the 
+master. The first stores the state (\textit{idle}, \textit{in-progress} or \textit{completed}) of each 
+\textit{map} and \textit{reduce} task, and the second stores the identity of the non-idle worker 
+nodes. In contrast, the worker nodes are responsible for the execution of \textit{map} or 
+\textit{reduce} tasks. A worker node assigned a \textit{map} task will parse out the key-value pairs 
+from the input file and pass it to the \textit{map} function. The intermediate keys produced by the 
+\textit{map} function are buffered in memory and periodically written to disk. The locations of these
+intermediate files are sent to the master who is responsible for forwarding these locations to the 
+reducer. A worker node assigned a \textit{reduce} task uses remote procedure calls to obtain the 
+intermediate local files created by the map tasks and sorts the keys such that identical keys are 
+grouped together. Each unique key and its list of values are then submitted to the the 
+\textit{reduce} task and the output of the \textit{reduce} task is stored in a final output file. Once all 
+\textit{reduce} tasks have completed, the master resumes the program and the 
+output is returned to the user.
 
-In order to ensure fault tolerance, the master regularly pings the worker nodes.
-Should a worker node not respond after a certain delay, the master labels the worker as failed.
-The \textit{map} tasks that were completed by the worker are all returned to their original state (\textit{idle}) as \textit{map} task data is stored locally.
-Both the completed and in-progress worker's \textit{map} tasks are subsequently reassigned to other workers.
-Only the failed worker's \textit{in-progress} \textit{reduce} tasks need to be reassigned.
+In order to ensure fault tolerance, the master regularly pings the worker nodes. Should a worker 
+node not respond after a certain delay, the master labels the worker as failed. The \textit{map} 
+tasks that were completed by the worker are all returned to their original state (\textit{idle}) as 
+\textit{map} task data is stored locally. Both the completed and in-progress worker's 
+\textit{map} tasks are subsequently reassigned to other workers. Only the failed worker's \textit{in-
+progress} \textit{reduce} tasks need to be reassigned.
 
-The master node undergoes periodic checkpoints, such that the master node can be reinitialized from its last checkpoint in case of failure. Master node failure is, however, unlikely.
+The master node undergoes periodic checkpoints, such that the master node can be reinitialized 
+from its last checkpoint in case of failure. Master node failure is, however, unlikely.
 
-As network bandwidth is a scarce resource, the master attempts to schedule \textit{map} tasks closest to where the data is located.
-it will first attempt to schedule a map task on the host that contains the data; should that option not be available, the master the will attempts to schedule the \textit{map} task on a host nearest to the data (e.g.
-same network switch)
+As network bandwidth is a scarce resource, the master attempts to schedule \textit{map} tasks 
+closest to where the data is located. it will first attempt to schedule a map task on the host that 
+contains the data; should that option not be available, the master the will attempts to 
+schedule the \textit{map} task on a host nearest to the data (e.g. same network switch)
 
-Ideally, the number of \textit{map} and \textit{reduce} tasks should be much larger than the number of available worker nodes.
-Having it as such improves both dynamic load balancing and recovery time.
-The number of \textit{map} tasks used in practice typically corresponds to the size of the input chunks, whereas the number of \textit{reduce} tasks are a small multiple of the number of worker nodes used.
+Ideally, the number of \textit{map} and \textit{reduce} tasks should be much larger than the number 
+of available worker nodes. Having it as such improves both dynamic load balancing and recovery 
+time. The number of \textit{map} tasks used in practice typically corresponds to the 
+size of the input chunks, whereas the number of \textit{reduce} tasks are a small multiple of the 
+number of worker nodes used.
 
-When nearing completion of the MapReduce program, it is possible for  ``stragglers" (\textit{in-progress} tasks running on nodes with below average performance) to delay completion.
-As a measure to counteract this, the master node assigns the same task to idle workers.
-The task is then marked as completed as soon as either the primary or the backups complete the task.
-This strategy has been found to significantly improve performance when used with very large datasets.
-\note{44\% speedup with their sort program}.
+When nearing completion of the MapReduce program, it is possible for  ``stragglers" (\textit{in-
+progress} tasks running on nodes with below average performance) to delay completion. As a 
+measure to counteract this, the master node assigns the same task to idle workers. The task is 
+then marked as completed as soon as either the primary or the backups complete the task. 
+This strategy has been found to significantly improve performance when used with very large 
+datasets. \note{44\% speedup with their sort program}.
 
-A few extensions have also been added to the library in order to improve user experience.
-An important extension is the introduction of a \textit{combiner}, which enables the commutative and associative reduce function to be applied to the same keys on the hosts where the \textit{map} task was executed.
-Like the \textit{map} task, the \textit{combiner} produces intermediate output which is sent to the \textit{reducer}.
-This alleviates the amount of key-value pairs that must be transferred over the network from the map host to the reduce host, and as a result, improves performance.
-Other extensions include the ability to create user-defined partitioning functions, a guaranteed increasing key sort order, the ability for the user to specify input types, an option to skip records that continuously result in execution failure, the ability to produce auxiliary output files in \textit{map} or \textit{reduce} task, live status updates stream to HTTP server,  the ability to run all code on a single local machine, and a \textit{Counter} class.
+A few extensions have also been added to the library in order to improve user experience. An 
+important extension is the introduction of a \textit{combiner}, which enables the commutative and 
+associative reduce function to be applied to the same keys on the hosts where the \textit{map} task 
+was executed. Like the \textit{map} task, the \textit{combiner} produces intermediate output which 
+is sent to the \textit{reducer}. This alleviates the amount of key-value pairs that must be transferred 
+over the network from the map host to the reduce host, and as a result, improves performance. 
+Other extensions include the ability to create user-defined partitioning functions, a guaranteed 
+increasing key sort order, the ability for the user to specify input types, an option to skip records 
+that continuously result in execution failure, the ability to produce auxiliary output files in 
+\textit{map} or \textit{reduce} task, live status updates stream to HTTP server,  the ability to run all 
+code on a single local machine, and a \textit{Counter} class.
 
-\section{Hadoop}
-\cite{hadoop}
+\note{Brief hadoop section??}
+
+\section{Hadoop Distributed File System}
+Summary of \cite{hadoop}
+
+The Hadoop Distributed File System (HDFS) is the filesystem component of Hadoop. The 
+metadata in HDFS is stored in on a dedicated server known as the NameNode, whereas 
+application data is distributed across many servers referred to as DataNode. To ensure fault 
+tolerance, HDFS replicates the data across multiple DataNodes (default replication factor: 
+3).
+Not only does this ensure that the data is not lost in the event of node failure, but it also increases data transfer bandwidth, as the data can be accessed from multiple nodes, and thus, there are more opportunities for computations to be performed nearest to where the data is located.
+
+The NameNode contains the namespace tree, a hierarchy of directories and their files, as well as the mapping of the split file blocks to DataNodes.
+The entire namespace is stored in memory.
+Information on the directories and files, such as modification and access times, namespace and disk quotes, are stored within \textit{inodes} on the NameNode.
+The name system's metadata, \textit{inode} and file block mapping, is known as the \textit{image}.
+Persistent record of the image that is stored on the local file system are known as \textit{checkpoint}.
+Locations of block replicas are not stored in the checkpoint as they may change over time.
+The \textit{journal} is a log of the modifications made to the image.
+Both the checkpoint and the journal may be copied across servers for increased durability.
+The journal is played back during the NameNode restart in order to restore the cluster.
+
+When a client requests to read data, it must first contact the NameNode.
+The NameNode provides it with the replica locations, and the client reads from the DataNode located closest to it.
+When the client requests to write a file, it contacts the NameNode which selects the DataNodes that will host the replicas.
+The client subsequently writes the data directly to the DataNodes in a pipeline fashion (The data gets propagated through each DataNode by being transferred from the nearest DataNode to it).
+
+There is only one NameNode assigned to each cluster, however, each cluster can have multiple clients and execute multiple tasks concurrently.
+
+A block replica stored on a DataNode consists of two files stored on the local file system: the metadata and the data.
+
+The DataNode connects to the NameNode during startup and performs handshake to for DataNode namespace ID and software version verification.
+Should either namespace ID or software version not match those of the NameNode, the DataNode shuts down.
+
+To ensure integrity of the system, a \textit{namespace ID} is assigned to all nodes during formatting of the namespace.
+DataNodes with different namespace IDs to the NameNode will not be permitted to join the cluster.
+Should the DataNode have not been assigned a namespace ID, it will be permitted to join the cluster.
+The cluster's namespace ID will then be assigned to that DataNode.
+
+Following the handshake, the DataNode register with the NameNode using their storage ID.
+The storage ID is a unique ID generated after initial registration to the NameNode and that is persistently stored on the DataNode.
+It permits identification of the DataNode in the event of an IP address or port change.
+
+inodes, stored on namenode, represent file and directories. inodes record attributes such as permissions, modifications and access times, namespace and disk quotas.
+
+namenode keeps track of namespace tree and mapping of file blocks to datanodes.
+
+The \textit{image} stores the \textit{inode} data and the list of blocks belonging to each file.
+
+The \textit{checkpoint} is a persistent record of the \textit{image} stored in the local filesystem.
+
+The \textit{journal} is the modification log stored on the local filesystem.
 
@@
 
+\section{Schedulers}
 \chapter{Pipelining engines used in neuroscience}

### 2018-01-05 paper.tex
created an outline, removed phd style

--- 2018-01-05 cuthesis_phd.tex
@@
 \documentclass{report}
-\usepackage{cuthesis}
+
 \usepackage{hyperref}
 \usepackage{xcolor}
-\author{Valerie Hayot-Sasson}
-\title{Thesis Title}
-\titleOfPhDAuthor{}          % or Ms., Mrs., Miss, etc. (only for PhD's)
-\PhD                            % Masters by default
-\degree{Computer Science}
-% \dept{Computer Science}       % default is Comp.Sci.
-\cosupervisor                   % if you also have a co-supervisor
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{document}
+\title{Pipeline systems and infrastructure for the efficient and open processing of Big neuroimaging Data}
+\author{Valerie Hayot-Sasson}
+\maketitle
 \begin{abstract}
@@
 \end{abstract}
-
-\begin{acknowledgments}
-  Text of acknowledgments.
-\end{acknowledgments}
-
+\tableofcontents
 \chapter{Introduction}
 intro here
-\chapter{Parallelization frameworks}
-\section{Databases}
-\section{Pipelining engines}
-\chapter{MapReduce}
+\chapter{BigData in neuroscience}
+    \note{Likely will not be split in two separate sections}
+    \section{large images}
+        \begin{itemize}  
+            \item BigBrain
+            \item Micro CT 
+            \item EM 
+        \end{itemize}
+    \section{large datasets}
+    	\note{off the top of my head, but need to double check}
+        \begin{itemize}
+            \item HCP
+            \item ADNI
+            \item OpenFMRI
+        \end{itemize}
+    An effective engine to process such data requires the following characteristics
+        \begin{itemize}
+            \item in-memory computing
+            \item data locality
+            \item lazy evaluation
+        \end{itemize}
+\chapter{Workflow engines for neuroimaging data}
+    \begin{itemize}
+        \item Nipype
+        \item PSOAM
+        \item SPM
+        \item these engines do not currently support processing big data
+      	\begin{itemize}
+        		\item PSOAM - can't do data locality
+		\item none do in-memory processing?
+	\end{itemize}
+    \end{itemize}
+\chapter{CBRAIN}
+\chapter{Domain nonspecific BigData frameworks}
+	\section{mapreduce}
+		\begin{itemize}
+			\item many computations can be described through map and reduce
+			\item fault tolerant
+		\end{itemize}
+	\section{spark}
+		\begin{itemize}
+			\item based on map-reduce
+			\item in-memory processing
+			\item lazy evaluation
+			\item general framework
+		\end{itemize}
+	\section{Hadoop Distributed file system}
+		\note{maybe should go in between mapreduce and spark}
+		\begin{itemize}
+			\item enables data-locality in Spark/ decouples data locality logic from mapreduce\
+			\item data replication
+			\item fault-tolerance
+		\end{itemize}
+\chapter{Current status of Big Data frameworks in neuroscience}
+	\section{Ariel's paper}
+	\section{Thunder}
+\chapter{Use of containers in Workflow engines}
+	\section{use of containers in nipype, psoam, spm}
+	\section{using containers w/ Spark}
+
+\chapter{Summaries}
+
+This section contains summaries and/or notes on the 5 main papers to aid with writing and structuring of ideas.
+\section{MapReduce}
 Summary of ~\cite{mapred}
@@
 \cite{spark}
-\chapter{MapReduce applications in neuroscience}
+\section{Thunder}
 \cite{thunder}
-
-\section{Schedulers}
-\chapter{Pipelining engines used in neuroscience}
+\section{NiPype}
 \cite{nipype}

### 2018-01-07 paper.tex
modifcations to outline and summaries removed

--- 2017-12-25 cuthesis_phd.tex
@@
 \documentclass{report}
-\usepackage{cuthesis}
+
 \usepackage{hyperref}
 \usepackage{xcolor}
-\author{Valerie Hayot-Sasson}
-\title{Thesis Title}
-\titleOfPhDAuthor{Ms.}          % or Ms., Mrs., Miss, etc. (only for PhD's)
-\PhD                            % Masters by default
-\degree{Computer Science}
-% \dept{Computer Science}       % default is Comp.Sci.
-\cosupervisor                   % if you also have a co-supervisor
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \end{abstract}
+\tableofcontents
+\chapter{Introduction}
+intro here
+\chapter{BigData in neuroimaging}
+    \note{Likely will not be split in two separate sections}
+    \section{large images}
+        \begin{itemize}  
+            \item BigBrain
+            \item Micro CT 
+            \item EM 
+        \end{itemize}
+    \section{large datasets}
+    	\note{off the top of my head, but need to double check}
+        \begin{itemize}
+            \item HCP
+            \item ADNI
+            \item OpenFMRI
+        \end{itemize}
+    An effective engine to process such data requires the following characteristics
+        \begin{itemize}
+            \item in-memory computing
+            \item data locality
+            \item lazy evaluation
+        \end{itemize}
+\chapter{Workflow engines for neuroimaging data}
+	\section{Definition of workflows/pipelines}
+	\section{Nipype}
+		\subsection{summary}
+		\subsection{limitations of nipype}
+	\section{PSOM}
+		\subsection{summary}
+		\subsection{limitations of PSOM}
+	\section{SPM}
+		\subsection{summary}
+		\subsection{limitations of SPM}
+	\section{Use of containers in Workflow engines}
+		\subsection{use of containers in nipype, psoam, spm}
+\chapter{Open Science in neuroimaging}
+	\section{define open-science (eg. characteristics to look for that can then be reapplied to spark)}
+	\section{CBRAIN}
+		\subsection{Summary}
+		\subsection{Limitations}
+	\section{OpenNEURO}
+		\subsection{Summary}
+		\subsection{Limitations}
+	\section{NeuroData}
+		\subsection{Summary}
+		\subsection{Limitations}
+\chapter{Domain nonspecific BigData frameworks}
+	\section{mapreduce}
+		\begin{itemize}
+			\item many computations can be described through map and reduce
+			\item fault tolerant
+		\end{itemize}
+	\section{spark}
+		\begin{itemize}
+			\item based on map-reduce
+			\item in-memory processing
+			\item lazy evaluation
+			\item general framework
+		\end{itemize}
+	\section{Use of containers with Big Data frameworks}
+	\section{Hadoop Distributed file system}
+		\note{maybe should go in between mapreduce and spark}
+		\begin{itemize}
+			\item enables data-locality in Spark/ decouples data locality logic from mapreduce\
+			\item data replication
+			\item fault-tolerance
+		\end{itemize}
+\chapter{High-Performance Computing and Grid Computing with Big Data}
+	\section{Definition}
+	\section{Why is use of such systems important}
+	\section{Scheduling}
+		\subsection{HPC schedulers: SLURM/QSUB}
+		\subsection{Big Data schedulers: YARN/MESOS}
+		\subsection{Multi-level scheduling w/ BigData schedulers}
+\chapter{Current status of Big Data frameworks in neuroscience}
+	\section{Ariel's paper}
+	\section{Thunder}
+\chapter{Conclusion}
 
-\begin{acknowledgments}
-  Text of acknowledgments.
-\end{acknowledgments}
-
-\chapter{Parallelization frameworks}
-\section{Databases}
-\section{Pipelining engines}
-\chapter{MapReduce}
-Summary of ~\cite{mapred}
-
-Typically simple computations need to be performed on increasingly growing large datasets. Processing of such large datasets require
-parallelization. As a result, these computations are complicated by details of parallelization, fault tolerance, data distribution and load
-balancing. The MapReduce library addresses this issue by allowing the expression of simple computations while abstracting the other
-details. The map and reduce paradigm was selected for this purpose as it was already commonly used in functional programming languages,
- such as Lisp, and it was observed that many computations could be expressed as such.
-
-The MapReduce library operates by first dividing the data into manageable chunks (16 to 64MB). It then creates multiple copies of the
-program and distributes them across nodes. There exists a single master node; the remainder are worker nodes. The master node is
-responsible for delegating map and reduce tasks to the idle workers. There are two data structures found in the master. The first stores the
-state (\textit{idle}, \textit{in-progress} or \textit{completed}) of each \textit{map} and \textit{reduce} task, and the second stores the identity of
-the non-idle worker nodes. In contrast, the worker nodes are responsible for the execution of \textit{map} or \textit{reduce} tasks. A worker
-node assigned a \textit{map} task will parse out the key-value pairs from the input file and pass it to the \textit{map} function. The
-intermediate keys produced by the \textit{map} function are buffered in memory and periodically written to disk. The locations of these
-intermediate files are sent to the master who is responsible for forwarding these locations to the reducer. A worker node assigned a 
-\textit{reduce} task uses remote procedure calls to obtain the intermediate local files created by the map tasks and sorts the keys such that 
-identical keys are grouped together. Each unique key and its list of values are then submitted to the the \textit{reduce} task and the output of 
-the \textit{reduce} task is stored in a final output file. Once all \textit{reduce} tasks have completed, the master resumes the program and the 
-output is returned to the user.
-
-In order to ensure fault tolerance, the master regularly pings the worker nodes. Should a worker node not respond after a certain delay, the 
-master labels the worker as failed. The \textit{map} tasks that were completed by the worker are all returned to their original state 
-(\textit{idle}) as \textit{map} task data is stored locally. Both the completed and in-progress worker's \textit{map} tasks are subsequently 
-reassigned to other workers. Only the failed worker's \textit{in-progress} \textit{reduce} tasks need to be reassigned.
-
-The master node undergoes periodic checkpoints, such that the master node can be reinitialized from its last checkpoint in case of failure. 
-Master node failure is, however, unlikely.
-
-As network bandwidth is a scarce resource, the master attempts to schedule \textit{map} tasks closest to where the data is located. it will first 
-attempt to schedule a map task on the host that contains the data; should that option not be available, the master the will attempts to 
-schedule the \textit{map} task on a host nearest to the data (e.g. same network switch)
-
-Ideally, the number of \textit{map} and \textit{reduce} tasks should be much larger than the number of available worker nodes. Having it as 
-such improves both dynamic load balancing and recovery time. The number of \textit{map} tasks used in practice typically corresponds to the 
-size of the input chunks, whereas the number of \textit{reduce} tasks are a small multiple of the number of worker nodes used.
-
-When nearing completion of the MapReduce program, it is possible for  ``stragglers" (\textit{in-progress} tasks running on nodes with below 
-average performance) to delay completion. As a measure to counteract this, the master node assigns the same task to idle workers. The 
-task is then marked as completed as soon as either the primary or the backups complete the task. This strategy has been found to 
-significantly improve performance when used with very large datasets. \note{44\% speedup with their sort program}.
-
-A few extensions have also been added to the library in order to improve user experience. An important extension is the introduction of a 
-\textit{combiner}, which enables the commutative and associative reduce function to be applied to the same keys on the hosts where the 
-\textit{map} task was executed. Like the \textit{map} task, the \textit{combiner} produces intermediate output which is sent to the 
-\textit{reducer}. This alleviates the amount of key-value pairs that must be transferred over the network from the map host to the reduce host, 
-and as a result, improves performance. Other extensions include the ability to create user-defined partitioning functions, a guaranteed 
-increasing key sort order, the ability for the user to specify input types, an option to skip records that continuously result in execution failure, 
-the ability to produce auxiliary output files in \textit{map} or \textit{reduce} task, live status updates stream to HTTP server,  the ability to run 
-all code on a single local machine, and a \textit{Counter} class.
-
-\section{Hadoop}
-\cite{hadoop}
-
-\section{Apache Spark}
-\cite{spark}
-\chapter{MapReduce applications in neuroscience}
-\cite{thunder}
-
-\chapter{Pipelining engines used in neuroscience}
-\cite{nipype}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-01-12 paper.tex
hpc and spark intro

@@
 intro here
-\chapter{BigData in neuroimaging}
+\chapter{BigData and Open science in neuroimaging}
     \note{Likely will not be split in two separate sections}
@@
         \end{itemize}
+        \section{open-science platforms for neuroimaging}
+        		\subsection{Introduction}
+		\subsection{CBRAIN}
+		\subsection{OpenNEURO}
+		\subsection{NeuroData}
 \chapter{Workflow engines for neuroimaging data}
@@
 		\subsection{use of containers in nipype, psoam, spm}
-\chapter{Open Science in neuroimaging}
-	\section{define open-science (eg. characteristics to look for that can then be reapplied to spark)}
-	\section{CBRAIN}
-		\subsection{Summary}
-		\subsection{Limitations}
-	\section{OpenNEURO}
-		\subsection{Summary}
-		\subsection{Limitations}
-	\section{NeuroData}
-		\subsection{Summary}
-		\subsection{Limitations}
+
 \chapter{Domain nonspecific BigData frameworks}
@@
 		\end{itemize}
-\chapter{High-Performance Computing and Grid Computing with Big Data}
-	\section{Definition}
+\chapter{Optimizing Big Data frameworks for High-Performance Computing}
+	\section{Introduction}
+	High-Performance Computing (HPC) clusters (Compute grids) are extensively used by the scientific community to process complex, computation intensive problems.
+Due to their importance in the advancement of scientific research, it is necessary that Big Data frameworks used to process scientific data are compatible with them.
+
+	Big Data frameworks were not designed with high-performance computing in mind.
+Big Data filesystems such as were intended to run in data centers with homogenous storage (ex.
+HDDs, SDDs.
+etc,.).
+In such infrastructures, data movement is expensive.
+To counteract this, strategies were employed to avoid moving data around (data locality).
+In contrast, HPC clusters use heterogenous storage (ex.
+a combination of HDD, SSD, Ram Disk and Parallel file systems) with compute nodes connected by high-performance interconnects.
+HPC nodes may therefore not all be equipped with the same type of storage devices, some of which may be faster than others.
+Unlike data centers, HPC compute nodes do not have local storage.
+Moreover, the high-performance interconnects ensure low latency and high throughput of the system.
+In some instances, it may be even more performant to move data around rather than ensuring data locality.
+When a single-node HPC cluster using the Luster parallel file system was compared to a workstation with local SSD, it was found that Spark performed about 4x slower on the HPC cluster \cite{Chaimov:2016}.
+As HDFS is not conscientious of the underlying infrastructure and expects a data center-type infrastructure, it cannot achieve its full performance potential.
+
+	Prior to examining how Big Data frameworks could be improved on HPC clusters, it is important to have a baseline of how it performs natively on an HPC cluster.
+Particularly, how it behave when processing scientific workflows.
+In a recent study done by \cite{valduriez}, they examined the scalability of Spark on an HPC cluster using a black-box scientific workflow.
+Five tests were performed in their study: 1) Measuring execution time when task and resources available increase proportionally, 2) Measuring the execution time of very short and short tasks, 3) Measuring the execution time of short (5s) to long tasks (120s), 4) Varying the number of tasks with fixed task durations, 5) Mixed task durations and varied number of tasks.
+Their results show Spark behaves as expected when tasks and resources are increased exponentially, and that Spark scales better than expected with tasks of longer durations.
+This is due to the fact that the scheduler is overloaded in the case of many short tasks in addition to their being a significant amount of I/O occurring between each task.
+Although I/O is problematic in both data center and HPC infrastructures, HPC is optimized for data transfer over the network rather than disk I/O, and therefore, such frequent I/O negatively impacts the system.
+However, many scientific workflows are composed of longer tasks, and therefore, Spark scales very well with scientific workflows in an HPC environment.
+
+	This chapter will examine the various efforts by the community to improve Big Data framework performance on HPC clusters.
+
+	\section{In-memory file systems for Spark}
+
+	Using Spark against a typical HPC parallel file system such a Luster has been found to impede scalability.
+This was found to be due the file system metadata latency, which limits Spark scalability to \textit{O}(10\textsuperscript{2}) cores \cite{Chaimov:2016}.
+
+	%%%Compute Grids, interconnected networks of computing resources, are heavily used in scientific research for the processing of data.
+Due to their use and importance to the scientific community, it is necessary that big data frameworks used for processing neuroimaging data are compatible with them.
+Compute grids typically utilize heterogenous storage, such as a mixture of SSD, HDD, RAM Disk and parallel file systems (e.g.
+Luster).
+However, big data frameworks, such as Spark are designed to work efficiently with homogenous storage.
+Such differences result in reduced performance and inefficient storage use when big data frameworks are used on HPC clusters.
+This chapter will review the difference techniques applied to big data frameworks as well as HPC and grids to improve collocation of both services.%%%
+
+	%%%\subsection{Anatomy of Grid storage}
+	%%%Typical grids consist of a cluster of data nodes possibly connected to local storage (e.g.
+RAM Disk, HDD, SDD), but is also connected through a network to another cluster representing a parallel file system such as Lustre.
+RAM Disk is the fastest storage as all data is located in memory, whereas SSD provides slightly slower, persistent storage, but is a good alternative for big data.
+HDDs are the least performant form of local storage.%%%
+	%%%Lustre is a stateful, object-based parallel file system.
+It consists of three components: 1) the meta data server (MDS), 2) the Object Storage Server (OSS) and 3) the Luster Network (LNET), which enables clients to communicate with each other.%%%
+
+	In an attempt to improve performance of HDFS on HPC clusters, Triple-H was created.
 	\section{Why is use of such systems important}

### 2018-02-02 paper.tex
Added some notes

@@
             \item OpenFMRI
+            \item BIDS/BIDS apps
         \end{itemize}
@@
         \end{itemize}
-        \section{open-science platforms for neuroimaging}
+        \section{open-science /data sharing platforms for neuroimaging}
+        		is open science just data sharing, computing. 
         		\subsection{Introduction}
@@
 	\section{Use of containers in Workflow engines}
-		\subsection{use of containers in nipype, psoam, spm}
-
+		\subsection{use of containers in nipype, psom, spm}
+\chapter{Scientific workflows in Spark}
+	\section{Introduction}
+	\section{ariel's paper and thunders astronomy - }
+	\section{Spark on hpc's or }
+	\note{large dataset for valduriez}
+	Prior to examining how Big Data frameworks could be improved on HPC clusters, it is important to have a baseline of how it performs natively on an HPC cluster.
+Particularly, how it behave when processing scientific workflows.
+In a recent study done by \cite{valduriez}, they examined the scalability of Spark on an HPC cluster using a black-box scientific workflow.
+Five tests were performed in their study: 1) Measuring execution time when task and resources available increase proportionally, 2) Measuring the execution time of very short and short tasks, 3) Measuring the execution time of short (5s) to long tasks (120s), 4) Varying the number of tasks with fixed task durations, 5) Mixed task durations and varied number of tasks.
+Their results show Spark behaves as expected when tasks and resources are increased exponentially, and that Spark scales better than expected with tasks of longer durations.
+This is due to the fact that the scheduler is overloaded in the case of many short tasks in addition to their being a significant amount of I/O occurring between each task.
+Although I/O is problematic in both data center and HPC infrastructures, HPC is optimized for data transfer over the network rather than disk I/O, and therefore, such frequent I/O negatively impacts the system.
+However, many scientific workflows are composed of longer tasks, and therefore, Spark scales very well with scientific workflows in an HPC environment.
 \chapter{Domain nonspecific BigData frameworks}
@@
 \chapter{Optimizing Big Data frameworks for High-Performance Computing}
-	\section{Introduction}
-	High-Performance Computing (HPC) clusters (Compute grids) are extensively used by the scientific community to process complex, computation intensive problems.
+
+	High-Performance Computing (HPC) clusters are extensively used by the scientific community to process complex, computation intensive problems.
 Due to their importance in the advancement of scientific research, it is necessary that Big Data frameworks used to process scientific data are compatible with them.
 
-	Big Data frameworks were not designed with high-performance computing in mind.
+	Big Data frameworks were optimized for a commodity cluster.
+Such clusters are made up of nodes with homogenous storage connected by lower-end network cables (e.g.
+Ethernet).
+As a result, moving data from local storage to memory was much less costly in these infrastructures than moving data across the network.
+Thus, Big Data frameworks implemented data locality in an attempt to reduce costly network (horizontal) transfers and favour more performant local storage to memory (vertical) transfers.
+HPC clusters, however, are not necessarily commodity clusters.
+Large HPC clusters are commonly made up of supercomputing nodes connected by a high bandwidth, low-latency network (e.g.
+InfiniBand).
+Nodes may contain their own local storage (e.g.
+HDD, SSD, tmpfs) and are all connected to a distributed file system (e.g GPFS, Lustre).
+Due to their employment of a high bandwidth low latency network, network transfers on such an HPC cluster may not be as costly as that of a typical data center, and it may in fact be more costly to do vertical transfers.
+Furthermore, HPC systems may enforce a batch submission system.
+This required users to specify number of nodes, cores and total processing time; requirements which are unknown to MapReduce users.
+
+	This chapter will examine the various efforts by the community to improve Big Data framework performance on HPC clusters.
+
+	\section{File system improvements}
+
+	In paper \note{Scaling Spark on HPC Systems} it was found that Spark on a Lustre backend performed significantly (4x) slower than that of a workstation using local fast SSDs for storage.
+This is due to the fact that file system metadata latency (more specifically, file open) is the determinant factor for performance on HPC, whereas network dominates performance in a typical data centre configuration.
+It was found by the authors that using either a local in-memory filesystem or a filesystem mounted to a single Lustre file improved performance up to the point where it matched that of a workstation.
+In addition, use of a cache was found to reduce file system metadata operations, and thus improve overall performance.
+	The issues related to using an in-memory filesystem is that it is limited by the amount of physical memory available, and a job will fail if physical memory runs out.
+Moreover, there is not persistence or resilience of data as it is not saved to non-volatile storage (disk).
+Spark fails with medium to large applications due to "lax garbage collection in block and shuffle managers".
+To resolve these issues, the authors first attempted to make all executors share a descriptor pool.
+This allowed files that were already opened by other executors to be accessed by the executor through the descriptor rather than having to reopen it again.
+However, the node OS image is subject to number of Inode constraints as well Lustre limits on number of open files for a given job.
+Due to Inode constraints, a livelock may occur when descriptor pool is at capacity and an executor is attempting to open a file.
+
+	.the authors created mounted a local filesystem backed by a Lustre file \note{requires admin privilege}.
+	\note{for each partition, a shuffle file is created and written to as many times as there are partitions.
+An index file is also created which contains shuffle file locations and offsets.
+Metadata access is therefore O(partitions2)}
+
+	%%%In other words, such frameworks would try to limit data movement, as network cables are expected to have low bandwidth and high latency.
+In addition, each node in a commodity cluster is expected to have access to its own local storage homogenous to that of the rest of the cluster, thus replication is necessary to ensure data availability.
+This is, however, not necessarily the case for computing grids.
+Computing grids may have access to Infinibands, which provide high bandwidth and low latency.
+Moreover, grids may use heterogenous storage in addition to a parallel file system such as Lustre.
+As a result, it may, in some cases, be faster to transfer data over the network than to access the nearest copy of the data.
+As well, data replication may burden a system with centralized storage as multiple copies of the data would be created, taking a significant amount of space.%%%
+
+	%%%Big Data frameworks were not designed with high-performance computing in mind.
 Big Data filesystems such as were intended to run in data centers with homogenous storage (ex.
@@
 As HDFS is not conscientious of the underlying infrastructure and expects a data center-type infrastructure, it cannot achieve its full performance potential.
+%%%
 
-	Prior to examining how Big Data frameworks could be improved on HPC clusters, it is important to have a baseline of how it performs natively on an HPC cluster.
-Particularly, how it behave when processing scientific workflows.
-In a recent study done by \cite{valduriez}, they examined the scalability of Spark on an HPC cluster using a black-box scientific workflow.
-Five tests were performed in their study: 1) Measuring execution time when task and resources available increase proportionally, 2) Measuring the execution time of very short and short tasks, 3) Measuring the execution time of short (5s) to long tasks (120s), 4) Varying the number of tasks with fixed task durations, 5) Mixed task durations and varied number of tasks.
-Their results show Spark behaves as expected when tasks and resources are increased exponentially, and that Spark scales better than expected with tasks of longer durations.
-This is due to the fact that the scheduler is overloaded in the case of many short tasks in addition to their being a significant amount of I/O occurring between each task.
-Although I/O is problematic in both data center and HPC infrastructures, HPC is optimized for data transfer over the network rather than disk I/O, and therefore, such frequent I/O negatively impacts the system.
-However, many scientific workflows are composed of longer tasks, and therefore, Spark scales very well with scientific workflows in an HPC environment.
-
-	This chapter will examine the various efforts by the community to improve Big Data framework performance on HPC clusters.
-
-	\section{In-memory file systems for Spark}
+	\section{Spark in the cloud}
 

### 2018-02-02 paper.tex
more notes

@@
 Spark fails with medium to large applications due to "lax garbage collection in block and shuffle managers".
-To resolve these issues, the authors first attempted to make all executors share a descriptor pool.
+To resolve these issues, the authors attempted to make all executors share a descriptor pool.
 This allowed files that were already opened by other executors to be accessed by the executor through the descriptor rather than having to reopen it again.
@@
 Due to Inode constraints, a livelock may occur when descriptor pool is at capacity and an executor is attempting to open a file.
-
-	.the authors created mounted a local filesystem backed by a Lustre file \note{requires admin privilege}.
+The authors also created mounted a local filesystem backed by a Lustre file (lustremount) to address the issues related to an in-memory file system \note{requires admin privilege}.
+It was found that the use of an in-memory file system improved performance by 7.7x and the lustremount improved performance by 6.6x.
+It was also found that implementation of the filepool improved performance.
+	Another issue observed by the authors is that of poor block management as a result of memory constraints resulting in a significant amount of vertical movement.
+When available memory is exhausted, the least recently used block is evicted by the block-manager.
+When a call is made to this removed block, it will need to be recomputed, and in the process blocks required for recomputation may also be evicted and necessitate recomputation of those blocks, thus leading to a chain of eviction and recomputation.
+To resolve this, marking intermediate RDDs as persistent, such that their results are saved to storage instead of requiring recomputation, may be applied.
+	\note{they didn't look at improving initial application I/O, though believe that it could be improved with lustremount and filepool. }
 	\note{for each partition, a shuffle file is created and written to as many times as there are partitions.
@@
 Metadata access is therefore O(partitions2)}
+
+	\subsection{tachyon paper}
+	Tachyon is an in-memory file system designed to improve read and write performance without impacting fault tolerance of the system.
+As write performance is impeded by the need to replicate data, Tachyon leverages the concept of lineage to recompute the output if data is lost.
+Tachyon exhibits 110x faster write throughput than in-memory HDFS and 4.4x faster end-to-end latency.
+	\subsection{triple-H}
 

### 2018-02-07 paper.tex
blurb on nipype

@@
 		\subsection{summary}
+		Nipype\cite{nipype} is a popular neuroimaging pipelinening framework written in python.
+		It provides users with uniform access to the rich ecosystem of neuroimaging software
+		libraries (e.g. SPM, FSL, Freesurfer) through its \textit{Interfaces}. Should users want to
+		use their own custom tool, it is also possible to create an interface for that. To execute 
+		python code in Nipype, the \textit{Function} interface can be used. The framework is 
+		also easy to use, and thus does not contribute to the steep learning curve new
+		researchers must face.
+
+		A pipeline in Nipype, consists of a data analysis \textit{Workflow} of connected \textit{interfaces}.
+The workflow is a directed acyclic graph that connects the outputs of one interface to be the inputs of another.
+A workflow may also consist of connected workflows.
+Workflows may be executed in parallel either locally or an a cluster through the use of a plugin in the Workflow's run function.
+For debugging, Nipype uses \textit{graphiz} to generate a static graph representing the Nodes and their relationships.
+
+		 Each \textit{interface} within the workflow must be contained within a \textit{Node} or a \textit{MapNode} object.
+The MapNode class being a subclass of the Node class.
+Wrapping an interface in such objects ensures that interfaces are executed within a uniquely named directories, which in turn enables provenance tracking.
+Node objects also hash inputs, provide the ability to iterate over inputs and cache results.
+Hash inputs are useful in the case of Workflow recomputation, where Nipype will only recompute Nodes whose inputs have changed since the previous run.
+
+		Inspired by the mapreduce paradigm, the MapNode copies the interface to each input and executes it independently.
+A reduce is subsequently performed by the MapNode to return the output in the form of a list.
+This differs from a Node object which can only execute the interface on a single input at a time.
+
+		To enable execution of different parameters on the input, Nodes have a property known as \textit{Iterables}.
+Similarly to a MapNode, a copy of the \textit{interface} will be made for each input.
+However, with \textit{iterables}, a copy of each dependent node will also be made.
+
 		\subsection{limitations of nipype}
+		Nipype, although meeting the various needs of researchers in Neuroinformatics, was not designed with the processing of Big Data in mind.
+Moreover, its performance relative to current Big Data frameworks remains unknown.
+		\note{nipype performs a lot of disk I/O. Recomputation of workflows vs RDD caching}
+
 	\section{PSOM}
@@
 		\subsection{use of containers in nipype, psom, spm}
-\chapter{Scientific workflows in Spark}
-	\section{Introduction}
-	\section{ariel's paper and thunders astronomy - }
-	\section{Spark on hpc's or }
-	\note{large dataset for valduriez}
-	Prior to examining how Big Data frameworks could be improved on HPC clusters, it is important to have a baseline of how it performs natively on an HPC cluster.
-Particularly, how it behave when processing scientific workflows.
-In a recent study done by \cite{valduriez}, they examined the scalability of Spark on an HPC cluster using a black-box scientific workflow.
-Five tests were performed in their study: 1) Measuring execution time when task and resources available increase proportionally, 2) Measuring the execution time of very short and short tasks, 3) Measuring the execution time of short (5s) to long tasks (120s), 4) Varying the number of tasks with fixed task durations, 5) Mixed task durations and varied number of tasks.
-Their results show Spark behaves as expected when tasks and resources are increased exponentially, and that Spark scales better than expected with tasks of longer durations.
-This is due to the fact that the scheduler is overloaded in the case of many short tasks in addition to their being a significant amount of I/O occurring between each task.
-Although I/O is problematic in both data center and HPC infrastructures, HPC is optimized for data transfer over the network rather than disk I/O, and therefore, such frequent I/O negatively impacts the system.
-However, many scientific workflows are composed of longer tasks, and therefore, Spark scales very well with scientific workflows in an HPC environment.
+
 \chapter{Domain nonspecific BigData frameworks}
@@
 		\end{itemize}
+\chapter{Scientific workflows in Spark}
+	\section{Introduction}
+	\section{ariel's paper and thunders astronomy - }
+	\section{Spark on hpc's or }
+	\note{large dataset for valduriez}
+	Prior to examining how Big Data frameworks could be improved on HPC clusters, it is important to have a baseline of how it performs natively on an HPC cluster.
+Particularly, how it behave when processing scientific workflows.
+In a recent study done by \cite{valduriez}, they examined the scalability of Spark on an HPC cluster using a black-box scientific workflow.
+Five tests were performed in their study: 1) Measuring execution time when task and resources available increase proportionally, 2) Measuring the execution time of very short and short tasks, 3) Measuring the execution time of short (5s) to long tasks (120s), 4) Varying the number of tasks with fixed task durations, 5) Mixed task durations and varied number of tasks.
+Their results show Spark behaves as expected when tasks and resources are increased exponentially, and that Spark scales better than expected with tasks of longer durations.
+This is due to the fact that the scheduler is overloaded in the case of many short tasks in addition to their being a significant amount of I/O occurring between each task.
+Although I/O is problematic in both data center and HPC infrastructures, HPC is optimized for data transfer over the network rather than disk I/O, and therefore, such frequent I/O negatively impacts the system.
+However, many scientific workflows are composed of longer tasks, and therefore, Spark scales very well with scientific workflows in an HPC environment.
+
 \chapter{Optimizing Big Data frameworks for High-Performance Computing}
 
+	\note{precise the systems that will be used. both have distributed memory. data centre is hpc cluster with slower network, because may span multiple racks and domains}
+	\note{variety of jobs on hpc whereas dedicated map-reduce jobs in a data center}
+	\note{talk about spark on cloud and the challenges associated with}
+	\note{talk about why this is important}
+	\note{talk }
 	High-Performance Computing (HPC) clusters are extensively used by the scientific community to process complex, computation intensive problems.
@@
 This is due to the fact that file system metadata latency (more specifically, file open) is the determinant factor for performance on HPC, whereas network dominates performance in a typical data centre configuration.
-It was found by the authors that using either a local in-memory filesystem or a filesystem mounted to a single Lustre file improved performance up to the point where it matched that of a workstation.
+It was found by the authors that using either a local in-memory filesystem or a filesystem mounted to a single Lustre file improved performance up to a point where it matched that of a workstation.
 In addition, use of a cache was found to reduce file system metadata operations, and thus improve overall performance.
@@
 	\subsection{tachyon paper}
-	Tachyon is an in-memory file system designed to improve read and write performance without impacting fault tolerance of the system.
+	I/O is a known bottleneck in Spark that is exacerbated in HPC systems in which data is stored in a separate cluster outside compute nodes.
+One possible solution used to mitigate the effects of the I/O bottleneck is to store data in an in-memory filesystem rather than on disk.
+Alluxio and Triple-H are two known examples of such filesystems.
+
+	Alluxio, formerly known as Tachyon, is an in-memory file system designed to improve read and write performance without impacting fault tolerance of the system.
 As write performance is impeded by the need to replicate data, Tachyon leverages the concept of lineage to recompute the output if data is lost.
@@
 
-	\section{Spark in the cloud}
+	%%%\section{Spark in the cloud}
 
-	Using Spark against a typical HPC parallel file system such a Luster has been found to impede scalability.
+	%%%Using Spark against a typical HPC parallel file system such a Luster has been found to impede scalability.
 This was found to be due the file system metadata latency, which limits Spark scalability to \textit{O}(10\textsuperscript{2}) cores \cite{Chaimov:2016}.

### 2018-02-09 paper.tex
Comments on nipype paragraph

@@
 \usepackage{xcolor}
+\usepackage{ulem}
 
 \newcommand{\note}[1]{\textcolor{green}{\textit{note}: #1}}
+\newcommand{\tristan}[1]{\textcolor{red}{TG: #1}}
+\newcommand{\weird}[1]{\uwave{#1}}
 
@@
 		\subsection{summary}
-		Nipype\cite{nipype} is a popular neuroimaging pipelinening framework written in python.
+		Nipype\cite{nipype} is a popular neuroimaging pipelinening framework written in Python.
 		It provides users with uniform access to the rich ecosystem of neuroimaging software
-		libraries (e.g. SPM, FSL, Freesurfer) through its \textit{Interfaces}. Should users want to
-		use their own custom tool, it is also possible to create an interface for that. To execute 
-		python code in Nipype, the \textit{Function} interface can be used. The framework is 
+		libraries (e.g. SPM, FSL, Freesurfer) through its \textit{Interfaces}. \weird{Should users want to
+		use their own custom tool, it is also possible to create an interface for that}. To execute 
+		Python code in Nipype, the \textit{Function} interface can be used \tristan{for arbitrary code?}. The framework is 
 		also easy to use, and thus does not contribute to the steep learning curve new
@@
 
-		A pipeline in Nipype, consists of a data analysis \textit{Workflow} of connected \textit{interfaces}.
-The workflow is a directed acyclic graph that connects the outputs of one interface to be the inputs of another.
-A workflow may also consist of connected workflows.
-Workflows may be executed in parallel either locally or an a cluster through the use of a plugin in the Workflow's run function.
-For debugging, Nipype uses \textit{graphiz} to generate a static graph representing the Nodes and their relationships.
+		A pipeline in Nipype, consists of a data analysis
+                \textit{Workflow} of connected
+                \textit{interfaces}. The workflow is a directed
+                acyclic graph that connects the outputs of one
+                interface to be the inputs of another. A workflow may
+                also consist of connected workflows.
 
-		 Each \textit{interface} within the workflow must be contained within a \textit{Node} or a \textit{MapNode} object.
-The MapNode class being a subclass of the Node class.
-Wrapping an interface in such objects ensures that interfaces are executed within a uniquely named directories, which in turn enables provenance tracking.
-Node objects also hash inputs, provide the ability to iterate over inputs and cache results.
-Hash inputs are useful in the case of Workflow recomputation, where Nipype will only recompute Nodes whose inputs have changed since the previous run.
+                \tristan{The text below is about workflow execution
+                  while the text above is about workflow description,
+                  this should be a new paragraph.}  Workflows may be
+                executed in parallel either locally or an a cluster
+                through the use of a plugin in the Workflow's run
+                function. For debugging, Nipype uses \textit{graphviz}
+                to generate a static graph representing the Nodes and
+                their relationships. \tristan{debugging might be a
+                  third theme, in addition to description and
+                  execution. Provenance might go in debugging.}
 
-		Inspired by the mapreduce paradigm, the MapNode copies the interface to each input and executes it independently.
-A reduce is subsequently performed by the MapNode to return the output in the form of a list.
-This differs from a Node object which can only execute the interface on a single input at a time.
+		 Each \textit{interface} within the workflow must be
+                 contained within a \textit{Node} or a
+                 \textit{MapNode} object, the MapNode class being a
+                 subclass of the Node class. Wrapping an interface in
+                 such objects ensures that interfaces are executed
+                 within a uniquely named directories, which in turn
+                 enables provenance tracking. Node objects also hash
+                 inputs, provide the ability to iterate over inputs
+                 and cache results. Hash inputs are useful in the case
+                 of Workflow recomputation, where Nipype will only
+                 recompute Nodes whose inputs have changed since the
+                 previous run.
 
-		To enable execution of different parameters on the input, Nodes have a property known as \textit{Iterables}.
-Similarly to a MapNode, a copy of the \textit{interface} will be made for each input.
-However, with \textit{iterables}, a copy of each dependent node will also be made.
+		Inspired by the MapReduce paradigm, the MapNode copies
+                the interface to each input and executes it
+                independently. \tristan{How are inputs defined/combined?} A reduce is subsequently performed by
+                the MapNode to return the output in the form of a
+                list. This differs from a Node object which can only
+                execute the interface on a single input at a time.
 
-		\subsection{limitations of nipype}
-		Nipype, although meeting the various needs of researchers in Neuroinformatics, was not designed with the processing of Big Data in mind.
-Moreover, its performance relative to current Big Data frameworks remains unknown.
-		\note{nipype performs a lot of disk I/O. Recomputation of workflows vs RDD caching}
+		To enable execution of different parameters on the
+                input, Nodes have a property known as
+                \textit{Iterables}. Similarly to a MapNode, a copy of
+                the \textit{interface} will be made for each
+                input. However, with \textit{iterables}, a copy of
+                each dependent node will also be made.
+
+		\subsection{Limitations}
+		Nipype, although meeting the various needs of
+                researchers in Neuroinformatics \tristan{vague}, was
+                not designed with the processing of Big Data in
+                mind. Moreover, its performance relative to current
+                Big Data frameworks remains unknown \tristan{You could
+                  refer to your MS thesis here}.  \note{nipype
+                  performs a lot of disk I/O. Recomputation of
+                  workflows vs RDD caching}
 

### 2018-02-09 paper.tex
reformat text

--- 2018-02-07 paper.tex
@@
 
-\usepackage{hyperref}
-\usepackage{xcolor}
+\usepackage{hyperref} \usepackage{xcolor}
 
@@
 
-\begin{document}
-\title{Pipeline systems and infrastructure for the efficient and open processing of Big neuroimaging Data}
-\author{Valerie Hayot-Sasson}
-\maketitle
-\begin{abstract}
-  Text of abstract.  
-\end{abstract}
-\tableofcontents
-\chapter{Introduction}
-intro here
-\chapter{BigData and Open science in neuroimaging}
-    \note{Likely will not be split in two separate sections}
-    \section{large images}
-        \begin{itemize}  
-            \item BigBrain
-            \item Micro CT 
-            \item EM 
-        \end{itemize}
-    \section{large datasets}
-    	\note{off the top of my head, but need to double check}
-        \begin{itemize}
-            \item HCP
-            \item ADNI
-            \item OpenFMRI
-            \item BIDS/BIDS apps
-        \end{itemize}
-    An effective engine to process such data requires the following characteristics
-        \begin{itemize}
-            \item in-memory computing
-            \item data locality
-            \item lazy evaluation
-        \end{itemize}
-        \section{open-science /data sharing platforms for neuroimaging}
-        		is open science just data sharing, computing. 
-        		\subsection{Introduction}
-		\subsection{CBRAIN}
-		\subsection{OpenNEURO}
-		\subsection{NeuroData}
-\chapter{Workflow engines for neuroimaging data}
-	\section{Definition of workflows/pipelines}
-	\section{Nipype}
-		\subsection{summary}
-		Nipype\cite{nipype} is a popular neuroimaging pipelinening framework written in python.
-		It provides users with uniform access to the rich ecosystem of neuroimaging software
-		libraries (e.g. SPM, FSL, Freesurfer) through its \textit{Interfaces}. Should users want to
-		use their own custom tool, it is also possible to create an interface for that. To execute 
-		python code in Nipype, the \textit{Function} interface can be used. The framework is 
-		also easy to use, and thus does not contribute to the steep learning curve new
-		researchers must face.
+\begin{document} \title{Pipeline systems and infrastructure for the efficient
+and open processing of Big neuroimaging Data} \author{Valerie Hayot-Sasson}
+\maketitle \begin{abstract} Text of abstract.  \end{abstract} \tableofcontents
+\chapter{Introduction} intro here \chapter{BigData and Open science in
+neuroimaging} \note{Likely will not be split in two separate sections}
+\section{large images} \begin{itemize}  \item BigBrain \item Micro CT \item EM
+\end{itemize} \section{large datasets} \note{off the top of my head, but need to
+double check} \begin{itemize} \item HCP \item ADNI \item OpenFMRI \item
+BIDS/BIDS apps \end{itemize} An effective engine to process such data requires
+the following characteristics \begin{itemize} \item in-memory computing \item
+data locality \item lazy evaluation \end{itemize} \section{open-science /data
+sharing platforms for neuroimaging} is open science just data sharing,
+computing.  \subsection{Introduction} \subsection{CBRAIN} \subsection{OpenNEURO}
+\subsection{NeuroData} \chapter{Workflow engines for neuroimaging data}
+\section{Definition of workflows/pipelines} \section{Nipype}
+\subsection{summary} Nipype\cite{nipype} is a popular neuroimaging pipelinening
+framework written in python.  It provides users with uniform access to the rich
+ecosystem of neuroimaging software libraries (e.g. SPM, FSL, Freesurfer) through
+its \textit{Interfaces}. Should users want to use their own custom tool, it is
+also possible to create an interface for that. To execute python code in Nipype,
+the \textit{Function} interface can be used. The framework is also easy to use,
+and thus does not contribute to the steep learning curve new researchers must
+face.
 
-		A pipeline in Nipype, consists of a data analysis \textit{Workflow} of connected \textit{interfaces}.
-The workflow is a directed acyclic graph that connects the outputs of one interface to be the inputs of another.
-A workflow may also consist of connected workflows.
-Workflows may be executed in parallel either locally or an a cluster through the use of a plugin in the Workflow's run function.
-For debugging, Nipype uses \textit{graphiz} to generate a static graph representing the Nodes and their relationships.
+        A pipeline in Nipype, consists of a data analysis \textit{Workflow} of
+connected \textit{interfaces}. The workflow is a directed acyclic graph that
+connects the outputs of one interface to be the inputs of another. A workflow
+may also consist of connected workflows. Workflows may be executed in parallel
+either locally or an a cluster through the use of a plugin in the Workflow's run
+function. For debugging, Nipype uses \textit{graphiz} to generate a static graph
+representing the Nodes and their relationships.
 
-		 Each \textit{interface} within the workflow must be contained within a \textit{Node} or a \textit{MapNode} object.
-The MapNode class being a subclass of the Node class.
-Wrapping an interface in such objects ensures that interfaces are executed within a uniquely named directories, which in turn enables provenance tracking.
-Node objects also hash inputs, provide the ability to iterate over inputs and cache results.
-Hash inputs are useful in the case of Workflow recomputation, where Nipype will only recompute Nodes whose inputs have changed since the previous run.
+         Each \textit{interface} within the workflow must be contained within a
+\textit{Node} or a \textit{MapNode} object. The MapNode class being a subclass
+of the Node class. Wrapping an interface in such objects ensures that interfaces
+are executed within a uniquely named directories, which in turn enables
+provenance tracking. Node objects also hash inputs, provide the ability to
+iterate over inputs and cache results. Hash inputs are useful in the case of
+Workflow recomputation, where Nipype will only recompute Nodes whose inputs have
+changed since the previous run.
 
-		Inspired by the mapreduce paradigm, the MapNode copies the interface to each input and executes it independently.
-A reduce is subsequently performed by the MapNode to return the output in the form of a list.
-This differs from a Node object which can only execute the interface on a single input at a time.
+        Inspired by the mapreduce paradigm, the MapNode copies the interface to
+each input and executes it independently. A reduce is subsequently performed by
+the MapNode to return the output in the form of a list. This differs from a Node
+object which can only execute the interface on a single input at a time.
 
-		To enable execution of different parameters on the input, Nodes have a property known as \textit{Iterables}.
-Similarly to a MapNode, a copy of the \textit{interface} will be made for each input.
-However, with \textit{iterables}, a copy of each dependent node will also be made.
+        To enable execution of different parameters on the input, Nodes have a
+property known as \textit{Iterables}. Similarly to a MapNode, a copy of the
+\textit{interface} will be made for each input. However, with
+\textit{iterables}, a copy of each dependent node will also be made.
 
-		\subsection{limitations of nipype}
-		Nipype, although meeting the various needs of researchers in Neuroinformatics, was not designed with the processing of Big Data in mind.
-Moreover, its performance relative to current Big Data frameworks remains unknown.
-		\note{nipype performs a lot of disk I/O. Recomputation of workflows vs RDD caching}
+        \subsection{limitations of nipype} Nipype, although meeting the various
+needs of researchers in Neuroinformatics, was not designed with the processing
+of Big Data in mind. Moreover, its performance relative to current Big Data
+frameworks remains unknown.  \note{nipype performs a lot of disk I/O.
+Recomputation of workflows vs RDD caching}
 
-	\section{PSOM}
-		\subsection{summary}
-		\subsection{limitations of PSOM}
-	\section{SPM}
-		\subsection{summary}
-		\subsection{limitations of SPM}
-	\section{Use of containers in Workflow engines}
-		\subsection{use of containers in nipype, psom, spm}
+    \section{PSOM} \subsection{summary} \subsection{limitations of PSOM}
+\section{SPM} \subsection{summary} \subsection{limitations of SPM} \section{Use
+of containers in Workflow engines} \subsection{use of containers in nipype,
+psom, spm}
 
-\chapter{Domain nonspecific BigData frameworks}
-	\section{mapreduce}
-		\begin{itemize}
-			\item many computations can be described through map and reduce
-			\item fault tolerant
-		\end{itemize}
-	\section{spark}
-		\begin{itemize}
-			\item based on map-reduce
-			\item in-memory processing
-			\item lazy evaluation
-			\item general framework
-		\end{itemize}
-	\section{Use of containers with Big Data frameworks}
-	\section{Hadoop Distributed file system}
-		\note{maybe should go in between mapreduce and spark}
-		\begin{itemize}
-			\item enables data-locality in Spark/ decouples data locality logic from mapreduce\
-			\item data replication
-			\item fault-tolerance
-		\end{itemize}
-\chapter{Scientific workflows in Spark}
-	\section{Introduction}
-	\section{ariel's paper and thunders astronomy - }
-	\section{Spark on hpc's or }
-	\note{large dataset for valduriez}
-	Prior to examining how Big Data frameworks could be improved on HPC clusters, it is important to have a baseline of how it performs natively on an HPC cluster.
-Particularly, how it behave when processing scientific workflows.
-In a recent study done by \cite{valduriez}, they examined the scalability of Spark on an HPC cluster using a black-box scientific workflow.
-Five tests were performed in their study: 1) Measuring execution time when task and resources available increase proportionally, 2) Measuring the execution time of very short and short tasks, 3) Measuring the execution time of short (5s) to long tasks (120s), 4) Varying the number of tasks with fixed task durations, 5) Mixed task durations and varied number of tasks.
-Their results show Spark behaves as expected when tasks and resources are increased exponentially, and that Spark scales better than expected with tasks of longer durations.
-This is due to the fact that the scheduler is overloaded in the case of many short tasks in addition to their being a significant amount of I/O occurring between each task.
-Although I/O is problematic in both data center and HPC infrastructures, HPC is optimized for data transfer over the network rather than disk I/O, and therefore, such frequent I/O negatively impacts the system.
-However, many scientific workflows are composed of longer tasks, and therefore, Spark scales very well with scientific workflows in an HPC environment.
+\chapter{Domain nonspecific BigData frameworks} \section{mapreduce}
+\begin{itemize} \item many computations can be described through map and reduce
+\item fault tolerant \end{itemize} \section{spark} \begin{itemize} \item based
+on map-reduce \item in-memory processing \item lazy evaluation \item general
+framework \end{itemize} \section{Use of containers with Big Data frameworks}
+\section{Hadoop Distributed file system} \note{maybe should go in between
+mapreduce and spark} \begin{itemize} \item enables data-locality in Spark/
+decouples data locality logic from mapreduce\ \item data replication \item
+fault-tolerance \end{itemize} \chapter{Scientific workflows in Spark}
+\section{Introduction} \section{ariel's paper and thunders astronomy - }
+\section{Spark on hpc's or } \note{large dataset for valduriez} Prior to
+examining how Big Data frameworks could be improved on HPC clusters, it is
+important to have a baseline of how it performs natively on an HPC cluster.
+Particularly, how it behave when processing scientific workflows. In a recent
+study done by \cite{valduriez}, they examined the scalability of Spark on an HPC
+cluster using a black-box scientific workflow. Five tests were performed in
+their study: 1) Measuring execution time when task and resources available
+increase proportionally, 2) Measuring the execution time of very short and short
+tasks, 3) Measuring the execution time of short (5s) to long tasks (120s), 4)
+Varying the number of tasks with fixed task durations, 5) Mixed task durations
+and varied number of tasks. Their results show Spark behaves as expected when
+tasks and resources are increased exponentially, and that Spark scales better
+than expected with tasks of longer durations. This is due to the fact that the
+scheduler is overloaded in the case of many short tasks in addition to their
+being a significant amount of I/O occurring between each task. Although I/O is
+problematic in both data center and HPC infrastructures, HPC is optimized for
+data transfer over the network rather than disk I/O, and therefore, such
+frequent I/O negatively impacts the system. However, many scientific workflows
+are composed of longer tasks, and therefore, Spark scales very well with
+scientific workflows in an HPC environment.
 
@@
 
-	\note{precise the systems that will be used. both have distributed memory. data centre is hpc cluster with slower network, because may span multiple racks and domains}
-	\note{variety of jobs on hpc whereas dedicated map-reduce jobs in a data center}
-	\note{talk about spark on cloud and the challenges associated with}
-	\note{talk about why this is important}
-	\note{talk }
-	High-Performance Computing (HPC) clusters are extensively used by the scientific community to process complex, computation intensive problems.
-Due to their importance in the advancement of scientific research, it is necessary that Big Data frameworks used to process scientific data are compatible with them.
+    \note{precise the systems that will be used. both have distributed memory.
+data centre is hpc cluster with slower network, because may span multiple racks
+and domains} \note{variety of jobs on hpc whereas dedicated map-reduce jobs in a
+data center} \note{talk about spark on cloud and the challenges associated with}
+\note{talk about why this is important} \note{talk } High-Performance Computing
+(HPC) clusters are extensively used by the scientific community to process
+complex, computation intensive problems. Due to their importance in the
+advancement of scientific research, it is necessary that Big Data frameworks
+used to process scientific data are compatible with them.
 
-	Big Data frameworks were optimized for a commodity cluster.
-Such clusters are made up of nodes with homogenous storage connected by lower-end network cables (e.g.
-Ethernet).
-As a result, moving data from local storage to memory was much less costly in these infrastructures than moving data across the network.
-Thus, Big Data frameworks implemented data locality in an attempt to reduce costly network (horizontal) transfers and favour more performant local storage to memory (vertical) transfers.
-HPC clusters, however, are not necessarily commodity clusters.
-Large HPC clusters are commonly made up of supercomputing nodes connected by a high bandwidth, low-latency network (e.g.
-InfiniBand).
-Nodes may contain their own local storage (e.g.
-HDD, SSD, tmpfs) and are all connected to a distributed file system (e.g GPFS, Lustre).
-Due to their employment of a high bandwidth low latency network, network transfers on such an HPC cluster may not be as costly as that of a typical data center, and it may in fact be more costly to do vertical transfers.
-Furthermore, HPC systems may enforce a batch submission system.
-This required users to specify number of nodes, cores and total processing time; requirements which are unknown to MapReduce users.
+    Big Data frameworks were optimized for a commodity cluster.  Such clusters
+are made up of nodes with homogenous storage connected by lower-end network
+cables (e.g. Ethernet). As a result, moving data from local storage to memory
+was much less costly in these infrastructures than moving data across the
+network. Thus, Big Data frameworks implemented data locality in an attempt to
+reduce costly network (horizontal) transfers and favour more performant local
+storage to memory (vertical) transfers. HPC clusters, however, are not
+necessarily commodity clusters. Large HPC clusters are commonly made up of
+supercomputing nodes connected by a high bandwidth, low-latency network (e.g.
+InfiniBand). Nodes may contain their own local storage (e.g. HDD, SSD, tmpfs)
+and are all connected to a distributed file system (e.g GPFS, Lustre).  Due to
+their employment of a high bandwidth low latency network, network transfers on
+such an HPC cluster may not be as costly as that of a typical data center, and
+it may in fact be more costly to do vertical transfers. Furthermore, HPC systems
+may enforce a batch submission system. This required users to specify number of
+nodes, cores and total processing time; requirements which are unknown to
+MapReduce users.
 
-	This chapter will examine the various efforts by the community to improve Big Data framework performance on HPC clusters.
+    This chapter will examine the various efforts by the community to improve
+Big Data framework performance on HPC clusters.
 
@@
 
-	In paper \note{Scaling Spark on HPC Systems} it was found that Spark on a Lustre backend performed significantly (4x) slower than that of a workstation using local fast SSDs for storage.
-This is due to the fact that file system metadata latency (more specifically, file open) is the determinant factor for performance on HPC, whereas network dominates performance in a typical data centre configuration.
-It was found by the authors that using either a local in-memory filesystem or a filesystem mounted to a single Lustre file improved performance up to a point where it matched that of a workstation.
-In addition, use of a cache was found to reduce file system metadata operations, and thus improve overall performance.
-	The issues related to using an in-memory filesystem is that it is limited by the amount of physical memory available, and a job will fail if physical memory runs out.
-Moreover, there is not persistence or resilience of data as it is not saved to non-volatile storage (disk).
-Spark fails with medium to large applications due to "lax garbage collection in block and shuffle managers".
-To resolve these issues, the authors attempted to make all executors share a descriptor pool.
-This allowed files that were already opened by other executors to be accessed by the executor through the descriptor rather than having to reopen it again.
-However, the node OS image is subject to number of Inode constraints as well Lustre limits on number of open files for a given job.
-Due to Inode constraints, a livelock may occur when descriptor pool is at capacity and an executor is attempting to open a file.
-The authors also created mounted a local filesystem backed by a Lustre file (lustremount) to address the issues related to an in-memory file system \note{requires admin privilege}.
-It was found that the use of an in-memory file system improved performance by 7.7x and the lustremount improved performance by 6.6x.
-It was also found that implementation of the filepool improved performance.
-	Another issue observed by the authors is that of poor block management as a result of memory constraints resulting in a significant amount of vertical movement.
-When available memory is exhausted, the least recently used block is evicted by the block-manager.
-When a call is made to this removed block, it will need to be recomputed, and in the process blocks required for recomputation may also be evicted and necessitate recomputation of those blocks, thus leading to a chain of eviction and recomputation.
-To resolve this, marking intermediate RDDs as persistent, such that their results are saved to storage instead of requiring recomputation, may be applied.
-	\note{they didn't look at improving initial application I/O, though believe that it could be improved with lustremount and filepool. }
-	\note{for each partition, a shuffle file is created and written to as many times as there are partitions.
-An index file is also created which contains shuffle file locations and offsets.
-Metadata access is therefore O(partitions2)}
+    In paper \note{Scaling Spark on HPC Systems} it was found that Spark on a
+Lustre backend performed significantly (4x) slower than that of a workstation
+using local fast SSDs for storage. This is due to the fact that file system
+metadata latency (more specifically, file open) is the determinant factor for
+performance on HPC, whereas network dominates performance in a typical data
+centre configuration. It was found by the authors that using either a local
+in-memory filesystem or a filesystem mounted to a single Lustre file improved
+performance up to a point where it matched that of a workstation. In addition,
+use of a cache was found to reduce file system metadata operations, and thus
+improve overall performance.  The issues related to using an in-memory
+filesystem is that it is limited by the amount of physical memory available, and
+a job will fail if physical memory runs out. Moreover, there is not persistence
+or resilience of data as it is not saved to non-volatile storage (disk). Spark
+fails with medium to large applications due to "lax garbage collection in block
+and shuffle managers". To resolve these issues, the authors attempted to make
+all executors share a descriptor pool. This allowed files that were already
+opened by other executors to be accessed by the executor through the descriptor
+rather than having to reopen it again. However, the node OS image is subject to
+number of Inode constraints as well Lustre limits on number of open files for a
+given job. Due to Inode constraints, a livelock may occur when descriptor pool
+is at capacity and an executor is attempting to open a file. The authors also
+created mounted a local filesystem backed by a Lustre file (lustremount) to
+address the issues related to an in-memory file system \note{requires admin
+privilege}. It was found that the use of an in-memory file system improved
+performance by 7.7x and the lustremount improved performance by 6.6x. It was
+also found that implementation of the filepool improved performance.  Another
+issue observed by the authors is that of poor block management as a result of
+memory constraints resulting in a significant amount of vertical movement. When
+available memory is exhausted, the least recently used block is evicted by the
+block-manager. When a call is made to this removed block, it will need to be
+recomputed, and in the process blocks required for recomputation may also be
+evicted and necessitate recomputation of those blocks, thus leading to a chain
+of eviction and recomputation. To resolve this, marking intermediate RDDs as
+persistent, such that their results are saved to storage instead of requiring
+recomputation, may be applied.  \note{they didn't look at improving initial
+application I/O, though believe that it could be improved with lustremount and
+filepool. } \note{for each partition, a shuffle file is created and written to
+as many times as there are partitions. An index file is also created which
+contains shuffle file locations and offsets. Metadata access is therefore
+O(partitions2)}
 
-	\subsection{tachyon paper}
-	I/O is a known bottleneck in Spark that is exacerbated in HPC systems in which data is stored in a separate cluster outside compute nodes.
-One possible solution used to mitigate the effects of the I/O bottleneck is to store data in an in-memory filesystem rather than on disk.
+    \subsection{tachyon paper} I/O is a known bottleneck in Spark that is
+exacerbated in HPC systems in which data is stored in a separate cluster outside
+compute nodes. One possible solution used to mitigate the effects of the I/O
+bottleneck is to store data in an in-memory filesystem rather than on disk.
 Alluxio and Triple-H are two known examples of such filesystems.
 
-	Alluxio, formerly known as Tachyon, is an in-memory file system designed to improve read and write performance without impacting fault tolerance of the system.
-As write performance is impeded by the need to replicate data, Tachyon leverages the concept of lineage to recompute the output if data is lost.
-Tachyon exhibits 110x faster write throughput than in-memory HDFS and 4.4x faster end-to-end latency.
+    Alluxio, formerly known as Tachyon, is an in-memory file system designed to
+improve read and write performance without impacting fault tolerance of the
+system. As write performance is impeded by the need to replicate data, Alluxio
+leverages the concept of lineage to recompute the output if data is lost.
+Tachyon exhibits 110x faster write throughput than in-memory HDFS and 4.4x
+faster end-to-end latency.
+
+    Alluxio has is made up of two layers: the lineage layer and the persistence
+layer. The lineage layer is the layer that performs recomputations to obtain the
+desired output. The persistance layer contains all the code used to regenerate
+the data, in addition to any data that cannot be recovered through lineage
+as it may be too large to fit in memory or has no lineage information stored.
+
     \subsection{triple-H}
 
-	%%%In other words, such frameworks would try to limit data movement, as network cables are expected to have low bandwidth and high latency.
-In addition, each node in a commodity cluster is expected to have access to its own local storage homogenous to that of the rest of the cluster, thus replication is necessary to ensure data availability.
-This is, however, not necessarily the case for computing grids.
-Computing grids may have access to Infinibands, which provide high bandwidth and low latency.
-Moreover, grids may use heterogenous storage in addition to a parallel file system such as Lustre.
-As a result, it may, in some cases, be faster to transfer data over the network than to access the nearest copy of the data.
-As well, data replication may burden a system with centralized storage as multiple copies of the data would be created, taking a significant amount of space.%%%
+    %%%In other words, such frameworks would try to limit data movement, as
+    %network cables are expected to have low bandwidth and high latency. In
+    %addition, each node in a commodity cluster is expected to have access to
+    %its own local storage homogenous to that of the rest of the cluster, thus
+    %replication is necessary to ensure data availability. This is, however, not
+    %necessarily the case for computing grids. Computing grids may have access
+    %to Infinibands, which provide high bandwidth and low latency. Moreover,
+    %grids may use heterogenous storage in addition to a parallel file system
+    %such as Lustre. As a result, it may, in some cases, be faster to transfer
+    %data over the network than to access the nearest copy of the data. As well,
+    %data replication may burden a system with centralized storage as multiple
+    %copies of the data would be created, taking a significant amount of
+    %space.%%%
 
-	%%%Big Data frameworks were not designed with high-performance computing in mind.
-Big Data filesystems such as were intended to run in data centers with homogenous storage (ex.
-HDDs, SDDs.
-etc,.).
-In such infrastructures, data movement is expensive.
-To counteract this, strategies were employed to avoid moving data around (data locality).
-In contrast, HPC clusters use heterogenous storage (ex.
-a combination of HDD, SSD, Ram Disk and Parallel file systems) with compute nodes connected by high-performance interconnects.
-HPC nodes may therefore not all be equipped with the same type of storage devices, some of which may be faster than others.
-Unlike data centers, HPC compute nodes do not have local storage.
-Moreover, the high-performance interconnects ensure low latency and high throughput of the system.
-In some instances, it may be even more performant to move data around rather than ensuring data locality.
-When a single-node HPC cluster using the Luster parallel file system was compared to a workstation with local SSD, it was found that Spark performed about 4x slower on the HPC cluster \cite{Chaimov:2016}.
-As HDFS is not conscientious of the underlying infrastructure and expects a data center-type infrastructure, it cannot achieve its full performance potential.
-%%%
+    %%%Big Data frameworks were not designed with high-performance computing in
+    %mind. Big Data filesystems such as were intended to run in data centers
+    %with homogenous storage (ex. HDDs, SDDs. etc,.). In such infrastructures,
+    %data movement is expensive. To counteract this, strategies were employed to
+    %avoid moving data around (data locality). In contrast, HPC clusters use
+    %heterogenous storage (ex. a combination of HDD, SSD, Ram Disk and Parallel
+    %file systems) with compute nodes connected by high-performance
+    %interconnects. HPC nodes may therefore not all be equipped with the same
+    %type of storage devices, some of which may be faster than others. Unlike
+    %data centers, HPC compute nodes do not have local storage. Moreover, the
+    %high-performance interconnects ensure low latency and high throughput of
+    %the system. In some instances, it may be even more performant to move data
+    %around rather than ensuring data locality. When a single-node HPC cluster
+    %using the Luster parallel file system was compared to a workstation with
+    %local SSD, it was found that Spark performed about 4x slower on the HPC
+    %cluster \cite{Chaimov:2016}.  As HDFS is not conscientious of the
+    %underlying infrastructure and expects a data center-type infrastructure, it
+    %cannot achieve its full performance potential. %%%
 
@@
 
-	%%%Using Spark against a typical HPC parallel file system such a Luster has been found to impede scalability.
-This was found to be due the file system metadata latency, which limits Spark scalability to \textit{O}(10\textsuperscript{2}) cores \cite{Chaimov:2016}.
+    %%%Using Spark against a typical HPC parallel file system such a Luster has
+    %been found to impede scalability. This was found to be due the file system
+    %metadata latency, which limits Spark scalability to
+    %\textit{O}(10\textsuperscript{2}) cores \cite{Chaimov:2016}.
 
-	%%%Compute Grids, interconnected networks of computing resources, are heavily used in scientific research for the processing of data.
-Due to their use and importance to the scientific community, it is necessary that big data frameworks used for processing neuroimaging data are compatible with them.
-Compute grids typically utilize heterogenous storage, such as a mixture of SSD, HDD, RAM Disk and parallel file systems (e.g.
-Luster).
-However, big data frameworks, such as Spark are designed to work efficiently with homogenous storage.
-Such differences result in reduced performance and inefficient storage use when big data frameworks are used on HPC clusters.
-This chapter will review the difference techniques applied to big data frameworks as well as HPC and grids to improve collocation of both services.%%%
+    %%%Compute Grids, interconnected networks of computing resources, are
+    %heavily used in scientific research for the processing of data. Due to
+    %their use and importance to the scientific community, it is necessary that
+    %big data frameworks used for processing neuroimaging data are compatible
+    %with them. Compute grids typically utilize heterogenous storage, such as a
+    %mixture of SSD, HDD, RAM Disk and parallel file systems (e.g. Luster).
+    %However, big data frameworks, such as Spark are designed to work
+    %efficiently with homogenous storage. Such differences result in reduced
+    %performance and inefficient storage use when big data frameworks are used
+    %on HPC clusters. This chapter will review the difference techniques applied
+    %to big data frameworks as well as HPC and grids to improve collocation of
+    %both services.%%%
 
-	%%%\subsection{Anatomy of Grid storage}
-	%%%Typical grids consist of a cluster of data nodes possibly connected to local storage (e.g.
-RAM Disk, HDD, SDD), but is also connected through a network to another cluster representing a parallel file system such as Lustre.
-RAM Disk is the fastest storage as all data is located in memory, whereas SSD provides slightly slower, persistent storage, but is a good alternative for big data.
-HDDs are the least performant form of local storage.%%%
-	%%%Lustre is a stateful, object-based parallel file system.
-It consists of three components: 1) the meta data server (MDS), 2) the Object Storage Server (OSS) and 3) the Luster Network (LNET), which enables clients to communicate with each other.%%%
+    %%%\subsection{Anatomy of Grid storage} %%Typical grids consist of a cluster
+    %of data nodes possibly connected to local storage (e.g. RAM Disk, HDD,
+    %SDD), but is also connected through a network to another cluster
+    %representing a parallel file system such as Lustre. RAM Disk is the fastest
+    %storage as all data is located in memory, whereas SSD provides slightly
+    %slower, persistent storage, but is a good alternative for big data. HDDs
+    %are the least performant form of local storage.%%% %%Lustre is a stateful,
+    %object-based parallel file system. It consists of three components: 1) the
+    %meta data server (MDS), 2) the Object Storage Server (OSS) and 3) the
+    %Luster Network (LNET), which enables clients to communicate with each
+    %other.%%%
 
-	In an attempt to improve performance of HDFS on HPC clusters, Triple-H was created.
-	\section{Why is use of such systems important}
-	\section{Scheduling}
-		\subsection{HPC schedulers: SLURM/QSUB}
-		\subsection{Big Data schedulers: YARN/MESOS}
-		\subsection{Multi-level scheduling w/ BigData schedulers}
-\chapter{Current status of Big Data frameworks in neuroscience}
-	\section{Ariel's paper}
-	\section{Thunder}
-\chapter{Conclusion}
+    In an attempt to improve performance of HDFS on HPC clusters, Triple-H was
+created.  \section{Why is use of such systems important} \section{Scheduling}
+\subsection{HPC schedulers: SLURM/QSUB} \subsection{Big Data schedulers:
+YARN/MESOS} \subsection{Multi-level scheduling w/ BigData schedulers}
+\chapter{Current status of Big Data frameworks in neuroscience} \section{Ariel's
+paper} \section{Thunder} \chapter{Conclusion}
 
@@
 
-\addcontentsline{toc}{chapter}{Bibliography}
-\bibliography{bibliography}
+\addcontentsline{toc}{chapter}{Bibliography} \bibliography{bibliography}
 \bibliographystyle{ieeetr}

### 2018-02-10 paper.tex

@@
 
-\usepackage{hyperref} \usepackage{xcolor}
+\usepackage{hyperref}
+\usepackage{xcolor}
+\usepackage{ulem}
 
 \newcommand{\note}[1]{\textcolor{green}{\textit{note}: #1}}
+\newcommand{\tristan}[1]{\textcolor{red}{TG: #1}}
+\newcommand{\weird}[1]{\uwave{#1}}
 
-\begin{document} \title{Pipeline systems and infrastructure for the efficient
-and open processing of Big neuroimaging Data} \author{Valerie Hayot-Sasson}
-\maketitle \begin{abstract} Text of abstract.  \end{abstract} \tableofcontents
-\chapter{Introduction} intro here \chapter{BigData and Open science in
-neuroimaging} \note{Likely will not be split in two separate sections}
-\section{large images} \begin{itemize}  \item BigBrain \item Micro CT \item EM
-\end{itemize} \section{large datasets} \note{off the top of my head, but need to
-double check} \begin{itemize} \item HCP \item ADNI \item OpenFMRI \item
-BIDS/BIDS apps \end{itemize} An effective engine to process such data requires
-the following characteristics \begin{itemize} \item in-memory computing \item
-data locality \item lazy evaluation \end{itemize} \section{open-science /data
-sharing platforms for neuroimaging} is open science just data sharing,
-computing.  \subsection{Introduction} \subsection{CBRAIN} \subsection{OpenNEURO}
-\subsection{NeuroData} \chapter{Workflow engines for neuroimaging data}
-\section{Definition of workflows/pipelines} \section{Nipype}
-\subsection{summary} Nipype\cite{nipype} is a popular neuroimaging pipelinening
-framework written in python.  It provides users with uniform access to the rich
-ecosystem of neuroimaging software libraries (e.g. SPM, FSL, Freesurfer) through
-its \textit{Interfaces}. Should users want to use their own custom tool, it is
-also possible to create an interface for that. To execute python code in Nipype,
-the \textit{Function} interface can be used. The framework is also easy to use,
-and thus does not contribute to the steep learning curve new researchers must
-face.
+\begin{document}
+\title{Pipeline systems and infrastructure for the efficient and open processing of Big neuroimaging Data}
+\author{Valerie Hayot-Sasson}
+\maketitle
+\begin{abstract}
+  Text of abstract.  
+\end{abstract}
+\tableofcontents
+\chapter{Introduction}
+intro here
+\chapter{BigData and Open science in neuroimaging}
+    \note{Likely will not be split in two separate sections}
+    \section{large images}
+        \begin{itemize}  
+            \item BigBrain
+            \item Micro CT 
+            \item EM 
+        \end{itemize}
+    \section{large datasets}
+    	\note{off the top of my head, but need to double check}
+        \begin{itemize}
+            \item HCP
+            \item ADNI
+            \item OpenFMRI
+            \item BIDS/BIDS apps
+        \end{itemize}
+    An effective engine to process such data requires the following characteristics
+        \begin{itemize}
+            \item in-memory computing
+            \item data locality
+            \item lazy evaluation
+        \end{itemize}
+        \section{open-science /data sharing platforms for neuroimaging}
+        		is open science just data sharing, computing. 
+        		\subsection{Introduction}
+		\subsection{CBRAIN}
+		\subsection{OpenNEURO}
+		\subsection{NeuroData}
+\chapter{Workflow engines for neuroimaging data}
+	\section{Definition of workflows/pipelines}
+	\section{Nipype}
+		\subsection{summary}
+		Nipype\cite{nipype} is a popular neuroimaging pipelinening framework written in Python.
+		It provides users with uniform access to the rich ecosystem of neuroimaging software
+		libraries (e.g. SPM, FSL, Freesurfer) through its \textit{Interfaces}. \weird{Should users want to
+		use their own custom tool, it is also possible to create an interface for that}. To execute 
+		Python code in Nipype, the \textit{Function} interface can be used \tristan{for arbitrary code?}. The framework is 
+		also easy to use, and thus does not contribute to the steep learning curve new
+		researchers must face.
 
-        A pipeline in Nipype, consists of a data analysis \textit{Workflow} of
-connected \textit{interfaces}. The workflow is a directed acyclic graph that
-connects the outputs of one interface to be the inputs of another. A workflow
-may also consist of connected workflows. Workflows may be executed in parallel
-either locally or an a cluster through the use of a plugin in the Workflow's run
-function. For debugging, Nipype uses \textit{graphiz} to generate a static graph
-representing the Nodes and their relationships.
+		A pipeline in Nipype, consists of a data analysis
+                \textit{Workflow} of connected
+                \textit{interfaces}. The workflow is a directed
+                acyclic graph that connects the outputs of one
+                interface to be the inputs of another. A workflow may
+                also consist of connected workflows.
 
-         Each \textit{interface} within the workflow must be contained within a
-\textit{Node} or a \textit{MapNode} object. The MapNode class being a subclass
-of the Node class. Wrapping an interface in such objects ensures that interfaces
-are executed within a uniquely named directories, which in turn enables
-provenance tracking. Node objects also hash inputs, provide the ability to
-iterate over inputs and cache results. Hash inputs are useful in the case of
-Workflow recomputation, where Nipype will only recompute Nodes whose inputs have
-changed since the previous run.
+                \tristan{The text below is about workflow execution
+                  while the text above is about workflow description,
+                  this should be a new paragraph.}  Workflows may be
+                executed in parallel either locally or an a cluster
+                through the use of a plugin in the Workflow's run
+                function. For debugging, Nipype uses \textit{graphviz}
+                to generate a static graph representing the Nodes and
+                their relationships. \tristan{debugging might be a
+                  third theme, in addition to description and
+                  execution. Provenance might go in debugging.}
 
-        Inspired by the mapreduce paradigm, the MapNode copies the interface to
-each input and executes it independently. A reduce is subsequently performed by
-the MapNode to return the output in the form of a list. This differs from a Node
-object which can only execute the interface on a single input at a time.
+		 Each \textit{interface} within the workflow must be
+                 contained within a \textit{Node} or a
+                 \textit{MapNode} object, the MapNode class being a
+                 subclass of the Node class. Wrapping an interface in
+                 such objects ensures that interfaces are executed
+                 within a uniquely named directories, which in turn
+                 enables provenance tracking. Node objects also hash
+                 inputs, provide the ability to iterate over inputs
+                 and cache results. Hash inputs are useful in the case
+                 of Workflow recomputation, where Nipype will only
+                 recompute Nodes whose inputs have changed since the
+                 previous run.
 
-        To enable execution of different parameters on the input, Nodes have a
-property known as \textit{Iterables}. Similarly to a MapNode, a copy of the
-\textit{interface} will be made for each input. However, with
-\textit{iterables}, a copy of each dependent node will also be made.
+		Inspired by the MapReduce paradigm, the MapNode copies
+                the interface to each input and executes it
+                independently. \tristan{How are inputs defined/combined?} A reduce is subsequently performed by
+                the MapNode to return the output in the form of a
+                list. This differs from a Node object which can only
+                execute the interface on a single input at a time.
 
-        \subsection{limitations of nipype} Nipype, although meeting the various
-needs of researchers in Neuroinformatics, was not designed with the processing
-of Big Data in mind. Moreover, its performance relative to current Big Data
-frameworks remains unknown.  \note{nipype performs a lot of disk I/O.
-Recomputation of workflows vs RDD caching}
+		To enable execution of different parameters on the
+                input, Nodes have a property known as
+                \textit{Iterables}. Similarly to a MapNode, a copy of
+                the \textit{interface} will be made for each
+                input. However, with \textit{iterables}, a copy of
+                each dependent node will also be made.
 
-    \section{PSOM} \subsection{summary} \subsection{limitations of PSOM}
-\section{SPM} \subsection{summary} \subsection{limitations of SPM} \section{Use
-of containers in Workflow engines} \subsection{use of containers in nipype,
-psom, spm}
+		\subsection{Limitations}
+		Nipype, although meeting the various needs of
+                researchers in Neuroinformatics \tristan{vague}, was
+                not designed with the processing of Big Data in
+                mind. Moreover, its performance relative to current
+                Big Data frameworks remains unknown \tristan{You could
+                  refer to your MS thesis here}.  \note{nipype
+                  performs a lot of disk I/O. Recomputation of
+                  workflows vs RDD caching}
 
-\chapter{Domain nonspecific BigData frameworks} \section{mapreduce}
-\begin{itemize} \item many computations can be described through map and reduce
-\item fault tolerant \end{itemize} \section{spark} \begin{itemize} \item based
-on map-reduce \item in-memory processing \item lazy evaluation \item general
-framework \end{itemize} \section{Use of containers with Big Data frameworks}
-\section{Hadoop Distributed file system} \note{maybe should go in between
-mapreduce and spark} \begin{itemize} \item enables data-locality in Spark/
-decouples data locality logic from mapreduce\ \item data replication \item
-fault-tolerance \end{itemize} \chapter{Scientific workflows in Spark}
-\section{Introduction} \section{ariel's paper and thunders astronomy - }
-\section{Spark on hpc's or } \note{large dataset for valduriez} Prior to
+    \section{PSOM} 
+        \subsection{summary} 
+        \subsection{limitations of PSOM}
+    \section{SPM} 
+        \subsection{summary} 
+        \subsection{limitations of SPM} 
+    \section{Use of containers in Workflow engines} 
+        \subsection{use of containers in nipype, psom, spm}
+
+\chapter{Domain nonspecific BigData frameworks} 
+    \section{mapreduce}
+        Summary of ~\cite{mapred}
+
+        Typically, simple computations need to be performed on 
+        increasingly growing large datasets. Processing of such 
+        large datasets require parallelization. As a result, 
+        these computations are complicated by details of 
+        parallelization, fault tolerance, data distribution and 
+        load balancing. The MapReduce library addresses this issue 
+        by allowing the expression of simple computations while 
+        abstracting the other details. The map and reduce paradigm 
+        was selected for this purpose as it was already commonly 
+        used in functional programming languages, such as Lisp, and 
+        it was observed that many computations could be expressed 
+        as such.
+
+        The MapReduce library operates by first dividing the data 
+        into manageable chunks (16 to 64MB). It then creates 
+        multiple copies of the program and distributes them across
+        nodes. There exists a single master node; the remainder are
+        worker nodes. The master node is responsible for delegating 
+        map and reduce tasks to the idle workers. There are two data
+        structures found in the master. The first stores the state 
+        (\textit{idle}, \textit{in-progress} or \textit{completed}) 
+        of each \textit{map} and \textit{reduce} task, and the second
+        stores the identity of the non-idle worker nodes. In contrast,
+        the worker nodes are responsible for the execution of 
+        \textit{map} or \textit{reduce} tasks. A worker node assigned
+        a \textit{map} task will parse out the key-value pairs from 
+        the input file and pass it to the \textit{map} function. The 
+        intermediate keys produced by the \textit{map} function are 
+        buffered in memory and periodically written to disk.The 
+        locations of these intermediate files are sent to the master 
+        who is responsible for forwarding these locations to the 
+        reducer. A worker node assigned a \textit{reduce} task uses 
+        remote procedure calls to obtain the intermediate local files
+        created by the map tasks and sorts the keys such that identical
+        keys are grouped together. Each unique key and its list of 
+        values are then submitted to the \textit{reduce} task and the 
+        output of the \textit{reduce} task is stored in a final output 
+        file. Once all \textit{reduce} tasks have completed, the master
+        resumes the program and the output is returned to the user.
+
+        In order to ensure fault tolerance, the master regularly pings 
+        the worker nodes. Should a worker node not respond after a certain
+        delay, the master labels the worker as failed. The \textit{map} 
+        tasks that were completed by the worker are all returned to their
+        original state (\textit{idle}) as \textit{map} task data is stored
+        locally. Both the completed and in-progress worker's \textit{map}
+        tasks are subsequently reassigned to other workers. Only the failed
+        worker's \textit{in-progress} \textit{reduce} tasks need to be 
+        reassigned.
+
+        The master node undergoes periodic checkpoints, such that the 
+        master node can be reinitialized from its last checkpoint in 
+        case of failure. Master node failure is, however, unlikely.
+
+        As network bandwidth is a scarce resource, the master attempts 
+        to schedule \textit{map} tasks closest to where the data is 
+        located. it will first attempt to schedule a map task on the host
+        that contains the data; should that option not be available, the
+        master will attempt to schedule the \textit{map} task on a host 
+        nearest to the data (e.g. same network switch)
+
+        Ideally, the number of \textit{map} and \textit{reduce} tasks 
+        should be much larger than the number of available worker nodes. 
+        Having it as such improves both dynamic load balancing and recovery 
+        time. The number of \textit{map} tasks used in practice typically 
+        corresponds to the size of the input chunks, whereas the number of 
+        \textit{reduce} tasks are a small multiple of the number of worker 
+        nodes used.
+
+        When nearing completion of the MapReduce program, it is possible for
+        ``stragglers" (\textit{in-progress} tasks running on nodes with below
+        average performance) to delay completion. As a measure to counteract 
+        this, the master node assigns the same task to idle workers. The task 
+        is then marked as completed as soon as either the primary or the 
+        backups complete the task. This strategy has been found to 
+        significantly improve performance when used with very large datasets. 
+        \note{44\% speedup with their sort program}.
+
+        A few extensions have also been added to the library in order to 
+        improve user experience. An important extension is the introduction 
+        of a \textit{combiner}, which enables the commutative and associative 
+        reduce function to be applied to the same keys on the hosts where the 
+        \textit{map} task was executed. Like the \textit{map} task, the 
+        \textit{combiner} produces intermediate output which is sent to the 
+        \textit{reducer}. This alleviates the amount of key-value pairs that 
+        must be transferred over the network from the map host to the reduce 
+        host, and as a result, improves performance. Other extensions include 
+        the ability to create user-defined partitioning functions, a guaranteed 
+        increasing key sort order, the ability for the user to specify input 
+        types, an option to skip records that continuously result in execution 
+        failure, the ability to produce auxiliary output files in \textit{map} 
+        or \textit{reduce} task, live status updates stream to HTTP server,
+        the ability to run all code on a single local machine, and a 
+        \textit{Counter} class.
+    \section{spark} 
+        \begin{itemize} 
+            \item based on map-reduce 
+            \item in-memory processing 
+            \item lazy evaluation 
+            \item general framework 
+        \end{itemize} 
+     \section{Use of containers with Big Data frameworks}
+     \section{Hadoop Distributed file system} 
+        \note{maybe should go in between mapreduce and spark}
+
+        The Hadoop Distributed File System (HDFS)\cite{hadoop} 
+        is the filesystem component of Hadoop. The metadata in 
+        HDFS is stored in on a dedicated server known as the 
+        NameNode, whereas application data is distributed across 
+        many servers referred to as DataNode. To ensure fault 
+        tolerance, HDFS replicates the data across multiple 
+        DataNodes (default replication factor: 3). Not only does 
+        this ensure that the data is not lost in the event of node
+        failure, but it also increases data transfer bandwidth, as
+        the data can be accessed from multiple nodes, and thus, 
+        there are more opportunities for computations to be 
+        performed nearest to where the data is located.
+
+        The NameNode contains the namespace tree, a hierarchy of 
+        directories and their files, as well as the mapping of the 
+        split file blocks to DataNodes. The entire namespace is 
+        stored in memory. Information on the directories and files,
+        such as modification and access times, namespace and disk 
+        quotes, are stored within \textit{inodes} on the NameNode. 
+        The name system's metadata, \textit{inode} and file block 
+        mapping, is known as the \textit{image}. Persistent record 
+        of the image that is stored on the local file system are 
+        known as \textit{checkpoint}. Locations of block replicas 
+        are not stored in the checkpoint as they may change over 
+        time. The \textit{journal} is a log of the modifications 
+        made to the image. Both the checkpoint and the journal may 
+        be copied across servers for increased durability. The 
+        journal is played back during the NameNode restart in order 
+        to restore the cluster.
+
+        When a client requests to read data, it must first contact 
+        the NameNode. The NameNode provides it with the replica 
+        locations, and the client reads from the DataNode located 
+        closest to it. When the client requests to write a file, it
+        contacts the NameNode which selects the DataNodes that will 
+        host the replicas. The client subsequently writes the data 
+        directly to the DataNodes in a pipeline fashion (The data 
+        gets propagated through each DataNode by being transferred
+        from the nearest DataNode to it).
+
+        There is only one NameNode assigned to each cluster, however, 
+        each cluster can have multiple clients and execute multiple 
+        tasks concurrently.
+
+        A block replica stored on a DataNode consists of two files 
+        stored on the local file system: the metadata and the data.
+
+        The DataNode connects to the NameNode during startup and 
+        performs handshake to for DataNode namespace ID and 
+        software version verification. Should either namespace ID 
+        or software version not match those of the NameNode, the 
+        DataNode shuts down.
+
+        To ensure integrity of the system, a \textit{namespace ID} 
+        is assigned to all nodes during formatting of the namespace. 
+        DataNodes with different namespace IDs to the NameNode will 
+        not be permitted to join the cluster. Should the DataNode
+        have not been assigned a namespace ID, it will be permitted 
+        to join the cluster. The cluster's namespace ID will then be 
+        assigned to that DataNode.
+
+        Following the handshake, the DataNode register with the 
+        NameNode using their storage ID. The storage ID is a unique 
+        ID generated after initial registration to the NameNode and 
+        that is persistently stored on the DataNode. It permits
+        identification of the DataNode in the event of an IP address 
+        or port change.
+
+\chapter{Scientific workflows in Spark}
+    \section{Introduction} 
+    \section{ariel's paper and thunders astronomy - }
+    \section{Spark on hpc's or } 
+        \note{large dataset for valduriez} Prior to
 examining how Big Data frameworks could be improved on HPC clusters, it is
@@
 layer. The lineage layer is the layer that performs recomputations to obtain the
-desired output. The persistance layer contains all the code used to regenerate
+desired output. The persistence layer contains all the code used to regenerate
 the data, in addition to any data that cannot be recovered through lineage

### 2018-02-17 paper.tex
psom pipeline implementation

@@
         \subsection{summary} 
+        While Python is a popular programming language amongst
+neuroinformaticians, it is not the only one that is commonly used. To address
+the needs of the Matlab/Octave community, PSOM (Pipeline System for Octave and
+Matlab) was developed. Similarly to Nipype, PSOM is easy-to-use, can run in
+distributed environments and can interface with tools written in other
+languages.
+
+        \note{pipeline description}
+        A pipeline in PSOM, is a series of connected jobs, which can be
+represented as a directed acyclic graph. The pipeline itself is implemented 
+using the Matlab/Octave data type. Each job within the pipeline
+is assigned a unique name. There is only one mandatory field in the job
+description (command). The remaining four optional fields include the list input
+files (files_in), the list output files (files_out), the list of files to be
+deleted after execution (files_clean) and any Octave or Matlab datatype
+containing any remaining optional parameters necessary for job execution (opt).
+Jobs in PSOM run in a protected environment where they only have access to the
+parameters that are passed.
+
+        In order to determine whether the opt field was correctly set, the
+generic function \textit{psom_struct_defaults} may be applied. This function
+will set the default values, ensure all mandatory fields were supplied and issue
+warnings for all unidentified attributes.
+
+        To define modules in PSOM, a special Octave/Matlab function type, known as "bricks"
+is used. Bricks employ the same parameters as jobs but lack the file_clean
+parameter. All bricks contain an opt.flag_test boolean flag, which, if set to
+true, allows the user to perform a dry-run of the brick. A dry-run of a brick
+consists of updating the default parameters and validating that the brick can
+successfully execute with the parameters provided. Bricks also have an option to
+specify the output folder (opt.output_folder) for brick outputs to be stored.
+
+        The pipeline generator is an easy way to create complete pipelines with
+minimal information provided. A pipeline generator may accept at most the
+file_in and opt arguments. It then builds the pipeline based on this
+information and the default parameters specified in each brick applied by the
+pipeline generator.
+
+        \note{pipeline execution}
+
+        \note{debugging and provenance}
+
         \subsection{limitations of PSOM}
@@
         \subsection{use of containers in nipype, psom, spm}
+
+       \section{other domains}
+       	\subsection{pegasus}
 

### 2018-02-17 paper.tex
minor sentence fix

@@
 deleted after execution (files_clean) and any Octave or Matlab datatype
-containing any remaining optional parameters necessary for job execution (opt).
+containing the remaining parameters necessary for job execution (opt).
 Jobs in PSOM run in a protected environment where they only have access to the

### 2018-02-17 paper.tex
fixed underscore bug

@@
 description (command). The remaining four optional fields include the list input
-files (files_in), the list output files (files_out), the list of files to be
-deleted after execution (files_clean) and any Octave or Matlab datatype
+files (files\_in), the list output files (files\_out), the list of files to be
+deleted after execution (files\_clean) and any Octave or Matlab datatype
 containing the remaining parameters necessary for job execution (opt).
@@
         In order to determine whether the opt field was correctly set, the
-generic function \textit{psom_struct_defaults} may be applied. This function
+generic function \textit{psom\_struct\_defaults} may be applied. This function
 will set the default values, ensure all mandatory fields were supplied and issue
@@
         To define modules in PSOM, a special Octave/Matlab function type, known as "bricks"
-is used. Bricks employ the same parameters as jobs but lack the file_clean
-parameter. All bricks contain an opt.flag_test boolean flag, which, if set to
+is used. Bricks employ the same parameters as jobs but lack the file\_clean
+parameter. All bricks contain an opt.flag\_test boolean flag, which, if set to
 true, allows the user to perform a dry-run of the brick. A dry-run of a brick
@@
 successfully execute with the parameters provided. Bricks also have an option to
-specify the output folder (opt.output_folder) for brick outputs to be stored.
+specify the output folder (opt.output\_folder) for brick outputs to be stored.
 
@@
 minimal information provided. A pipeline generator may accept at most the
-file_in and opt arguments. It then builds the pipeline based on this
+file\_in and opt arguments. It then builds the pipeline based on this
 information and the default parameters specified in each brick applied by the
@@
         \note{pipeline execution}
+
+        To execute a pipeline, the user must provide the pipeline structure and
+the pipeline configuration options. The latter consists of the log
+directory and the pipepline execution modes (e.g., session, background, batch,
+qsub, msub).
 

### 2018-02-17 paper.tex
psom execution

@@
 the pipeline configuration options. The latter consists of the log
-directory and the pipepline execution modes (e.g., session, background, batch,
-qsub, msub).
+directory, which must be provided, and the pipepline execution modes (e.g., session, background, batch,
+qsub, msub), in which the default mode is \textit{background}. Pipeline
+execution is achieved through the execution of the \textit{psom\_run\_pipeline}
+command.
+
+        Pipeline execution is broken down into three distinct modules: 1)
+initialization stage, 2) pipeline manager and 3) job manager. The initialization
+stage performs some checks to ensure that the pipeline can execute successfully
+(e.g., verify that pipeline is a DAG, check that output files are generated only
+once, check that all input files not generated by the pipeline exist on disk,
+delete any already existing output files), as well as performance checks (e.g.
+do not recompute pipeline steps that have already been executed, create
+necessary output folders to avoid creating them for each individual job). The
+initialization manager can determine which jobs need to be executed by comparing
+the current pipeline jobs with the details of those previously executed stored
+in the logs folder. The jobs that have been executed can have the following
+states: 1) none, 2) finished and 3) failed. The initialization manager stage the jobs 
+with a \textit{none} or \textit{failed} status for recomputation. For jobs whose
+description has changed since the previous \textit{finished} execution, the
+initialization will also stage it for recomputation. Should the user want to
+reexecution a previously executed \textit{finished} job, the user may force
+reexecution by setting \textit{opt\_pipe.restart} parameter to include all jobs
+to be reexecuted.
+
+        The pipeline manager is responsible for monitoring the state of the
+pipeline and submitting jobs for execution. Its duties begin after the
+initialization manager has completed execution. Jobs can be submitted by the
+pipeline manager once all their dependencies have been satisfied. The
+dependencies of a job are designated as met if there are none or when its parent jobs within the
+dependency graph are tagged as \textit{finished}. In addition, a job is
+considered to be successfully completely when all the dependencies of its
+children jobs have been satisfied. Depending on the pipeline configuration,
+there may be a limit to how many jobs can run in parallel. Once a job has
+finished executing, an empty tag file is created with a filename containing the
+final state in the logs folder. The pipeline manager continuously parses through
+the logs folder in search of these state files to delete and update the history
+file with the job state.
+
+        Every time a job is submitted by the pipeline manager for execution, a
+job manager is started. It is a Matlab function responsible for creating the job
+profile and generating the job-specific logs and the job exit tags. When
+the job terminates, the job manager will verify that all output has been
+correctly generated and assigns a job tagged based on this verification that
+will be reported back to the pipeline manager.
 

### 2018-02-17 paper.tex
psom debugging and provenance

@@
 the logs folder in search of these state files to delete and update the history
-file with the job state.
+file with the job state. Both the pipeline manager and the jobs can be executed
+within the current session, locally in a independent session, or in an
+independent session on a cluster
 
@@
 
+        The information stored in the logs folder is complete with details
+allowing for both debugging and provenance. The logs folder contains six files:
+1) PIPE\_history.txt, 2) PIPE\_jobs.mat, 3) PIPE\_status.mat, 4) PIPE\_logs.mat,
+5) PIPE\_profile.mat and 6) PIPE.mat. PIPE\_history.txt includes the execution
+history of the pipeline manager. PIPE\_jobs.mat and PIPE\_profile.mat are both
+Matlab/Octave files containing each executed job as a Matlab variable. They
+differ in that PIPE\_jobs.mat contains the latest executed jobs and can be used
+to reexecute the pipeline at any moment in time and PIPE\_profile.mat contains
+the execution time of each job \note{i think both can be used to reexecute
+pipeline from scratch. Also unclear whether PIPE\_profile.mat contains more than
+just latest jobs}. It is also possible to access any of the previously 
+used parameters through PIPE\_jobs.mat. PIPE\_status.mat stores the status of each 
+job as a string variable, whereas PIPE\_log.mat stores each job's log as a string 
+variable. Job logs, as generated by the log manager, consist of a text copy of the 
+execution, in addition to the user name, date, time and operating system used to 
+execute the job. Finally, all configuration details are stored in PIPE.mat.
+
         \subsection{limitations of PSOM}
@@
         \subsection{limitations of SPM}
+    \section{Nexflow}
+        \subsection{summary}
+    \section{OpenMOLE}
+        \subsection{summary} 
     \section{Use of containers in Workflow engines} 

### 2018-02-18 paper.tex
a not great summary of SPM

@@
         A pipeline in PSOM, is a series of connected jobs, which can be
-represented as a directed acyclic graph. The pipeline itself is implemented 
+represented as a (static?) directed acyclic graph. The pipeline itself is implemented 
 using the Matlab/Octave data type. Each job within the pipeline
@@
         \subsection{limitations of PSOM}
-    \section{SPM} 
+    \section{SPM and matlabbatch} 
+        \note{This section may not make much sense. Have not really yet fully
+understood matlabbatch}
         \subsection{summary} 
+        SPM is a very popular neuroimaging toolkit, and as a result,
+neuroimaging pipeline engines such as Nipype provide interfaces for it. SPM is
+also equipped with its own neuroimaging pipeline engine known as
+\textit{matlabbatch}. Similarly to PSOM and as its name suggest, it is written
+in Matlab.
+
+        \note{pipeline description}
+        Pipelines in matlabbatch consist of a series of connected data
+processing steps known as "modules" \note{DAG?? probably but not explicitly
+listed anywhere}. All SPM modules, helper modules and their respective
+dependencies can be defined using SPM's GUI \note{can probably also be done
+through command-line}.
+
+        For any matlabbatch job, it is possible to create a skeleton of it. A
+skeleton is a Matlab script containing a commented list of necessary inputs at
+the top and all code necessary to initialize and run the script. Before running,
+the skeleton will need to be configured with the input names and other
+necessary parameters.
+
+        All configuration options can be set/changed by modifying the
+\textit{cfg\_<type>} files.
+
+        Defaults can be set within the configuration files, through a default
+file, or in a \textit{.def} field for leaf nodes \note{terminal jobs???}. The
+latter updates at the start of the new job execution whereas the former are
+configured at the beginning of the SPM/matlabbatch startup.
+
+        Virtual outputs represent the dependencies between jobs \note{unsure}.
+They contain three paramaters of which only one is optional: sname, src\_output,
+tgt\_spec. sname is the display name that will be stored in the dependecies
+list, whereas src\_output is a subscript reference used to refer to the output
+that is produced at runtime. The optional parameter, tgt\_spec, is a description
+of the kind of inputs the outputs produced can be a dependency for. The virtual
+output callbacks are evaluated when cfg\_(ex)branch, cfg\_choice and cfg\_repeat 
+have the correct number of child nodes within all their in-tree nodes.
+
+        \note{pipeline execution}
+        Pipeline execution can be achieved either through the provided GUI or by
+the command-line interface. The GUI, however, requires significant user
+interaction, as the user must repeat the "Runs" command as many times as
+necessary to reexecute the command on multiple data. The CLI, however, avoids
+this with the help of its \textit{spm\_jobman} interface.
+
+        The first step to running a batch pipeline is to perform the
+initialization. This may include configuring the \textit{spm\_defaults.m} code
+to run in command line and executing the \textit{spm(’defaults’, MODALITY)}
+(where MODALITY represents the modality used in the pipeline) and
+\textit{spm\_jobman(’initcfg’)}.
+
+        After the intialization is complete, a job may be executed using the
+\textit{spm\_jobman(’run’, job[, input1, input2 ...])} command. This command
+requires the job names as well as its necessary inputs specified in the correct
+order. The job can be specified as a variable, the name of a script that will
+generate a job variable or a cell list containing a mixture of scripts and job
+variables.
+
+        \note{debugging and provenance}
+        All batch scripts can be saved as a Matlab scripts. These scripts not
+only contain the code that was executed, but also all input parameters. Should
+any script fail in the process of executing, an error will be printed to the
+Matlab console. If the user is using the GUI, the error output may also be saved
+to a Matlab file.
+
         \subsection{limitations of SPM}

### 2018-02-18 paper.tex
fixed special char bug
Summary of Nextflow

@@
         All configuration options can be set/changed by modifying the
-\textit{cfg\_<type>} files.
+\textit{cfg\_$\langle type \rangle$ } files.
 
@@
         \subsection{limitations of SPM}
-    \section{Nexflow}
+    \section{Nextflow}
         \subsection{summary}
+        Nextflow is a POSIX-compatible pipeline engine for scientific workflows.
+It extends the Groovy programming language to provide an  intuitive domain-specific language (DSL) 
+that enables users to easily describe their workflows. Unlike pipeline engines such as Nipype and
+PSOM, Netflow does not employ DAG. It instead adheres to the
+\textit{Dataflow} programming model which decides how data will flow through the pipeline at
+runtime \note{fix this sentence}. Nextflow prides itself on allowing the
+integration of containerized tools to address the issue of numerical stability
+that may arise from different environment setups.
+
+        \note{Pipeline description}
+        A Nextflow pipeline is represented by the flow of data between processes. These
+processes can be written in any Linux-compatible scripting language. As in Nipype, 
+these processes are executed in an isolated environment. The only means
+of communication between processes is through asynchronous FIFO queues known as
+\textit{channels}. Each process can contain any number of input and output
+channels. Through the use of channels, input and output can be stored not only
+on disk but also in-memory \note{Nipype can also return outputs that are not on
+disk, so this is nothing special}.
+
+        As with other pipeline engines (nipype, psom, etc.,), the order in which
+the processes appear is irrelevant. Execution order is determined by the dependencies.
+
+        \note{Pipeline execution}
+        The \textit{executor} is responsible for the execution of a Nextflow
+pipeline. By default, it assumes that the pipeline is to be run locally. The
+local executor will use multithreading in order to exploit available cores
+within the local environment. To configure the executor to run in a remote distributed environment, it is
+necessary to modify the configuration file to specify the environment. By
+following this approach, Nextflow users can seamlessly run their code on various
+processing platforms with a simple modification to a configuration file. In
+addition to local, Nextflow currently has executors for SGE, LSF, SLURM,
+PBS/Torque, NQSII, HTCondor, DRMAA, Ignite, Kubernetes and AWS Batch.
+
+        \note{Debugging and Provenance}
+        When executing a pipeline, it is always possible to specify the
+generation of a execution report. The report is written in HTML and consists of three sections:
+Summary, Resources and Tasks. Summary provides details on the execution command,
+execution status, overall execution time, in addtion to other metadata. The
+Resources tab, on the other hand, contains several HighChart plots representing
+CPU, memory, time and disk read/write rate of each pipeline process. The final
+section, Tasks, contains many metrics including the status and command script of
+each task.
+
+    Other reports that can be generated at runtime include the Trace report and
+the Timeline report. The Trace report include details such as CPU usage, start
+time, submission time, etc, of each process that was executed, whereas, the
+Timeline report generates an HTML timeline of all processes that have been
+executed.
+
+    \note{doesn't seem to be anything on provenance...}
+
+    \section{FastR}
+        \subsection{summary}
+        \note{Pipeline description}
+        \note{Pipeline execution}
+        \note{Debugging and Provenance}
     \section{OpenMOLE}
         \subsection{summary}
+        \note{Pipeline description}
+        \note{Pipeline execution}
+        \note{Debugging and Provenance}
+    \section{LONI}
+    \section{Automatic Analysis}
+    \section{Dipy, Make, Pydiper} 
     \section{Use of containers in Workflow engines} 

### 2018-02-23 paper.tex
Start of FastR

@@
 	\section{Definition of workflows/pipelines}
+
+    Broad definition of activities - between the tool and the engine.
 	\section{Nipype}
@@
 
-		A pipeline in Nipype, consists of a data analysis
+		A pipeline in Nipype, consists of a data analysis. explicit
                 \textit{Workflow} of connected
@@
 
-        \note{pipeline description}
+        \subsection{pipeline description}
         A pipeline in PSOM, is a series of connected jobs, which can be
@@
 
-        \note{pipeline execution}
+        \subsection{pipeline execution}
 
@@
 
-        \note{debugging and provenance}
+        \subsection{debugging and provenance}
 
@@
 
-        \note{pipeline description}
+        \subsection{pipeline description}
         Pipelines in matlabbatch consist of a series of connected data
@@
 
-        \note{pipeline execution}
+        \subsection{pipeline execution}
         Pipeline execution can be achieved either through the provided GUI or by
@@
 
-        \note{debugging and provenance}
+        \subsection{debugging and provenance}
         All batch scripts can be saved as a Matlab scripts. These scripts not
@@
 
-        \note{Pipeline description}
+        \subsection{Pipeline description}
         A Nextflow pipeline is represented by the flow of data between processes. These
@@
 
-        \note{Pipeline execution}
+        \subsection{Pipeline execution}
         The \textit{executor} is responsible for the execution of a Nextflow
@@
 
-        \note{Debugging and Provenance}
+        \subsection{Debugging and Provenance}
         When executing a pipeline, it is always possible to specify the
@@
         \subsection{summary}
-        \note{Pipeline description}
-        \note{Pipeline execution}
-        \note{Debugging and Provenance}
+        FastR is another Python-based neuroimaging pipeline framework that
+provides added features such including access to different versions of the same
+command-line tool \note{criticizes nipype for not having such a feature, but
+nipype can access boutiques tools...so...} and compatibility with various
+possible filesystems. In addition to its added features, FastR also guarantees
+a simple-to-use framework for building pipelines written in any languages, has
+built-in provenance support, provides the seamless ability to parallelize
+pipelines with the use of plugins and avoids the numerical instability issue with
+the simple integration of containerized pipelines. Similary to many other
+pipeline framework, it also employs a DAG to describe data flow.
+
+        \subsection{Pipeline description}
+        A pipeline in FastR is a \textit{Network} of connected \textit{Nodes} in
+which the directed edges represent the flow of data.  Each \textit{node}
+contains a JSON or XML template known as a \textit{Tool}.
+
+\textit{Tools}
+describe the details of the underlying application that is to be executed at the
+node. Tools consist of three components: general metadata, target and interface.
+The general metadata provides details on the Tools such as name, version,
+authors, etc, whereas the target specifies details on the execution environment.
+The interface, on the other hand, provides details on the inputs and outputs of
+the tool.
+
+A JSON schema is required to validate the python object generated by the Tool. 
+Fastr currently provides two schemas for object validation: the \textit{Tool} schema
+and the \textit{FastrInterface} schema. It is, however, possible to add a custom
+schema to FastR and define an \textit{Interface} for it.
+
+The \textit{Interface} class is responsible for executing a call to the command line tool.
+It dictates the tags that must be defined within the interface tag of the
+\textit{Tool}. For instance, the generic interface class requires details on the
+input and output's id, cardinality (number of input values) and DataType (a
+python object denoting the details of the Input or Output's data type).
+Verification is done by FastR to determine if the linked Input and Output data
+types are compatible \note{data types can also be grouped, but did not really
+understand that}.
+
+In FastR, there are four types of \textit{Nodes}: normal Nodes, Sink Nodes,
+Source Nodes and Constant Nodes. Data enters the network through Source or
+Constant Nodes and exit through Sink Nodes. Constant nodes differ from Source
+Nodes in that their input data are defined by the network \note{Not sure if
+saying this correctly}. Normal nodes are responsible for processing the
+data. Nodes are connected to each other through links that highlight the
+dependencies and form the \textit{Network}. Once the \textit{Netwokr} is
+defined, it can be executed.
+
+        \subsection{Pipeline execution}
+        \subsection{Debugging and Provenance}
     \section{OpenMOLE}
         \subsection{summary}
-        \note{Pipeline description}
-        \note{Pipeline execution}
-        \note{Debugging and Provenance}
+        \subsection{Pipeline description}
+        \subsection{Pipeline execution}
+        \subsection{Debugging and Provenance}
     \section{LONI}
@@
     \section{Introduction}
+    emphasis on data rather than on interface. transformation describe how
+interfaces interact with the data. 
     \section{ariel's paper and thunders astronomy - }

### 2018-02-24 paper.tex
FastR pipeline description

@@
 \chapter{Introduction}
-intro here
+intro here -mention hpc, data centre, cloud. 
 \chapter{BigData and Open science in neuroimaging}
@@
 runtime \note{fix this sentence}. Nextflow prides itself on allowing the
-integration of containerized tools to address the issue of numerical stability
+integration of containerized tools to address the issue of numerical instability
 that may arise from different environment setups.
@@
 python object denoting the details of the Input or Output's data type).
+DataTypes can also be grouped, allowing an inputs containing various files types
+to be inputted to applications which can support the file types.
 Verification is done by FastR to determine if the linked Input and Output data
-types are compatible \note{data types can also be grouped, but did not really
-understand that}.
+types are compatible.
+
+Input samples can consist of one or more values, in which the output is also a
+sample consisting of the same number of values. During the instances in which it
+is desirable to apply the same task to multiple inputs, a multidimensional array of samples, known as a collection, can be supplied to a
+\textit{Node} for processing. If there are multiple input samples, they can be
+passed to a node in pairwise behaviour (e.g. 1-1) or cross-product behaviour
+(e.g. nxn) \textit{need to better explain this}.
+
+Output samples may need to be expanded (i.e. a single sample of values becomes separate
+samples) or collapsed (i.e. multiple samples become a single sample of values).
+FastR provides data link flow directives for such operations.
 
@@
 data. Nodes are connected to each other through links that highlight the
-dependencies and form the \textit{Network}. Once the \textit{Netwokr} is
+dependencies and form the \textit{Network}. Once the \textit{Network} is
 defined, it can be executed.
 
+The data type of the Source Node and Sink Node must be defined in the
+\textit{Network} definition.
+The data is define at runtime by evaluation of the uniform resource identifier
+(URI). Multiple URIs can be defined through the use of wildcards and searches. Plugins are used to read and write data to
+different filesystems specified in the URI scheme. Currently, FastR supports
+virtual file system (VSF), csv files and XNAT. Plugins to support other
+filesystems can easily be added to FastR. In addition, plugins that can only
+read, write, or perform searches can also be created.
+
         \subsection{Pipeline execution}
+
         \subsection{Debugging and Provenance}

### 2018-02-24 paper.tex
FastR pipeline execution

@@
         \subsection{Pipeline description}
+
         A pipeline in FastR is a \textit{Network} of connected \textit{Nodes} in
@@
         \subsection{Pipeline execution}
+        To commence pipeline execution, the \textit{Network}'s execution
+function is called on the Submission host. FastR then analyses the network and
+its dependencies and breaks
+down the Network into chunks that can be further processed. From the chunks, the
+execution order of the \textit{Nodes} is determined.
+
+During \textit{Node} execution, a job is created for each input/output
+combination. Jobs are then executed on the execution host (i.e. local, cluster,
+cloud), as determined by the execution engine plugin. FastR currently only
+offers support for local and cluster plugins, however can easily be extended to
+support other plugins.
+
+While the job is executing, input URIs are translated to their actual values,
+the execution environment is set up by the \textit{Tool} and the \textit{Tool}
+command is executed by the \textit{Interface}. When the \textit{Interface}
+returns the results, the
+output data is validated and translated to URI format.
+
+After job completion, the results are returned to the Submission Host which
+reads the output and updates the network accordingly. Should a chunk be
+completed, the next chunk will be processed with the updated information, until
+all chunks have been processed and network execution is complete.
 

### 2018-02-24 paper.tex
FastR provenance

@@
 
-While the job is executing, input URIs are translated to their actual values,
+While the job is executing, input URIs are translated to their actual path/values,
 the execution environment is set up by the \textit{Tool} and the \textit{Tool}
@@
         \subsection{Debugging and Provenance}
+        In FastR, a provenance document is generated for each resulting file
+generated in the Prov-N format. Data recorded in the provenance documents include \textit{Tool}
+versions, input/output paths/values and their respective checksums, start and
+end time of execution for each \textit{job}, stdout and stderr logs, the end
+status and the complete environment description.
+
+    Similarly to Nipype, the FastR network can be visualized using graphviz.
+
     \section{OpenMOLE}

### 2018-02-25 paper.tex
openMOLE summary

@@
         \subsection{summary}
+        OpenMOLE (Open MOdeL Experiment) is a domain agnostic pipelining
+framework that focuses on simplifying workflow creation through its Domain
+Specific Language (DSL) built on top of Scala. Through the use of its DSL, much
+of the backend is abstracted from users. Similarly to other existing pipelining
+frameworks, OpenMOLE enables parallelization through plugins, meaning that the
+workflow need not be changed between different distribution environments.
+Moreover, OpenMOLE supports the deployment of tasks or groups of tasks on different
+computing environments.
+Another key feature of OpenMOLE is that it attempts to address the
+reproducibility issue through the integration of the CARE archiver. CARE allows
+for the execution of command-line applications, written in any language, and
+attempts to simulate the environment in which the application was designed on.
+OpenMOLE also provides embedded methods for parameter exploration.
+
         \subsection{Pipeline description}
+       An OpenMOLE workflow consists of tasks connected by transitions. Each
+task contains details such as input, output and optional parameters, in addition
+to embedding the application that will be applied to the data. Input values are
+generally supplied by the dataflow through an entry point known as
+\textit{Sources} and output values are returned to the
+dataflow through \textit{Hooks}.
+
+      Variables in OpenMOLE can be native types (e.g. int, string), files,
+directories, any Java or Scala types, and user-defined types (i.e. Java or Scala
+classes). The output type of an input supplied to another task should be the
+same as the input type, or be in a child class of the input type.
+
+      There exist five types of transitions in OpenMOLE: flat, divergent,
+convergent, explorative and aggregative. A flat transition is one in which the output of one
+task is the input to a single other task. On the other hand, a divergent
+transition is one in which the output of one task serves as the input to several
+other tasks. The inverse of a divergent task is a convergent task, in which
+several tasks supply input to a single task. Explorative transitions allow for
+multiple excutions of the same task on the sampling outputs produced by an
+exploration task. Conversly, an aggregative transition gathers all the streams
+produced by an explorative transition. It is also possible to add a stopping
+condition to terminate all streams once a condition is satisfied.
+
+      Tasks are not directly connected to each other by transitions. They are
+encapsulated within \textit{capsules} who are connected to each other by
+transitions. This permits the same task to be used multiple times in the
+workflow by encapsulating them in different capsules.
+
+    There also exists the notion of \textit{Input Slots} in OpenMOLE, which enables
+cycles in the workflow. A task may have several input slots and will execute
+each time that all the transitions belonging to the same input slots has passed
+through.
+
         \subsection{Pipeline execution}
+
+      In order for an OpenMOLE workflow to be executed, there must exist a
+single encapsulated task. A task will execute once all its dependencies have
+been met.
+
+    During execution of a task, the CARETask will decompress the CARE archive
+containing the application and its dependencies and the application is deployed
+within the confines of the decompressed archive.
+
+OpenMOLE relies on the exit code returned by the application in order to
+determine whether to reschedule the task. It is possible to prevent task
+rescheduling by setting the boolean flag \textit{errorOnReturnValue} to false.
+
+      OpenMOLE workflows can be executed on a variety of platforms such as
+remote servers, clusters and compute grids. As these execution hosts may not have all
+the required dependencies to execute a task, tasks are archived using CARE. This
+ensures that all the required applications and their dependencies can be shipped
+and executed at runtime on any Linux host.
+
         \subsection{Debugging and Provenance}
+    A CARETasks standard output and standard error are normally printed out to
+OpenMOLES's console. For distributed executions, this might not be a practical
+solution, in which case such information may be passed through
+the dataflow using OpenMOLE variables.
+
     \section{LONI}

### 2018-02-25 paper.tex
start summary of LONI

@@
 Specific Language (DSL) built on top of Scala. Through the use of its DSL, much
-of the backend is abstracted from users. Similarly to other existing pipelining
+of the backend is abstracted from users. It also offers a GUI for users to build
+workflows. Similarly to other existing pipelining
 frameworks, OpenMOLE enables parallelization through plugins, meaning that the
@@
     \section{LONI}
+        \subsection{Summary}
+        The LONI pipeline processing environment facilitates the development,
+sharing and execution of pipelines. Workflow Activities, known as \textit{Modules} in
+LONI, are described using Extensible Markup Language (XML), which can easily
+shared between user groups. LONI runs on the Java platform, allowing it to be
+platform independent and has support DRMAA clusters and symmetric
+multiprocessing (SMP) systems. In order to
+capture provenance information, LONI is also equipped with a provenance manager.
+
+        \subsection{Pipeline description}
+        LONI uses the dataflow programming model to represent the pipeline. Data
+flows between \textit{Modules} through \textit{data pipes}. Sub-pipelines are
+known and \textit{Pipelets}. Modules are defined by the input arguments/streams
+and output arguments/stream. Arguments in LONI may be specific
+filetypes, flags or values listed in the order in which the application
+requires. The provided arguments may be required or optional and can also be
+duplicated, if necessary. List of inputs may also be provided to process
+multiple inputs with copies of the same module in parallel (batch processing). Conversely, it is
+also possible to group list of inputs so as to process them all by the single
+module \note{Need to express this better}. Input/Output arguments may be added
+to the table of global variables for easy access throughout the pipeline. It is
+possible to apply variable modifiers to the variables stored within the table.   
+        \subsection{Pipeline execution}
+        \subsection{Debugging and provenance}
     \section{Automatic Analysis}
-    \section{Dipy, Make, Pydiper} 
+        \subsection{Summary}
+        \subsection{Pipeline description}
+        \subsection{Pipeline execution}
+        \subsection{Debugging and provenance}
+    \section{Dipy}
+        \subsection{Summary}
+        \subsection{Pipeline description}
+        \subsection{Pipeline execution}
+        \subsection{Debugging and provenance}
+    \section{Make}
+        \subsection{Summary}
+        \subsection{Pipeline description}
+        \subsection{Pipeline execution}
+        \subsection{Debugging and provenance}
+    \section{Pydiper}
+        \subsection{Summary}
+        \subsection{Pipeline description}
+        \subsection{Pipeline execution}
+        \subsection{Debugging and provenance} 
     \section{Use of containers in Workflow engines} 

### 2018-02-26 paper.tex
loni summary

@@
 requires. The provided arguments may be required or optional and can also be
-duplicated, if necessary. List of inputs may also be provided to process
+duplicated, if necessary. File inputs in the form of URLs are accepted by LONI
+and will be downloaded at runtime. List of inputs may also be provided to process
 multiple inputs with copies of the same module in parallel (batch processing). Conversely, it is
@@
 possible to apply variable modifiers to the variables stored within the table.
+
+Intermediary outputs as well as the file output are treated as temporary by
+LONI. They are scheduled for deletion before a new pipeline is executed. It is,
+however, possible to preserve these files.
+
         \subsection{Pipeline execution}
+        As a
+module completes it execution, its output gets transmitted through the dataflow
+and the dependent modules are signaled that their input data has arrived.
+Modules who have met all their dependency requirements are ready for execution.
+Should a module have no dependecies, it can be executed immediately. Modules
+ready to be executed are placed in a FIFO launch queue to avoid straining the
+platform.
+
+       LONI offers a built in client/server strategy that uses Java's remote
+method invocation (RMI) libraries to communicate between servers. This enables
+modules, pipelets or pipelines to be executed on servers different from the
+client. It also enables full pipelines to consist of connected modules and
+pipelets located on different servers. It is favorable to have such a
+configuration in LONI, as the tools required for processing may not be available
+on client server, but located in another. The client is responsible for
+dictating all file transfers and module launches that are to occur. The client
+authenticates with the servers through a one-way hash. Only files produced by
+the pipeline are accessible to the client and only temporary files created by
+the server may be written to the server.
+
         \subsection{Debugging and provenance}
+        For each dataset processed in a LONI pipeline, there exists a corresponding
+provenance (.prov) XML file. The \textit{Provenance Editor}, a standalone
+platform-independent application, reads the metadata from the image headers,
+extracts relevant provenance information from them and creates the provenance files. 
+This data can be modified by the user in the to correct inaccuracies or append
+additional details before being saved. During pipeline execution, the provenance
+file will be edited by the LONI pipeline to include details on each processing
+step, the environment in which the data was processed and tool binaries used. 
     \section{Automatic Analysis}

### 2018-02-27 paper.tex
start to automatic analysis

@@
 step, the environment in which the data was processed and tool binaries used.
+
     \section{Automatic Analysis}
         \subsection{Summary}
+        Automatic analysis (aa) is another Matlab-based modular workflow engine,
+having started of as a pipelining system for SPM. User may define their pipeline
+through a Matlab user script and specify the tasks in addition to their input
+and output types through an XML module interface. Modules used in pipelines must
+be integrated into aa which can be easily accomplished through XML interfaces. It is also possible to easily
+parallelize the code in qsub, if required. 
         \subsection{Pipeline description}
+        aa pipelines have two main components: 1) the Matlab user script and 2)
+the XML tasklist. The user script is a script that defines which data will be
+processed, any necessary parameter, and how it will be processed. aa provides an example user script,
+which can be modified to meet user's needs. The entire pipeline in aa is defined
+in a single structure, known as aap, which makes the task of record-keeping
+simple.
+
+        The XML tasklist is an ordered list of all the modules that will be applied to
+the data, that is read by the user script. This list does not specify any
+dependencies as those dependendecies can be inferred by the pipeline through
+analysis of the module interfaces. Modules can be separated into two
+sections: initialization and main. Modules listed in the initialization section
+are to be executed for every task. Conversely, those modules liste in the main
+section are only executed once for every task, unless otherwise specified.
+
+        In order to ensure that the correct output data is being operated on,
+each task is provided with its own unique folder. The aa engine will ensure that
+the necessary input data is located in the folder before task execution and also
+acquiring the outputs after task completion.
+
         \subsection{Pipeline execution}

### 2018-02-28 paper.tex
aa summary

@@
         Automatic analysis (aa) is another Matlab-based modular workflow engine,
-having started of as a pipelining system for SPM. User may define their pipeline
+having started of as a workflow engine for SPM. User may define their pipeline
 through a Matlab user script and specify the tasks in addition to their input
 and output types through an XML module interface. Modules used in pipelines must
-be integrated into aa which can be easily accomplished through XML interfaces. It is also possible to easily
-parallelize the code in qsub, if required. 
+be integrated into aa which can be easily accomplished through XML interfaces.
+aa workflows can be executed locally, or in parallel through qsub or condor.
+
         \subsection{Pipeline description}
@@
         In order to ensure that the correct output data is being operated on,
-each task is provided with its own unique folder. The aa engine will ensure that
-the necessary input data is located in the folder before task execution and also
-acquiring the outputs after task completion.
+each task is provided with its own unique folder. The aa engine will transmit
+the necessary input data to the folder before task execution and also
+acquiring the outputs after task completion. Output folders are numbered such
+that if a given module is invoked multiple times, each of those folders are
+assigned an incremental ordering corresponding to the order in which they have been
+executed \note{sentence might not make sense}. Should all data be required for
+each task in the tasklist, all the data will be copied to each task folder.
+
+    Similarly to a pipeline, each module is comprised of an XML interface and a
+Matlab source file. A module's XML file consist of specifications on required input
+and output modalities, as well as the desired domain. One of the most important
+properties include the domain at which the module operates. While there exist mainy domains for
+modules and new ones can easily be created and integrated by the user, the most
+prominently used ones include: "study", "subject", "session". As their name suggests, a
+module with the domain "study" will only be executed once for each time it
+appears in the tasklist, whereas "subject" executes the module for each subject
+within the task and "session" executes once for each session within a subject.
+
+    In addition to module domains, another important module property is the
+input and output data stream types. Data stream types denote what input and
+output data type each module produces. Commonly used data stream
+types include "epi", "structural" and "dicom\_header".
+
+    Optional fields specified in a module include qsub, for estimates of
+requirements for parallel processing,
+and \textit{performanceoutput}, for garbage collection.
+
+   Analysis parameters can easily be configured in aa. Any variable defined by
+the \textit{aaprecipe} command may be modified. Parameters can also be defined
+for an instance of a module within the XML tasklist. Alternatively, it is also
+possible to create an XML file that inherits parameters from another XML file
+and overrides some or all of the parameters.
+
+   To set defaults and site-specific configurations, it is possible to either
+create a local config file or by creating another parameter default file which
+inherits the original one.
+
+    aa also supports the creation of branched tasklist. This allows a section of
+the pipeline steps to undergo different tasks or the same tasks in a different
+order in one pipeline rather than having to create separate pipelines.
 
         \subsection{Pipeline execution}
+        Once the definition of the user-script, XML tasklist and modules is
+complete, the pipeline is ready for execution. Dependencies in aa are
+established based on input and output data stream types. The aa engine attempts
+to match the last task executed with a corresponding data stream type to the
+current type. It is, however, possible to specify which module provides the
+input to the current module by using fully qualified stream references.
+
+        During execution, a map of all the modules that need to be executed and
+their respective dependencies is generated. Modules that have completed contain
+a flag file indicating their completion within their instance's root directory. Such
+modules will not be re-executed by the engine.
+Additionally, if it is required that a previous module is re-executed, the done flag is
+removed from all dependent modules.
+
+        Should part of the processing be performed on a different server it is
+possible to connect the pipelines together through an analysis script. The aa engine will be reponsible for
+matching the data streams between the servers in order to avoid the copying of
+data and modules.
+
         \subsection{Debugging and provenance}
-    \section{Dipy}
-        \subsection{Summary}
-        \subsection{Pipeline description}
-        \subsection{Pipeline execution}
-        \subsection{Debugging and provenance}
+        In order to retrace the steps of previous executions, aa keeps a
+detailed record of the events that occurred. These detailes are stored in a
+Matlab object \note{similarly to PSOM}, which permits reexcution of a previous
+run or extraction of specific parameter values.
+
+        Additionally, to verify the correctness of the results produced, aa
+implements tools for quality control and comes with a dedicated module for
+low-level quality control.
+
     \section{Make}

### 2018-03-09 paper.tex
make summary

@@
         \subsection{Summary}
+        As the creation of workflow activities requires a certain level of
+programming expertise, Make was proposed as a simpler alternative. Make is a
+tool used for the development of directed acyclic dependency graphs. It is
+available in all Linux and UNIX distributions, which relieves the users from
+worflow engine installation. Additionally, Make is capable of offering other
+features necessary to workflow engines, such as, reproducibility,
+parallelization, fault-tolerance and quality control. The motivation behind Make
+is essentially that the tools available for simple and efficient pipeline
+creation already exist and there is no need to 'reinvent the wheel'. In
+addition, the use of Make is not limited to the neuroimaging community, and thus
+support for it is widely available. Make has also been adapted for use in other
+programming languages such as SnakeMake for python
+
         \subsection{Pipeline description}
+        Make pipelines are described in makefiles, which consist of a set of
+rules which describe the flow of data. Rules defined in make can be define by three
+components: targets, dependecies and recipes. The target is the output
+(typically file) produced by execution of the recipe, whereas the dependecy
+would be the input file from which the target would be created from. The target
+is only modified if its last modified date is early than that of a depencies's.
+A recipe, in make, is the set of shell commands used to create the target from
+its dependencies.
+
+In order connect the rules together, it is possible to define "phony" targets.
+Phony targets do not correspond to any actual file and will be generated once
+all of their dependencies (targets of other rules) have been
+created. Chaining multiple rules together will automatically generate the directed acyclic
+graph that is the pipeline. This is contrary to other pipelines such as Nipype
+and LONI, in which the user must explicitly state all connections \note{as
+stated in paper}. Through the use of rules, it is easily possible to reexecute
+only the desired parts of the pipeline.
+
+Cleanup after rule execution can be defined using the phony target
+clean.
+
+To create different targets for the same dependency files organized in different
+ways (e.g. subject/timpepoint timepoint/session), it is possible to use symbolic
+links for the dependencies, however this may complexify the code. Otherwise, it
+may be simpler to copy subject and session makefiles.
+
         \subsection{Pipeline execution}
+
+        Compared to the previous workflow engines, Make pipeline variables are
+only evaluated at time of use (lazy evaluation). In addition, targets are not
+executed in the order they are specified (similarly to other pipeline engines).
+
+Make files are also well-equipped to handle workflows with conditionals. Should
+a conditional exist in a workflow, it will be evaluated during the creation of
+the DAG (i.e. not at runtime) and only the required components will be executed.
+
+        As parallelism can be inferred from the dependency graph, parallel execution of a make pipeline can occur simply through the
+addition of the -j flag for multicore machines, or by using qmake command to run
+the pipeline on a cluster.
+
+        By default, make only rebuilds targets whose last modified preceedes
+that of its dependencies. As such, when a task fails and is re-executed, only the targets that need
+to be rebuilt will be rebuilt. It is possible to specify reexcution in Make such
+that the task will be reexecuted in case of failure.
+
         \subsection{Debugging and provenance}
+        As users of a preexisting Make pipeline may not be aware of the proper
+usage of each target, it is necessary to provide a `help` module that is
+adaptable to targets being added removed or modified. It is possible to define
+in Make a macro for such a purpose that will provide users with usage
+information when requested (e.g. calling make help).
+
+    Debugging in Make can be rather difficult, particularly when executing code
+in parallel. However, pipelines in Make are broken down into many targets which
+can each be tested individually. In addition, Make provides built-in flags to
+aid in debugging. These flags include -n, for printing out the commands without
+any execution and -p to print out all the rules. In the instance where pattern
+substistution is used, -p will also include the results obtained from the
+pattern substitution. More recent versions of GNU Make also include the --trace
+and --output-sync flags which display why each recipe was executed and print the
+parallelized output in sequential order, respectively.
+
+    Neuroimaging tools are typically equipped with quality assurance modules. As
+a result, it is possible to utilize the quality assurance modules from these
+tools to provide a complete report for the entire pipeline. The proposed
+solution suggests to write all the quality assurance outputs from each tool to a single HTML
+report with R Markdown or a script. The R Markdown or script, with the metrics
+and images are listed as dependencies to the QA report target in Make.
+
+    Only a basic form of data provenance has been implemented by the authors.
+The implemented data provenance consisted of a makefile that calls a bash to
+extract acquisition parameters from the raw input images. The makefile also
+obtains tool versions from the installation directories and pipeline options
+from their respective makefiles. All of these details are stored in a CSV file.
+The makefile will then call an R Markdown script which will then generate an
+HTML provenance report from the corresponding CSV file.
+
     \section{Pydiper}
         \subsection{Summary}
+        atoms -  wrappers for distinct operations
+        modules - link atoms together
+        applications - command line interface to the pipeline
         \subsection{Pipeline description}
@@
         \end{itemize}
-     \section{Use of containers with Big Data frameworks}
-     \section{Hadoop Distributed file system} 
-        \note{maybe should go in between mapreduce and spark}
-
-        The Hadoop Distributed File System (HDFS)\cite{hadoop} 
-        is the filesystem component of Hadoop. The metadata in 
-        HDFS is stored in on a dedicated server known as the 
-        NameNode, whereas application data is distributed across 
-        many servers referred to as DataNode. To ensure fault 
-        tolerance, HDFS replicates the data across multiple 
-        DataNodes (default replication factor: 3). Not only does 
-        this ensure that the data is not lost in the event of node
-        failure, but it also increases data transfer bandwidth, as
-        the data can be accessed from multiple nodes, and thus, 
-        there are more opportunities for computations to be 
-        performed nearest to where the data is located.
-
-        The NameNode contains the namespace tree, a hierarchy of 
-        directories and their files, as well as the mapping of the 
-        split file blocks to DataNodes. The entire namespace is 
-        stored in memory. Information on the directories and files,
-        such as modification and access times, namespace and disk 
-        quotes, are stored within \textit{inodes} on the NameNode. 
-        The name system's metadata, \textit{inode} and file block 
-        mapping, is known as the \textit{image}. Persistent record 
-        of the image that is stored on the local file system are 
-        known as \textit{checkpoint}. Locations of block replicas 
-        are not stored in the checkpoint as they may change over 
-        time. The \textit{journal} is a log of the modifications 
-        made to the image. Both the checkpoint and the journal may 
-        be copied across servers for increased durability. The 
-        journal is played back during the NameNode restart in order 
-        to restore the cluster.
-
-        When a client requests to read data, it must first contact 
-        the NameNode. The NameNode provides it with the replica 
-        locations, and the client reads from the DataNode located 
-        closest to it. When the client requests to write a file, it
-        contacts the NameNode which selects the DataNodes that will 
-        host the replicas. The client subsequently writes the data 
-        directly to the DataNodes in a pipeline fashion (The data 
-        gets propagated through each DataNode by being transferred
-        from the nearest DataNode to it).
-
-        There is only one NameNode assigned to each cluster, however, 
-        each cluster can have multiple clients and execute multiple 
-        tasks concurrently.
-
-        A block replica stored on a DataNode consists of two files 
-        stored on the local file system: the metadata and the data.
-
-        The DataNode connects to the NameNode during startup and 
-        performs handshake to for DataNode namespace ID and 
-        software version verification. Should either namespace ID 
-        or software version not match those of the NameNode, the 
-        DataNode shuts down.
-
-        To ensure integrity of the system, a \textit{namespace ID} 
-        is assigned to all nodes during formatting of the namespace. 
-        DataNodes with different namespace IDs to the NameNode will 
-        not be permitted to join the cluster. Should the DataNode
-        have not been assigned a namespace ID, it will be permitted 
-        to join the cluster. The cluster's namespace ID will then be 
-        assigned to that DataNode.
-
-        Following the handshake, the DataNode register with the 
-        NameNode using their storage ID. The storage ID is a unique 
-        ID generated after initial registration to the NameNode and 
-        that is persistently stored on the DataNode. It permits
-        identification of the DataNode in the event of an IP address 
-        or port change.
-
-\chapter{Scientific workflows in Spark}
-    \section{Introduction}
+
+    \subsection{Scientific workflows in Spark}
+        \subsubsection{Introduction}
         emphasis on data rather than on interface. transformation describe how
     interfaces interact with the data. 
-    \section{ariel's paper and thunders astronomy - }
-    \section{Spark on hpc's or } 
+        \subsubsection{ariel's paper and thunders astronomy - }
+        \subsubsection{Spark on hpc's or } 
             \note{large dataset for valduriez} Prior to
@@
     scientific workflows in an HPC environment.
+     \section{Use of containers with Big Data frameworks}
+     \section{Hadoop Distributed file system} 
+        \note{maybe should go in between mapreduce and spark}
+
+        The Hadoop Distributed File System (HDFS)\cite{hadoop} 
+        is the filesystem component of Hadoop. The metadata in 
+        HDFS is stored in on a dedicated server known as the 
+        NameNode, whereas application data is distributed across 
+        many servers referred to as DataNode. To ensure fault 
+        tolerance, HDFS replicates the data across multiple 
+        DataNodes (default replication factor: 3). Not only does 
+        this ensure that the data is not lost in the event of node
+        failure, but it also increases data transfer bandwidth, as
+        the data can be accessed from multiple nodes, and thus, 
+        there are more opportunities for computations to be 
+        performed nearest to where the data is located.
+
+        The NameNode contains the namespace tree, a hierarchy of 
+        directories and their files, as well as the mapping of the 
+        split file blocks to DataNodes. The entire namespace is 
+        stored in memory. Information on the directories and files,
+        such as modification and access times, namespace and disk 
+        quotes, are stored within \textit{inodes} on the NameNode. 
+        The name system's metadata, \textit{inode} and file block 
+        mapping, is known as the \textit{image}. Persistent record 
+        of the image that is stored on the local file system are 
+        known as \textit{checkpoint}. Locations of block replicas 
+        are not stored in the checkpoint as they may change over 
+        time. The \textit{journal} is a log of the modifications 
+        made to the image. Both the checkpoint and the journal may 
+        be copied across servers for increased durability. The 
+        journal is played back during the NameNode restart in order 
+        to restore the cluster.
+
+        When a client requests to read data, it must first contact 
+        the NameNode. The NameNode provides it with the replica 
+        locations, and the client reads from the DataNode located 
+        closest to it. When the client requests to write a file, it
+        contacts the NameNode which selects the DataNodes that will 
+        host the replicas. The client subsequently writes the data 
+        directly to the DataNodes in a pipeline fashion (The data 
+        gets propagated through each DataNode by being transferred
+        from the nearest DataNode to it).
+
+        There is only one NameNode assigned to each cluster, however, 
+        each cluster can have multiple clients and execute multiple 
+        tasks concurrently.
+
+        A block replica stored on a DataNode consists of two files 
+        stored on the local file system: the metadata and the data.
+
+        The DataNode connects to the NameNode during startup and 
+        performs handshake to for DataNode namespace ID and 
+        software version verification. Should either namespace ID 
+        or software version not match those of the NameNode, the 
+        DataNode shuts down.
+
+        To ensure integrity of the system, a \textit{namespace ID} 
+        is assigned to all nodes during formatting of the namespace. 
+        DataNodes with different namespace IDs to the NameNode will 
+        not be permitted to join the cluster. Should the DataNode
+        have not been assigned a namespace ID, it will be permitted 
+        to join the cluster. The cluster's namespace ID will then be 
+        assigned to that DataNode.
+
+        Following the handshake, the DataNode register with the 
+        NameNode using their storage ID. The storage ID is a unique 
+        ID generated after initial registration to the NameNode and 
+        that is persistently stored on the DataNode. It permits
+        identification of the DataNode in the event of an IP address 
+        or port change.
 
@@
 YARN/MESOS} \subsection{Multi-level scheduling w/ BigData schedulers}
-\chapter{Current status of Big Data frameworks in neuroscience} \section{Ariel's
-paper} \section{Thunder} \chapter{Conclusion}
+\chapter{Conclusion}
 

### 2018-07-25 paper.tex
section on pydpiper

@@
         \subsection{Summary}
-        atoms -  wrappers for distinct operations
-        modules - link atoms together
-        applications - command line interface to the pipeline
+
+        Pydiper is a neuroimaging registration workflow engine implemented in
+Python. This differs from other workflow engines in that it is specific to the
+process of image registration, but can, however, be extended for use with any
+analysis process. Its main features include being easy to use, particularly for
+non-developpers, undertaking a modular approach, giving the ability to access
+output files at any stage, and eliminating duplicate stages. Pydiper also
+provides the added advantage of being equipped with four commonly used
+registration pipelines, such that unexperienced researchers can use such
+pipelines out-of-the-box.
+
+        \subsection{Pipeline description}
+        Pydiper workflows can be broken down into three components: atoms,
+modules and applications. An application is the command line application that
+encapsulates the pipeline which consists of modules and atoms. Atoms are wrappers for individual operations,
+typically frquently used MINC tools. Modules are reponsible for linking
+atoms together.
+
+        Pydpiper provides three main classes to use to develop pipelines. These
+include the \textit{PipelineStage}, \textit{CmdStage} and \textit{Pipeline} classes. The
+\textit{PipelineStage} is the base class wrapper for all pipeline tools.
+Pipeline tools can be written in Python or can be a command-line application.
+For the tools that are command-line applications, the \textit{CmdStage} class, a
+child class of \textit{PipelineStage}, must be used. Input to \textit{CmdStage}
+is in a adequately parsed array format containing the specific command-line
+tool and its arguments. Reusable subclasses of the \textit{CmdStage} class can be
+created, or \textit{CmdStage} can be called directly.
+
+        To combine pipelines, the \textit{AddPipeline} function of the \textit{Pipeline}
+        class may be used.
+
+        The base class common to all Pydpiper applications is \textit{AbstractApplication}. 
+Classes that inherit from \textit{AbstractApplication} are command-line tools, with their 
+options set by the \textit{AbstractApplication} class. The class 
+is also responsible for initializing or resuming the pipeline, setting up a logger, and 
+launching the pipeline daemon.
+
+        Pydpiper atoms inherit from the \textit{CmdStage} class in order to wrap the MINC Tools. 
+They require as input a MINC file, and may additionally also require a target MINC file as second argument. These arguments
+can be in the form of a string filename or a file handler, however, the same format must be used for the two.
+An atom may also accepts various optional arguments.
+Using the inputs provided, the atom will put together the command to be executed.
+
+        \subsection{Pipeline execution}
+
+        In Pydpiper, there are two main classes,\textit{Pipeline} and \textit{PipelineExecutor}, 
+that monitor and manage pipeline execution. The \textit{Pipeline} class's main responsibility
+is to infer dependencies between pipeline stages from their inputs and outputs. It also manages
+a queue of runnable stages (stages in which all their dependencies have been met), ensuring
+that any runnable stage is placed in the queue. It also keeps track of the status of the runnable
+stages (running, finished, or failed). Should an already existing stage be added to the pipeline, it
+is automatically skipped by the \textit{Pipeline} class.
+
+        As stages may be added to incomplete pipelines, Pydpiper provides the functionality of resuming
+pipelines. This is achieved through a pickling mechanism implemented by the \textit{Pipeline} class. The
+pickled components include the directed acyclic graph, the array of pipeline stages, the stage counter,
+the hashes of all stages and each of their output files, and the status of each stage.
+
+        The \textit{PipelineExecutor} class, on the other hand, is responsible for the efficient processing
+of the pipeline through parallelization. Pydpiper's architecture mimics that of client/server, in which the 
+server is the pipeline daemon started by the command-line executable and the executors undertaking the role of 
+the clients. The pipeline's executors each consist of a thread pool in which each
+thread is responsible for running a single pipeline stage from the queue of runnable stages. To
+obtain a stage for execution, the executor polls the server to obtain the next available stage. Should
+the executor have enough memory and processors to execute the stage, it will be executed by the thread
+and release memory and processors and repoll the server once stage has completed.
+Otherwise, the executors will sleep for certain duration and repoll the server again. 
+Should there be any failed stages, the pipeline will be shut down after all possible stages have finished 
+executing. If additional executors are required by the pipeline, they may be added at any time through
+the command-line. Details related to
+the number of threads, number of threads per executor and the amount of memory can be specified through the
+command-line options.
+
+        \subsection{Debugging and provenance}
+
+        As mentioned previously, Pydpiper uses a pickling mechanism to gather important pipeline details
+        during execution in order to allow for the pipeline to be resumed should any stages have failed.
+        In addition, to keep track pipeline stage interdependencies in addition to file versioning, Pydpiper
+        utilizes the helper class \textit{RegistrationPipeFH} and its parent \textit{RegistrationFHBase} to create
+        file handlers of the files.
+
+        \note{from github wiki}
+        For debugging, a Pydpiper pipeline creats its own log file and executors typically also each create 
+        their own log file that are stored in a \textit{logs/}
+        subdirectory. Verbosity of logging information can be managed through the setting of the PYRO\_LOGLEVEL
+        environment variable. In addition, individual stage stdout/stderr is also redirected to a log file.
+
+        To verify the status of an executing pipeline, the check\_pipeline\_status.py may be executed.
+
+\chapter{Other domains and general scientific worflow engines}
+    \section{Taverna}
+    Taverne is a scientific workflow engine designed to allow users to combine distributed web services, other service
+    types (e.g. BioMart queries or R-scripts) and local scripts into complex analysis pipelines. The main goal of
+    Taverna is to give a unified access to the plethora of services already existing on the web without requiring
+    the user to download the services locally or to have knowledge on all of the platforms and programming languages
+    the services are hosted under.
+
         \subsection{Pipeline description}
@@
         \subsection{Debugging and provenance}
-    \section{Use of containers in Workflow engines} 
-        \subsection{use of containers in nipype, psom, spm}
-
-       \section{other domains}
-       	\subsection{pegasus}
-
-\chapter{Domain nonspecific BigData frameworks} 
+    \section{Pegasus}
+        \subsection{Pipeline description}
+        \subsection{Pipeline execution}
+        \subsection{Debugging and provenance}
+    \section{Geneflow}
+        \subsection{Pipeline description}
+        \subsection{Pipeline execution}
+        \subsection{Debugging and provenance}
+    \section{Toil}
+        \subsection{Pipeline description}
+        \subsection{Pipeline execution}
+        \subsection{Debugging and provenance}
+    \section{Thunder}
+
+        Thunder is a Python-based modular set of tools for the analysis of imaging and timeseries data belonging
+        to various domains including neuroscience. Although it is not specifically a pipelining framework, 
+        it can run on top of the Apache
+        Spark framework enabling the Thunder pipeline to be extended and customized. By running on top of Spark,
+        Thunder leverages its features to improve performance on large scale studies. Thunder does not need to
+        be executed on top of a Spark cluster and can easily be executed, with the same script, an a local system
+        with no Spark installation.
+
+        \note{The following was extracted from the documenation}
+        \subsection{Pipeline description}
+        The API for Thunder, regardless of if backed by an Apache Spark cluster or not, remains the same. If not 
+        using an Spark Resilient Distributed Dataset (RDD), the data can be backed by a \textit{numpy} array or a
+        distributed \textit{bolt} array. Thunder is broken down into several packages, in which the core package
+        consists of tools for reading and writing data, in addition to defining common data structures. Users may
+        choose to install other Thunder packages for added functionality, or define their own using any Python
+        library.
+
+        Thunder contains two data structures: \textit{images} and \textit{series}. As implied by its name, the 
+        \textit{images} data type may represent a collection of images or volumes. 
+        The \textit{series} data type,
+        however, represents a one-dimensional collection that shares a common index. Both \textit{images} and 
+        \textit{series} can be loaded from a list, ndarray or binary file type, however \textit{series} can also 
+        be loaded from a text file and \textit{images} can be loaded from png and tiff files.
+
+        Methods that are common to both data types are, such as \textit{map} and \textit{reduce}, are found in the 
+        \textit{data} class which is a subclass of the
+        \textit{base} class. In order to convert between the \textit{images} and \textit{series} data types, the
+        intermediary data type \textit{blocks}, child of \textit{base} class,  is used. 
+        Using the \textit{blocks} class, it is also possible to convert the data type to an ndarray for use with other
+        Python modules.
+
+        \note{not really anything on execution, debugging and provenance in the paper nor in the documentation}
+        \subsection{Pipeline execution}
+        In order to use Spark within Thunder, a SparkContext must be started and passed to the Thunder API. Should
+        the user need to switch between local and Spark frequently, it is possible to use the \textit{station} package
+        as a context manager for the various contexts.
+
+        As Thunder is an extension to Spark, numpy and \textit{bolt}, much of the execution details are likely differed 
+        to the 
+        framework that it is being executed on.
+
+        \subsection{Debugging and provenance}
+        Thunder does not explicitly provide any details on debugging and provenance tracking. Similarly to execution,
+        it is expected that much of the debugging and provenance details are not obtained from Thunder itself, but
+        rather, from the frameworks it is being executed on. For instance, should Thunder be exected on a Spark cluster,
+        it is likely that much of the debugging and provenance information can be obtained from the Spark logs and RDD
+        lineage.
+
+\chapter{MapReduce-based BigData frameworks and related components} 
     \section{mapreduce}

### 2018-09-08 paper.tex

@@
             \item HCP
+            \item UK Biobank
             \item ADNI
@@
     \section{Taverna}
-    Taverne is a scientific workflow engine designed to allow users to combine distributed web services, other service
+    Taverna is an open-source bioinformatics workflow engine, designed to allow users to combine distributed web services, other service
     types (e.g. BioMart queries or R-scripts) and local scripts into complex analysis pipelines. The main goal of
@@
         \subsection{Pipeline description}
+        Workflows in Taverna are built by combining web services, local scripts and other services together. Web services compatible with
+        Taverna must comply to the Web Service Description Language (WSDL) using the Simple Object Access Protocol (SOAP protocol). 
+        These services can be added to the BioCatalogue or the Biodiveristy catalogue for sharing and reuse of the web services.
+
+        Workflows can be built in the Taverna workbench. Taverna, through the help of the BioCatalogue and the Biodiversity Catalogue is
+        able to provide the user with a variety of webservices to combine into workflows in a drag-and-drop fashion. Workflows may also be composed
+        of sub-workflows. As service inputs and outputs may not match between services, shims are provided.
+        Similarly to the services, the user-generated workflows can be uploaded the myExperiment repository to be shared with other user.
+
         \subsection{Pipeline execution}
+        The workflows can be executed remotely utilizing the Taverna Server. The Taverna Server is a an application that provides the full library of possible workflows
+        through webservices. It has an interface, known as the Taverna Player, that enables workflow execution to occur through a web-interface or through
+        third-party clients. The Taverna Server, however, is limited in that it can only execute workflows available on the Server. Taverna Lite is an alternative to
+        the Taverna Server which allows users to upload and execute workflows from myExperiment or elsewhere.
+
         \subsection{Debugging and provenance}
+         Provenance capture is achieved through the Taverna Provenance Suite. Taverna does not capture provenance of the workflow definition, but generates a new
+        UUID after each definition change. Workflow run provenance captures information on "individual processor iterations" and their respective inputs and outputs and
+        stores this information in a database. The results can then be exported as a workflow trace or as a PROV-O RDF graph using the Taverna-PROV Plugin. The model format used
+        for exported provenance uses both the W3C PROV model and Open Provenance Model format.
+
     \section{Pegasus}

### 2018-09-10 paper.tex
unfinished pegasus summary

@@
 		\subsection{NeuroData}
+        \subsection{Apache Airvata}
 \chapter{Workflow engines for neuroimaging data}
@@
     \section{Pegasus}
+        Pegasus is a general-purpose scientific workflow engine that specializes in mapping abstract descriptions of workflows to resources for efficient processing. 
+        Similarly to other workflows, a Pegasus workflow can be represented by a DAG. This DAG is described by the user in DAX format (Directed Acyclic graph in XML).
+        As the DAX abstract workflow is resource-independent, Pegasus can convert the abstract workflow into an execution workflow that is catered to the resources available.
+
         \subsection{Pipeline description}
+        Within the DAX, the nodes are the tasks and the execution order is represented by the edges.
+        In the instance of nested graphs, the nodes in the DAX may also represent a DAX. Pipeline inputs and outputs, as well as pipeline executables,
+         are represented by logical identifiers.
+
+         Various APIs are provided by Pegasus in order for the user to generate the
+        DAX using their preferred language of choice. Available APIs include Perl, Python and Java. 
         \subsection{Pipeline execution}
+        With the effecient execution of pipelines being Pegasus' main focus, there exist various components that ensure optimal processing. The Mapper's responsability is to map
+        abstract workflow tasks to available resources. In order for the Mapper to successfully convert an abstract workflow into an execution workflow, it requires details on the execution
+        environment. The Mapper will also query various catalogs to determine the location of executables and input and output files. These catalogs can be either file-based or in
+        database format. With all the necessary information provided, the Mapper is equipped to convert abstract tasks into executable tasks. Data nodes are also added for input and output
+        staging. Should the data required for a task already exist in the catalog, the task and all its parents will be removed by the Mapper from the workflow. This functionality is optional,
+        and in some instances, it may be found that recomputation is preferable. If the executables required by the task are not pre-installed on the node, the Mapper will query the 
+        \textit{Transformation Catalog} for the stageable binaries.
+
+        The default, Pegasus relies on the HTCondor framework as both the workflow engine and scheduler. The executable workflow, using HTCondor DAGMan, consists of nodes representing jobs.
+        These jobs each have a corresponding job submit file that details how they are meant to be executed. A job that is ready for execution is released by the DAG and placed into a Condor
+        queue that is managed by the HTCondor Schedd daemon. The default scheduling is done using a local HTCondor pool, but can also be achieved through built-in interfaces with remote 
+        job-management protocols, or through SSH to directly submit jobs to remote resource managers.
+
+        Using HTCondor, Pegasus is able to achieve portability. Pegasus is compatible with local, HPC, distributed and cloud execution environments. As distributed environments may
+        not employ a shared filesystem, it is recommended to use HTCondor glideins to ensure that tasks are mapped to nodes where the resources are located. Should a user running their
+        pipeline locally without HTCondor installed, it is possible through \textit{submit host}. Using this method, however, executes the pipeline in serial order. Should the user wish to exploit
+        the pipeline's parallelism, HTCondor is required. Pegasus pipelines are not limited to a single execution environment, and can combine multiple environments to select the best resource
+        configuration to execute each job.
+
+        In order to manage tasks and their dependencies on remote resources, three different execution engines have been developed. The first and simplest execution engine is the 
+        \textit{Pegasus Cluster}, which is 
+        compatible with all the execution environments. It runs on a single-thread, meaning that it can only executed a single task at a time using an ordered-list of tasks.
+        The \textit{Pegasus Cluster} is particularly useful for the efficient execution of short tasks and the reduction of data transfers.
+
+        For execution environments lacking a shared filesystem, the \textit{PegasusLite} execution engine can be used. In order for the Pegasus to execute in such environments, the Mapper first
+        adds data transfer nodes to the pipeline to move the data to a data staging server associated with the execution node. Once a job is assigned to a node by the remote cluster scheduler,
+        \textit{PegasusLite} is deployed.
+
+        The final execution engine provided by Pegasus is the \textit{Pegasus MPI Cluster} (PMC). This execution engine is designed for petascale systems for which Pegasus' other strategies 
+        leveraging Condor Glideins are unsuitable. It employs the Master-Worker architecture, launching a single Master and many Workers, and using MPI for communication between the 
+        Master and Workers. The main limitation of this strategy is that a task must use less resources than that available on a single node. Jobs are not limited to a single execution engine,
+        and thus, multiple can be used in a job to optimize performance.
+
+        As executing a workflow on remote resources can be slow due to job queuing times and data transfers, the Mapper will group short-running task together to reduce to improve runtime. 
+        When tasks located at the same distance from the DAG root have similar runtimes, level-based clustering can be used. When runtimes of same-level tasks may
+        vary, it may be preferable to use job runtime-based horizontal clustering, to achieve load-balancing of clusters. Alternatively, the user may opt for label-based clustering, in which
+        the user can define the task clusters themselves.
+
+        To avoid running out of disk space during data-intensive workflows, the Mapper can add cleanup nodes that remove all unecessary data after a job has completed.
+
+        \note{Dynamic workflows permitted}
+
         \subsection{Debugging and provenance}
+         Job monitoring, in Pegasus, is achieved through \textit{pegasus-kickstart} (kickstart). Every job within a pegasus workflow has a kickstart wrapper that collects runtime provenance
+        and performance. Kickstart relays this information to the submit host after job completion.
+
     \section{Geneflow}

### 2018-09-13 paper.tex
finished pegasus summary

@@
          Job monitoring, in Pegasus, is achieved through \textit{pegasus-kickstart} (kickstart). Every job within a pegasus workflow has a kickstart wrapper that collects runtime provenance
-        and performance. Kickstart relays this information to the submit host after job completion.
+        and performance. Kickstart relays this information to the submit host after job completion. This data is ultimately collected and stored in a relational database.
+
+        The Mapper Logs contain details on the abstract workflow, the jobs in the executable workflow 
+        and the mapping of tasks to jobs. Using the mapper logs, Pegasus can generate the provenance of the results. Should it be required, Pegasus can also generate the provenance of the 
+        mapping process.
+
+        In addition to the Mapper Logs, there also exists the DAGMan logs, generated by HTCondor DAGMan and updated in near real-time. These logs capture details on the status of the jobs
+        and their associated pre- and post- execute scripts.
+
+        All provenance and debugging data is accessibly through either the \textit{pegasus-analyzer} tool or through the provided lightweight web interface.
 

### 2018-09-18 paper.tex
some restructuring
start of toil

@@
         \subsection{limitations of SPM}
-    \section{Nextflow}
-        \subsection{summary}
-        Nextflow is a POSIX-compatible pipeline engine for scientific workflows.
-It extends the Groovy programming language to provide an  intuitive domain-specific language (DSL) 
-that enables users to easily describe their workflows. Unlike pipeline engines such as Nipype and
-PSOM, Netflow does not employ DAG. It instead adheres to the
-\textit{Dataflow} programming model which decides how data will flow through the pipeline at
-runtime \note{fix this sentence}. Nextflow prides itself on allowing the
-integration of containerized tools to address the issue of numerical instability
-that may arise from different environment setups.
-
-        \subsection{Pipeline description}
-        A Nextflow pipeline is represented by the flow of data between processes. These
-processes can be written in any Linux-compatible scripting language. As in Nipype, 
-these processes are executed in an isolated environment. The only means
-of communication between processes is through asynchronous FIFO queues known as
-\textit{channels}. Each process can contain any number of input and output
-channels. Through the use of channels, input and output can be stored not only
-on disk but also in-memory \note{Nipype can also return outputs that are not on
-disk, so this is nothing special}.
-
-        As with other pipeline engines (nipype, psom, etc.,), the order in which
-the processes appear is irrelevant. Execution order is determined by the dependencies.
-
-        \subsection{Pipeline execution}
-        The \textit{executor} is responsible for the execution of a Nextflow
-pipeline. By default, it assumes that the pipeline is to be run locally. The
-local executor will use multithreading in order to exploit available cores
-within the local environment. To configure the executor to run in a remote distributed environment, it is
-necessary to modify the configuration file to specify the environment. By
-following this approach, Nextflow users can seamlessly run their code on various
-processing platforms with a simple modification to a configuration file. In
-addition to local, Nextflow currently has executors for SGE, LSF, SLURM,
-PBS/Torque, NQSII, HTCondor, DRMAA, Ignite, Kubernetes and AWS Batch.
-
-        \subsection{Debugging and Provenance}
-        When executing a pipeline, it is always possible to specify the
-generation of a execution report. The report is written in HTML and consists of three sections:
-Summary, Resources and Tasks. Summary provides details on the execution command,
-execution status, overall execution time, in addtion to other metadata. The
-Resources tab, on the other hand, contains several HighChart plots representing
-CPU, memory, time and disk read/write rate of each pipeline process. The final
-section, Tasks, contains many metrics including the status and command script of
-each task.
-
-    Other reports that can be generated at runtime include the Trace report and
-the Timeline report. The Trace report include details such as CPU usage, start
-time, submission time, etc, of each process that was executed, whereas, the
-Timeline report generates an HTML timeline of all processes that have been
-executed.
-
-    \note{doesn't seem to be anything on provenance...}
 
@@
     Similarly to Nipype, the FastR network can be visualized using graphviz.
+
+    \section{LONI}
+        \subsection{Summary}
+        The LONI pipeline processing environment facilitates the development,
+sharing and execution of pipelines. Workflow Activities, known as \textit{Modules} in
+LONI, are described using Extensible Markup Language (XML), which can easily
+shared between user groups. LONI runs on the Java platform, allowing it to be
+platform independent and has support DRMAA clusters and symmetric
+multiprocessing (SMP) systems. In order to
+capture provenance information, LONI is also equipped with a provenance manager.
+
+        \subsection{Pipeline description}
+        LONI uses the dataflow programming model to represent the pipeline. Data
+flows between \textit{Modules} through \textit{data pipes}. Sub-pipelines are
+known and \textit{Pipelets}. Modules are defined by the input arguments/streams
+and output arguments/stream. Arguments in LONI may be specific
+filetypes, flags or values listed in the order in which the application
+requires. The provided arguments may be required or optional and can also be
+duplicated, if necessary. File inputs in the form of URLs are accepted by LONI
+and will be downloaded at runtime. List of inputs may also be provided to process
+multiple inputs with copies of the same module in parallel (batch processing). Conversely, it is
+also possible to group list of inputs so as to process them all by the single
+module \note{Need to express this better}. Input/Output arguments may be added
+to the table of global variables for easy access throughout the pipeline. It is
+possible to apply variable modifiers to the variables stored within the table.
+
+Intermediary outputs as well as the file output are treated as temporary by
+LONI. They are scheduled for deletion before a new pipeline is executed. It is,
+however, possible to preserve these files.
+
+        \subsection{Pipeline execution}
+        As a
+module completes it execution, its output gets transmitted through the dataflow
+and the dependent modules are signaled that their input data has arrived.
+Modules who have met all their dependency requirements are ready for execution.
+Should a module have no dependecies, it can be executed immediately. Modules
+ready to be executed are placed in a FIFO launch queue to avoid straining the
+platform.
+
+       LONI offers a built in client/server strategy that uses Java's remote
+method invocation (RMI) libraries to communicate between servers. This enables
+modules, pipelets or pipelines to be executed on servers different from the
+client. It also enables full pipelines to consist of connected modules and
+pipelets located on different servers. It is favorable to have such a
+configuration in LONI, as the tools required for processing may not be available
+on client server, but located in another. The client is responsible for
+dictating all file transfers and module launches that are to occur. The client
+authenticates with the servers through a one-way hash. Only files produced by
+the pipeline are accessible to the client and only temporary files created by
+the server may be written to the server.
+
+        \subsection{Debugging and provenance}
+        For each dataset processed in a LONI pipeline, there exists a corresponding
+provenance (.prov) XML file. The \textit{Provenance Editor}, a standalone
+platform-independent application, reads the metadata from the image headers,
+extracts relevant provenance information from them and creates the provenance files. 
+This data can be modified by the user in the to correct inaccuracies or append
+additional details before being saved. During pipeline execution, the provenance
+file will be edited by the LONI pipeline to include details on each processing
+step, the environment in which the data was processed and tool binaries used.
+
+    \section{Automatic Analysis}
+        \subsection{Summary}
+        Automatic analysis (aa) is another Matlab-based modular workflow engine,
+having started of as a workflow engine for SPM. User may define their pipeline
+through a Matlab user script and specify the tasks in addition to their input
+and output types through an XML module interface. Modules used in pipelines must
+be integrated into aa which can be easily accomplished through XML interfaces.
+aa workflows can be executed locally, or in parallel through qsub or condor.
+
+        \subsection{Pipeline description}
+        aa pipelines have two main components: 1) the Matlab user script and 2)
+the XML tasklist. The user script is a script that defines which data will be
+processed, any necessary parameter, and how it will be processed. aa provides an example user script,
+which can be modified to meet user's needs. The entire pipeline in aa is defined
+in a single structure, known as aap, which makes the task of record-keeping
+simple.
+
+        The XML tasklist is an ordered list of all the modules that will be applied to
+the data, that is read by the user script. This list does not specify any
+dependencies as those dependendecies can be inferred by the pipeline through
+analysis of the module interfaces. Modules can be separated into two
+sections: initialization and main. Modules listed in the initialization section
+are to be executed for every task. Conversely, those modules liste in the main
+section are only executed once for every task, unless otherwise specified.
+
+        In order to ensure that the correct output data is being operated on,
+each task is provided with its own unique folder. The aa engine will transmit
+the necessary input data to the folder before task execution and also
+acquiring the outputs after task completion. Output folders are numbered such
+that if a given module is invoked multiple times, each of those folders are
+assigned an incremental ordering corresponding to the order in which they have been
+executed \note{sentence might not make sense}. Should all data be required for
+each task in the tasklist, all the data will be copied to each task folder.
+
+    Similarly to a pipeline, each module is comprised of an XML interface and a
+Matlab source file. A module's XML file consist of specifications on required input
+and output modalities, as well as the desired domain. One of the most important
+properties include the domain at which the module operates. While there exist mainy domains for
+modules and new ones can easily be created and integrated by the user, the most
+prominently used ones include: "study", "subject", "session". As their name suggests, a
+module with the domain "study" will only be executed once for each time it
+appears in the tasklist, whereas "subject" executes the module for each subject
+within the task and "session" executes once for each session within a subject.
+
+    In addition to module domains, another important module property is the
+input and output data stream types. Data stream types denote what input and
+output data type each module produces. Commonly used data stream
+types include "epi", "structural" and "dicom\_header".
+
+    Optional fields specified in a module include qsub, for estimates of
+requirements for parallel processing,
+and \textit{performanceoutput}, for garbage collection.
+
+   Analysis parameters can easily be configured in aa. Any variable defined by
+the \textit{aaprecipe} command may be modified. Parameters can also be defined
+for an instance of a module within the XML tasklist. Alternatively, it is also
+possible to create an XML file that inherits parameters from another XML file
+and overrides some or all of the parameters.
+
+   To set defaults and site-specific configurations, it is possible to either
+create a local config file or by creating another parameter default file which
+inherits the original one.
+
+    aa also supports the creation of branched tasklist. This allows a section of
+the pipeline steps to undergo different tasks or the same tasks in a different
+order in one pipeline rather than having to create separate pipelines.
+
+        \subsection{Pipeline execution}
+        Once the definition of the user-script, XML tasklist and modules is
+complete, the pipeline is ready for execution. Dependencies in aa are
+established based on input and output data stream types. The aa engine attempts
+to match the last task executed with a corresponding data stream type to the
+current type. It is, however, possible to specify which module provides the
+input to the current module by using fully qualified stream references.
+
+        During execution, a map of all the modules that need to be executed and
+their respective dependencies is generated. Modules that have completed contain
+a flag file indicating their completion within their instance's root directory. Such
+modules will not be re-executed by the engine.
+Additionally, if it is required that a previous module is re-executed, the done flag is
+removed from all dependent modules.
+
+        Should part of the processing be performed on a different server it is
+possible to connect the pipelines together through an analysis script. The aa engine will be reponsible for
+matching the data streams between the servers in order to avoid the copying of
+data and modules.
+
+        \subsection{Debugging and provenance}
+        In order to retrace the steps of previous executions, aa keeps a
+detailed record of the events that occurred. These detailes are stored in a
+Matlab object \note{similarly to PSOM}, which permits reexcution of a previous
+run or extraction of specific parameter values.
+
+        Additionally, to verify the correctness of the results produced, aa
+implements tools for quality control and comes with a dedicated module for
+low-level quality control.
+
+    \section{Pydiper}
+        \subsection{Summary}
+
+        Pydiper is a neuroimaging registration workflow engine implemented in
+Python. This differs from other workflow engines in that it is specific to the
+process of image registration, but can, however, be extended for use with any
+analysis process. Its main features include being easy to use, particularly for
+non-developpers, undertaking a modular approach, giving the ability to access
+output files at any stage, and eliminating duplicate stages. Pydiper also
+provides the added advantage of being equipped with four commonly used
+registration pipelines, such that unexperienced researchers can use such
+pipelines out-of-the-box.
+
+        \subsection{Pipeline description}
+        Pydiper workflows can be broken down into three components: atoms,
+modules and applications. An application is the command line application that
+encapsulates the pipeline which consists of modules and atoms. Atoms are wrappers for individual operations,
+typically frquently used MINC tools. Modules are reponsible for linking
+atoms together.
+
+        Pydpiper provides three main classes to use to develop pipelines. These
+include the \textit{PipelineStage}, \textit{CmdStage} and \textit{Pipeline} classes. The
+\textit{PipelineStage} is the base class wrapper for all pipeline tools.
+Pipeline tools can be written in Python or can be a command-line application.
+For the tools that are command-line applications, the \textit{CmdStage} class, a
+child class of \textit{PipelineStage}, must be used. Input to \textit{CmdStage}
+is in a adequately parsed array format containing the specific command-line
+tool and its arguments. Reusable subclasses of the \textit{CmdStage} class can be
+created, or \textit{CmdStage} can be called directly.
+
+        To combine pipelines, the \textit{AddPipeline} function of the \textit{Pipeline}
+        class may be used.
+
+        The base class common to all Pydpiper applications is \textit{AbstractApplication}. 
+Classes that inherit from \textit{AbstractApplication} are command-line tools, with their 
+options set by the \textit{AbstractApplication} class. The class 
+is also responsible for initializing or resuming the pipeline, setting up a logger, and 
+launching the pipeline daemon.
+
+        Pydpiper atoms inherit from the \textit{CmdStage} class in order to wrap the MINC Tools. 
+They require as input a MINC file, and may additionally also require a target MINC file as second argument. These arguments
+can be in the form of a string filename or a file handler, however, the same format must be used for the two.
+An atom may also accepts various optional arguments.
+Using the inputs provided, the atom will put together the command to be executed.
+
+        \subsection{Pipeline execution}
+
+        In Pydpiper, there are two main classes,\textit{Pipeline} and \textit{PipelineExecutor}, 
+that monitor and manage pipeline execution. The \textit{Pipeline} class's main responsibility
+is to infer dependencies between pipeline stages from their inputs and outputs. It also manages
+a queue of runnable stages (stages in which all their dependencies have been met), ensuring
+that any runnable stage is placed in the queue. It also keeps track of the status of the runnable
+stages (running, finished, or failed). Should an already existing stage be added to the pipeline, it
+is automatically skipped by the \textit{Pipeline} class.
+
+        As stages may be added to incomplete pipelines, Pydpiper provides the functionality of resuming
+pipelines. This is achieved through a pickling mechanism implemented by the \textit{Pipeline} class. The
+pickled components include the directed acyclic graph, the array of pipeline stages, the stage counter,
+the hashes of all stages and each of their output files, and the status of each stage.
+
+        The \textit{PipelineExecutor} class, on the other hand, is responsible for the efficient processing
+of the pipeline through parallelization. Pydpiper's architecture mimics that of client/server, in which the 
+server is the pipeline daemon started by the command-line executable and the executors undertaking the role of 
+the clients. The pipeline's executors each consist of a thread pool in which each
+thread is responsible for running a single pipeline stage from the queue of runnable stages. To
+obtain a stage for execution, the executor polls the server to obtain the next available stage. Should
+the executor have enough memory and processors to execute the stage, it will be executed by the thread
+and release memory and processors and repoll the server once stage has completed.
+Otherwise, the executors will sleep for certain duration and repoll the server again. 
+Should there be any failed stages, the pipeline will be shut down after all possible stages have finished 
+executing. If additional executors are required by the pipeline, they may be added at any time through
+the command-line. Details related to
+the number of threads, number of threads per executor and the amount of memory can be specified through the
+command-line options.
+
+        \subsection{Debugging and provenance}
+
+        As mentioned previously, Pydpiper uses a pickling mechanism to gather important pipeline details
+        during execution in order to allow for the pipeline to be resumed should any stages have failed.
+        In addition, to keep track pipeline stage interdependencies in addition to file versioning, Pydpiper
+        utilizes the helper class \textit{RegistrationPipeFH} and its parent \textit{RegistrationFHBase} to create
+        file handlers of the files.
+
+        \note{from github wiki}
+        For debugging, a Pydpiper pipeline creats its own log file and executors typically also each create 
+        their own log file that are stored in a \textit{logs/}
+        subdirectory. Verbosity of logging information can be managed through the setting of the PYRO\_LOGLEVEL
+        environment variable. In addition, individual stage stdout/stderr is also redirected to a log file.
+
+        To verify the status of an executing pipeline, the check\_pipeline\_status.py may be executed.
+
+\chapter{Other domains and general scientific worflow engines}
+    \section{Taverna}
+    Taverna is an open-source bioinformatics workflow engine, designed to allow users to combine distributed web services, other service
+    types (e.g. BioMart queries or R-scripts) and local scripts into complex analysis pipelines. The main goal of
+    Taverna is to give a unified access to the plethora of services already existing on the web without requiring
+    the user to download the services locally or to have knowledge on all of the platforms and programming languages
+    the services are hosted under.
+
+        \subsection{Pipeline description}
+        Workflows in Taverna are built by combining web services, local scripts and other services together. Web services compatible with
+        Taverna must comply to the Web Service Description Language (WSDL) using the Simple Object Access Protocol (SOAP protocol). 
+        These services can be added to the BioCatalogue or the Biodiveristy catalogue for sharing and reuse of the web services.
+
+        Workflows can be built in the Taverna workbench. Taverna, through the help of the BioCatalogue and the Biodiversity Catalogue is
+        able to provide the user with a variety of webservices to combine into workflows in a drag-and-drop fashion. Workflows may also be composed
+        of sub-workflows. As service inputs and outputs may not match between services, shims are provided.
+        Similarly to the services, the user-generated workflows can be uploaded the myExperiment repository to be shared with other user.
+
+        \subsection{Pipeline execution}
+        The workflows can be executed remotely utilizing the Taverna Server. The Taverna Server is a an application that provides the full library of possible workflows
+        through webservices. It has an interface, known as the Taverna Player, that enables workflow execution to occur through a web-interface or through
+        third-party clients. The Taverna Server, however, is limited in that it can only execute workflows available on the Server. Taverna Lite is an alternative to
+        the Taverna Server which allows users to upload and execute workflows from myExperiment or elsewhere.
+
+        \subsection{Debugging and provenance}
+         Provenance capture is achieved through the Taverna Provenance Suite. Taverna does not capture provenance of the workflow definition, but generates a new
+        UUID after each definition change. Workflow run provenance captures information on "individual processor iterations" and their respective inputs and outputs and
+        stores this information in a database. The results can then be exported as a workflow trace or as a PROV-O RDF graph using the Taverna-PROV Plugin. The model format used
+        for exported provenance uses both the W3C PROV model and Open Provenance Model format.
+
+    \section{Pegasus}
+        Pegasus is a general-purpose scientific workflow engine that specializes in mapping abstract descriptions of workflows to resources for efficient processing. 
+        Similarly to other workflows, a Pegasus workflow can be represented by a DAG. This DAG is described by the user in DAX format (Directed Acyclic graph in XML).
+        As the DAX abstract workflow is resource-independent, Pegasus can convert the abstract workflow into an execution workflow that is catered to the resources available.
+
+        \subsection{Pipeline description}
+        Within the DAX, the nodes are the tasks and the execution order is represented by the edges.
+        In the instance of nested graphs, the nodes in the DAX may also represent a DAX. Pipeline inputs and outputs, as well as pipeline executables,
+         are represented by logical identifiers.
+
+         Various APIs are provided by Pegasus in order for the user to generate the
+        DAX using their preferred language of choice. Available APIs include Perl, Python and Java. 
+        \subsection{Pipeline execution}
+        With the effecient execution of pipelines being Pegasus' main focus, there exist various components that ensure optimal processing. The Mapper's responsability is to map
+        abstract workflow tasks to available resources. In order for the Mapper to successfully convert an abstract workflow into an execution workflow, it requires details on the execution
+        environment. The Mapper will also query various catalogs to determine the location of executables and input and output files. These catalogs can be either file-based or in
+        database format. With all the necessary information provided, the Mapper is equipped to convert abstract tasks into executable tasks. Data nodes are also added for input and output
+        staging. Should the data required for a task already exist in the catalog, the task and all its parents will be removed by the Mapper from the workflow. This functionality is optional,
+        and in some instances, it may be found that recomputation is preferable. If the executables required by the task are not pre-installed on the node, the Mapper will query the 
+        \textit{Transformation Catalog} for the stageable binaries.
+
+        The default, Pegasus relies on the HTCondor framework as both the workflow engine and scheduler. The executable workflow, using HTCondor DAGMan, consists of nodes representing jobs.
+        These jobs each have a corresponding job submit file that details how they are meant to be executed. A job that is ready for execution is released by the DAG and placed into a Condor
+        queue that is managed by the HTCondor Schedd daemon. The default scheduling is done using a local HTCondor pool, but can also be achieved through built-in interfaces with remote 
+        job-management protocols, or through SSH to directly submit jobs to remote resource managers.
+
+        Using HTCondor, Pegasus is able to achieve portability. Pegasus is compatible with local, HPC, distributed and cloud execution environments. As distributed environments may
+        not employ a shared filesystem, it is recommended to use HTCondor glideins to ensure that tasks are mapped to nodes where the resources are located. Should a user running their
+        pipeline locally without HTCondor installed, it is possible through \textit{submit host}. Using this method, however, executes the pipeline in serial order. Should the user wish to exploit
+        the pipeline's parallelism, HTCondor is required. Pegasus pipelines are not limited to a single execution environment, and can combine multiple environments to select the best resource
+        configuration to execute each job.
+
+        In order to manage tasks and their dependencies on remote resources, three different execution engines have been developed. The first and simplest execution engine is the 
+        \textit{Pegasus Cluster}, which is 
+        compatible with all the execution environments. It runs on a single-thread, meaning that it can only executed a single task at a time using an ordered-list of tasks.
+        The \textit{Pegasus Cluster} is particularly useful for the efficient execution of short tasks and the reduction of data transfers.
+
+        For execution environments lacking a shared filesystem, the \textit{PegasusLite} execution engine can be used. In order for the Pegasus to execute in such environments, the Mapper first
+        adds data transfer nodes to the pipeline to move the data to a data staging server associated with the execution node. Once a job is assigned to a node by the remote cluster scheduler,
+        \textit{PegasusLite} is deployed.
+
+        The final execution engine provided by Pegasus is the \textit{Pegasus MPI Cluster} (PMC). This execution engine is designed for petascale systems for which Pegasus' other strategies 
+        leveraging Condor Glideins are unsuitable. It employs the Master-Worker architecture, launching a single Master and many Workers, and using MPI for communication between the 
+        Master and Workers. The main limitation of this strategy is that a task must use less resources than that available on a single node. Jobs are not limited to a single execution engine,
+        and thus, multiple can be used in a job to optimize performance.
+
+        As executing a workflow on remote resources can be slow due to job queuing times and data transfers, the Mapper will group short-running task together to reduce to improve runtime. 
+        When tasks located at the same distance from the DAG root have similar runtimes, level-based clustering can be used. When runtimes of same-level tasks may
+        vary, it may be preferable to use job runtime-based horizontal clustering, to achieve load-balancing of clusters. Alternatively, the user may opt for label-based clustering, in which
+        the user can define the task clusters themselves.
+
+        To avoid running out of disk space during data-intensive workflows, the Mapper can add cleanup nodes that remove all unecessary data after a job has completed.
+
+        \note{Dynamic workflows permitted}
+
+        \subsection{Debugging and provenance}
+         Job monitoring, in Pegasus, is achieved through \textit{pegasus-kickstart} (kickstart). Every job within a pegasus workflow has a kickstart wrapper that collects runtime provenance
+        and performance. Kickstart relays this information to the submit host after job completion. This data is ultimately collected and stored in a relational database.
+
+        The Mapper Logs contain details on the abstract workflow, the jobs in the executable workflow 
+        and the mapping of tasks to jobs. Using the mapper logs, Pegasus can generate the provenance of the results. Should it be required, Pegasus can also generate the provenance of the 
+        mapping process.
+
+        In addition to the Mapper Logs, there also exists the DAGMan logs, generated by HTCondor DAGMan and updated in near real-time. These logs capture details on the status of the jobs
+        and their associated pre- and post- execute scripts.
+
+        All provenance and debugging data is accessibly through either the \textit{pegasus-analyzer} tool or through the provided lightweight web interface.
+
+           \section{Make}
+        \subsection{Summary}
+        As the creation of workflow activities requires a certain level of
+programming expertise, Make was proposed as a simpler alternative. Make is a
+tool used for the development of directed acyclic dependency graphs. It is
+available in all Linux and UNIX distributions, which relieves the users from
+worflow engine installation. Additionally, Make is capable of offering other
+features necessary to workflow engines, such as, reproducibility,
+parallelization, fault-tolerance and quality control. The motivation behind Make
+is essentially that the tools available for simple and efficient pipeline
+creation already exist and there is no need to 'reinvent the wheel'. In
+addition, the use of Make is not limited to the neuroimaging community, and thus
+support for it is widely available. Make has also been adapted for use in other
+programming languages such as SnakeMake for python
+
+        \subsection{Pipeline description}
+        Make pipelines are described in makefiles, which consist of a set of
+rules which describe the flow of data. Rules defined in make can be define by three
+components: targets, dependecies and recipes. The target is the output
+(typically file) produced by execution of the recipe, whereas the dependecy
+would be the input file from which the target would be created from. The target
+is only modified if its last modified date is early than that of a depencies's.
+A recipe, in make, is the set of shell commands used to create the target from
+its dependencies.
+
+In order connect the rules together, it is possible to define "phony" targets.
+Phony targets do not correspond to any actual file and will be generated once
+all of their dependencies (targets of other rules) have been
+created. Chaining multiple rules together will automatically generate the directed acyclic
+graph that is the pipeline. This is contrary to other pipelines such as Nipype
+and LONI, in which the user must explicitly state all connections \note{as
+stated in paper}. Through the use of rules, it is easily possible to reexecute
+only the desired parts of the pipeline.
+
+Cleanup after rule execution can be defined using the phony target
+clean.
+
+To create different targets for the same dependency files organized in different
+ways (e.g. subject/timpepoint timepoint/session), it is possible to use symbolic
+links for the dependencies, however this may complexify the code. Otherwise, it
+may be simpler to copy subject and session makefiles.
+
+        \subsection{Pipeline execution}
+
+        Compared to the previous workflow engines, Make pipeline variables are
+only evaluated at time of use (lazy evaluation). In addition, targets are not
+executed in the order they are specified (similarly to other pipeline engines).
+
+Make files are also well-equipped to handle workflows with conditionals. Should
+a conditional exist in a workflow, it will be evaluated during the creation of
+the DAG (i.e. not at runtime) and only the required components will be executed.
+
+        As parallelism can be inferred from the dependency graph, parallel execution of a make pipeline can occur simply through the
+addition of the -j flag for multicore machines, or by using qmake command to run
+the pipeline on a cluster.
+
+        By default, make only rebuilds targets whose last modified preceedes
+that of its dependencies. As such, when a task fails and is re-executed, only the targets that need
+to be rebuilt will be rebuilt. It is possible to specify reexcution in Make such
+that the task will be reexecuted in case of failure.
+
+        \subsection{Debugging and provenance}
+        As users of a preexisting Make pipeline may not be aware of the proper
+usage of each target, it is necessary to provide a `help` module that is
+adaptable to targets being added removed or modified. It is possible to define
+in Make a macro for such a purpose that will provide users with usage
+information when requested (e.g. calling make help).
+
+    Debugging in Make can be rather difficult, particularly when executing code
+in parallel. However, pipelines in Make are broken down into many targets which
+can each be tested individually. In addition, Make provides built-in flags to
+aid in debugging. These flags include -n, for printing out the commands without
+any execution and -p to print out all the rules. In the instance where pattern
+substistution is used, -p will also include the results obtained from the
+pattern substitution. More recent versions of GNU Make also include the --trace
+and --output-sync flags which display why each recipe was executed and print the
+parallelized output in sequential order, respectively.
+
+    Neuroimaging tools are typically equipped with quality assurance modules. As
+a result, it is possible to utilize the quality assurance modules from these
+tools to provide a complete report for the entire pipeline. The proposed
+solution suggests to write all the quality assurance outputs from each tool to a single HTML
+report with R Markdown or a script. The R Markdown or script, with the metrics
+and images are listed as dependencies to the QA report target in Make.
+
+    Only a basic form of data provenance has been implemented by the authors.
+The implemented data provenance consisted of a makefile that calls a bash to
+extract acquisition parameters from the raw input images. The makefile also
+obtains tool versions from the installation directories and pipeline options
+from their respective makefiles. All of these details are stored in a CSV file.
+The makefile will then call an R Markdown script which will then generate an
+HTML provenance report from the corresponding CSV file.
 
@@
 
-    \section{LONI}
-        \subsection{Summary}
-        The LONI pipeline processing environment facilitates the development,
-sharing and execution of pipelines. Workflow Activities, known as \textit{Modules} in
-LONI, are described using Extensible Markup Language (XML), which can easily
-shared between user groups. LONI runs on the Java platform, allowing it to be
-platform independent and has support DRMAA clusters and symmetric
-multiprocessing (SMP) systems. In order to
-capture provenance information, LONI is also equipped with a provenance manager.
+        \section{Nextflow}
+        \subsection{summary}
+        Nextflow is a POSIX-compatible pipeline engine for scientific workflows.
+It extends the Groovy programming language to provide an  intuitive domain-specific language (DSL) 
+that enables users to easily describe their workflows. Unlike pipeline engines such as Nipype and
+PSOM, Netflow does not employ DAG. It instead adheres to the
+\textit{Dataflow} programming model which decides how data will flow through the pipeline at
+runtime \note{fix this sentence}. Nextflow prides itself on allowing the
+integration of containerized tools to address the issue of numerical instability
+that may arise from different environment setups.
 
         \subsection{Pipeline description}
-        LONI uses the dataflow programming model to represent the pipeline. Data
-flows between \textit{Modules} through \textit{data pipes}. Sub-pipelines are
-known and \textit{Pipelets}. Modules are defined by the input arguments/streams
-and output arguments/stream. Arguments in LONI may be specific
-filetypes, flags or values listed in the order in which the application
-requires. The provided arguments may be required or optional and can also be
-duplicated, if necessary. File inputs in the form of URLs are accepted by LONI
-and will be downloaded at runtime. List of inputs may also be provided to process
-multiple inputs with copies of the same module in parallel (batch processing). Conversely, it is
-also possible to group list of inputs so as to process them all by the single
-module \note{Need to express this better}. Input/Output arguments may be added
-to the table of global variables for easy access throughout the pipeline. It is
-possible to apply variable modifiers to the variables stored within the table.
-
-Intermediary outputs as well as the file output are treated as temporary by
-LONI. They are scheduled for deletion before a new pipeline is executed. It is,
-however, possible to preserve these files.
+        A Nextflow pipeline is represented by the flow of data between processes. These
+processes can be written in any Linux-compatible scripting language. As in Nipype, 
+these processes are executed in an isolated environment. The only means
+of communication between processes is through asynchronous FIFO queues known as
+\textit{channels}. Each process can contain any number of input and output
+channels. Through the use of channels, input and output can be stored not only
+on disk but also in-memory \note{Nipype can also return outputs that are not on
+disk, so this is nothing special}.
+
+        As with other pipeline engines (nipype, psom, etc.,), the order in which
+the processes appear is irrelevant. Execution order is determined by the dependencies.
 
         \subsection{Pipeline execution}
-        As a
-module completes it execution, its output gets transmitted through the dataflow
-and the dependent modules are signaled that their input data has arrived.
-Modules who have met all their dependency requirements are ready for execution.
-Should a module have no dependecies, it can be executed immediately. Modules
-ready to be executed are placed in a FIFO launch queue to avoid straining the
-platform.
-
-       LONI offers a built in client/server strategy that uses Java's remote
-method invocation (RMI) libraries to communicate between servers. This enables
-modules, pipelets or pipelines to be executed on servers different from the
-client. It also enables full pipelines to consist of connected modules and
-pipelets located on different servers. It is favorable to have such a
-configuration in LONI, as the tools required for processing may not be available
-on client server, but located in another. The client is responsible for
-dictating all file transfers and module launches that are to occur. The client
-authenticates with the servers through a one-way hash. Only files produced by
-the pipeline are accessible to the client and only temporary files created by
-the server may be written to the server.
-
-        \subsection{Debugging and provenance}
-        For each dataset processed in a LONI pipeline, there exists a corresponding
-provenance (.prov) XML file. The \textit{Provenance Editor}, a standalone
-platform-independent application, reads the metadata from the image headers,
-extracts relevant provenance information from them and creates the provenance files. 
-This data can be modified by the user in the to correct inaccuracies or append
-additional details before being saved. During pipeline execution, the provenance
-file will be edited by the LONI pipeline to include details on each processing
-step, the environment in which the data was processed and tool binaries used.
-
-    \section{Automatic Analysis}
-        \subsection{Summary}
-        Automatic analysis (aa) is another Matlab-based modular workflow engine,
-having started of as a workflow engine for SPM. User may define their pipeline
-through a Matlab user script and specify the tasks in addition to their input
-and output types through an XML module interface. Modules used in pipelines must
-be integrated into aa which can be easily accomplished through XML interfaces.
-aa workflows can be executed locally, or in parallel through qsub or condor.
-
-        \subsection{Pipeline description}
-        aa pipelines have two main components: 1) the Matlab user script and 2)
-the XML tasklist. The user script is a script that defines which data will be
-processed, any necessary parameter, and how it will be processed. aa provides an example user script,
-which can be modified to meet user's needs. The entire pipeline in aa is defined
-in a single structure, known as aap, which makes the task of record-keeping
-simple.
-
-        The XML tasklist is an ordered list of all the modules that will be applied to
-the data, that is read by the user script. This list does not specify any
-dependencies as those dependendecies can be inferred by the pipeline through
-analysis of the module interfaces. Modules can be separated into two
-sections: initialization and main. Modules listed in the initialization section
-are to be executed for every task. Conversely, those modules liste in the main
-section are only executed once for every task, unless otherwise specified.
-
-        In order to ensure that the correct output data is being operated on,
-each task is provided with its own unique folder. The aa engine will transmit
-the necessary input data to the folder before task execution and also
-acquiring the outputs after task completion. Output folders are numbered such
-that if a given module is invoked multiple times, each of those folders are
-assigned an incremental ordering corresponding to the order in which they have been
-executed \note{sentence might not make sense}. Should all data be required for
-each task in the tasklist, all the data will be copied to each task folder.
-
-    Similarly to a pipeline, each module is comprised of an XML interface and a
-Matlab source file. A module's XML file consist of specifications on required input
-and output modalities, as well as the desired domain. One of the most important
-properties include the domain at which the module operates. While there exist mainy domains for
-modules and new ones can easily be created and integrated by the user, the most
-prominently used ones include: "study", "subject", "session". As their name suggests, a
-module with the domain "study" will only be executed once for each time it
-appears in the tasklist, whereas "subject" executes the module for each subject
-within the task and "session" executes once for each session within a subject.
-
-    In addition to module domains, another important module property is the
-input and output data stream types. Data stream types denote what input and
-output data type each module produces. Commonly used data stream
-types include "epi", "structural" and "dicom\_header".
-
-    Optional fields specified in a module include qsub, for estimates of
-requirements for parallel processing,
-and \textit{performanceoutput}, for garbage collection.
-
-   Analysis parameters can easily be configured in aa. Any variable defined by
-the \textit{aaprecipe} command may be modified. Parameters can also be defined
-for an instance of a module within the XML tasklist. Alternatively, it is also
-possible to create an XML file that inherits parameters from another XML file
-and overrides some or all of the parameters.
-
-   To set defaults and site-specific configurations, it is possible to either
-create a local config file or by creating another parameter default file which
-inherits the original one.
-
-    aa also supports the creation of branched tasklist. This allows a section of
-the pipeline steps to undergo different tasks or the same tasks in a different
-order in one pipeline rather than having to create separate pipelines.
-
-        \subsection{Pipeline execution}
-        Once the definition of the user-script, XML tasklist and modules is
-complete, the pipeline is ready for execution. Dependencies in aa are
-established based on input and output data stream types. The aa engine attempts
-to match the last task executed with a corresponding data stream type to the
-current type. It is, however, possible to specify which module provides the
-input to the current module by using fully qualified stream references.
-
-        During execution, a map of all the modules that need to be executed and
-their respective dependencies is generated. Modules that have completed contain
-a flag file indicating their completion within their instance's root directory. Such
-modules will not be re-executed by the engine.
-Additionally, if it is required that a previous module is re-executed, the done flag is
-removed from all dependent modules.
-
-        Should part of the processing be performed on a different server it is
-possible to connect the pipelines together through an analysis script. The aa engine will be reponsible for
-matching the data streams between the servers in order to avoid the copying of
-data and modules.
-
-        \subsection{Debugging and provenance}
-        In order to retrace the steps of previous executions, aa keeps a
-detailed record of the events that occurred. These detailes are stored in a
-Matlab object \note{similarly to PSOM}, which permits reexcution of a previous
-run or extraction of specific parameter values.
-
-        Additionally, to verify the correctness of the results produced, aa
-implements tools for quality control and comes with a dedicated module for
-low-level quality control.
-
-    \section{Make}
-        \subsection{Summary}
-        As the creation of workflow activities requires a certain level of
-programming expertise, Make was proposed as a simpler alternative. Make is a
-tool used for the development of directed acyclic dependency graphs. It is
-available in all Linux and UNIX distributions, which relieves the users from
-worflow engine installation. Additionally, Make is capable of offering other
-features necessary to workflow engines, such as, reproducibility,
-parallelization, fault-tolerance and quality control. The motivation behind Make
-is essentially that the tools available for simple and efficient pipeline
-creation already exist and there is no need to 'reinvent the wheel'. In
-addition, the use of Make is not limited to the neuroimaging community, and thus
-support for it is widely available. Make has also been adapted for use in other
-programming languages such as SnakeMake for python
-
-        \subsection{Pipeline description}
-        Make pipelines are described in makefiles, which consist of a set of
-rules which describe the flow of data. Rules defined in make can be define by three
-components: targets, dependecies and recipes. The target is the output
-(typically file) produced by execution of the recipe, whereas the dependecy
-would be the input file from which the target would be created from. The target
-is only modified if its last modified date is early than that of a depencies's.
-A recipe, in make, is the set of shell commands used to create the target from
-its dependencies.
-
-In order connect the rules together, it is possible to define "phony" targets.
-Phony targets do not correspond to any actual file and will be generated once
-all of their dependencies (targets of other rules) have been
-created. Chaining multiple rules together will automatically generate the directed acyclic
-graph that is the pipeline. This is contrary to other pipelines such as Nipype
-and LONI, in which the user must explicitly state all connections \note{as
-stated in paper}. Through the use of rules, it is easily possible to reexecute
-only the desired parts of the pipeline.
-
-Cleanup after rule execution can be defined using the phony target
-clean.
-
-To create different targets for the same dependency files organized in different
-ways (e.g. subject/timpepoint timepoint/session), it is possible to use symbolic
-links for the dependencies, however this may complexify the code. Otherwise, it
-may be simpler to copy subject and session makefiles.
-
-        \subsection{Pipeline execution}
-
-        Compared to the previous workflow engines, Make pipeline variables are
-only evaluated at time of use (lazy evaluation). In addition, targets are not
-executed in the order they are specified (similarly to other pipeline engines).
-
-Make files are also well-equipped to handle workflows with conditionals. Should
-a conditional exist in a workflow, it will be evaluated during the creation of
-the DAG (i.e. not at runtime) and only the required components will be executed.
-
-        As parallelism can be inferred from the dependency graph, parallel execution of a make pipeline can occur simply through the
-addition of the -j flag for multicore machines, or by using qmake command to run
-the pipeline on a cluster.
-
-        By default, make only rebuilds targets whose last modified preceedes
-that of its dependencies. As such, when a task fails and is re-executed, only the targets that need
-to be rebuilt will be rebuilt. It is possible to specify reexcution in Make such
-that the task will be reexecuted in case of failure.
-
-        \subsection{Debugging and provenance}
-        As users of a preexisting Make pipeline may not be aware of the proper
-usage of each target, it is necessary to provide a `help` module that is
-adaptable to targets being added removed or modified. It is possible to define
-in Make a macro for such a purpose that will provide users with usage
-information when requested (e.g. calling make help).
-
-    Debugging in Make can be rather difficult, particularly when executing code
-in parallel. However, pipelines in Make are broken down into many targets which
-can each be tested individually. In addition, Make provides built-in flags to
-aid in debugging. These flags include -n, for printing out the commands without
-any execution and -p to print out all the rules. In the instance where pattern
-substistution is used, -p will also include the results obtained from the
-pattern substitution. More recent versions of GNU Make also include the --trace
-and --output-sync flags which display why each recipe was executed and print the
-parallelized output in sequential order, respectively.
-
-    Neuroimaging tools are typically equipped with quality assurance modules. As
-a result, it is possible to utilize the quality assurance modules from these
-tools to provide a complete report for the entire pipeline. The proposed
-solution suggests to write all the quality assurance outputs from each tool to a single HTML
-report with R Markdown or a script. The R Markdown or script, with the metrics
-and images are listed as dependencies to the QA report target in Make.
-
-    Only a basic form of data provenance has been implemented by the authors.
-The implemented data provenance consisted of a makefile that calls a bash to
-extract acquisition parameters from the raw input images. The makefile also
-obtains tool versions from the installation directories and pipeline options
-from their respective makefiles. All of these details are stored in a CSV file.
-The makefile will then call an R Markdown script which will then generate an
-HTML provenance report from the corresponding CSV file.
-
-    \section{Pydiper}
-        \subsection{Summary}
-
-        Pydiper is a neuroimaging registration workflow engine implemented in
-Python. This differs from other workflow engines in that it is specific to the
-process of image registration, but can, however, be extended for use with any
-analysis process. Its main features include being easy to use, particularly for
-non-developpers, undertaking a modular approach, giving the ability to access
-output files at any stage, and eliminating duplicate stages. Pydiper also
-provides the added advantage of being equipped with four commonly used
-registration pipelines, such that unexperienced researchers can use such
-pipelines out-of-the-box.
-
-        \subsection{Pipeline description}
-        Pydiper workflows can be broken down into three components: atoms,
-modules and applications. An application is the command line application that
-encapsulates the pipeline which consists of modules and atoms. Atoms are wrappers for individual operations,
-typically frquently used MINC tools. Modules are reponsible for linking
-atoms together.
-
-        Pydpiper provides three main classes to use to develop pipelines. These
-include the \textit{PipelineStage}, \textit{CmdStage} and \textit{Pipeline} classes. The
-\textit{PipelineStage} is the base class wrapper for all pipeline tools.
-Pipeline tools can be written in Python or can be a command-line application.
-For the tools that are command-line applications, the \textit{CmdStage} class, a
-child class of \textit{PipelineStage}, must be used. Input to \textit{CmdStage}
-is in a adequately parsed array format containing the specific command-line
-tool and its arguments. Reusable subclasses of the \textit{CmdStage} class can be
-created, or \textit{CmdStage} can be called directly.
-
-        To combine pipelines, the \textit{AddPipeline} function of the \textit{Pipeline}
-        class may be used.
-
-        The base class common to all Pydpiper applications is \textit{AbstractApplication}. 
-Classes that inherit from \textit{AbstractApplication} are command-line tools, with their 
-options set by the \textit{AbstractApplication} class. The class 
-is also responsible for initializing or resuming the pipeline, setting up a logger, and 
-launching the pipeline daemon.
-
-        Pydpiper atoms inherit from the \textit{CmdStage} class in order to wrap the MINC Tools. 
-They require as input a MINC file, and may additionally also require a target MINC file as second argument. These arguments
-can be in the form of a string filename or a file handler, however, the same format must be used for the two.
-An atom may also accepts various optional arguments.
-Using the inputs provided, the atom will put together the command to be executed.
-
-        \subsection{Pipeline execution}
-
-        In Pydpiper, there are two main classes,\textit{Pipeline} and \textit{PipelineExecutor}, 
-that monitor and manage pipeline execution. The \textit{Pipeline} class's main responsibility
-is to infer dependencies between pipeline stages from their inputs and outputs. It also manages
-a queue of runnable stages (stages in which all their dependencies have been met), ensuring
-that any runnable stage is placed in the queue. It also keeps track of the status of the runnable
-stages (running, finished, or failed). Should an already existing stage be added to the pipeline, it
-is automatically skipped by the \textit{Pipeline} class.
-
-        As stages may be added to incomplete pipelines, Pydpiper provides the functionality of resuming
-pipelines. This is achieved through a pickling mechanism implemented by the \textit{Pipeline} class. The
-pickled components include the directed acyclic graph, the array of pipeline stages, the stage counter,
-the hashes of all stages and each of their output files, and the status of each stage.
-
-        The \textit{PipelineExecutor} class, on the other hand, is responsible for the efficient processing
-of the pipeline through parallelization. Pydpiper's architecture mimics that of client/server, in which the 
-server is the pipeline daemon started by the command-line executable and the executors undertaking the role of 
-the clients. The pipeline's executors each consist of a thread pool in which each
-thread is responsible for running a single pipeline stage from the queue of runnable stages. To
-obtain a stage for execution, the executor polls the server to obtain the next available stage. Should
-the executor have enough memory and processors to execute the stage, it will be executed by the thread
-and release memory and processors and repoll the server once stage has completed.
-Otherwise, the executors will sleep for certain duration and repoll the server again. 
-Should there be any failed stages, the pipeline will be shut down after all possible stages have finished 
-executing. If additional executors are required by the pipeline, they may be added at any time through
-the command-line. Details related to
-the number of threads, number of threads per executor and the amount of memory can be specified through the
-command-line options.
-
-        \subsection{Debugging and provenance}
-
-        As mentioned previously, Pydpiper uses a pickling mechanism to gather important pipeline details
-        during execution in order to allow for the pipeline to be resumed should any stages have failed.
-        In addition, to keep track pipeline stage interdependencies in addition to file versioning, Pydpiper
-        utilizes the helper class \textit{RegistrationPipeFH} and its parent \textit{RegistrationFHBase} to create
-        file handlers of the files.
-
-        \note{from github wiki}
-        For debugging, a Pydpiper pipeline creats its own log file and executors typically also each create 
-        their own log file that are stored in a \textit{logs/}
-        subdirectory. Verbosity of logging information can be managed through the setting of the PYRO\_LOGLEVEL
-        environment variable. In addition, individual stage stdout/stderr is also redirected to a log file.
-
-        To verify the status of an executing pipeline, the check\_pipeline\_status.py may be executed.
-
-\chapter{Other domains and general scientific worflow engines}
-    \section{Taverna}
-    Taverna is an open-source bioinformatics workflow engine, designed to allow users to combine distributed web services, other service
-    types (e.g. BioMart queries or R-scripts) and local scripts into complex analysis pipelines. The main goal of
-    Taverna is to give a unified access to the plethora of services already existing on the web without requiring
-    the user to download the services locally or to have knowledge on all of the platforms and programming languages
-    the services are hosted under.
-
-        \subsection{Pipeline description}
-        Workflows in Taverna are built by combining web services, local scripts and other services together. Web services compatible with
-        Taverna must comply to the Web Service Description Language (WSDL) using the Simple Object Access Protocol (SOAP protocol). 
-        These services can be added to the BioCatalogue or the Biodiveristy catalogue for sharing and reuse of the web services.
-
-        Workflows can be built in the Taverna workbench. Taverna, through the help of the BioCatalogue and the Biodiversity Catalogue is
-        able to provide the user with a variety of webservices to combine into workflows in a drag-and-drop fashion. Workflows may also be composed
-        of sub-workflows. As service inputs and outputs may not match between services, shims are provided.
-        Similarly to the services, the user-generated workflows can be uploaded the myExperiment repository to be shared with other user.
-
-        \subsection{Pipeline execution}
-        The workflows can be executed remotely utilizing the Taverna Server. The Taverna Server is a an application that provides the full library of possible workflows
-        through webservices. It has an interface, known as the Taverna Player, that enables workflow execution to occur through a web-interface or through
-        third-party clients. The Taverna Server, however, is limited in that it can only execute workflows available on the Server. Taverna Lite is an alternative to
-        the Taverna Server which allows users to upload and execute workflows from myExperiment or elsewhere.
-
-        \subsection{Debugging and provenance}
-         Provenance capture is achieved through the Taverna Provenance Suite. Taverna does not capture provenance of the workflow definition, but generates a new
-        UUID after each definition change. Workflow run provenance captures information on "individual processor iterations" and their respective inputs and outputs and
-        stores this information in a database. The results can then be exported as a workflow trace or as a PROV-O RDF graph using the Taverna-PROV Plugin. The model format used
-        for exported provenance uses both the W3C PROV model and Open Provenance Model format.
-
-    \section{Pegasus}
-        Pegasus is a general-purpose scientific workflow engine that specializes in mapping abstract descriptions of workflows to resources for efficient processing. 
-        Similarly to other workflows, a Pegasus workflow can be represented by a DAG. This DAG is described by the user in DAX format (Directed Acyclic graph in XML).
-        As the DAX abstract workflow is resource-independent, Pegasus can convert the abstract workflow into an execution workflow that is catered to the resources available.
-
-        \subsection{Pipeline description}
-        Within the DAX, the nodes are the tasks and the execution order is represented by the edges.
-        In the instance of nested graphs, the nodes in the DAX may also represent a DAX. Pipeline inputs and outputs, as well as pipeline executables,
-         are represented by logical identifiers.
-
-         Various APIs are provided by Pegasus in order for the user to generate the
-        DAX using their preferred language of choice. Available APIs include Perl, Python and Java. 
-        \subsection{Pipeline execution}
-        With the effecient execution of pipelines being Pegasus' main focus, there exist various components that ensure optimal processing. The Mapper's responsability is to map
-        abstract workflow tasks to available resources. In order for the Mapper to successfully convert an abstract workflow into an execution workflow, it requires details on the execution
-        environment. The Mapper will also query various catalogs to determine the location of executables and input and output files. These catalogs can be either file-based or in
-        database format. With all the necessary information provided, the Mapper is equipped to convert abstract tasks into executable tasks. Data nodes are also added for input and output
-        staging. Should the data required for a task already exist in the catalog, the task and all its parents will be removed by the Mapper from the workflow. This functionality is optional,
-        and in some instances, it may be found that recomputation is preferable. If the executables required by the task are not pre-installed on the node, the Mapper will query the 
-        \textit{Transformation Catalog} for the stageable binaries.
-
-        The default, Pegasus relies on the HTCondor framework as both the workflow engine and scheduler. The executable workflow, using HTCondor DAGMan, consists of nodes representing jobs.
-        These jobs each have a corresponding job submit file that details how they are meant to be executed. A job that is ready for execution is released by the DAG and placed into a Condor
-        queue that is managed by the HTCondor Schedd daemon. The default scheduling is done using a local HTCondor pool, but can also be achieved through built-in interfaces with remote 
-        job-management protocols, or through SSH to directly submit jobs to remote resource managers.
-
-        Using HTCondor, Pegasus is able to achieve portability. Pegasus is compatible with local, HPC, distributed and cloud execution environments. As distributed environments may
-        not employ a shared filesystem, it is recommended to use HTCondor glideins to ensure that tasks are mapped to nodes where the resources are located. Should a user running their
-        pipeline locally without HTCondor installed, it is possible through \textit{submit host}. Using this method, however, executes the pipeline in serial order. Should the user wish to exploit
-        the pipeline's parallelism, HTCondor is required. Pegasus pipelines are not limited to a single execution environment, and can combine multiple environments to select the best resource
-        configuration to execute each job.
-
-        In order to manage tasks and their dependencies on remote resources, three different execution engines have been developed. The first and simplest execution engine is the 
-        \textit{Pegasus Cluster}, which is 
-        compatible with all the execution environments. It runs on a single-thread, meaning that it can only executed a single task at a time using an ordered-list of tasks.
-        The \textit{Pegasus Cluster} is particularly useful for the efficient execution of short tasks and the reduction of data transfers.
-
-        For execution environments lacking a shared filesystem, the \textit{PegasusLite} execution engine can be used. In order for the Pegasus to execute in such environments, the Mapper first
-        adds data transfer nodes to the pipeline to move the data to a data staging server associated with the execution node. Once a job is assigned to a node by the remote cluster scheduler,
-        \textit{PegasusLite} is deployed.
-
-        The final execution engine provided by Pegasus is the \textit{Pegasus MPI Cluster} (PMC). This execution engine is designed for petascale systems for which Pegasus' other strategies 
-        leveraging Condor Glideins are unsuitable. It employs the Master-Worker architecture, launching a single Master and many Workers, and using MPI for communication between the 
-        Master and Workers. The main limitation of this strategy is that a task must use less resources than that available on a single node. Jobs are not limited to a single execution engine,
-        and thus, multiple can be used in a job to optimize performance.
-
-        As executing a workflow on remote resources can be slow due to job queuing times and data transfers, the Mapper will group short-running task together to reduce to improve runtime. 
-        When tasks located at the same distance from the DAG root have similar runtimes, level-based clustering can be used. When runtimes of same-level tasks may
-        vary, it may be preferable to use job runtime-based horizontal clustering, to achieve load-balancing of clusters. Alternatively, the user may opt for label-based clustering, in which
-        the user can define the task clusters themselves.
-
-        To avoid running out of disk space during data-intensive workflows, the Mapper can add cleanup nodes that remove all unecessary data after a job has completed.
-
-        \note{Dynamic workflows permitted}
-
-        \subsection{Debugging and provenance}
-         Job monitoring, in Pegasus, is achieved through \textit{pegasus-kickstart} (kickstart). Every job within a pegasus workflow has a kickstart wrapper that collects runtime provenance
-        and performance. Kickstart relays this information to the submit host after job completion. This data is ultimately collected and stored in a relational database.
-
-        The Mapper Logs contain details on the abstract workflow, the jobs in the executable workflow 
-        and the mapping of tasks to jobs. Using the mapper logs, Pegasus can generate the provenance of the results. Should it be required, Pegasus can also generate the provenance of the 
-        mapping process.
-
-        In addition to the Mapper Logs, there also exists the DAGMan logs, generated by HTCondor DAGMan and updated in near real-time. These logs capture details on the status of the jobs
-        and their associated pre- and post- execute scripts.
-
-        All provenance and debugging data is accessibly through either the \textit{pegasus-analyzer} tool or through the provided lightweight web interface.
+        The \textit{executor} is responsible for the execution of a Nextflow
+pipeline. By default, it assumes that the pipeline is to be run locally. The
+local executor will use multithreading in order to exploit available cores
+within the local environment. To configure the executor to run in a remote distributed environment, it is
+necessary to modify the configuration file to specify the environment. By
+following this approach, Nextflow users can seamlessly run their code on various
+processing platforms with a simple modification to a configuration file. In
+addition to local, Nextflow currently has executors for SGE, LSF, SLURM,
+PBS/Torque, NQSII, HTCondor, DRMAA, Ignite, Kubernetes and AWS Batch.
+
+        \subsection{Debugging and Provenance}
+        When executing a pipeline, it is always possible to specify the
+generation of a execution report. The report is written in HTML and consists of three sections:
+Summary, Resources and Tasks. Summary provides details on the execution command,
+execution status, overall execution time, in addtion to other metadata. The
+Resources tab, on the other hand, contains several HighChart plots representing
+CPU, memory, time and disk read/write rate of each pipeline process. The final
+section, Tasks, contains many metrics including the status and command script of
+each task.
+
+    Other reports that can be generated at runtime include the Trace report and
+the Timeline report. The Trace report include details such as CPU usage, start
+time, submission time, etc, of each process that was executed, whereas, the
+Timeline report generates an HTML timeline of all processes that have been
+executed.
+
+    \note{doesn't seem to be anything on provenance...}
 
@@
     \section{Toil}
+        Toil is another general workflow framework aimed at the processing of scientific big data, with its main focus being on
+        genomics workflows. It prides itself in not only its ability to not only scale the processing of very large datasets,
+        but also its portability, allowing it to be executed on any environment. Toil pipelines can be written in CWL, WDL (Workflow Description Language) and Python,
+        with its Python API permitting the dynamic declaration of workflows. The Toil workflow framework is also open-sourced, making it a community-driven effort \note{as is the case with most}
         \subsection{Pipeline description}
+
         \subsection{Pipeline execution}

### 2018-09-21 paper.tex
completed toil summary

--- 2018-09-13 paper.tex
@@
 
-\usepackage{hyperref}
-\usepackage{xcolor}
-\usepackage{ulem}
-
-\newcommand{\note}[1]{\textcolor{green}{\textit{note}: #1}}
+\usepackage{hyperref} \usepackage{xcolor} \usepackage{ulem}
+
+\newcommand{\note}[1]{\textcolor{blue}{\textit{note}: #1}}
 \newcommand{\tristan}[1]{\textcolor{red}{TG: #1}}
@@
 
-\begin{document}
-\title{Pipeline systems and infrastructure for the efficient and open processing of Big neuroimaging Data}
-\author{Valerie Hayot-Sasson}
-\maketitle
-\begin{abstract}
-  Text of abstract.  
-\end{abstract}
-\tableofcontents
-\chapter{Introduction}
-intro here -mention hpc, data centre, cloud. 
-\chapter{BigData and Open science in neuroimaging}
-    \note{Likely will not be split in two separate sections}
-    \section{large images}
-        \begin{itemize}  
-            \item BigBrain
-            \item Micro CT 
-            \item EM 
-        \end{itemize}
-    \section{large datasets}
-    	\note{off the top of my head, but need to double check}
-        \begin{itemize}
-            \item HCP
-            \item UK Biobank
-            \item ADNI
-            \item OpenFMRI
-            \item BIDS/BIDS apps
-        \end{itemize}
-    An effective engine to process such data requires the following characteristics
-        \begin{itemize}
-            \item in-memory computing
-            \item data locality
-            \item lazy evaluation
-        \end{itemize}
-        \section{open-science /data sharing platforms for neuroimaging}
-        		is open science just data sharing, computing. 
-        		\subsection{Introduction}
-		\subsection{CBRAIN}
-		\subsection{OpenNEURO}
-		\subsection{NeuroData}
-        \subsection{Apache Airvata}
-\chapter{Workflow engines for neuroimaging data}
-	\section{Definition of workflows/pipelines}
+\begin{document} \title{Pipeline systems and infrastructure for the efficient
+and open processing of Big neuroimaging Data} \author{Valerie Hayot-Sasson}
+\maketitle \begin{abstract} Text of abstract.  \end{abstract} \tableofcontents
+    \chapter{Introduction} intro here -mention hpc, data centre, cloud.
+\chapter{BigData and Open science in neuroimaging} \note{Likely will not be
+split in two separate sections} \section{large images} \begin{itemize}  \item
+        BigBrain \item Micro CT \item EM \end{itemize} \section{large datasets}
+        \note{off the top of my head, but need to double check} \begin{itemize}
+        \item HCP \item UK Biobank \item ADNI \item OpenFMRI \item BIDS/BIDS
+        apps \end{itemize} An effective engine to process such data requires the
+        following characteristics \begin{itemize} \item in-memory computing
+        \item data locality \item lazy evaluation \end{itemize}
+            \section{open-science /data sharing platforms for neuroimaging} is
+            open science just data sharing, computing.
+            \subsection{Introduction} \subsection{CBRAIN} \subsection{OpenNEURO}
+            \subsection{NeuroData} \subsection{Apache Airvata} \chapter{Workflow
+            engines for neuroimaging data} \section{Definition of
+            workflows/pipelines}
 
     Broad definition of activities - between the tool and the engine.
-	\section{Nipype}
-		\subsection{summary}
-		Nipype\cite{nipype} is a popular neuroimaging pipelinening framework written in Python.
-		It provides users with uniform access to the rich ecosystem of neuroimaging software
-		libraries (e.g. SPM, FSL, Freesurfer) through its \textit{Interfaces}. \weird{Should users want to
-		use their own custom tool, it is also possible to create an interface for that}. To execute 
-		Python code in Nipype, the \textit{Function} interface can be used \tristan{for arbitrary code?}. The framework is 
-		also easy to use, and thus does not contribute to the steep learning curve new
+    \section{Nipype} \subsection{summary} Nipype\cite{nipype} is a popular
+    neuroimaging pipelinening framework written in Python.  It provides users
+    with uniform access to the rich ecosystem of neuroimaging software libraries
+    (e.g. SPM, FSL, Freesurfer) through its \textit{Interfaces}. \weird{Should
+    users want to use their own custom tool, it is also possible to create an
+    interface for that}. To execute Python code in Nipype, the \textit{Function}
+    interface can be used \tristan{for arbitrary code?}. The framework is also
+    easy to use, and thus does not contribute to the steep learning curve new
     researchers must face.
@@
         A pipeline in Nipype, consists of a data analysis. explicit
-                \textit{Workflow} of connected
-                \textit{interfaces}. The workflow is a directed
-                acyclic graph that connects the outputs of one
-                interface to be the inputs of another. A workflow may
-                also consist of connected workflows.
-
-                \tristan{The text below is about workflow execution
-                  while the text above is about workflow description,
-                  this should be a new paragraph.}  Workflows may be
-                executed in parallel either locally or an a cluster
-                through the use of a plugin in the Workflow's run
-                function. For debugging, Nipype uses \textit{graphviz}
-                to generate a static graph representing the Nodes and
-                their relationships. \tristan{debugging might be a
-                  third theme, in addition to description and
-                  execution. Provenance might go in debugging.}
-
-		 Each \textit{interface} within the workflow must be
-                 contained within a \textit{Node} or a
-                 \textit{MapNode} object, the MapNode class being a
-                 subclass of the Node class. Wrapping an interface in
-                 such objects ensures that interfaces are executed
-                 within a uniquely named directories, which in turn
-                 enables provenance tracking. Node objects also hash
-                 inputs, provide the ability to iterate over inputs
-                 and cache results. Hash inputs are useful in the case
-                 of Workflow recomputation, where Nipype will only
-                 recompute Nodes whose inputs have changed since the
-                 previous run.
-
-		Inspired by the MapReduce paradigm, the MapNode copies
-                the interface to each input and executes it
-                independently. \tristan{How are inputs defined/combined?} A reduce is subsequently performed by
-                the MapNode to return the output in the form of a
-                list. This differs from a Node object which can only
-                execute the interface on a single input at a time.
-
-		To enable execution of different parameters on the
-                input, Nodes have a property known as
-                \textit{Iterables}. Similarly to a MapNode, a copy of
-                the \textit{interface} will be made for each
-                input. However, with \textit{iterables}, a copy of
-                each dependent node will also be made.
-
-		\subsection{Limitations}
-		Nipype, although meeting the various needs of
-                researchers in Neuroinformatics \tristan{vague}, was
-                not designed with the processing of Big Data in
-                mind. Moreover, its performance relative to current
-                Big Data frameworks remains unknown \tristan{You could
-                  refer to your MS thesis here}.  \note{nipype
-                  performs a lot of disk I/O. Recomputation of
-                  workflows vs RDD caching}
-
-    \section{PSOM} 
-        \subsection{summary} 
-        While Python is a popular programming language amongst
-neuroinformaticians, it is not the only one that is commonly used. To address
-the needs of the Matlab/Octave community, PSOM (Pipeline System for Octave and
-Matlab) was developed. Similarly to Nipype, PSOM is easy-to-use, can run in
-distributed environments and can interface with tools written in other
-languages.
-
-        \subsection{pipeline description}
-        A pipeline in PSOM, is a series of connected jobs, which can be
-represented as a (static?) directed acyclic graph. The pipeline itself is implemented 
-using the Matlab/Octave data type. Each job within the pipeline
-is assigned a unique name. There is only one mandatory field in the job
-description (command). The remaining four optional fields include the list input
-files (files\_in), the list output files (files\_out), the list of files to be
-deleted after execution (files\_clean) and any Octave or Matlab datatype
-containing the remaining parameters necessary for job execution (opt).
-Jobs in PSOM run in a protected environment where they only have access to the
+        \textit{Workflow} of connected \textit{interfaces}. The workflow is a
+        directed acyclic graph that connects the outputs of one interface to be
+        the inputs of another. A workflow may also consist of connected
+        workflows.
+
+                \tristan{The text below is about workflow execution while the
+                text above is about workflow description, this should be a new
+                paragraph.}  Workflows may be executed in parallel either
+                locally or an a cluster through the use of a plugin in the
+                Workflow's run function. For debugging, Nipype uses
+                \textit{graphviz} to generate a static graph representing the
+                Nodes and their relationships. \tristan{debugging might be a
+                third theme, in addition to description and execution.
+                Provenance might go in debugging.}
+
+         Each \textit{interface} within the workflow must be contained within a
+         \textit{Node} or a \textit{MapNode} object, the MapNode class being a
+         subclass of the Node class. Wrapping an interface in such objects
+         ensures that interfaces are executed within a uniquely named
+         directories, which in turn enables provenance tracking. Node objects
+         also hash inputs, provide the ability to iterate over inputs and cache
+         results. Hash inputs are useful in the case of Workflow recomputation,
+         where Nipype will only recompute Nodes whose inputs have changed since
+         the previous run.
+
+        Inspired by the MapReduce paradigm, the MapNode copies the interface to
+        each input and executes it independently. \tristan{How are inputs
+        defined/combined?} A reduce is subsequently performed by the MapNode to
+        return the output in the form of a list. This differs from a Node object
+        which can only execute the interface on a single input at a time.
+
+        To enable execution of different parameters on the input, Nodes have a
+        property known as \textit{Iterables}. Similarly to a MapNode, a copy of
+        the \textit{interface} will be made for each input. However, with
+        \textit{iterables}, a copy of each dependent node will also be made.
+
+        \subsection{Limitations} Nipype, although meeting the various needs of
+        researchers in Neuroinformatics \tristan{vague}, was not designed with
+        the processing of Big Data in mind. Moreover, its performance relative
+        to current Big Data frameworks remains unknown \tristan{You could refer
+        to your MS thesis here}.  \note{nipype performs a lot of disk I/O.
+        Recomputation of workflows vs RDD caching}
+
+    \section{PSOM} \subsection{summary} While Python is a popular programming
+    language amongst neuroinformaticians, it is not the only one that is
+    commonly used. To address the needs of the Matlab/Octave community, PSOM
+    (Pipeline System for Octave and Matlab) was developed. Similarly to Nipype,
+    PSOM is easy-to-use, can run in distributed environments and can interface
+    with tools written in other languages.
+
+        \subsection{pipeline description} A pipeline in PSOM, is a series of
+        connected jobs, which can be represented as a (static?) directed acyclic
+        graph. The pipeline itself is implemented using the Matlab/Octave data
+        type. Each job within the pipeline is assigned a unique name. There is
+        only one mandatory field in the job description (command). The remaining
+        four optional fields include the list input files (files\_in), the list
+        output files (files\_out), the list of files to be deleted after
+        execution (files\_clean) and any Octave or Matlab datatype containing
+        the remaining parameters necessary for job execution (opt).  Jobs in
+        PSOM run in a protected environment where they only have access to the
         parameters that are passed.
@@
         In order to determine whether the opt field was correctly set, the
-generic function \textit{psom\_struct\_defaults} may be applied. This function
-will set the default values, ensure all mandatory fields were supplied and issue
-warnings for all unidentified attributes.
-
-        To define modules in PSOM, a special Octave/Matlab function type, known as "bricks"
-is used. Bricks employ the same parameters as jobs but lack the file\_clean
-parameter. All bricks contain an opt.flag\_test boolean flag, which, if set to
-true, allows the user to perform a dry-run of the brick. A dry-run of a brick
-consists of updating the default parameters and validating that the brick can
-successfully execute with the parameters provided. Bricks also have an option to
-specify the output folder (opt.output\_folder) for brick outputs to be stored.
+        generic function \textit{psom\_struct\_defaults} may be applied. This
+        function will set the default values, ensure all mandatory fields were
+        supplied and issue warnings for all unidentified attributes.
+
+        To define modules in PSOM, a special Octave/Matlab function type, known
+        as "bricks" is used. Bricks employ the same parameters as jobs but lack
+        the file\_clean parameter. All bricks contain an opt.flag\_test boolean
+        flag, which, if set to true, allows the user to perform a dry-run of the
+        brick. A dry-run of a brick consists of updating the default parameters
+        and validating that the brick can successfully execute with the
+        parameters provided. Bricks also have an option to specify the output
+        folder (opt.output\_folder) for brick outputs to be stored.
 
         The pipeline generator is an easy way to create complete pipelines with
-minimal information provided. A pipeline generator may accept at most the
-file\_in and opt arguments. It then builds the pipeline based on this
-information and the default parameters specified in each brick applied by the
-pipeline generator.
+        minimal information provided. A pipeline generator may accept at most
+        the file\_in and opt arguments. It then builds the pipeline based on
+        this information and the default parameters specified in each brick
+        applied by the pipeline generator.
 
@@
         the pipeline configuration options. The latter consists of the log
-directory, which must be provided, and the pipepline execution modes (e.g., session, background, batch,
-qsub, msub), in which the default mode is \textit{background}. Pipeline
-execution is achieved through the execution of the \textit{psom\_run\_pipeline}
-command.
+        directory, which must be provided, and the pipepline execution modes
+        (e.g., session, background, batch, qsub, msub), in which the default
+        mode is \textit{background}. Pipeline execution is achieved through the
+        execution of the \textit{psom\_run\_pipeline} command.
 
         Pipeline execution is broken down into three distinct modules: 1)
-initialization stage, 2) pipeline manager and 3) job manager. The initialization
-stage performs some checks to ensure that the pipeline can execute successfully
-(e.g., verify that pipeline is a DAG, check that output files are generated only
-once, check that all input files not generated by the pipeline exist on disk,
-delete any already existing output files), as well as performance checks (e.g.
-do not recompute pipeline steps that have already been executed, create
-necessary output folders to avoid creating them for each individual job). The
-initialization manager can determine which jobs need to be executed by comparing
-the current pipeline jobs with the details of those previously executed stored
-in the logs folder. The jobs that have been executed can have the following
-states: 1) none, 2) finished and 3) failed. The initialization manager stage the jobs 
-with a \textit{none} or \textit{failed} status for recomputation. For jobs whose
-description has changed since the previous \textit{finished} execution, the
-initialization will also stage it for recomputation. Should the user want to
-reexecution a previously executed \textit{finished} job, the user may force
-reexecution by setting \textit{opt\_pipe.restart} parameter to include all jobs
-to be reexecuted.
+        initialization stage, 2) pipeline manager and 3) job manager. The
+        initialization stage performs some checks to ensure that the pipeline
+        can execute successfully (e.g., verify that pipeline is a DAG, check
+        that output files are generated only once, check that all input files
+        not generated by the pipeline exist on disk, delete any already existing
+        output files), as well as performance checks (e.g.  do not recompute
+        pipeline steps that have already been executed, create necessary output
+        folders to avoid creating them for each individual job). The
+        initialization manager can determine which jobs need to be executed by
+        comparing the current pipeline jobs with the details of those previously
+        executed stored in the logs folder. The jobs that have been executed can
+        have the following states: 1) none, 2) finished and 3) failed. The
+        initialization manager stage the jobs with a \textit{none} or
+        \textit{failed} status for recomputation. For jobs whose description has
+        changed since the previous \textit{finished} execution, the
+        initialization will also stage it for recomputation. Should the user
+        want to reexecution a previously executed \textit{finished} job, the
+        user may force reexecution by setting \textit{opt\_pipe.restart}
+        parameter to include all jobs to be reexecuted.
 
@@
         pipeline and submitting jobs for execution. Its duties begin after the
-initialization manager has completed execution. Jobs can be submitted by the
-pipeline manager once all their dependencies have been satisfied. The
-dependencies of a job are designated as met if there are none or when its parent jobs within the
-dependency graph are tagged as \textit{finished}. In addition, a job is
-considered to be successfully completely when all the dependencies of its
-children jobs have been satisfied. Depending on the pipeline configuration,
-there may be a limit to how many jobs can run in parallel. Once a job has
-finished executing, an empty tag file is created with a filename containing the
-final state in the logs folder. The pipeline manager continuously parses through
-the logs folder in search of these state files to delete and update the history
-file with the job state. Both the pipeline manager and the jobs can be executed
-within the current session, locally in a independent session, or in an
-independent session on a cluster
+        initialization manager has completed execution. Jobs can be submitted by
+        the pipeline manager once all their dependencies have been satisfied.
+        The dependencies of a job are designated as met if there are none or
+        when its parent jobs within the dependency graph are tagged as
+        \textit{finished}. In addition, a job is considered to be successfully
+        completely when all the dependencies of its children jobs have been
+        satisfied. Depending on the pipeline configuration, there may be a limit
+        to how many jobs can run in parallel. Once a job has finished executing,
+        an empty tag file is created with a filename containing the final state
+        in the logs folder. The pipeline manager continuously parses through the
+        logs folder in search of these state files to delete and update the
+        history file with the job state. Both the pipeline manager and the jobs
+        can be executed within the current session, locally in a independent
+        session, or in an independent session on a cluster
 
         Every time a job is submitted by the pipeline manager for execution, a
-job manager is started. It is a Matlab function responsible for creating the job
-profile and generating the job-specific logs and the job exit tags. When
-the job terminates, the job manager will verify that all output has been
-correctly generated and assigns a job tagged based on this verification that
-will be reported back to the pipeline manager.
+        job manager is started. It is a Matlab function responsible for creating
+        the job profile and generating the job-specific logs and the job exit
+        tags. When the job terminates, the job manager will verify that all
+        output has been correctly generated and assigns a job tagged based on
+        this verification that will be reported back to the pipeline manager.
 
@@
         The information stored in the logs folder is complete with details
-allowing for both debugging and provenance. The logs folder contains six files:
-1) PIPE\_history.txt, 2) PIPE\_jobs.mat, 3) PIPE\_status.mat, 4) PIPE\_logs.mat,
-5) PIPE\_profile.mat and 6) PIPE.mat. PIPE\_history.txt includes the execution
-history of the pipeline manager. PIPE\_jobs.mat and PIPE\_profile.mat are both
-Matlab/Octave files containing each executed job as a Matlab variable. They
-differ in that PIPE\_jobs.mat contains the latest executed jobs and can be used
-to reexecute the pipeline at any moment in time and PIPE\_profile.mat contains
-the execution time of each job \note{i think both can be used to reexecute
-pipeline from scratch. Also unclear whether PIPE\_profile.mat contains more than
-just latest jobs}. It is also possible to access any of the previously 
-used parameters through PIPE\_jobs.mat. PIPE\_status.mat stores the status of each 
-job as a string variable, whereas PIPE\_log.mat stores each job's log as a string 
-variable. Job logs, as generated by the log manager, consist of a text copy of the 
-execution, in addition to the user name, date, time and operating system used to 
-execute the job. Finally, all configuration details are stored in PIPE.mat.
-
-        \subsection{limitations of PSOM}
-    \section{SPM and matlabbatch} 
+        allowing for both debugging and provenance. The logs folder contains six
+        files: 1) PIPE\_history.txt, 2) PIPE\_jobs.mat, 3) PIPE\_status.mat, 4)
+        PIPE\_logs.mat, 5) PIPE\_profile.mat and 6) PIPE.mat. PIPE\_history.txt
+        includes the execution history of the pipeline manager. PIPE\_jobs.mat
+        and PIPE\_profile.mat are both Matlab/Octave files containing each
+        executed job as a Matlab variable. They differ in that PIPE\_jobs.mat
+        contains the latest executed jobs and can be used to reexecute the
+        pipeline at any moment in time and PIPE\_profile.mat contains the
+        execution time of each job \note{i think both can be used to reexecute
+        pipeline from scratch. Also unclear whether PIPE\_profile.mat contains
+        more than just latest jobs}. It is also possible to access any of the
+        previously used parameters through PIPE\_jobs.mat. PIPE\_status.mat
+        stores the status of each job as a string variable, whereas
+        PIPE\_log.mat stores each job's log as a string variable. Job logs, as
+        generated by the log manager, consist of a text copy of the execution,
+        in addition to the user name, date, time and operating system used to
+        execute the job. Finally, all configuration details are stored in
+        PIPE.mat.
+
+        \subsection{limitations of PSOM} \section{SPM and matlabbatch}
         \note{This section may not make much sense. Have not really yet fully
-understood matlabbatch}
-        \subsection{summary} 
-        SPM is a very popular neuroimaging toolkit, and as a result,
-neuroimaging pipeline engines such as Nipype provide interfaces for it. SPM is
-also equipped with its own neuroimaging pipeline engine known as
-\textit{matlabbatch}. Similarly to PSOM and as its name suggest, it is written
-in Matlab.
-
-        \subsection{pipeline description}
-        Pipelines in matlabbatch consist of a series of connected data
-processing steps known as "modules" \note{DAG?? probably but not explicitly
-listed anywhere}. All SPM modules, helper modules and their respective
-dependencies can be defined using SPM's GUI \note{can probably also be done
-through command-line}.
+        understood matlabbatch} \subsection{summary} SPM is a very popular
+        neuroimaging toolkit, and as a result, neuroimaging pipeline engines
+        such as Nipype provide interfaces for it. SPM is also equipped with its
+        own neuroimaging pipeline engine known as \textit{matlabbatch}.
+        Similarly to PSOM and as its name suggest, it is written in Matlab.
+
+        \subsection{pipeline description} Pipelines in matlabbatch consist of a
+        series of connected data processing steps known as "modules" \note{DAG??
+        probably but not explicitly listed anywhere}. All SPM modules, helper
+        modules and their respective dependencies can be defined using SPM's GUI
+        \note{can probably also be done through command-line}.
 
         For any matlabbatch job, it is possible to create a skeleton of it. A
-skeleton is a Matlab script containing a commented list of necessary inputs at
-the top and all code necessary to initialize and run the script. Before running,
-the skeleton will need to be configured with the input names and other
-necessary parameters.
+        skeleton is a Matlab script containing a commented list of necessary
+        inputs at the top and all code necessary to initialize and run the
+        script. Before running, the skeleton will need to be configured with the
+        input names and other necessary parameters.
 
@@
         Defaults can be set within the configuration files, through a default
-file, or in a \textit{.def} field for leaf nodes \note{terminal jobs???}. The
-latter updates at the start of the new job execution whereas the former are
-configured at the beginning of the SPM/matlabbatch startup.
+        file, or in a \textit{.def} field for leaf nodes \note{terminal
+        jobs???}. The latter updates at the start of the new job execution
+        whereas the former are configured at the beginning of the
+        SPM/matlabbatch startup.
 
         Virtual outputs represent the dependencies between jobs \note{unsure}.
-They contain three paramaters of which only one is optional: sname, src\_output,
-tgt\_spec. sname is the display name that will be stored in the dependecies
-list, whereas src\_output is a subscript reference used to refer to the output
-that is produced at runtime. The optional parameter, tgt\_spec, is a description
-of the kind of inputs the outputs produced can be a dependency for. The virtual
-output callbacks are evaluated when cfg\_(ex)branch, cfg\_choice and cfg\_repeat 
-have the correct number of child nodes within all their in-tree nodes.
-
-        \subsection{pipeline execution}
-        Pipeline execution can be achieved either through the provided GUI or by
-the command-line interface. The GUI, however, requires significant user
-interaction, as the user must repeat the "Runs" command as many times as
-necessary to reexecute the command on multiple data. The CLI, however, avoids
-this with the help of its \textit{spm\_jobman} interface.
+        They contain three paramaters of which only one is optional: sname,
+        src\_output, tgt\_spec. sname is the display name that will be stored in
+        the dependecies list, whereas src\_output is a subscript reference used
+        to refer to the output that is produced at runtime. The optional
+        parameter, tgt\_spec, is a description of the kind of inputs the outputs
+        produced can be a dependency for. The virtual output callbacks are
+        evaluated when cfg\_(ex)branch, cfg\_choice and cfg\_repeat have the
+        correct number of child nodes within all their in-tree nodes.
+
+        \subsection{pipeline execution} Pipeline execution can be achieved
+        either through the provided GUI or by the command-line interface. The
+        GUI, however, requires significant user interaction, as the user must
+        repeat the "Runs" command as many times as necessary to reexecute the
+        command on multiple data. The CLI, however, avoids this with the help of
+        its \textit{spm\_jobman} interface.
 
         The first step to running a batch pipeline is to perform the
-initialization. This may include configuring the \textit{spm\_defaults.m} code
-to run in command line and executing the \textit{spm(’defaults’, MODALITY)}
-(where MODALITY represents the modality used in the pipeline) and
-\textit{spm\_jobman(’initcfg’)}.
+        initialization. This may include configuring the
+        \textit{spm\_defaults.m} code to run in command line and executing the
+        \textit{spm(’defaults’, MODALITY)} (where MODALITY represents the
+        modality used in the pipeline) and \textit{spm\_jobman(’initcfg’)}.
 
         After the intialization is complete, a job may be executed using the
-\textit{spm\_jobman(’run’, job[, input1, input2 ...])} command. This command
-requires the job names as well as its necessary inputs specified in the correct
-order. The job can be specified as a variable, the name of a script that will
-generate a job variable or a cell list containing a mixture of scripts and job
-variables.
-
-        \subsection{debugging and provenance}
-        All batch scripts can be saved as a Matlab scripts. These scripts not
-only contain the code that was executed, but also all input parameters. Should
-any script fail in the process of executing, an error will be printed to the
-Matlab console. If the user is using the GUI, the error output may also be saved
-to a Matlab file.
+        \textit{spm\_jobman(’run’, job[, input1, input2 ...])} command. This
+        command requires the job names as well as its necessary inputs specified
+        in the correct order. The job can be specified as a variable, the name
+        of a script that will generate a job variable or a cell list containing
+        a mixture of scripts and job variables.
+
+        \subsection{debugging and provenance} All batch scripts can be saved as
+        a Matlab scripts. These scripts not only contain the code that was
+        executed, but also all input parameters. Should any script fail in the
+        process of executing, an error will be printed to the Matlab console. If
+        the user is using the GUI, the error output may also be saved to a
+        Matlab file.
 
         \subsection{limitations of SPM}
-    \section{Nextflow}
-        \subsection{summary}
-        Nextflow is a POSIX-compatible pipeline engine for scientific workflows.
-It extends the Groovy programming language to provide an  intuitive domain-specific language (DSL) 
-that enables users to easily describe their workflows. Unlike pipeline engines such as Nipype and
-PSOM, Netflow does not employ DAG. It instead adheres to the
-\textit{Dataflow} programming model which decides how data will flow through the pipeline at
-runtime \note{fix this sentence}. Nextflow prides itself on allowing the
-integration of containerized tools to address the issue of numerical instability
-that may arise from different environment setups.
-
-        \subsection{Pipeline description}
-        A Nextflow pipeline is represented by the flow of data between processes. These
-processes can be written in any Linux-compatible scripting language. As in Nipype, 
-these processes are executed in an isolated environment. The only means
-of communication between processes is through asynchronous FIFO queues known as
-\textit{channels}. Each process can contain any number of input and output
-channels. Through the use of channels, input and output can be stored not only
-on disk but also in-memory \note{Nipype can also return outputs that are not on
-disk, so this is nothing special}.
-
-        As with other pipeline engines (nipype, psom, etc.,), the order in which
-the processes appear is irrelevant. Execution order is determined by the dependencies.
-
-        \subsection{Pipeline execution}
-        The \textit{executor} is responsible for the execution of a Nextflow
-pipeline. By default, it assumes that the pipeline is to be run locally. The
-local executor will use multithreading in order to exploit available cores
-within the local environment. To configure the executor to run in a remote distributed environment, it is
-necessary to modify the configuration file to specify the environment. By
-following this approach, Nextflow users can seamlessly run their code on various
-processing platforms with a simple modification to a configuration file. In
-addition to local, Nextflow currently has executors for SGE, LSF, SLURM,
-PBS/Torque, NQSII, HTCondor, DRMAA, Ignite, Kubernetes and AWS Batch.
-
-        \subsection{Debugging and Provenance}
-        When executing a pipeline, it is always possible to specify the
-generation of a execution report. The report is written in HTML and consists of three sections:
-Summary, Resources and Tasks. Summary provides details on the execution command,
-execution status, overall execution time, in addtion to other metadata. The
-Resources tab, on the other hand, contains several HighChart plots representing
-CPU, memory, time and disk read/write rate of each pipeline process. The final
-section, Tasks, contains many metrics including the status and command script of
-each task.
-
-    Other reports that can be generated at runtime include the Trace report and
-the Timeline report. The Trace report include details such as CPU usage, start
-time, submission time, etc, of each process that was executed, whereas, the
-Timeline report generates an HTML timeline of all processes that have been
-executed.
-
-    \note{doesn't seem to be anything on provenance...}
-
-    \section{FastR}
-        \subsection{summary}
-        FastR is another Python-based neuroimaging pipeline framework that
-provides added features such including access to different versions of the same
-command-line tool \note{criticizes nipype for not having such a feature, but
-nipype can access boutiques tools...so...} and compatibility with various
-possible filesystems. In addition to its added features, FastR also guarantees
-a simple-to-use framework for building pipelines written in any languages, has
-built-in provenance support, provides the seamless ability to parallelize
-pipelines with the use of plugins and avoids the numerical instability issue with
-the simple integration of containerized pipelines. Similary to many other
+
+    \section{FastR} \subsection{summary} FastR is another Python-based
+    neuroimaging pipeline framework that provides added features such including
+    access to different versions of the same command-line tool \note{criticizes
+    nipype for not having such a feature, but nipype can access boutiques
+    tools...so...} and compatibility with various possible filesystems. In
+    addition to its added features, FastR also guarantees a simple-to-use
+    framework for building pipelines written in any languages, has built-in
+    provenance support, provides the seamless ability to parallelize pipelines
+    with the use of plugins and avoids the numerical instability issue with the
+    simple integration of containerized pipelines. Similary to many other
     pipeline framework, it also employs a DAG to describe data flow.
@@
 
-\textit{Tools}
-describe the details of the underlying application that is to be executed at the
-node. Tools consist of three components: general metadata, target and interface.
-The general metadata provides details on the Tools such as name, version,
-authors, etc, whereas the target specifies details on the execution environment.
-The interface, on the other hand, provides details on the inputs and outputs of
-the tool.
+\textit{Tools} describe the details of the underlying application that is to be
+executed at the node. Tools consist of three components: general metadata,
+target and interface.  The general metadata provides details on the Tools such
+as name, version, authors, etc, whereas the target specifies details on the
+execution environment.  The interface, on the other hand, provides details on
+the inputs and outputs of the tool.
 
 A JSON schema is required to validate the python object generated by the Tool.
-Fastr currently provides two schemas for object validation: the \textit{Tool} schema
-and the \textit{FastrInterface} schema. It is, however, possible to add a custom
-schema to FastR and define an \textit{Interface} for it.
-
-The \textit{Interface} class is responsible for executing a call to the command line tool.
-It dictates the tags that must be defined within the interface tag of the
-\textit{Tool}. For instance, the generic interface class requires details on the
-input and output's id, cardinality (number of input values) and DataType (a
-python object denoting the details of the Input or Output's data type).
+Fastr currently provides two schemas for object validation: the \textit{Tool}
+schema and the \textit{FastrInterface} schema. It is, however, possible to add a
+custom schema to FastR and define an \textit{Interface} for it.
+
+The \textit{Interface} class is responsible for executing a call to the command
+line tool.  It dictates the tags that must be defined within the interface tag
+of the \textit{Tool}. For instance, the generic interface class requires details
+on the input and output's id, cardinality (number of input values) and DataType
+(a python object denoting the details of the Input or Output's data type).
 DataTypes can also be grouped, allowing an inputs containing various files types
-to be inputted to applications which can support the file types.
-Verification is done by FastR to determine if the linked Input and Output data
-types are compatible.
+to be inputted to applications which can support the file types.  Verification
+is done by FastR to determine if the linked Input and Output data types are
+compatible.
 
@@
 sample consisting of the same number of values. During the instances in which it
-is desirable to apply the same task to multiple inputs, a multidimensional array of samples, known as a collection, can be supplied to a
-\textit{Node} for processing. If there are multiple input samples, they can be
-passed to a node in pairwise behaviour (e.g. 1-1) or cross-product behaviour
-(e.g. nxn) \textit{need to better explain this}.
-
-Output samples may need to be expanded (i.e. a single sample of values becomes separate
-samples) or collapsed (i.e. multiple samples become a single sample of values).
-FastR provides data link flow directives for such operations.
+is desirable to apply the same task to multiple inputs, a multidimensional array
+of samples, known as a collection, can be supplied to a \textit{Node} for
+processing. If there are multiple input samples, they can be passed to a node in
+pairwise behaviour (e.g. 1-1) or cross-product behaviour (e.g. nxn) \textit{need
+to better explain this}.
+
+Output samples may need to be expanded (i.e. a single sample of values becomes
+separate samples) or collapsed (i.e. multiple samples become a single sample of
+values).  FastR provides data link flow directives for such operations.
 
@@
 Nodes in that their input data are defined by the network \note{Not sure if
-saying this correctly}. Normal nodes are responsible for processing the
-data. Nodes are connected to each other through links that highlight the
-dependencies and form the \textit{Network}. Once the \textit{Network} is
-defined, it can be executed.
+saying this correctly}. Normal nodes are responsible for processing the data.
+Nodes are connected to each other through links that highlight the dependencies
+and form the \textit{Network}. Once the \textit{Network} is defined, it can be
+executed.
 
 The data type of the Source Node and Sink Node must be defined in the
-\textit{Network} definition.
-The data is define at runtime by evaluation of the uniform resource identifier
-(URI). Multiple URIs can be defined through the use of wildcards and searches. Plugins are used to read and write data to
-different filesystems specified in the URI scheme. Currently, FastR supports
-virtual file system (VSF), csv files and XNAT. Plugins to support other
-filesystems can easily be added to FastR. In addition, plugins that can only
-read, write, or perform searches can also be created.
-
-        \subsection{Pipeline execution}
-        To commence pipeline execution, the \textit{Network}'s execution
-function is called on the Submission host. FastR then analyses the network and
-its dependencies and breaks
-down the Network into chunks that can be further processed. From the chunks, the
+\textit{Network} definition.  The data is define at runtime by evaluation of the
+uniform resource identifier (URI). Multiple URIs can be defined through the use
+of wildcards and searches. Plugins are used to read and write data to different
+filesystems specified in the URI scheme. Currently, FastR supports virtual file
+system (VSF), csv files and XNAT. Plugins to support other filesystems can
+easily be added to FastR. In addition, plugins that can only read, write, or
+perform searches can also be created.
+
+        \subsection{Pipeline execution} To commence pipeline execution, the
+        \textit{Network}'s execution function is called on the Submission host.
+        FastR then analyses the network and its dependencies and breaks down the
+        Network into chunks that can be further processed. From the chunks, the
         execution order of the \textit{Nodes} is determined.
@@
 
-While the job is executing, input URIs are translated to their actual path/values,
-the execution environment is set up by the \textit{Tool} and the \textit{Tool}
-command is executed by the \textit{Interface}. When the \textit{Interface}
-returns the results, the
-output data is validated and translated to URI format.
+While the job is executing, input URIs are translated to their actual
+path/values, the execution environment is set up by the \textit{Tool} and the
+\textit{Tool} command is executed by the \textit{Interface}. When the
+\textit{Interface} returns the results, the output data is validated and
+translated to URI format.
 
@@
 
-        \subsection{Debugging and Provenance}
-        In FastR, a provenance document is generated for each resulting file
-generated in the Prov-N format. Data recorded in the provenance documents include \textit{Tool}
-versions, input/output paths/values and their respective checksums, start and
-end time of execution for each \textit{job}, stdout and stderr logs, the end
+        \subsection{Debugging and Provenance} In FastR, a provenance document is
+        generated for each resulting file generated in the Prov-N format. Data
+        recorded in the provenance documents include \textit{Tool} versions,
+        input/output paths/values and their respective checksums, start and end
+        time of execution for each \textit{job}, stdout and stderr logs, the end
         status and the complete environment description.
@@
 
-    \section{OpenMOLE}
-        \subsection{summary}
-        OpenMOLE (Open MOdeL Experiment) is a domain agnostic pipelining
-framework that focuses on simplifying workflow creation through its Domain
-Specific Language (DSL) built on top of Scala. Through the use of its DSL, much
-of the backend is abstracted from users. It also offers a GUI for users to build
-workflows. Similarly to other existing pipelining
-frameworks, OpenMOLE enables parallelization through plugins, meaning that the
-workflow need not be changed between different distribution environments.
-Moreover, OpenMOLE supports the deployment of tasks or groups of tasks on different
-computing environments.
-Another key feature of OpenMOLE is that it attempts to address the
-reproducibility issue through the integration of the CARE archiver. CARE allows
-for the execution of command-line applications, written in any language, and
-attempts to simulate the environment in which the application was designed on.
-OpenMOLE also provides embedded methods for parameter exploration.
-
-        \subsection{Pipeline description}
-       An OpenMOLE workflow consists of tasks connected by transitions. Each
-task contains details such as input, output and optional parameters, in addition
-to embedding the application that will be applied to the data. Input values are
-generally supplied by the dataflow through an entry point known as
-\textit{Sources} and output values are returned to the
-dataflow through \textit{Hooks}.
+    \section{LONI} \subsection{Summary} The LONI pipeline processing environment
+    facilitates the development, sharing and execution of pipelines. Workflow
+    Activities, known as \textit{Modules} in LONI, are described using
+    Extensible Markup Language (XML), which can easily shared between user
+    groups. LONI runs on the Java platform, allowing it to be platform
+    independent and has support DRMAA clusters and symmetric multiprocessing
+    (SMP) systems. In order to capture provenance information, LONI is also
+    equipped with a provenance manager.
+
+        \subsection{Pipeline description} LONI uses the dataflow programming
+        model to represent the pipeline. Data flows between \textit{Modules}
+        through \textit{data pipes}. Sub-pipelines are known and
+        \textit{Pipelets}. Modules are defined by the input arguments/streams
+        and output arguments/stream. Arguments in LONI may be specific
+        filetypes, flags or values listed in the order in which the application
+        requires. The provided arguments may be required or optional and can
+        also be duplicated, if necessary. File inputs in the form of URLs are
+        accepted by LONI and will be downloaded at runtime. List of inputs may
+        also be provided to process multiple inputs with copies of the same
+        module in parallel (batch processing). Conversely, it is also possible
+        to group list of inputs so as to process them all by the single module
+        \note{Need to express this better}. Input/Output arguments may be added
+        to the table of global variables for easy access throughout the
+        pipeline. It is possible to apply variable modifiers to the variables
+        stored within the table.
+
+Intermediary outputs as well as the file output are treated as temporary by
+LONI. They are scheduled for deletion before a new pipeline is executed. It is,
+however, possible to preserve these files.
+
+        \subsection{Pipeline execution} As a module completes it execution, its
+        output gets transmitted through the dataflow and the dependent modules
+        are signaled that their input data has arrived.  Modules who have met
+        all their dependency requirements are ready for execution.  Should a
+        module have no dependecies, it can be executed immediately. Modules
+        ready to be executed are placed in a FIFO launch queue to avoid
+        straining the platform.
+
+       LONI offers a built in client/server strategy that uses Java's remote
+       method invocation (RMI) libraries to communicate between servers. This
+       enables modules, pipelets or pipelines to be executed on servers
+       different from the client. It also enables full pipelines to consist of
+       connected modules and pipelets located on different servers. It is
+       favorable to have such a configuration in LONI, as the tools required for
+       processing may not be available on client server, but located in another.
+       The client is responsible for dictating all file transfers and module
+       launches that are to occur. The client authenticates with the servers
+       through a one-way hash. Only files produced by the pipeline are
+       accessible to the client and only temporary files created by the server
+       may be written to the server.
+
+        \subsection{Debugging and provenance} For each dataset processed in a
+        LONI pipeline, there exists a corresponding provenance (.prov) XML file.
+        The \textit{Provenance Editor}, a standalone platform-independent
+        application, reads the metadata from the image headers, extracts
+        relevant provenance information from them and creates the provenance
+        files.  This data can be modified by the user in the to correct
+        inaccuracies or append additional details before being saved. During
+        pipeline execution, the provenance file will be edited by the LONI
+        pipeline to include details on each processing step, the environment in
+        which the data was processed and tool binaries used.
+
+    \section{Automatic Analysis} \subsection{Summary} Automatic analysis (aa) is
+    another Matlab-based modular workflow engine, having started of as a
+    workflow engine for SPM. User may define their pipeline through a Matlab
+    user script and specify the tasks in addition to their input and output
+    types through an XML module interface. Modules used in pipelines must be
+    integrated into aa which can be easily accomplished through XML interfaces.
+    aa workflows can be executed locally, or in parallel through qsub or condor.
+
+        \subsection{Pipeline description} aa pipelines have two main components:
+        1) the Matlab user script and 2) the XML tasklist. The user script is a
+        script that defines which data will be processed, any necessary
+        parameter, and how it will be processed. aa provides an example user
+        script, which can be modified to meet user's needs. The entire pipeline
+        in aa is defined in a single structure, known as aap, which makes the
+        task of record-keeping simple.
+
+        The XML tasklist is an ordered list of all the modules that will be
+        applied to the data, that is read by the user script. This list does not
+        specify any dependencies as those dependendecies can be inferred by the
+        pipeline through analysis of the module interfaces. Modules can be
+        separated into two sections: initialization and main. Modules listed in
+        the initialization section are to be executed for every task.
+        Conversely, those modules liste in the main section are only executed
+        once for every task, unless otherwise specified.
+
+        In order to ensure that the correct output data is being operated on,
+        each task is provided with its own unique folder. The aa engine will
+        transmit the necessary input data to the folder before task execution
+        and also acquiring the outputs after task completion. Output folders are
+        numbered such that if a given module is invoked multiple times, each of
+        those folders are assigned an incremental ordering corresponding to the
+        order in which they have been executed \note{sentence might not make
+        sense}. Should all data be required for each task in the tasklist, all
+        the data will be copied to each task folder.
+
+    Similarly to a pipeline, each module is comprised of an XML interface and a
+    Matlab source file. A module's XML file consist of specifications on
+    required input and output modalities, as well as the desired domain. One of
+    the most important properties include the domain at which the module
+    operates. While there exist mainy domains for modules and new ones can
+    easily be created and integrated by the user, the most prominently used ones
+    include: "study", "subject", "session". As their name suggests, a module
+    with the domain "study" will only be executed once for each time it appears
+    in the tasklist, whereas "subject" executes the module for each subject
+    within the task and "session" executes once for each session within a
+    subject.
+
+    In addition to module domains, another important module property is the
+    input and output data stream types. Data stream types denote what input and
+    output data type each module produces. Commonly used data stream types
+    include "epi", "structural" and "dicom\_header".
+
+    Optional fields specified in a module include qsub, for estimates of
+    requirements for parallel processing, and \textit{performanceoutput}, for
+    garbage collection.
+
+   Analysis parameters can easily be configured in aa. Any variable defined by
+   the \textit{aaprecipe} command may be modified. Parameters can also be
+   defined for an instance of a module within the XML tasklist. Alternatively,
+   it is also possible to create an XML file that inherits parameters from
+   another XML file and overrides some or all of the parameters.
+
+   To set defaults and site-specific configurations, it is possible to either
+   create a local config file or by creating another parameter default file
+   which inherits the original one.
+
+    aa also supports the creation of branched tasklist. This allows a section of
+    the pipeline steps to undergo different tasks or the same tasks in a
+    different order in one pipeline rather than having to create separate
+    pipelines.
+
+        \subsection{Pipeline execution} Once the definition of the user-script,
+        XML tasklist and modules is complete, the pipeline is ready for
+        execution. Dependencies in aa are established based on input and output
+        data stream types. The aa engine attempts to match the last task
+        executed with a corresponding data stream type to the current type. It
+        is, however, possible to specify which module provides the input to the
+        current module by using fully qualified stream references.
+
+        During execution, a map of all the modules that need to be executed and
+        their respective dependencies is generated. Modules that have completed
+        contain a flag file indicating their completion within their instance's
+        root directory. Such modules will not be re-executed by the engine.
+        Additionally, if it is required that a previous module is re-executed,
+        the done flag is removed from all dependent modules.
+
+        Should part of the processing be performed on a different server it is
+        possible to connect the pipelines together through an analysis script.
+        The aa engine will be reponsible for matching the data streams between
+        the servers in order to avoid the copying of data and modules.
+
+        \subsection{Debugging and provenance} In order to retrace the steps of
+        previous executions, aa keeps a detailed record of the events that
+        occurred. These detailes are stored in a Matlab object \note{similarly
+        to PSOM}, which permits reexcution of a previous run or extraction of
+        specific parameter values.
+
+        Additionally, to verify the correctness of the results produced, aa
+        implements tools for quality control and comes with a dedicated module
+        for low-level quality control.
+
+    \section{Pydiper} \subsection{Summary}
+
+        Pydiper is a neuroimaging registration workflow engine implemented in
+        Python. This differs from other workflow engines in that it is specific
+        to the process of image registration, but can, however, be extended for
+        use with any analysis process. Its main features include being easy to
+        use, particularly for non-developpers, undertaking a modular approach,
+        giving the ability to access output files at any stage, and eliminating
+        duplicate stages. Pydiper also provides the added advantage of being
+        equipped with four commonly used registration pipelines, such that
+        unexperienced researchers can use such pipelines out-of-the-box.
+
+        \subsection{Pipeline description} Pydiper workflows can be broken down
+        into three components: atoms, modules and applications. An application
+        is the command line application that encapsulates the pipeline which
+        consists of modules and atoms. Atoms are wrappers for individual
+        operations, typically frquently used MINC tools. Modules are reponsible
+        for linking atoms together.
+
+        Pydpiper provides three main classes to use to develop pipelines. These
+        include the \textit{PipelineStage}, \textit{CmdStage} and
+        \textit{Pipeline} classes. The \textit{PipelineStage} is the base class
+        wrapper for all pipeline tools.  Pipeline tools can be written in Python
+        or can be a command-line application.  For the tools that are
+        command-line applications, the \textit{CmdStage} class, a child class of
+        \textit{PipelineStage}, must be used. Input to \textit{CmdStage} is in a
+        adequately parsed array format containing the specific command-line tool
+        and its arguments. Reusable subclasses of the \textit{CmdStage} class
+        can be created, or \textit{CmdStage} can be called directly.
+
+        To combine pipelines, the \textit{AddPipeline} function of the
+        \textit{Pipeline} class may be used.
+
+        The base class common to all Pydpiper applications is
+        \textit{AbstractApplication}.  Classes that inherit from
+        \textit{AbstractApplication} are command-line tools, with their options
+        set by the \textit{AbstractApplication} class. The class is also
+        responsible for initializing or resuming the pipeline, setting up a
+        logger, and launching the pipeline daemon.
+
+        Pydpiper atoms inherit from the \textit{CmdStage} class in order to wrap
+        the MINC Tools.  They require as input a MINC file, and may additionally
+        also require a target MINC file as second argument. These arguments can
+        be in the form of a string filename or a file handler, however, the same
+        format must be used for the two. An atom may also accepts various
+        optional arguments. Using the inputs provided, the atom will put
+        together the command to be executed.
+
+        \subsection{Pipeline execution}
+
+        In Pydpiper, there are two main classes,\textit{Pipeline} and
+        \textit{PipelineExecutor}, that monitor and manage pipeline execution.
+        The \textit{Pipeline} class's main responsibility is to infer
+        dependencies between pipeline stages from their inputs and outputs. It
+        also manages a queue of runnable stages (stages in which all their
+        dependencies have been met), ensuring that any runnable stage is placed
+        in the queue. It also keeps track of the status of the runnable stages
+        (running, finished, or failed). Should an already existing stage be
+        added to the pipeline, it is automatically skipped by the
+        \textit{Pipeline} class.
+
+        As stages may be added to incomplete pipelines, Pydpiper provides the
+        functionality of resuming pipelines. This is achieved through a pickling
+        mechanism implemented by the \textit{Pipeline} class. The pickled
+        components include the directed acyclic graph, the array of pipeline
+        stages, the stage counter, the hashes of all stages and each of their
+        output files, and the status of each stage.
+
+        The \textit{PipelineExecutor} class, on the other hand, is responsible
+        for the efficient processing of the pipeline through parallelization.
+        Pydpiper's architecture mimics that of client/server, in which the
+        server is the pipeline daemon started by the command-line executable and
+        the executors undertaking the role of the clients. The pipeline's
+        executors each consist of a thread pool in which each thread is
+        responsible for running a single pipeline stage from the queue of
+        runnable stages. To obtain a stage for execution, the executor polls the
+        server to obtain the next available stage. Should the executor have
+        enough memory and processors to execute the stage, it will be executed
+        by the thread and release memory and processors and repoll the server
+        once stage has completed.  Otherwise, the executors will sleep for
+        certain duration and repoll the server again.  Should there be any
+        failed stages, the pipeline will be shut down after all possible stages
+        have finished executing. If additional executors are required by the
+        pipeline, they may be added at any time through the command-line.
+        Details related to the number of threads, number of threads per executor
+        and the amount of memory can be specified through the command-line
+        options.
+
+        \subsection{Debugging and provenance}
+
+        As mentioned previously, Pydpiper uses a pickling mechanism to gather
+        important pipeline details during execution in order to allow for the
+        pipeline to be resumed should any stages have failed.  In addition, to
+        keep track pipeline stage interdependencies in addition to file
+        versioning, Pydpiper utilizes the helper class
+        \textit{RegistrationPipeFH} and its parent \textit{RegistrationFHBase}
+        to create file handlers of the files.
+
+        \note{from github wiki} For debugging, a Pydpiper pipeline creats its
+        own log file and executors typically also each create their own log file
+        that are stored in a \textit{logs/} subdirectory. Verbosity of logging
+        information can be managed through the setting of the PYRO\_LOGLEVEL
+        environment variable. In addition, individual stage stdout/stderr is
+        also redirected to a log file.
+
+        To verify the status of an executing pipeline, the
+        check\_pipeline\_status.py may be executed.
+
+\chapter{Other domains and general scientific worflow engines} \section{Taverna}
+Taverna is an open-source bioinformatics workflow engine, designed to allow
+users to combine distributed web services, other service types (e.g. BioMart
+queries or R-scripts) and local scripts into complex analysis pipelines. The
+main goal of Taverna is to give a unified access to the plethora of services
+already existing on the web without requiring the user to download the services
+locally or to have knowledge on all of the platforms and programming languages
+the services are hosted under.
+
+        \subsection{Pipeline description} Workflows in Taverna are built by
+        combining web services, local scripts and other services together. Web
+        services compatible with Taverna must comply to the Web Service
+        Description Language (WSDL) using the Simple Object Access Protocol
+        (SOAP protocol).  These services can be added to the BioCatalogue or the
+        Biodiveristy catalogue for sharing and reuse of the web services.
+
+        Workflows can be built in the Taverna workbench. Taverna, through the
+        help of the BioCatalogue and the Biodiversity Catalogue is able to
+        provide the user with a variety of webservices to combine into workflows
+        in a drag-and-drop fashion. Workflows may also be composed of
+        sub-workflows. As service inputs and outputs may not match between
+        services, shims are provided.  Similarly to the services, the
+        user-generated workflows can be uploaded the myExperiment repository to
+        be shared with other user.
+
+        \subsection{Pipeline execution} The workflows can be executed remotely
+        utilizing the Taverna Server. The Taverna Server is a an application
+        that provides the full library of possible workflows through
+        webservices. It has an interface, known as the Taverna Player, that
+        enables workflow execution to occur through a web-interface or through
+        third-party clients. The Taverna Server, however, is limited in that it
+        can only execute workflows available on the Server. Taverna Lite is an
+        alternative to the Taverna Server which allows users to upload and
+        execute workflows from myExperiment or elsewhere.
+
+        \subsection{Debugging and provenance} Provenance capture is achieved
+        through the Taverna Provenance Suite. Taverna does not capture
+        provenance of the workflow definition, but generates a new UUID after
+        each definition change. Workflow run provenance captures information on
+        "individual processor iterations" and their respective inputs and
+        outputs and stores this information in a database. The results can then
+        be exported as a workflow trace or as a PROV-O RDF graph using the
+        Taverna-PROV Plugin. The model format used for exported provenance uses
+        both the W3C PROV model and Open Provenance Model format.
+
+    \section{Pegasus} Pegasus is a general-purpose scientific workflow engine
+    that specializes in mapping abstract descriptions of workflows to resources
+    for efficient processing.  Similarly to other workflows, a Pegasus workflow
+    can be represented by a DAG. This DAG is described by the user in DAX format
+    (Directed Acyclic graph in XML).  As the DAX abstract workflow is
+    resource-independent, Pegasus can convert the abstract workflow into an
+    execution workflow that is catered to the resources available.
+
+        \subsection{Pipeline description} Within the DAX, the nodes are the
+        tasks and the execution order is represented by the edges.  In the
+        instance of nested graphs, the nodes in the DAX may also represent a
+        DAX. Pipeline inputs and outputs, as well as pipeline executables, are
+        represented by logical identifiers.
+
+         Various APIs are provided by Pegasus in order for the user to generate
+         the DAX using their preferred language of choice. Available APIs
+         include Perl, Python and Java.  \subsection{Pipeline execution} With
+         the effecient execution of pipelines being Pegasus' main focus, there
+         exist various components that ensure optimal processing. The Mapper's
+         responsability is to map abstract workflow tasks to available
+         resources. In order for the Mapper to successfully convert an abstract
+         workflow into an execution workflow, it requires details on the
+         execution environment. The Mapper will also query various catalogs to
+         determine the location of executables and input and output files. These
+         catalogs can be either file-based or in database format. With all the
+         necessary information provided, the Mapper is equipped to convert
+         abstract tasks into executable tasks. Data nodes are also added for
+         input and output staging. Should the data required for a task already
+         exist in the catalog, the task and all its parents will be removed by
+         the Mapper from the workflow. This functionality is optional, and in
+         some instances, it may be found that recomputation is preferable. If
+         the executables required by the task are not pre-installed on the node,
+         the Mapper will query the \textit{Transformation Catalog} for the
+         stageable binaries.
+
+        The default, Pegasus relies on the HTCondor framework as both the
+        workflow engine and scheduler. The executable workflow, using HTCondor
+        DAGMan, consists of nodes representing jobs.  These jobs each have a
+        corresponding job submit file that details how they are meant to be
+        executed. A job that is ready for execution is released by the DAG and
+        placed into a Condor queue that is managed by the HTCondor Schedd
+        daemon. The default scheduling is done using a local HTCondor pool, but
+        can also be achieved through built-in interfaces with remote
+        job-management protocols, or through SSH to directly submit jobs to
+        remote resource managers.
+
+        Using HTCondor, Pegasus is able to achieve portability. Pegasus is
+        compatible with local, HPC, distributed and cloud execution
+        environments. As distributed environments may not employ a shared
+        filesystem, it is recommended to use HTCondor glideins to ensure that
+        tasks are mapped to nodes where the resources are located. Should a user
+        running their pipeline locally without HTCondor installed, it is
+        possible through \textit{submit host}. Using this method, however,
+        executes the pipeline in serial order. Should the user wish to exploit
+        the pipeline's parallelism, HTCondor is required. Pegasus pipelines are
+        not limited to a single execution environment, and can combine multiple
+        environments to select the best resource configuration to execute each
+        job.
+
+        In order to manage tasks and their dependencies on remote resources,
+        three different execution engines have been developed. The first and
+        simplest execution engine is the \textit{Pegasus Cluster}, which is
+        compatible with all the execution environments. It runs on a
+        single-thread, meaning that it can only executed a single task at a time
+        using an ordered-list of tasks.  The \textit{Pegasus Cluster} is
+        particularly useful for the efficient execution of short tasks and the
+        reduction of data transfers.
+
+        For execution environments lacking a shared filesystem, the
+        \textit{PegasusLite} execution engine can be used. In order for the
+        Pegasus to execute in such environments, the Mapper first adds data
+        transfer nodes to the pipeline to move the data to a data staging server
+        associated with the execution node. Once a job is assigned to a node by
+        the remote cluster scheduler, \textit{PegasusLite} is deployed.
+
+        The final execution engine provided by Pegasus is the \textit{Pegasus
+        MPI Cluster} (PMC). This execution engine is designed for petascale
+        systems for which Pegasus' other strategies leveraging Condor Glideins
+        are unsuitable. It employs the Master-Worker architecture, launching a
+        single Master and many Workers, and using MPI for communication between
+        the Master and Workers. The main limitation of this strategy is that a
+        task must use less resources than that available on a single node. Jobs
+        are not limited to a single execution engine, and thus, multiple can be
+        used in a job to optimize performance.
+
+        As executing a workflow on remote resources can be slow due to job
+        queuing times and data transfers, the Mapper will group short-running
+        task together to reduce to improve runtime.  When tasks located at the
+        same distance from the DAG root have similar runtimes, level-based
+        clustering can be used. When runtimes of same-level tasks may vary, it
+        may be preferable to use job runtime-based horizontal clustering, to
+        achieve load-balancing of clusters. Alternatively, the user may opt for
+        label-based clustering, in which the user can define the task clusters
+        themselves.
+
+        To avoid running out of disk space during data-intensive workflows, the
+        Mapper can add cleanup nodes that remove all unecessary data after a job
+        has completed.
+
+        \note{Dynamic workflows permitted}
+
+        \subsection{Debugging and provenance} Job monitoring, in Pegasus, is
+        achieved through \textit{pegasus-kickstart} (kickstart). Every job
+        within a pegasus workflow has a kickstart wrapper that collects runtime
+        provenance and performance. Kickstart relays this information to the
+        submit host after job completion. This data is ultimately collected and
+        stored in a relational database.
+
+        The Mapper Logs contain details on the abstract workflow, the jobs in
+        the executable workflow and the mapping of tasks to jobs. Using the
+        mapper logs, Pegasus can generate the provenance of the results. Should
+        it be required, Pegasus can also generate the provenance of the mapping
+        process.
+
+        In addition to the Mapper Logs, there also exists the DAGMan logs,
+        generated by HTCondor DAGMan and updated in near real-time. These logs
+        capture details on the status of the jobs and their associated pre- and
+        post- execute scripts.
+
+        All provenance and debugging data is accessibly through either the
+        \textit{pegasus-analyzer} tool or through the provided lightweight web
+        interface.
+
+           \section{Make} \subsection{Summary} As the creation of workflow
+           activities requires a certain level of programming expertise, Make
+           was proposed as a simpler alternative. Make is a tool used for the
+           development of directed acyclic dependency graphs. It is available in
+           all Linux and UNIX distributions, which relieves the users from
+           worflow engine installation. Additionally, Make is capable of
+           offering other features necessary to workflow engines, such as,
+           reproducibility, parallelization, fault-tolerance and quality
+           control. The motivation behind Make is essentially that the tools
+           available for simple and efficient pipeline creation already exist
+           and there is no need to 'reinvent the wheel'. In addition, the use of
+           Make is not limited to the neuroimaging community, and thus support
+           for it is widely available. Make has also been adapted for use in
+           other programming languages such as SnakeMake for python
+
+        \subsection{Pipeline description} Make pipelines are described in
+        makefiles, which consist of a set of rules which describe the flow of
+        data. Rules defined in make can be define by three components: targets,
+        dependecies and recipes. The target is the output (typically file)
+        produced by execution of the recipe, whereas the dependecy would be the
+        input file from which the target would be created from. The target is
+        only modified if its last modified date is early than that of a
+        depencies's.  A recipe, in make, is the set of shell commands used to
+        create the target from its dependencies.
+
+In order connect the rules together, it is possible to define "phony" targets.
+Phony targets do not correspond to any actual file and will be generated once
+all of their dependencies (targets of other rules) have been created. Chaining
+multiple rules together will automatically generate the directed acyclic graph
+that is the pipeline. This is contrary to other pipelines such as Nipype and
+LONI, in which the user must explicitly state all connections \note{as stated in
+paper}. Through the use of rules, it is easily possible to reexecute only the
+desired parts of the pipeline.
+
+Cleanup after rule execution can be defined using the phony target clean.
+
+To create different targets for the same dependency files organized in different
+ways (e.g. subject/timpepoint timepoint/session), it is possible to use symbolic
+links for the dependencies, however this may complexify the code. Otherwise, it
+may be simpler to copy subject and session makefiles.
+
+        \subsection{Pipeline execution}
+
+        Compared to the previous workflow engines, Make pipeline variables are
+        only evaluated at time of use (lazy evaluation). In addition, targets
+        are not executed in the order they are specified (similarly to other
+        pipeline engines).
+
+Make files are also well-equipped to handle workflows with conditionals. Should
+a conditional exist in a workflow, it will be evaluated during the creation of
+the DAG (i.e. not at runtime) and only the required components will be executed.
+
+        As parallelism can be inferred from the dependency graph, parallel
+        execution of a make pipeline can occur simply through the addition of
+        the -j flag for multicore machines, or by using qmake command to run the
+        pipeline on a cluster.
+
+        By default, make only rebuilds targets whose last modified preceedes
+        that of its dependencies. As such, when a task fails and is re-executed,
+        only the targets that need to be rebuilt will be rebuilt. It is possible
+        to specify reexcution in Make such that the task will be reexecuted in
+        case of failure.
+
+        \subsection{Debugging and provenance} As users of a preexisting Make
+        pipeline may not be aware of the proper usage of each target, it is
+        necessary to provide a `help` module that is adaptable to targets being
+        added removed or modified. It is possible to define in Make a macro for
+        such a purpose that will provide users with usage information when
+        requested (e.g. calling make help).
+
+    Debugging in Make can be rather difficult, particularly when executing code
+    in parallel. However, pipelines in Make are broken down into many targets
+    which can each be tested individually. In addition, Make provides built-in
+    flags to aid in debugging. These flags include -n, for printing out the
+    commands without any execution and -p to print out all the rules. In the
+    instance where pattern substistution is used, -p will also include the
+    results obtained from the pattern substitution. More recent versions of GNU
+    Make also include the --trace and --output-sync flags which display why each
+    recipe was executed and print the parallelized output in sequential order,
+    respectively.
+
+    Neuroimaging tools are typically equipped with quality assurance modules. As
+    a result, it is possible to utilize the quality assurance modules from these
+    tools to provide a complete report for the entire pipeline. The proposed
+    solution suggests to write all the quality assurance outputs from each tool
+    to a single HTML report with R Markdown or a script. The R Markdown or
+    script, with the metrics and images are listed as dependencies to the QA
+    report target in Make.
+
+    Only a basic form of data provenance has been implemented by the authors.
+    The implemented data provenance consisted of a makefile that calls a bash to
+    extract acquisition parameters from the raw input images. The makefile also
+    obtains tool versions from the installation directories and pipeline options
+    from their respective makefiles. All of these details are stored in a CSV
+    file.  The makefile will then call an R Markdown script which will then
+    generate an HTML provenance report from the corresponding CSV file.
+
+         \section{OpenMOLE} \subsection{summary} OpenMOLE (Open MOdeL
+         Experiment) is a domain agnostic pipelining framework that focuses on
+         simplifying workflow creation through its Domain Specific Language
+         (DSL) built on top of Scala. Through the use of its DSL, much of the
+         backend is abstracted from users. It also offers a GUI for users to
+         build workflows. Similarly to other existing pipelining frameworks,
+         OpenMOLE enables parallelization through plugins, meaning that the
+         workflow need not be changed between different distribution
+         environments.  Moreover, OpenMOLE supports the deployment of tasks or
+         groups of tasks on different computing environments.  Another key
+         feature of OpenMOLE is that it attempts to address the reproducibility
+         issue through the integration of the CARE archiver. CARE allows for the
+         execution of command-line applications, written in any language, and
+         attempts to simulate the environment in which the application was
+         designed on.  OpenMOLE also provides embedded methods for parameter
+         exploration.
+
+        \subsection{Pipeline description} An OpenMOLE workflow consists of tasks
+        connected by transitions. Each task contains details such as input,
+        output and optional parameters, in addition to embedding the application
+        that will be applied to the data. Input values are generally supplied by
+        the dataflow through an entry point known as \textit{Sources} and output
+        values are returned to the dataflow through \textit{Hooks}.
 
       Variables in OpenMOLE can be native types (e.g. int, string), files,
-directories, any Java or Scala types, and user-defined types (i.e. Java or Scala
-classes). The output type of an input supplied to another task should be the
-same as the input type, or be in a child class of the input type.
+      directories, any Java or Scala types, and user-defined types (i.e. Java or
+      Scala classes). The output type of an input supplied to another task
+      should be the same as the input type, or be in a child class of the input
+      type.
 
       There exist five types of transitions in OpenMOLE: flat, divergent,
-convergent, explorative and aggregative. A flat transition is one in which the output of one
-task is the input to a single other task. On the other hand, a divergent
-transition is one in which the output of one task serves as the input to several
-other tasks. The inverse of a divergent task is a convergent task, in which
-several tasks supply input to a single task. Explorative transitions allow for
-multiple excutions of the same task on the sampling outputs produced by an
-exploration task. Conversly, an aggregative transition gathers all the streams
-produced by an explorative transition. It is also possible to add a stopping
-condition to terminate all streams once a condition is satisfied.
+      convergent, explorative and aggregative. A flat transition is one in which
+      the output of one task is the input to a single other task. On the other
+      hand, a divergent transition is one in which the output of one task serves
+      as the input to several other tasks. The inverse of a divergent task is a
+      convergent task, in which several tasks supply input to a single task.
+      Explorative transitions allow for multiple excutions of the same task on
+      the sampling outputs produced by an exploration task. Conversly, an
+      aggregative transition gathers all the streams produced by an explorative
+      transition. It is also possible to add a stopping condition to terminate
+      all streams once a condition is satisfied.
 
@@
 
-    There also exists the notion of \textit{Input Slots} in OpenMOLE, which enables
-cycles in the workflow. A task may have several input slots and will execute
-each time that all the transitions belonging to the same input slots has passed
-through.
+    There also exists the notion of \textit{Input Slots} in OpenMOLE, which
+    enables cycles in the workflow. A task may have several input slots and will
+    execute each time that all the transitions belonging to the same input slots
+    has passed through.
 
@@
       In order for an OpenMOLE workflow to be executed, there must exist a
-single encapsulated task. A task will execute once all its dependencies have
-been met.
+      single encapsulated task. A task will execute once all its dependencies
+      have been met.
 
     During execution of a task, the CARETask will decompress the CARE archive
-containing the application and its dependencies and the application is deployed
-within the confines of the decompressed archive.
+    containing the application and its dependencies and the application is
+    deployed within the confines of the decompressed archive.
 
@@
       OpenMOLE workflows can be executed on a variety of platforms such as
-remote servers, clusters and compute grids. As these execution hosts may not have all
-the required dependencies to execute a task, tasks are archived using CARE. This
-ensures that all the required applications and their dependencies can be shipped
-and executed at runtime on any Linux host.
-
-        \subsection{Debugging and Provenance}
-    A CARETasks standard output and standard error are normally printed out to
-OpenMOLES's console. For distributed executions, this might not be a practical
-solution, in which case such information may be passed through
-the dataflow using OpenMOLE variables.
-
-    \section{LONI}
-        \subsection{Summary}
-        The LONI pipeline processing environment facilitates the development,
-sharing and execution of pipelines. Workflow Activities, known as \textit{Modules} in
-LONI, are described using Extensible Markup Language (XML), which can easily
-shared between user groups. LONI runs on the Java platform, allowing it to be
-platform independent and has support DRMAA clusters and symmetric
-multiprocessing (SMP) systems. In order to
-capture provenance information, LONI is also equipped with a provenance manager.
-
-        \subsection{Pipeline description}
-        LONI uses the dataflow programming model to represent the pipeline. Data
-flows between \textit{Modules} through \textit{data pipes}. Sub-pipelines are
-known and \textit{Pipelets}. Modules are defined by the input arguments/streams
-and output arguments/stream. Arguments in LONI may be specific
-filetypes, flags or values listed in the order in which the application
-requires. The provided arguments may be required or optional and can also be
-duplicated, if necessary. File inputs in the form of URLs are accepted by LONI
-and will be downloaded at runtime. List of inputs may also be provided to process
-multiple inputs with copies of the same module in parallel (batch processing). Conversely, it is
-also possible to group list of inputs so as to process them all by the single
-module \note{Need to express this better}. Input/Output arguments may be added
-to the table of global variables for easy access throughout the pipeline. It is
-possible to apply variable modifiers to the variables stored within the table.
-
-Intermediary outputs as well as the file output are treated as temporary by
-LONI. They are scheduled for deletion before a new pipeline is executed. It is,
-however, possible to preserve these files.
-
-        \subsection{Pipeline execution}
-        As a
-module completes it execution, its output gets transmitted through the dataflow
-and the dependent modules are signaled that their input data has arrived.
-Modules who have met all their dependency requirements are ready for execution.
-Should a module have no dependecies, it can be executed immediately. Modules
-ready to be executed are placed in a FIFO launch queue to avoid straining the
-platform.
-
-       LONI offers a built in client/server strategy that uses Java's remote
-method invocation (RMI) libraries to communicate between servers. This enables
-modules, pipelets or pipelines to be executed on servers different from the
-client. It also enables full pipelines to consist of connected modules and
-pipelets located on different servers. It is favorable to have such a
-configuration in LONI, as the tools required for processing may not be available
-on client server, but located in another. The client is responsible for
-dictating all file transfers and module launches that are to occur. The client
-authenticates with the servers through a one-way hash. Only files produced by
-the pipeline are accessible to the client and only temporary files created by
-the server may be written to the server.
-
-        \subsection{Debugging and provenance}
-        For each dataset processed in a LONI pipeline, there exists a corresponding
-provenance (.prov) XML file. The \textit{Provenance Editor}, a standalone
-platform-independent application, reads the metadata from the image headers,
-extracts relevant provenance information from them and creates the provenance files. 
-This data can be modified by the user in the to correct inaccuracies or append
-additional details before being saved. During pipeline execution, the provenance
-file will be edited by the LONI pipeline to include details on each processing
-step, the environment in which the data was processed and tool binaries used.
-
-    \section{Automatic Analysis}
-        \subsection{Summary}
-        Automatic analysis (aa) is another Matlab-based modular workflow engine,
-having started of as a workflow engine for SPM. User may define their pipeline
-through a Matlab user script and specify the tasks in addition to their input
-and output types through an XML module interface. Modules used in pipelines must
-be integrated into aa which can be easily accomplished through XML interfaces.
-aa workflows can be executed locally, or in parallel through qsub or condor.
-
-        \subsection{Pipeline description}
-        aa pipelines have two main components: 1) the Matlab user script and 2)
-the XML tasklist. The user script is a script that defines which data will be
-processed, any necessary parameter, and how it will be processed. aa provides an example user script,
-which can be modified to meet user's needs. The entire pipeline in aa is defined
-in a single structure, known as aap, which makes the task of record-keeping
-simple.
-
-        The XML tasklist is an ordered list of all the modules that will be applied to
-the data, that is read by the user script. This list does not specify any
-dependencies as those dependendecies can be inferred by the pipeline through
-analysis of the module interfaces. Modules can be separated into two
-sections: initialization and main. Modules listed in the initialization section
-are to be executed for every task. Conversely, those modules liste in the main
-section are only executed once for every task, unless otherwise specified.
-
-        In order to ensure that the correct output data is being operated on,
-each task is provided with its own unique folder. The aa engine will transmit
-the necessary input data to the folder before task execution and also
-acquiring the outputs after task completion. Output folders are numbered such
-that if a given module is invoked multiple times, each of those folders are
-assigned an incremental ordering corresponding to the order in which they have been
-executed \note{sentence might not make sense}. Should all data be required for
-each task in the tasklist, all the data will be copied to each task folder.
-
-    Similarly to a pipeline, each module is comprised of an XML interface and a
-Matlab source file. A module's XML file consist of specifications on required input
-and output modalities, as well as the desired domain. One of the most important
-properties include the domain at which the module operates. While there exist mainy domains for
-modules and new ones can easily be created and integrated by the user, the most
-prominently used ones include: "study", "subject", "session". As their name suggests, a
-module with the domain "study" will only be executed once for each time it
-appears in the tasklist, whereas "subject" executes the module for each subject
-within the task and "session" executes once for each session within a subject.
-
-    In addition to module domains, another important module property is the
-input and output data stream types. Data stream types denote what input and
-output data type each module produces. Commonly used data stream
-types include "epi", "structural" and "dicom\_header".
-
-    Optional fields specified in a module include qsub, for estimates of
-requirements for parallel processing,
-and \textit{performanceoutput}, for garbage collection.
-
-   Analysis parameters can easily be configured in aa. Any variable defined by
-the \textit{aaprecipe} command may be modified. Parameters can also be defined
-for an instance of a module within the XML tasklist. Alternatively, it is also
-possible to create an XML file that inherits parameters from another XML file
-and overrides some or all of the parameters.
-
-   To set defaults and site-specific configurations, it is possible to either
-create a local config file or by creating another parameter default file which
-inherits the original one.
-
-    aa also supports the creation of branched tasklist. This allows a section of
-the pipeline steps to undergo different tasks or the same tasks in a different
-order in one pipeline rather than having to create separate pipelines.
-
-        \subsection{Pipeline execution}
-        Once the definition of the user-script, XML tasklist and modules is
-complete, the pipeline is ready for execution. Dependencies in aa are
-established based on input and output data stream types. The aa engine attempts
-to match the last task executed with a corresponding data stream type to the
-current type. It is, however, possible to specify which module provides the
-input to the current module by using fully qualified stream references.
-
-        During execution, a map of all the modules that need to be executed and
-their respective dependencies is generated. Modules that have completed contain
-a flag file indicating their completion within their instance's root directory. Such
-modules will not be re-executed by the engine.
-Additionally, if it is required that a previous module is re-executed, the done flag is
-removed from all dependent modules.
-
-        Should part of the processing be performed on a different server it is
-possible to connect the pipelines together through an analysis script. The aa engine will be reponsible for
-matching the data streams between the servers in order to avoid the copying of
-data and modules.
-
-        \subsection{Debugging and provenance}
-        In order to retrace the steps of previous executions, aa keeps a
-detailed record of the events that occurred. These detailes are stored in a
-Matlab object \note{similarly to PSOM}, which permits reexcution of a previous
-run or extraction of specific parameter values.
-
-        Additionally, to verify the correctness of the results produced, aa
-implements tools for quality control and comes with a dedicated module for
-low-level quality control.
-
-    \section{Make}
-        \subsection{Summary}
-        As the creation of workflow activities requires a certain level of
-programming expertise, Make was proposed as a simpler alternative. Make is a
-tool used for the development of directed acyclic dependency graphs. It is
-available in all Linux and UNIX distributions, which relieves the users from
-worflow engine installation. Additionally, Make is capable of offering other
-features necessary to workflow engines, such as, reproducibility,
-parallelization, fault-tolerance and quality control. The motivation behind Make
-is essentially that the tools available for simple and efficient pipeline
-creation already exist and there is no need to 'reinvent the wheel'. In
-addition, the use of Make is not limited to the neuroimaging community, and thus
-support for it is widely available. Make has also been adapted for use in other
-programming languages such as SnakeMake for python
-
-        \subsection{Pipeline description}
-        Make pipelines are described in makefiles, which consist of a set of
-rules which describe the flow of data. Rules defined in make can be define by three
-components: targets, dependecies and recipes. The target is the output
-(typically file) produced by execution of the recipe, whereas the dependecy
-would be the input file from which the target would be created from. The target
-is only modified if its last modified date is early than that of a depencies's.
-A recipe, in make, is the set of shell commands used to create the target from
-its dependencies.
-
-In order connect the rules together, it is possible to define "phony" targets.
-Phony targets do not correspond to any actual file and will be generated once
-all of their dependencies (targets of other rules) have been
-created. Chaining multiple rules together will automatically generate the directed acyclic
-graph that is the pipeline. This is contrary to other pipelines such as Nipype
-and LONI, in which the user must explicitly state all connections \note{as
-stated in paper}. Through the use of rules, it is easily possible to reexecute
-only the desired parts of the pipeline.
-
-Cleanup after rule execution can be defined using the phony target
-clean.
-
-To create different targets for the same dependency files organized in different
-ways (e.g. subject/timpepoint timepoint/session), it is possible to use symbolic
-links for the dependencies, however this may complexify the code. Otherwise, it
-may be simpler to copy subject and session makefiles.
-
-        \subsection{Pipeline execution}
-
-        Compared to the previous workflow engines, Make pipeline variables are
-only evaluated at time of use (lazy evaluation). In addition, targets are not
-executed in the order they are specified (similarly to other pipeline engines).
-
-Make files are also well-equipped to handle workflows with conditionals. Should
-a conditional exist in a workflow, it will be evaluated during the creation of
-the DAG (i.e. not at runtime) and only the required components will be executed.
-
-        As parallelism can be inferred from the dependency graph, parallel execution of a make pipeline can occur simply through the
-addition of the -j flag for multicore machines, or by using qmake command to run
-the pipeline on a cluster.
-
-        By default, make only rebuilds targets whose last modified preceedes
-that of its dependencies. As such, when a task fails and is re-executed, only the targets that need
-to be rebuilt will be rebuilt. It is possible to specify reexcution in Make such
-that the task will be reexecuted in case of failure.
-
-        \subsection{Debugging and provenance}
-        As users of a preexisting Make pipeline may not be aware of the proper
-usage of each target, it is necessary to provide a `help` module that is
-adaptable to targets being added removed or modified. It is possible to define
-in Make a macro for such a purpose that will provide users with usage
-information when requested (e.g. calling make help).
-
-    Debugging in Make can be rather difficult, particularly when executing code
-in parallel. However, pipelines in Make are broken down into many targets which
-can each be tested individually. In addition, Make provides built-in flags to
-aid in debugging. These flags include -n, for printing out the commands without
-any execution and -p to print out all the rules. In the instance where pattern
-substistution is used, -p will also include the results obtained from the
-pattern substitution. More recent versions of GNU Make also include the --trace
-and --output-sync flags which display why each recipe was executed and print the
-parallelized output in sequential order, respectively.
-
-    Neuroimaging tools are typically equipped with quality assurance modules. As
-a result, it is possible to utilize the quality assurance modules from these
-tools to provide a complete report for the entire pipeline. The proposed
-solution suggests to write all the quality assurance outputs from each tool to a single HTML
-report with R Markdown or a script. The R Markdown or script, with the metrics
-and images are listed as dependencies to the QA report target in Make.
-
-    Only a basic form of data provenance has been implemented by the authors.
-The implemented data provenance consisted of a makefile that calls a bash to
-extract acquisition parameters from the raw input images. The makefile also
-obtains tool versions from the installation directories and pipeline options
-from their respective makefiles. All of these details are stored in a CSV file.
-The makefile will then call an R Markdown script which will then generate an
-HTML provenance report from the corresponding CSV file.
-
-    \section{Pydiper}
-        \subsection{Summary}
-
-        Pydiper is a neuroimaging registration workflow engine implemented in
-Python. This differs from other workflow engines in that it is specific to the
-process of image registration, but can, however, be extended for use with any
-analysis process. Its main features include being easy to use, particularly for
-non-developpers, undertaking a modular approach, giving the ability to access
-output files at any stage, and eliminating duplicate stages. Pydiper also
-provides the added advantage of being equipped with four commonly used
-registration pipelines, such that unexperienced researchers can use such
-pipelines out-of-the-box.
-
-        \subsection{Pipeline description}
-        Pydiper workflows can be broken down into three components: atoms,
-modules and applications. An application is the command line application that
-encapsulates the pipeline which consists of modules and atoms. Atoms are wrappers for individual operations,
-typically frquently used MINC tools. Modules are reponsible for linking
-atoms together.
-
-        Pydpiper provides three main classes to use to develop pipelines. These
-include the \textit{PipelineStage}, \textit{CmdStage} and \textit{Pipeline} classes. The
-\textit{PipelineStage} is the base class wrapper for all pipeline tools.
-Pipeline tools can be written in Python or can be a command-line application.
-For the tools that are command-line applications, the \textit{CmdStage} class, a
-child class of \textit{PipelineStage}, must be used. Input to \textit{CmdStage}
-is in a adequately parsed array format containing the specific command-line
-tool and its arguments. Reusable subclasses of the \textit{CmdStage} class can be
-created, or \textit{CmdStage} can be called directly.
-
-        To combine pipelines, the \textit{AddPipeline} function of the \textit{Pipeline}
-        class may be used.
-
-        The base class common to all Pydpiper applications is \textit{AbstractApplication}. 
-Classes that inherit from \textit{AbstractApplication} are command-line tools, with their 
-options set by the \textit{AbstractApplication} class. The class 
-is also responsible for initializing or resuming the pipeline, setting up a logger, and 
-launching the pipeline daemon.
-
-        Pydpiper atoms inherit from the \textit{CmdStage} class in order to wrap the MINC Tools. 
-They require as input a MINC file, and may additionally also require a target MINC file as second argument. These arguments
-can be in the form of a string filename or a file handler, however, the same format must be used for the two.
-An atom may also accepts various optional arguments.
-Using the inputs provided, the atom will put together the command to be executed.
-
-        \subsection{Pipeline execution}
-
-        In Pydpiper, there are two main classes,\textit{Pipeline} and \textit{PipelineExecutor}, 
-that monitor and manage pipeline execution. The \textit{Pipeline} class's main responsibility
-is to infer dependencies between pipeline stages from their inputs and outputs. It also manages
-a queue of runnable stages (stages in which all their dependencies have been met), ensuring
-that any runnable stage is placed in the queue. It also keeps track of the status of the runnable
-stages (running, finished, or failed). Should an already existing stage be added to the pipeline, it
-is automatically skipped by the \textit{Pipeline} class.
-
-        As stages may be added to incomplete pipelines, Pydpiper provides the functionality of resuming
-pipelines. This is achieved through a pickling mechanism implemented by the \textit{Pipeline} class. The
-pickled components include the directed acyclic graph, the array of pipeline stages, the stage counter,
-the hashes of all stages and each of their output files, and the status of each stage.
-
-        The \textit{PipelineExecutor} class, on the other hand, is responsible for the efficient processing
-of the pipeline through parallelization. Pydpiper's architecture mimics that of client/server, in which the 
-server is the pipeline daemon started by the command-line executable and the executors undertaking the role of 
-the clients. The pipeline's executors each consist of a thread pool in which each
-thread is responsible for running a single pipeline stage from the queue of runnable stages. To
-obtain a stage for execution, the executor polls the server to obtain the next available stage. Should
-the executor have enough memory and processors to execute the stage, it will be executed by the thread
-and release memory and processors and repoll the server once stage has completed.
-Otherwise, the executors will sleep for certain duration and repoll the server again. 
-Should there be any failed stages, the pipeline will be shut down after all possible stages have finished 
-executing. If additional executors are required by the pipeline, they may be added at any time through
-the command-line. Details related to
-the number of threads, number of threads per executor and the amount of memory can be specified through the
-command-line options.
-
-        \subsection{Debugging and provenance}
-
-        As mentioned previously, Pydpiper uses a pickling mechanism to gather important pipeline details
-        during execution in order to allow for the pipeline to be resumed should any stages have failed.
-        In addition, to keep track pipeline stage interdependencies in addition to file versioning, Pydpiper
-        utilizes the helper class \textit{RegistrationPipeFH} and its parent \textit{RegistrationFHBase} to create
-        file handlers of the files.
-
-        \note{from github wiki}
-        For debugging, a Pydpiper pipeline creats its own log file and executors typically also each create 
-        their own log file that are stored in a \textit{logs/}
-        subdirectory. Verbosity of logging information can be managed through the setting of the PYRO\_LOGLEVEL
-        environment variable. In addition, individual stage stdout/stderr is also redirected to a log file.
-
-        To verify the status of an executing pipeline, the check\_pipeline\_status.py may be executed.
-
-\chapter{Other domains and general scientific worflow engines}
-    \section{Taverna}
-    Taverna is an open-source bioinformatics workflow engine, designed to allow users to combine distributed web services, other service
-    types (e.g. BioMart queries or R-scripts) and local scripts into complex analysis pipelines. The main goal of
-    Taverna is to give a unified access to the plethora of services already existing on the web without requiring
-    the user to download the services locally or to have knowledge on all of the platforms and programming languages
-    the services are hosted under.
-
-        \subsection{Pipeline description}
-        Workflows in Taverna are built by combining web services, local scripts and other services together. Web services compatible with
-        Taverna must comply to the Web Service Description Language (WSDL) using the Simple Object Access Protocol (SOAP protocol). 
-        These services can be added to the BioCatalogue or the Biodiveristy catalogue for sharing and reuse of the web services.
-
-        Workflows can be built in the Taverna workbench. Taverna, through the help of the BioCatalogue and the Biodiversity Catalogue is
-        able to provide the user with a variety of webservices to combine into workflows in a drag-and-drop fashion. Workflows may also be composed
-        of sub-workflows. As service inputs and outputs may not match between services, shims are provided.
-        Similarly to the services, the user-generated workflows can be uploaded the myExperiment repository to be shared with other user.
-
-        \subsection{Pipeline execution}
-        The workflows can be executed remotely utilizing the Taverna Server. The Taverna Server is a an application that provides the full library of possible workflows
-        through webservices. It has an interface, known as the Taverna Player, that enables workflow execution to occur through a web-interface or through
-        third-party clients. The Taverna Server, however, is limited in that it can only execute workflows available on the Server. Taverna Lite is an alternative to
-        the Taverna Server which allows users to upload and execute workflows from myExperiment or elsewhere.
-
-        \subsection{Debugging and provenance}
-         Provenance capture is achieved through the Taverna Provenance Suite. Taverna does not capture provenance of the workflow definition, but generates a new
-        UUID after each definition change. Workflow run provenance captures information on "individual processor iterations" and their respective inputs and outputs and
-        stores this information in a database. The results can then be exported as a workflow trace or as a PROV-O RDF graph using the Taverna-PROV Plugin. The model format used
-        for exported provenance uses both the W3C PROV model and Open Provenance Model format.
-
-    \section{Pegasus}
-        Pegasus is a general-purpose scientific workflow engine that specializes in mapping abstract descriptions of workflows to resources for efficient processing. 
-        Similarly to other workflows, a Pegasus workflow can be represented by a DAG. This DAG is described by the user in DAX format (Directed Acyclic graph in XML).
-        As the DAX abstract workflow is resource-independent, Pegasus can convert the abstract workflow into an execution workflow that is catered to the resources available.
-
-        \subsection{Pipeline description}
-        Within the DAX, the nodes are the tasks and the execution order is represented by the edges.
-        In the instance of nested graphs, the nodes in the DAX may also represent a DAX. Pipeline inputs and outputs, as well as pipeline executables,
-         are represented by logical identifiers.
-
-         Various APIs are provided by Pegasus in order for the user to generate the
-        DAX using their preferred language of choice. Available APIs include Perl, Python and Java. 
-        \subsection{Pipeline execution}
-        With the effecient execution of pipelines being Pegasus' main focus, there exist various components that ensure optimal processing. The Mapper's responsability is to map
-        abstract workflow tasks to available resources. In order for the Mapper to successfully convert an abstract workflow into an execution workflow, it requires details on the execution
-        environment. The Mapper will also query various catalogs to determine the location of executables and input and output files. These catalogs can be either file-based or in
-        database format. With all the necessary information provided, the Mapper is equipped to convert abstract tasks into executable tasks. Data nodes are also added for input and output
-        staging. Should the data required for a task already exist in the catalog, the task and all its parents will be removed by the Mapper from the workflow. This functionality is optional,
-        and in some instances, it may be found that recomputation is preferable. If the executables required by the task are not pre-installed on the node, the Mapper will query the 
-        \textit{Transformation Catalog} for the stageable binaries.
-
-        The default, Pegasus relies on the HTCondor framework as both the workflow engine and scheduler. The executable workflow, using HTCondor DAGMan, consists of nodes representing jobs.
-        These jobs each have a corresponding job submit file that details how they are meant to be executed. A job that is ready for execution is released by the DAG and placed into a Condor
-        queue that is managed by the HTCondor Schedd daemon. The default scheduling is done using a local HTCondor pool, but can also be achieved through built-in interfaces with remote 
-        job-management protocols, or through SSH to directly submit jobs to remote resource managers.
-
-        Using HTCondor, Pegasus is able to achieve portability. Pegasus is compatible with local, HPC, distributed and cloud execution environments. As distributed environments may
-        not employ a shared filesystem, it is recommended to use HTCondor glideins to ensure that tasks are mapped to nodes where the resources are located. Should a user running their
-        pipeline locally without HTCondor installed, it is possible through \textit{submit host}. Using this method, however, executes the pipeline in serial order. Should the user wish to exploit
-        the pipeline's parallelism, HTCondor is required. Pegasus pipelines are not limited to a single execution environment, and can combine multiple environments to select the best resource
-        configuration to execute each job.
-
-        In order to manage tasks and their dependencies on remote resources, three different execution engines have been developed. The first and simplest execution engine is the 
-        \textit{Pegasus Cluster}, which is 
-        compatible with all the execution environments. It runs on a single-thread, meaning that it can only executed a single task at a time using an ordered-list of tasks.
-        The \textit{Pegasus Cluster} is particularly useful for the efficient execution of short tasks and the reduction of data transfers.
-
-        For execution environments lacking a shared filesystem, the \textit{PegasusLite} execution engine can be used. In order for the Pegasus to execute in such environments, the Mapper first
-        adds data transfer nodes to the pipeline to move the data to a data staging server associated with the execution node. Once a job is assigned to a node by the remote cluster scheduler,
-        \textit{PegasusLite} is deployed.
-
-        The final execution engine provided by Pegasus is the \textit{Pegasus MPI Cluster} (PMC). This execution engine is designed for petascale systems for which Pegasus' other strategies 
-        leveraging Condor Glideins are unsuitable. It employs the Master-Worker architecture, launching a single Master and many Workers, and using MPI for communication between the 
-        Master and Workers. The main limitation of this strategy is that a task must use less resources than that available on a single node. Jobs are not limited to a single execution engine,
-        and thus, multiple can be used in a job to optimize performance.
-
-        As executing a workflow on remote resources can be slow due to job queuing times and data transfers, the Mapper will group short-running task together to reduce to improve runtime. 
-        When tasks located at the same distance from the DAG root have similar runtimes, level-based clustering can be used. When runtimes of same-level tasks may
-        vary, it may be preferable to use job runtime-based horizontal clustering, to achieve load-balancing of clusters. Alternatively, the user may opt for label-based clustering, in which
-        the user can define the task clusters themselves.
-
-        To avoid running out of disk space during data-intensive workflows, the Mapper can add cleanup nodes that remove all unecessary data after a job has completed.
-
-        \note{Dynamic workflows permitted}
-
-        \subsection{Debugging and provenance}
-         Job monitoring, in Pegasus, is achieved through \textit{pegasus-kickstart} (kickstart). Every job within a pegasus workflow has a kickstart wrapper that collects runtime provenance
-        and performance. Kickstart relays this information to the submit host after job completion. This data is ultimately collected and stored in a relational database.
-
-        The Mapper Logs contain details on the abstract workflow, the jobs in the executable workflow 
-        and the mapping of tasks to jobs. Using the mapper logs, Pegasus can generate the provenance of the results. Should it be required, Pegasus can also generate the provenance of the 
-        mapping process.
-
-        In addition to the Mapper Logs, there also exists the DAGMan logs, generated by HTCondor DAGMan and updated in near real-time. These logs capture details on the status of the jobs
-        and their associated pre- and post- execute scripts.
-
-        All provenance and debugging data is accessibly through either the \textit{pegasus-analyzer} tool or through the provided lightweight web interface.
+      remote servers, clusters and compute grids. As these execution hosts may
+      not have all the required dependencies to execute a task, tasks are
+      archived using CARE. This ensures that all the required applications and
+      their dependencies can be shipped and executed at runtime on any Linux
+      host.
+
+        \subsection{Debugging and Provenance} A CARETasks standard output and
+        standard error are normally printed out to OpenMOLES's console. For
+        distributed executions, this might not be a practical solution, in which
+        case such information may be passed through the dataflow using OpenMOLE
+        variables.
+
+        \section{Nextflow} \subsection{summary} Nextflow is a POSIX-compatible
+        pipeline engine for scientific workflows.  It extends the Groovy
+        programming language to provide an  intuitive domain-specific language
+        (DSL) that enables users to easily describe their workflows. Unlike
+        pipeline engines such as Nipype and PSOM, Netflow does not employ DAG.
+        It instead adheres to the \textit{Dataflow} programming model which
+        decides how data will flow through the pipeline at runtime \note{fix
+        this sentence}. Nextflow prides itself on allowing the integration of
+        containerized tools to address the issue of numerical instability that
+        may arise from different environment setups.
+
+        \subsection{Pipeline description} A Nextflow pipeline is represented by
+        the flow of data between processes. These processes can be written in
+        any Linux-compatible scripting language. As in Nipype, these processes
+        are executed in an isolated environment. The only means of communication
+        between processes is through asynchronous FIFO queues known as
+        \textit{channels}. Each process can contain any number of input and
+        output channels. Through the use of channels, input and output can be
+        stored not only on disk but also in-memory \note{Nipype can also return
+        outputs that are not on disk, so this is nothing special}.
+
+        As with other pipeline engines (nipype, psom, etc.,), the order in which
+        the processes appear is irrelevant. Execution order is determined by the
+        dependencies.
+
+        \subsection{Pipeline execution} The \textit{executor} is responsible for
+        the execution of a Nextflow pipeline. By default, it assumes that the
+        pipeline is to be run locally. The local executor will use
+        multithreading in order to exploit available cores within the local
+        environment. To configure the executor to run in a remote distributed
+        environment, it is necessary to modify the configuration file to specify
+        the environment. By following this approach, Nextflow users can
+        seamlessly run their code on various processing platforms with a simple
+        modification to a configuration file. In addition to local, Nextflow
+        currently has executors for SGE, LSF, SLURM, PBS/Torque, NQSII,
+        HTCondor, DRMAA, Ignite, Kubernetes and AWS Batch.
+
+        \subsection{Debugging and Provenance} When executing a pipeline, it is
+        always possible to specify the generation of a execution report. The
+        report is written in HTML and consists of three sections: Summary,
+        Resources and Tasks. Summary provides details on the execution command,
+        execution status, overall execution time, in addtion to other metadata.
+        The Resources tab, on the other hand, contains several HighChart plots
+        representing CPU, memory, time and disk read/write rate of each pipeline
+        process. The final section, Tasks, contains many metrics including the
+        status and command script of each task.
+
+    Other reports that can be generated at runtime include the Trace report and
+    the Timeline report. The Trace report include details such as CPU usage,
+    start time, submission time, etc, of each process that was executed,
+    whereas, the Timeline report generates an HTML timeline of all processes
+    that have been executed.
+
+    \note{doesn't seem to be anything on provenance...}
 
@@
     \section{Toil}
+        \note{obtained from paper's supplementary documentation}
+        Toil is another general workflow framework aimed at the processing of 
+        scientific big data, with its main focus being on genomics workflows. It 
+        prides itself in not only its ability to not only scale the processing 
+        of very large datasets, but also its portability, allowing it to be 
+        executed on any environment. Toil pipelines can be written in CWL, WDL 
+        (Workflow Description Language) and Python. The Toil workflow 
+        framework is also open-sourced, making it a community-driven effort. 
+        \note{as is the case with most}
+
         \subsection{Pipeline description}
+           A Toil workflow consists of a DAG of Jobs connected to each other
+           through input and output dependencies. All custom user scripts must
+           inherit from the Job class. In order to absolve the user from having
+           to manually create these classes for each function, Toil provides
+           wrapper classes that convert Python functions to Toil jobs directly.
+           A Toil job need not be part of the static workflow, it can also be
+           added at runtime.
+
+           There exists three types of relationships between Toil jobs: parent, 
+           child and follow-on jobs. Child jobs begin execution immediately
+           after their Parent completes their execution. Follow-on jobs, however
+           , begin execution after the Parent, associated child job and their
+           successors complete execution. They are particularly useful as 
+           cleanup jobs. Both child jobs and follow-on jobs are executed in
+           parallel.
+
+           Variables may be passed at runtime to dynamically added child or
+           follow-on jobs, or they can be referenced statically using promises.
+           Promises enable the reference to the result to be passed prior to 
+           runtime, evaluating the result only at runtime.
+
+           Workflows can be executed or 
+           resumed through the Runner class, or alternatively, invoked by the
+           Toil class. The latter differing from the Runner class as it serves 
+           as the workflow's context manager. The Toil class performs
+           a preliminary configuration of the workflow (e.g. file staging).
+           All Toil workflows begin with a single job. The value of this root 
+           job is returned at the end of the workflow.
+
+            The FileStore class manages the permanent and temporary files
+            created by the workflow. This class ensures that all necessary files
+            will be cleaned up after job execution, but also that workflows
+            can resume from the last successful state. Both temporary 
+            directories and temporary files can be created at a user's desired
+            location (e.g. local file system, Amazon S3, etc.,). Should files
+            and directories need to persist during workflow execution and
+            be accessible to all jobs, the FileStore
+            can add these files to a global store. All files stored in the 
+            global store are immutable, preventing any updates to these files
+            from occuring other than deletion. There also does not exist any
+            kind of filesystem hierarchy in the global store.\note{allows use
+            of various different object stores/ enables caching of files as they
+            never change}
+
+            In order to add files to a workflow, it is necessary to stage them
+            prior to execution. This is achieved using the \textit{Toil} context
+            manager as it is running on the leader prior to execution. It is 
+            possible to stage the file as a shared file, if necessary. In the 
+            instance of a job failure, the staged file behaves like any other 
+            file in the job store. If the file is not cleaned up post execution,
+            the file will not need to be staged again during resumption.
+
+            Toil also provides the functionality to startup services, such as
+            databases or servers, during pipeline execution. Multiple services
+            can be define per job and services can have associated child
+            services. This functionality allows, for instance, Apache Spark 
+            clusters to be spawned during workflow execution. As services 
+            complicate workflow resumption post failure, Toil enables the
+            checkpointing of services such that the workflow will resume from a
+            checkpoint if any of the successors to the service fail to restart.
+
         \subsection{Pipeline execution} 
+            \note{obtained from website documentation}
+            Toil is composed of three components: 1) the Job Store, 2) the Batch
+            System and 3) the Provisioner. The Job Store centralizes all the 
+            files necessary for the workflow. During a workflow re-execution,
+            the job store is used to resume the workflow from its last
+            successful task. Currently, there are two types of job stores 
+            supported: the file job store and the cloud job store.
+
+            The Batch System specifies which type of scheduler will be used.
+            A Toil workflow can either be executed locally, on HPC clusters or 
+            the cloud.
+
+            The Provisioner, on the other hand, is used to execute Toil
+            workflows on specific cloud platforms. Using the Cluster Utilities
+            command line tools, a user can provision nodes for their desired 
+            cloud platform.
+
+            Toil executes workflows using a leader-worker architecture. The
+            leader is a single-threaded read-only process responsible for 
+            determining which jobs are ready to be executed. Workers, on the 
+            other hand, are created by the batch-system and are responsible
+            for the execution of a job. To alleviate the single-threaded leader
+            from having to do an excessive amount of work, the worker maintains
+            track of its status (i.e. success or failure) and reports its status
+            back to the leader by writing it to the job-store. The worker may
+            also choose to immediately execute any child-jobs defined by the 
+            job it has executed.
+
+            In the instance when a job or a subset of jobs fail, the workflow
+            will continue to execute until it can proceed no more and will throw
+            an error. The user is given the option to resolve the bug and resume
+            the workflow or delete the job store and reexecute the pipeline 
+            entirely.
         \subsection{Debugging and provenance}
+
+        By default, workflow standard input and output is hidden to the user.
+        It is however possible to make the log data visible to the user during
+        execution or to redirect this data to a logfile.
+
+        Toil also provide a fuctionality, stats, to capture statistics on the 
+        workflow in the job store. Examples of the statistics it captures
+        includes cpu, memory and job duration.
+
     \section{Thunder}
 
-        Thunder is a Python-based modular set of tools for the analysis of imaging and timeseries data belonging
-        to various domains including neuroscience. Although it is not specifically a pipelining framework, 
-        it can run on top of the Apache
-        Spark framework enabling the Thunder pipeline to be extended and customized. By running on top of Spark,
-        Thunder leverages its features to improve performance on large scale studies. Thunder does not need to
-        be executed on top of a Spark cluster and can easily be executed, with the same script, an a local system
-        with no Spark installation.
+        Thunder is a Python-based modular set of tools for the analysis of
+        imaging and timeseries data belonging to various domains including
+        neuroscience. Although it is not specifically a pipelining framework, it
+        can run on top of the Apache Spark framework enabling the Thunder
+        pipeline to be extended and customized. By running on top of Spark,
+        Thunder leverages its features to improve performance on large scale
+        studies. Thunder does not need to be executed on top of a Spark cluster
+        and can easily be executed, with the same script, an a local system with
+        no Spark installation.
 
         \note{The following was extracted from the documenation}
-        \subsection{Pipeline description}
-        The API for Thunder, regardless of if backed by an Apache Spark cluster or not, remains the same. If not 
-        using an Spark Resilient Distributed Dataset (RDD), the data can be backed by a \textit{numpy} array or a
-        distributed \textit{bolt} array. Thunder is broken down into several packages, in which the core package
-        consists of tools for reading and writing data, in addition to defining common data structures. Users may
-        choose to install other Thunder packages for added functionality, or define their own using any Python
-        library.
-
-        Thunder contains two data structures: \textit{images} and \textit{series}. As implied by its name, the 
-        \textit{images} data type may represent a collection of images or volumes. 
-        The \textit{series} data type,
-        however, represents a one-dimensional collection that shares a common index. Both \textit{images} and 
-        \textit{series} can be loaded from a list, ndarray or binary file type, however \textit{series} can also 
-        be loaded from a text file and \textit{images} can be loaded from png and tiff files.
-
-        Methods that are common to both data types are, such as \textit{map} and \textit{reduce}, are found in the 
-        \textit{data} class which is a subclass of the
-        \textit{base} class. In order to convert between the \textit{images} and \textit{series} data types, the
-        intermediary data type \textit{blocks}, child of \textit{base} class,  is used. 
-        Using the \textit{blocks} class, it is also possible to convert the data type to an ndarray for use with other
-        Python modules.
-
-        \note{not really anything on execution, debugging and provenance in the paper nor in the documentation}
-        \subsection{Pipeline execution}
-        In order to use Spark within Thunder, a SparkContext must be started and passed to the Thunder API. Should
-        the user need to switch between local and Spark frequently, it is possible to use the \textit{station} package
-        as a context manager for the various contexts.
-
-        As Thunder is an extension to Spark, numpy and \textit{bolt}, much of the execution details are likely differed 
-        to the 
-        framework that it is being executed on.
-
-        \subsection{Debugging and provenance}
-        Thunder does not explicitly provide any details on debugging and provenance tracking. Similarly to execution,
-        it is expected that much of the debugging and provenance details are not obtained from Thunder itself, but
-        rather, from the frameworks it is being executed on. For instance, should Thunder be exected on a Spark cluster,
-        it is likely that much of the debugging and provenance information can be obtained from the Spark logs and RDD
+        \subsection{Pipeline description} The API for Thunder, regardless of if
+        backed by an Apache Spark cluster or not, remains the same. If not using
+        an Spark Resilient Distributed Dataset (RDD), the data can be backed by
+        a \textit{numpy} array or a distributed \textit{bolt} array. Thunder is
+        broken down into several packages, in which the core package consists of
+        tools for reading and writing data, in addition to defining common data
+        structures. Users may choose to install other Thunder packages for added
+        functionality, or define their own using any Python library.
+
+        Thunder contains two data structures: \textit{images} and
+        \textit{series}. As implied by its name, the \textit{images} data type
+        may represent a collection of images or volumes.  The \textit{series}
+        data type, however, represents a one-dimensional collection that shares
+        a common index. Both \textit{images} and \textit{series} can be loaded
+        from a list, ndarray or binary file type, however \textit{series} can
+        also be loaded from a text file and \textit{images} can be loaded from
+        png and tiff files.
+
+        Methods that are common to both data types are, such as \textit{map} and
+        \textit{reduce}, are found in the \textit{data} class which is a
+        subclass of the \textit{base} class. In order to convert between the
+        \textit{images} and \textit{series} data types, the intermediary data
+        type \textit{blocks}, child of \textit{base} class,  is used.  Using the
+        \textit{blocks} class, it is also possible to convert the data type to
+        an ndarray for use with other Python modules.
+
+        \note{not really anything on execution, debugging and provenance in the
+        paper nor in the documentation} \subsection{Pipeline execution} In order
+        to use Spark within Thunder, a SparkContext must be started and passed
+        to the Thunder API. Should the user need to switch between local and
+        Spark frequently, it is possible to use the \textit{station} package as
+        a context manager for the various contexts.
+
+        As Thunder is an extension to Spark, numpy and \textit{bolt}, much of
+        the execution details are likely differed to the framework that it is
+        being executed on.
+
+        \subsection{Debugging and provenance} Thunder does not explicitly
+        provide any details on debugging and provenance tracking. Similarly to
+        execution, it is expected that much of the debugging and provenance
+        details are not obtained from Thunder itself, but rather, from the
+        frameworks it is being executed on. For instance, should Thunder be
+        exected on a Spark cluster, it is likely that much of the debugging and
+        provenance information can be obtained from the Spark logs and RDD
         lineage.
@@
 \chapter{MapReduce-based BigData frameworks and related components}
-    \section{mapreduce}
-        Summary of ~\cite{mapred}
-
-        Typically, simple computations need to be performed on 
-        increasingly growing large datasets. Processing of such 
-        large datasets require parallelization. As a result, 
-        these computations are complicated by details of 
-        parallelization, fault tolerance, data distribution and 
-        load balancing. The MapReduce library addresses this issue 
-        by allowing the expression of simple computations while 
-        abstracting the other details. The map and reduce paradigm 
-        was selected for this purpose as it was already commonly 
-        used in functional programming languages, such as Lisp, and 
-        it was observed that many computations could be expressed 
-        as such.
-
-        The MapReduce library operates by first dividing the data 
-        into manageable chunks (16 to 64MB). It then creates 
-        multiple copies of the program and distributes them across
-        nodes. There exists a single master node; the remainder are
-        worker nodes. The master node is responsible for delegating 
-        map and reduce tasks to the idle workers. There are two data
+\section{mapreduce} Summary of ~\cite{mapred}
+
+        Typically, simple computations need to be performed on increasingly
+        growing large datasets. Processing of such large datasets require
+        parallelization. As a result, these computations are complicated by
+        details of parallelization, fault tolerance, data distribution and load
+        balancing. The MapReduce library addresses this issue by allowing the
+        expression of simple computations while abstracting the other details.
+        The map and reduce paradigm was selected for this purpose as it was
+        already commonly used in functional programming languages, such as Lisp,
+        and it was observed that many computations could be expressed as such.
+
+        The MapReduce library operates by first dividing the data into
+        manageable chunks (16 to 64MB). It then creates multiple copies of the
+        program and distributes them across nodes. There exists a single master
+        node; the remainder are worker nodes. The master node is responsible for
+        delegating map and reduce tasks to the idle workers. There are two data
         structures found in the master. The first stores the state
-        (\textit{idle}, \textit{in-progress} or \textit{completed}) 
-        of each \textit{map} and \textit{reduce} task, and the second
-        stores the identity of the non-idle worker nodes. In contrast,
-        the worker nodes are responsible for the execution of 
-        \textit{map} or \textit{reduce} tasks. A worker node assigned
-        a \textit{map} task will parse out the key-value pairs from 
-        the input file and pass it to the \textit{map} function. The 
-        intermediate keys produced by the \textit{map} function are 
-        buffered in memory and periodically written to disk.The 
-        locations of these intermediate files are sent to the master 
-        who is responsible for forwarding these locations to the 
-        reducer. A worker node assigned a \textit{reduce} task uses 
-        remote procedure calls to obtain the intermediate local files
-        created by the map tasks and sorts the keys such that identical
-        keys are grouped together. Each unique key and its list of 
-        values are then submitted to the \textit{reduce} task and the 
-        output of the \textit{reduce} task is stored in a final output 
-        file. Once all \textit{reduce} tasks have completed, the master
-        resumes the program and the output is returned to the user.
-
-        In order to ensure fault tolerance, the master regularly pings 
-        the worker nodes. Should a worker node not respond after a certain
-        delay, the master labels the worker as failed. The \textit{map} 
-        tasks that were completed by the worker are all returned to their
-        original state (\textit{idle}) as \textit{map} task data is stored
-        locally. Both the completed and in-progress worker's \textit{map}
-        tasks are subsequently reassigned to other workers. Only the failed
-        worker's \textit{in-progress} \textit{reduce} tasks need to be 
-        reassigned.
-
-        The master node undergoes periodic checkpoints, such that the 
-        master node can be reinitialized from its last checkpoint in 
-        case of failure. Master node failure is, however, unlikely.
-
-        As network bandwidth is a scarce resource, the master attempts 
-        to schedule \textit{map} tasks closest to where the data is 
-        located. it will first attempt to schedule a map task on the host
-        that contains the data; should that option not be available, the
-        master will attempt to schedule the \textit{map} task on a host 
-        nearest to the data (e.g. same network switch)
-
-        Ideally, the number of \textit{map} and \textit{reduce} tasks 
-        should be much larger than the number of available worker nodes. 
-        Having it as such improves both dynamic load balancing and recovery 
-        time. The number of \textit{map} tasks used in practice typically 
-        corresponds to the size of the input chunks, whereas the number of 
-        \textit{reduce} tasks are a small multiple of the number of worker 
-        nodes used.
+        (\textit{idle}, \textit{in-progress} or \textit{completed}) of each
+        \textit{map} and \textit{reduce} task, and the second stores the
+        identity of the non-idle worker nodes. In contrast, the worker nodes are
+        responsible for the execution of \textit{map} or \textit{reduce} tasks.
+        A worker node assigned a \textit{map} task will parse out the key-value
+        pairs from the input file and pass it to the \textit{map} function. The
+        intermediate keys produced by the \textit{map} function are buffered in
+        memory and periodically written to disk.The locations of these
+        intermediate files are sent to the master who is responsible for
+        forwarding these locations to the reducer. A worker node assigned a
+        \textit{reduce} task uses remote procedure calls to obtain the
+        intermediate local files created by the map tasks and sorts the keys
+        such that identical keys are grouped together. Each unique key and its
+        list of values are then submitted to the \textit{reduce} task and the
+        output of the \textit{reduce} task is stored in a final output file.
+        Once all \textit{reduce} tasks have completed, the master resumes the
+        program and the output is returned to the user.
+
+        In order to ensure fault tolerance, the master regularly pings the
+        worker nodes. Should a worker node not respond after a certain delay,
+        the master labels the worker as failed. The \textit{map} tasks that were
+        completed by the worker are all returned to their original state
+        (\textit{idle}) as \textit{map} task data is stored locally. Both the
+        completed and in-progress worker's \textit{map} tasks are subsequently
+        reassigned to other workers. Only the failed worker's
+        \textit{in-progress} \textit{reduce} tasks need to be reassigned.
+
+        The master node undergoes periodic checkpoints, such that the master
+        node can be reinitialized from its last checkpoint in case of failure.
+        Master node failure is, however, unlikely.
+
+        As network bandwidth is a scarce resource, the master attempts to
+        schedule \textit{map} tasks closest to where the data is located. it
+        will first attempt to schedule a map task on the host that contains the
+        data; should that option not be available, the master will attempt to
+        schedule the \textit{map} task on a host nearest to the data (e.g. same
+        network switch)
+
+        Ideally, the number of \textit{map} and \textit{reduce} tasks should be
+        much larger than the number of available worker nodes.  Having it as
+        such improves both dynamic load balancing and recovery time. The number
+        of \textit{map} tasks used in practice typically corresponds to the size
+        of the input chunks, whereas the number of \textit{reduce} tasks are a
+        small multiple of the number of worker nodes used.
 
@@
         average performance) to delay completion. As a measure to counteract
-        this, the master node assigns the same task to idle workers. The task 
-        is then marked as completed as soon as either the primary or the 
-        backups complete the task. This strategy has been found to 
-        significantly improve performance when used with very large datasets. 
-        \note{44\% speedup with their sort program}.
-
-        A few extensions have also been added to the library in order to 
-        improve user experience. An important extension is the introduction 
-        of a \textit{combiner}, which enables the commutative and associative 
-        reduce function to be applied to the same keys on the hosts where the 
+        this, the master node assigns the same task to idle workers. The task is
+        then marked as completed as soon as either the primary or the backups
+        complete the task. This strategy has been found to significantly improve
+        performance when used with very large datasets.  \note{44\% speedup with
+        their sort program}.
+
+        A few extensions have also been added to the library in order to improve
+        user experience. An important extension is the introduction of a
+        \textit{combiner}, which enables the commutative and associative reduce
+        function to be applied to the same keys on the hosts where the
         \textit{map} task was executed. Like the \textit{map} task, the
@@
         failure, the ability to produce auxiliary output files in \textit{map}
-        or \textit{reduce} task, live status updates stream to HTTP server,
-        the ability to run all code on a single local machine, and a 
-        \textit{Counter} class.
-    \section{spark} 
-        \begin{itemize} 
-            \item based on map-reduce 
-            \item in-memory processing 
-            \item lazy evaluation 
-            \item general framework 
-        \end{itemize}
-
-    \subsection{Scientific workflows in Spark}
-        \subsubsection{Introduction}
+        or \textit{reduce} task, live status updates stream to HTTP server, the
+        ability to run all code on a single local machine, and a
+        \textit{Counter} class.  \section{spark} \begin{itemize} \item based on
+        map-reduce \item in-memory processing \item lazy evaluation \item
+        general framework \end{itemize}
+
+    \subsection{Scientific workflows in Spark} \subsubsection{Introduction}
     emphasis on data rather than on interface. transformation describe how
-    interfaces interact with the data. 
-        \subsubsection{ariel's paper and thunders astronomy - }
-        \subsubsection{Spark on hpc's or } 
-            \note{large dataset for valduriez} Prior to
-    examining how Big Data frameworks could be improved on HPC clusters, it is
-    important to have a baseline of how it performs natively on an HPC cluster.
-    Particularly, how it behave when processing scientific workflows. In a recent
-    study done by \cite{valduriez}, they examined the scalability of Spark on an HPC
-    cluster using a black-box scientific workflow. Five tests were performed in
-    their study: 1) Measuring execution time when task and resources available
-    increase proportionally, 2) Measuring the execution time of very short and short
-    tasks, 3) Measuring the execution time of short (5s) to long tasks (120s), 4)
-    Varying the number of tasks with fixed task durations, 5) Mixed task durations
-    and varied number of tasks. Their results show Spark behaves as expected when
-    tasks and resources are increased exponentially, and that Spark scales better
-    than expected with tasks of longer durations. This is due to the fact that the
+    interfaces interact with the data.  \subsubsection{ariel's paper and
+    thunders astronomy - } \subsubsection{Spark on hpc's or } \note{large
+    dataset for valduriez} Prior to examining how Big Data frameworks could be
+    improved on HPC clusters, it is important to have a baseline of how it
+    performs natively on an HPC cluster.  Particularly, how it behave when
+    processing scientific workflows. In a recent study done by \cite{valduriez},
+    they examined the scalability of Spark on an HPC cluster using a black-box
+    scientific workflow. Five tests were performed in their study: 1) Measuring
+    execution time when task and resources available increase proportionally, 2)
+    Measuring the execution time of very short and short tasks, 3) Measuring the
+    execution time of short (5s) to long tasks (120s), 4) Varying the number of
+    tasks with fixed task durations, 5) Mixed task durations and varied number
+    of tasks. Their results show Spark behaves as expected when tasks and
+    resources are increased exponentially, and that Spark scales better than
+    expected with tasks of longer durations. This is due to the fact that the
     scheduler is overloaded in the case of many short tasks in addition to their
-    being a significant amount of I/O occurring between each task. Although I/O is
-    problematic in both data center and HPC infrastructures, HPC is optimized for
-    data transfer over the network rather than disk I/O, and therefore, such
-    frequent I/O negatively impacts the system. However, many scientific workflows
-    are composed of longer tasks, and therefore, Spark scales very well with
-    scientific workflows in an HPC environment.
-     \section{Use of containers with Big Data frameworks}
-     \section{Hadoop Distributed file system} 
-        \note{maybe should go in between mapreduce and spark}
-
-        The Hadoop Distributed File System (HDFS)\cite{hadoop} 
-        is the filesystem component of Hadoop. The metadata in 
-        HDFS is stored in on a dedicated server known as the 
-        NameNode, whereas application data is distributed across 
-        many servers referred to as DataNode. To ensure fault 
-        tolerance, HDFS replicates the data across multiple 
-        DataNodes (default replication factor: 3). Not only does 
-        this ensure that the data is not lost in the event of node
-        failure, but it also increases data transfer bandwidth, as
-        the data can be accessed from multiple nodes, and thus, 
-        there are more opportunities for computations to be 
-        performed nearest to where the data is located.
-
-        The NameNode contains the namespace tree, a hierarchy of 
-        directories and their files, as well as the mapping of the 
-        split file blocks to DataNodes. The entire namespace is 
-        stored in memory. Information on the directories and files,
-        such as modification and access times, namespace and disk 
-        quotes, are stored within \textit{inodes} on the NameNode. 
-        The name system's metadata, \textit{inode} and file block 
-        mapping, is known as the \textit{image}. Persistent record 
-        of the image that is stored on the local file system are 
-        known as \textit{checkpoint}. Locations of block replicas 
-        are not stored in the checkpoint as they may change over 
-        time. The \textit{journal} is a log of the modifications 
-        made to the image. Both the checkpoint and the journal may 
-        be copied across servers for increased durability. The 
-        journal is played back during the NameNode restart in order 
-        to restore the cluster.
-
-        When a client requests to read data, it must first contact 
-        the NameNode. The NameNode provides it with the replica 
-        locations, and the client reads from the DataNode located 
-        closest to it. When the client requests to write a file, it
-        contacts the NameNode which selects the DataNodes that will 
-        host the replicas. The client subsequently writes the data 
-        directly to the DataNodes in a pipeline fashion (The data 
-        gets propagated through each DataNode by being transferred
-        from the nearest DataNode to it).
-
-        There is only one NameNode assigned to each cluster, however, 
-        each cluster can have multiple clients and execute multiple 
-        tasks concurrently.
-
-        A block replica stored on a DataNode consists of two files 
-        stored on the local file system: the metadata and the data.
-
-        The DataNode connects to the NameNode during startup and 
-        performs handshake to for DataNode namespace ID and 
-        software version verification. Should either namespace ID 
-        or software version not match those of the NameNode, the 
-        DataNode shuts down.
-
-        To ensure integrity of the system, a \textit{namespace ID} 
-        is assigned to all nodes during formatting of the namespace. 
-        DataNodes with different namespace IDs to the NameNode will 
-        not be permitted to join the cluster. Should the DataNode
-        have not been assigned a namespace ID, it will be permitted 
-        to join the cluster. The cluster's namespace ID will then be 
-        assigned to that DataNode.
-
-        Following the handshake, the DataNode register with the 
-        NameNode using their storage ID. The storage ID is a unique 
-        ID generated after initial registration to the NameNode and 
-        that is persistently stored on the DataNode. It permits
-        identification of the DataNode in the event of an IP address 
-        or port change.
+    being a significant amount of I/O occurring between each task. Although I/O
+    is problematic in both data center and HPC infrastructures, HPC is optimized
+    for data transfer over the network rather than disk I/O, and therefore, such
+    frequent I/O negatively impacts the system. However, many scientific
+    workflows are composed of longer tasks, and therefore, Spark scales very
+    well with scientific workflows in an HPC environment.  \section{Use of
+    containers with Big Data frameworks} \section{Hadoop Distributed file
+    system} \note{maybe should go in between mapreduce and spark}
+
+        The Hadoop Distributed File System (HDFS)\cite{hadoop} is the filesystem
+        component of Hadoop. The metadata in HDFS is stored in on a dedicated
+        server known as the NameNode, whereas application data is distributed
+        across many servers referred to as DataNode. To ensure fault tolerance,
+        HDFS replicates the data across multiple DataNodes (default replication
+        factor: 3). Not only does this ensure that the data is not lost in the
+        event of node failure, but it also increases data transfer bandwidth, as
+        the data can be accessed from multiple nodes, and thus, there are more
+        opportunities for computations to be performed nearest to where the data
+        is located.
+
+        The NameNode contains the namespace tree, a hierarchy of directories and
+        their files, as well as the mapping of the split file blocks to
+        DataNodes. The entire namespace is stored in memory. Information on the
+        directories and files, such as modification and access times, namespace
+        and disk quotes, are stored within \textit{inodes} on the NameNode.  The
+        name system's metadata, \textit{inode} and file block mapping, is known
+        as the \textit{image}. Persistent record of the image that is stored on
+        the local file system are known as \textit{checkpoint}. Locations of
+        block replicas are not stored in the checkpoint as they may change over
+        time. The \textit{journal} is a log of the modifications made to the
+        image. Both the checkpoint and the journal may be copied across servers
+        for increased durability. The journal is played back during the NameNode
+        restart in order to restore the cluster.
+
+        When a client requests to read data, it must first contact the NameNode.
+        The NameNode provides it with the replica locations, and the client
+        reads from the DataNode located closest to it. When the client requests
+        to write a file, it contacts the NameNode which selects the DataNodes
+        that will host the replicas. The client subsequently writes the data
+        directly to the DataNodes in a pipeline fashion (The data gets
+        propagated through each DataNode by being transferred from the nearest
+        DataNode to it).
+
+        There is only one NameNode assigned to each cluster, however, each
+        cluster can have multiple clients and execute multiple tasks
+        concurrently.
+
+        A block replica stored on a DataNode consists of two files stored on the
+        local file system: the metadata and the data.
+
+        The DataNode connects to the NameNode during startup and performs
+        handshake to for DataNode namespace ID and software version
+        verification. Should either namespace ID or software version not match
+        those of the NameNode, the DataNode shuts down.
+
+        To ensure integrity of the system, a \textit{namespace ID} is assigned
+        to all nodes during formatting of the namespace.  DataNodes with
+        different namespace IDs to the NameNode will not be permitted to join
+        the cluster. Should the DataNode have not been assigned a namespace ID,
+        it will be permitted to join the cluster. The cluster's namespace ID
+        will then be assigned to that DataNode.
+
+        Following the handshake, the DataNode register with the NameNode using
+        their storage ID. The storage ID is a unique ID generated after initial
+        registration to the NameNode and that is persistently stored on the
+        DataNode. It permits identification of the DataNode in the event of an
+        IP address or port change.
 
@@
     \note{precise the systems that will be used. both have distributed memory.
-data centre is hpc cluster with slower network, because may span multiple racks
-and domains} \note{variety of jobs on hpc whereas dedicated map-reduce jobs in a
-data center} \note{talk about spark on cloud and the challenges associated with}
-\note{talk about why this is important} \note{talk } High-Performance Computing
-(HPC) clusters are extensively used by the scientific community to process
-complex, computation intensive problems. Due to their importance in the
-advancement of scientific research, it is necessary that Big Data frameworks
-used to process scientific data are compatible with them.
+    data centre is hpc cluster with slower network, because may span multiple
+    racks and domains} \note{variety of jobs on hpc whereas dedicated map-reduce
+    jobs in a data center} \note{talk about spark on cloud and the challenges
+    associated with} \note{talk about why this is important} \note{talk }
+    High-Performance Computing (HPC) clusters are extensively used by the
+    scientific community to process complex, computation intensive problems. Due
+    to their importance in the advancement of scientific research, it is
+    necessary that Big Data frameworks used to process scientific data are
+    compatible with them.
 
@@
     are made up of nodes with homogenous storage connected by lower-end network
-cables (e.g. Ethernet). As a result, moving data from local storage to memory
-was much less costly in these infrastructures than moving data across the
-network. Thus, Big Data frameworks implemented data locality in an attempt to
-reduce costly network (horizontal) transfers and favour more performant local
-storage to memory (vertical) transfers. HPC clusters, however, are not
-necessarily commodity clusters. Large HPC clusters are commonly made up of
-supercomputing nodes connected by a high bandwidth, low-latency network (e.g.
-InfiniBand). Nodes may contain their own local storage (e.g. HDD, SSD, tmpfs)
-and are all connected to a distributed file system (e.g GPFS, Lustre).  Due to
-their employment of a high bandwidth low latency network, network transfers on
-such an HPC cluster may not be as costly as that of a typical data center, and
-it may in fact be more costly to do vertical transfers. Furthermore, HPC systems
-may enforce a batch submission system. This required users to specify number of
-nodes, cores and total processing time; requirements which are unknown to
-MapReduce users.
+    cables (e.g. Ethernet). As a result, moving data from local storage to
+    memory was much less costly in these infrastructures than moving data across
+    the network. Thus, Big Data frameworks implemented data locality in an
+    attempt to reduce costly network (horizontal) transfers and favour more
+    performant local storage to memory (vertical) transfers. HPC clusters,
+    however, are not necessarily commodity clusters. Large HPC clusters are
+    commonly made up of supercomputing nodes connected by a high bandwidth,
+    low-latency network (e.g.  InfiniBand). Nodes may contain their own local
+    storage (e.g. HDD, SSD, tmpfs) and are all connected to a distributed file
+    system (e.g GPFS, Lustre).  Due to their employment of a high bandwidth low
+    latency network, network transfers on such an HPC cluster may not be as
+    costly as that of a typical data center, and it may in fact be more costly
+    to do vertical transfers. Furthermore, HPC systems may enforce a batch
+    submission system. This required users to specify number of nodes, cores and
+    total processing time; requirements which are unknown to MapReduce users.
 
@@
     In paper \note{Scaling Spark on HPC Systems} it was found that Spark on a
-Lustre backend performed significantly (4x) slower than that of a workstation
-using local fast SSDs for storage. This is due to the fact that file system
-metadata latency (more specifically, file open) is the determinant factor for
-performance on HPC, whereas network dominates performance in a typical data
-centre configuration. It was found by the authors that using either a local
-in-memory filesystem or a filesystem mounted to a single Lustre file improved
-performance up to a point where it matched that of a workstation. In addition,
-use of a cache was found to reduce file system metadata operations, and thus
-improve overall performance.  The issues related to using an in-memory
-filesystem is that it is limited by the amount of physical memory available, and
-a job will fail if physical memory runs out. Moreover, there is not persistence
-or resilience of data as it is not saved to non-volatile storage (disk). Spark
-fails with medium to large applications due to "lax garbage collection in block
-and shuffle managers". To resolve these issues, the authors attempted to make
-all executors share a descriptor pool. This allowed files that were already
-opened by other executors to be accessed by the executor through the descriptor
-rather than having to reopen it again. However, the node OS image is subject to
-number of Inode constraints as well Lustre limits on number of open files for a
-given job. Due to Inode constraints, a livelock may occur when descriptor pool
-is at capacity and an executor is attempting to open a file. The authors also
-created mounted a local filesystem backed by a Lustre file (lustremount) to
-address the issues related to an in-memory file system \note{requires admin
-privilege}. It was found that the use of an in-memory file system improved
-performance by 7.7x and the lustremount improved performance by 6.6x. It was
-also found that implementation of the filepool improved performance.  Another
-issue observed by the authors is that of poor block management as a result of
-memory constraints resulting in a significant amount of vertical movement. When
-available memory is exhausted, the least recently used block is evicted by the
-block-manager. When a call is made to this removed block, it will need to be
-recomputed, and in the process blocks required for recomputation may also be
-evicted and necessitate recomputation of those blocks, thus leading to a chain
-of eviction and recomputation. To resolve this, marking intermediate RDDs as
-persistent, such that their results are saved to storage instead of requiring
-recomputation, may be applied.  \note{they didn't look at improving initial
-application I/O, though believe that it could be improved with lustremount and
-filepool. } \note{for each partition, a shuffle file is created and written to
-as many times as there are partitions. An index file is also created which
-contains shuffle file locations and offsets. Metadata access is therefore
-O(partitions2)}
+    Lustre backend performed significantly (4x) slower than that of a
+    workstation using local fast SSDs for storage. This is due to the fact that
+    file system metadata latency (more specifically, file open) is the
+    determinant factor for performance on HPC, whereas network dominates
+    performance in a typical data centre configuration. It was found by the
+    authors that using either a local in-memory filesystem or a filesystem
+    mounted to a single Lustre file improved performance up to a point where it
+    matched that of a workstation. In addition, use of a cache was found to
+    reduce file system metadata operations, and thus improve overall
+    performance.  The issues related to using an in-memory filesystem is that it
+    is limited by the amount of physical memory available, and a job will fail
+    if physical memory runs out. Moreover, there is not persistence or
+    resilience of data as it is not saved to non-volatile storage (disk). Spark
+    fails with medium to large applications due to "lax garbage collection in
+    block and shuffle managers". To resolve these issues, the authors attempted
+    to make all executors share a descriptor pool. This allowed files that were
+    already opened by other executors to be accessed by the executor through the
+    descriptor rather than having to reopen it again. However, the node OS image
+    is subject to number of Inode constraints as well Lustre limits on number of
+    open files for a given job. Due to Inode constraints, a livelock may occur
+    when descriptor pool is at capacity and an executor is attempting to open a
+    file. The authors also created mounted a local filesystem backed by a Lustre
+    file (lustremount) to address the issues related to an in-memory file system
+    \note{requires admin privilege}. It was found that the use of an in-memory
+    file system improved performance by 7.7x and the lustremount improved
+    performance by 6.6x. It was also found that implementation of the filepool
+    improved performance.  Another issue observed by the authors is that of poor
+    block management as a result of memory constraints resulting in a
+    significant amount of vertical movement. When available memory is exhausted,
+    the least recently used block is evicted by the block-manager. When a call
+    is made to this removed block, it will need to be recomputed, and in the
+    process blocks required for recomputation may also be evicted and
+    necessitate recomputation of those blocks, thus leading to a chain of
+    eviction and recomputation. To resolve this, marking intermediate RDDs as
+    persistent, such that their results are saved to storage instead of
+    requiring recomputation, may be applied.  \note{they didn't look at
+    improving initial application I/O, though believe that it could be improved
+    with lustremount and filepool. } \note{for each partition, a shuffle file is
+    created and written to as many times as there are partitions. An index file
+    is also created which contains shuffle file locations and offsets. Metadata
+    access is therefore O(partitions2)}
 
     \subsection{tachyon paper} I/O is a known bottleneck in Spark that is
-exacerbated in HPC systems in which data is stored in a separate cluster outside
-compute nodes. One possible solution used to mitigate the effects of the I/O
-bottleneck is to store data in an in-memory filesystem rather than on disk.
-Alluxio and Triple-H are two known examples of such filesystems.
+    exacerbated in HPC systems in which data is stored in a separate cluster
+    outside compute nodes. One possible solution used to mitigate the effects of
+    the I/O bottleneck is to store data in an in-memory filesystem rather than
+    on disk.  Alluxio and Triple-H are two known examples of such filesystems.
 
@@
     improve read and write performance without impacting fault tolerance of the
-system. As write performance is impeded by the need to replicate data, Alluxio
-leverages the concept of lineage to recompute the output if data is lost.
-Tachyon exhibits 110x faster write throughput than in-memory HDFS and 4.4x
-faster end-to-end latency.
+    system. As write performance is impeded by the need to replicate data,
+    Alluxio leverages the concept of lineage to recompute the output if data is
+    lost.  Tachyon exhibits 110x faster write throughput than in-memory HDFS and
+    4.4x faster end-to-end latency.
 
     Alluxio has is made up of two layers: the lineage layer and the persistence
-layer. The lineage layer is the layer that performs recomputations to obtain the
-desired output. The persistence layer contains all the code used to regenerate
-the data, in addition to any data that cannot be recovered through lineage
-as it may be too large to fit in memory or has no lineage information stored.
+    layer. The lineage layer is the layer that performs recomputations to obtain
+    the desired output. The persistence layer contains all the code used to
+    regenerate the data, in addition to any data that cannot be recovered
+    through lineage as it may be too large to fit in memory or has no lineage
+    information stored.
 
@@
     In an attempt to improve performance of HDFS on HPC clusters, Triple-H was
-created.  \section{Why is use of such systems important} \section{Scheduling}
-\subsection{HPC schedulers: SLURM/QSUB} \subsection{Big Data schedulers:
-YARN/MESOS} \subsection{Multi-level scheduling w/ BigData schedulers}
-\chapter{Conclusion}
+    created.  \section{Why is use of such systems important}
+    \section{Scheduling} \subsection{HPC schedulers: SLURM/QSUB} \subsection{Big
+    Data schedulers: YARN/MESOS} \subsection{Multi-level scheduling w/ BigData
+    schedulers} \chapter{Conclusion}
 

### 2018-09-22 paper.tex
added dask summary -- to be fixed

@@
 
-    \section{Geneflow} 
+    \section{Galaxy}
+        Galaxy is another open-source genomics web-platformed designed to 
+        address the
+        reproducibility crisis in genomics in addition to facilitating workflow
+        creation process for non-programmers and ensuring the transparency of 
+        the workflows. It differs from workflow engines in that it is a full
+        platform designed for researchers to create and share their workflows
+        and data with the community. Building workflows in Galaxy is
+        entirely GUI-based enabling users with little to no programming
+        knowledge to build workflows. Metadata is extracted from the workflows
+        during executions and pipelines can be annoted and shared with other 
+        users.
+
         \subsection{Pipeline description}
+        In order to make workflow creation accessible to researchers with little
+        -to-no programming skills, Galaxy provides a GUI. 
+        The GUI allows users to drag-and-drop CLI tools to the screen to create
+        workflows. In order for new tools to be added to Galaxy, the user must 
+        write a configuration file which contains details on input and output
+        specifications and how to execute the tool. During workflow creation, 
+        the workflow editor will determine if the inputs and outputs of 
+        connected tools are compatible.
+
+        Alternatively, users may also create workflows by editing workflows that
+        have been published to Galaxy or using their own analysis history to 
+        generate a workflow. Any workflow created by a user can be published to
+        the Galaxy server and made available to all other users.
+
         \subsection{Pipeline execution}
+
+        \note{no details on workflow failure}
+        As Galaxy is meant to be easy-to-use for researchers, all that is 
+        required to execute pipelines is access to their webpage. Users have 
+        access to detailed records of the executed workflow's metadata and 
+        provenance indicating the outcome of the executed workflow.
+
+        Should users desired to host their own Galaxy server, it is possible to
+        install Galaxy locally, on HPC with SLURM or PBS installed or the cloud.
+
         \subsection{Debugging and provenance} 
+        Provenance is a big part of Galaxy as it aims to improve accessibility, 
+        reproducibility and transparency of workflows. During workflow execution
+        , Galaxy automatically produces metadata for each of the workflow tasks.
+        This metadata includes details on the input and output datasets, tools 
+        user and parameter values.
+
+        In addition to Galaxy's metadata capture, it is also possible for the 
+        user to annotate and tag their workflow. This enables the user to 
+        communicate with others the intent of the analysis steps in addition to 
+        incorporating search capabilities to their workflows.
+
+        Galaxy also provides the Pages capability. Pages is a web-based document
+        that enables users to automatically generate detailed documentation of 
+        their analyses.
+
     \section{Toil}
@@
         ability to run all code on a single local machine, and a
-        \textit{Counter} class.  \section{spark} \begin{itemize} \item based on
+        \textit{Counter} class.
+
+    \section{spark} \begin{itemize} \item based on
         map-reduce \item in-memory processing \item lazy evaluation \item
@@
         IP address or port change.
+
+        \section{Dask}
+        Dask contains three primary abstractions: the \textit{dask.array}, 
+        \textit{dask.bag} and \textit{dask.dataframe}. The \textit{dask.array} 
+        is an abstraction that enables parallel NumPy-like operations using task
+        scheduling. It can additionally be used on data that does not fit in 
+        memory. The \textit{dask.array} used blocked array algorithms to break 
+        down large commonly-used linear algebra functions into smaller ones that
+        can be partially parallelized.
+
+        Dask array functions all create an \textit{Array} object which contains 
+        the dask graph \textit{.dask}, details on the overall shape and the 
+        shape of each subdivided portion (chunk) of the array (\textit{.chunks})
+        , the \textit{.name} representing which keys map to the results, and the
+        data type.
+
+        A limitation of the Dask array is that it must know the shape of the 
+        array at graph creation time.
+
+        The Dask bag is similar to that of a Python list, however, order cannot 
+        be guaranteed. As Dask bags typically operate on Python objects, 
+        performance is bound by the Global Interpreter Lock (GIL), therefore 
+        multiprocessing schedulers must be used instead of multithreading ones.
+
+        The Dask bag contains functions like \textit{map} and \textit{filter}, 
+        in addition to many other functions that can be found in the PyToolz API
+        . It is particularly well suited for the handling of data dumps.
+
+        The Dask dataframe consists of a collection of Pandas DataFrames, and 
+        thus implements many of the functionalities found in an Pandas DataFrame
+        . The dataframe is well suited for running on partioned datasets, where 
+        partitions are clearly defined by an index.
+
+        In order to execute the Dask graphs in an efficient manner, proper 
+        scheduling is required. Dask uses dynamic scheduling for all their 
+        schedulers. When a worker reports completion of a task, the runtime 
+        state is updated to reflect that the task has been completed, tasks that
+        can now be executed are marked and the data which can be released is 
+        determined. To ensure that workers complete execution of related tasks 
+        before commencing unrelated ones, a \textit{last in first out} policy is
+        used. As Dask graphs are entirely decoupled from the scheduler that they
+        use, Dask users are free to design their own scheduler that best suits 
+        their needs.
 

### 2018-09-22 paper.tex
some reformatting

@@
 
-    \section{spark} \begin{itemize} \item based on
-        map-reduce \item in-memory processing \item lazy evaluation \item
-        general framework \end{itemize}
-
-    \subsection{Scientific workflows in Spark} \subsubsection{Introduction}
-    emphasis on data rather than on interface. transformation describe how
-    interfaces interact with the data.  \subsubsection{ariel's paper and
-    thunders astronomy - } \subsubsection{Spark on hpc's or } \note{large
-    dataset for valduriez} Prior to examining how Big Data frameworks could be
-    improved on HPC clusters, it is important to have a baseline of how it
-    performs natively on an HPC cluster.  Particularly, how it behave when
-    processing scientific workflows. In a recent study done by \cite{valduriez},
-    they examined the scalability of Spark on an HPC cluster using a black-box
-    scientific workflow. Five tests were performed in their study: 1) Measuring
-    execution time when task and resources available increase proportionally, 2)
-    Measuring the execution time of very short and short tasks, 3) Measuring the
-    execution time of short (5s) to long tasks (120s), 4) Varying the number of
-    tasks with fixed task durations, 5) Mixed task durations and varied number
-    of tasks. Their results show Spark behaves as expected when tasks and
-    resources are increased exponentially, and that Spark scales better than
-    expected with tasks of longer durations. This is due to the fact that the
-    scheduler is overloaded in the case of many short tasks in addition to their
-    being a significant amount of I/O occurring between each task. Although I/O
-    is problematic in both data center and HPC infrastructures, HPC is optimized
-    for data transfer over the network rather than disk I/O, and therefore, such
-    frequent I/O negatively impacts the system. However, many scientific
-    workflows are composed of longer tasks, and therefore, Spark scales very
-    well with scientific workflows in an HPC environment.  \section{Use of
-    containers with Big Data frameworks} \section{Hadoop Distributed file
-    system} \note{maybe should go in between mapreduce and spark}
+    \section{spark} 
+        \begin{itemize} 
+            \item based on map-reduce 
+            \item in-memory processing 
+            \item lazy evaluation 
+            \item general framework 
+        \end{itemize}
+
+    \subsection{Scientific workflows in Spark} 
+        \subsubsection{Introduction}
+            emphasis on data rather than on interface. transformation describe 
+            how interfaces interact with the data.  
+        \subsubsection{ariel's paper and thunders astronomy - } 
+        \subsubsection{Spark on hpc's or } \note{large dataset for valduriez} 
+            Prior to examining how Big Data frameworks could be improved on HPC 
+            clusters, it is important to have a baseline of how it performs 
+            natively on an HPC cluster.  Particularly, how it behave when
+            processing scientific workflows. In a recent study done by 
+            \cite{valduriez}, they examined the scalability of Spark on an HPC 
+            cluster using a black-box scientific workflow. Five tests were 
+            performed in their study: 1) Measuring execution time when task and 
+            resources available increase proportionally, 2) Measuring the 
+            execution time of very short and short tasks, 3) Measuring the
+            execution time of short (5s) to long tasks (120s), 4) Varying the 
+            number of tasks with fixed task durations, 5) Mixed task durations 
+            and varied number of tasks. Their results show Spark behaves as 
+            expected when tasks and resources are increased exponentially, and 
+            that Spark scales better than expected with tasks of longer 
+            durations. This is due to the fact that the scheduler is overloaded 
+            in the case of many short tasks in addition to their being a 
+            significant amount of I/O occurring between each task. Although I/O
+            is problematic in both data center and HPC infrastructures, HPC is 
+            optimized for data transfer over the network rather than disk I/O, 
+            and therefore, such frequent I/O negatively impacts the system. 
+            However, many scientific workflows are composed of longer tasks, 
+            and therefore, Spark scales very well with scientific workflows in 
+            an HPC environment.  
+    \section{Use of containers with Big Data frameworks} 
+    \section{Hadoop Distributed file system} 
+        \note{maybe should go in between mapreduce and spark}
 

### 2018-09-22 paper.tex
more formatting

@@
             Dask contains three primary abstractions: the \textit{dask.array}, 
-        \textit{dask.bag} and \textit{dask.dataframe}. The \textit{dask.array} 
-        is an abstraction that enables parallel NumPy-like operations using task
-        scheduling. It can additionally be used on data that does not fit in 
-        memory. The \textit{dask.array} used blocked array algorithms to break 
-        down large commonly-used linear algebra functions into smaller ones that
-        can be partially parallelized.
-
-        Dask array functions all create an \textit{Array} object which contains 
-        the dask graph \textit{.dask}, details on the overall shape and the 
-        shape of each subdivided portion (chunk) of the array (\textit{.chunks})
-        , the \textit{.name} representing which keys map to the results, and the
-        data type.
+            \textit{dask.bag} and \textit{dask.dataframe}. The 
+            \textit{dask.array} is an abstraction that enables parallel 
+            NumPy-like operations using task scheduling. It can additionally be 
+            used on data that does not fit in memory. The \textit{dask.array} 
+            used blocked array algorithms to break down large commonly-used 
+            linear algebra functions into smaller ones that can be partially 
+            parallelized.
+
+            Dask array functions all create an \textit{Array} object which 
+            contains the dask graph \textit{.dask}, details on the overall shape
+            and the shape of each subdivided portion (chunk) of the array 
+            (\textit{.chunks}), the \textit{.name} representing which keys map 
+            to the results, and the data type.
 
@@
 
-        The Dask bag is similar to that of a Python list, however, order cannot 
-        be guaranteed. As Dask bags typically operate on Python objects, 
-        performance is bound by the Global Interpreter Lock (GIL), therefore 
-        multiprocessing schedulers must be used instead of multithreading ones.
-
-        The Dask bag contains functions like \textit{map} and \textit{filter}, 
-        in addition to many other functions that can be found in the PyToolz API
-        . It is particularly well suited for the handling of data dumps.
-
-        The Dask dataframe consists of a collection of Pandas DataFrames, and 
-        thus implements many of the functionalities found in an Pandas DataFrame
-        . The dataframe is well suited for running on partioned datasets, where 
-        partitions are clearly defined by an index.
+            The Dask bag is similar to that of a Python list, however, order 
+            cannot be guaranteed. As Dask bags typically operate on Python 
+            objects, performance is bound by the Global Interpreter Lock (GIL), 
+            therefore multiprocessing schedulers must be used instead of 
+            multithreading ones.
+
+            The Dask bag contains functions like \textit{map} and 
+            \textit{filter}, in addition to many other functions that can be 
+            found in the PyToolz API. It is particularly well suited for the 
+            handling of data dumps.
+
+            The Dask dataframe consists of a collection of Pandas DataFrames, 
+            and thus implements many of the functionalities found in an Pandas 
+            DataFrame. The dataframe is well suited for running on partioned 
+            datasets, where partitions are clearly defined by an index.
 
@@
             schedulers. When a worker reports completion of a task, the runtime 
-        state is updated to reflect that the task has been completed, tasks that
-        can now be executed are marked and the data which can be released is 
-        determined. To ensure that workers complete execution of related tasks 
-        before commencing unrelated ones, a \textit{last in first out} policy is
-        used. As Dask graphs are entirely decoupled from the scheduler that they
-        use, Dask users are free to design their own scheduler that best suits 
-        their needs.
+            state is updated to reflect that the task has been completed, tasks 
+            that can now be executed are marked and the data which can be 
+            released is determined. To ensure that workers complete execution of
+            related tasks before commencing unrelated ones, a 
+            \textit{last in first out} policy is used. As Dask graphs are 
+            entirely decoupled from the scheduler that they use, Dask users are 
+            free to design their own scheduler that best suits their needs.
 

### 2018-09-25 oldpaper.tex
restructured report

--- 2018-09-22 paper.tex
@@
         \end{itemize}
+
+        Apache Spark is a unified BigData processing framework aimed at enabling
+        the description of a varietly of workflows encompassing different
+        domains. It was inspired by the mapreduce programming paradigm and 
+        incorporates many aspects present in the hadoop framework, but extends 
+        this model to support interactive queries and iterative workflows. 
+        As having a unified engined can limit performance, Spark 
+        leverages techniques such as in-memory processing to ensure that 
+        processing times remain optimal.
+
+        Spark's main abstraction is the Resilient Distributed Dataset (RDD). 
+        The RDD is an
+        immutable fault-tolerant collection of objects distributed across nodes.
+        It incorporates lineage capture, lazy evaluation, data-locality, caching
+        and persistance to meet Spark's goals. In BigData frameworks prior to 
+        Spark, fault-tolerance was achieved through data replication on disk.
+        As writing data to disk is a rather costly procedure, Spark instead
+        maintains the data in-memory and achieves fault-tolerance of partitions
+        by recording lineage. Should the RDD be too large to be maintained 
+        in-memory, it is possible
 

### 2018-09-25 paper.tex
restructured report

@@
 
-\usepackage{hyperref} \usepackage{xcolor} \usepackage{ulem}
+\usepackage{titlesec}
+\titleformat{\chapter}
+  {\normalfont\LARGE\bfseries}{\thechapter}{1em}{}
+\titlespacing*{\chapter}{0pt}{3.5ex plus 1ex minus .2ex}{2.3ex plus .2ex}
+\usepackage[left=2cm, right=2cm, top=2cm, bottom=2cm]{geometry}
+\usepackage{hyperref} 
+\usepackage{xcolor} 
+\usepackage{ulem}
 
@@
 
-\begin{document} \title{Pipeline systems and infrastructure for the efficient
-and open processing of Big neuroimaging Data} \author{Valerie Hayot-Sasson}
-\maketitle \begin{abstract} Text of abstract.  \end{abstract} \tableofcontents
-    \chapter{Introduction} intro here -mention hpc, data centre, cloud.
-\chapter{BigData and Open science in neuroimaging} \note{Likely will not be
-split in two separate sections} \section{large images} \begin{itemize}  \item
-        BigBrain \item Micro CT \item EM \end{itemize} \section{large datasets}
-        \note{off the top of my head, but need to double check} \begin{itemize}
-        \item HCP \item UK Biobank \item ADNI \item OpenFMRI \item BIDS/BIDS
-        apps \end{itemize} An effective engine to process such data requires the
-        following characteristics \begin{itemize} \item in-memory computing
-        \item data locality \item lazy evaluation \end{itemize}
-            \section{open-science /data sharing platforms for neuroimaging} is
-            open science just data sharing, computing.
-            \subsection{Introduction} \subsection{CBRAIN} \subsection{OpenNEURO}
-            \subsection{NeuroData} \subsection{Apache Airvata} \chapter{Workflow
-            engines for neuroimaging data} \section{Definition of
-            workflows/pipelines}
+\begin{document} 
+    \title{Pipeline systems and infrastructure for the efficient
+            and open processing of Big neuroimaging Data} 
+    \author{Valerie Hayot-Sasson}
+    \maketitle
 
-    Broad definition of activities - between the tool and the engine.
-    \section{Nipype} \subsection{summary} Nipype\cite{nipype} is a popular
-    neuroimaging pipelinening framework written in Python.  It provides users
-    with uniform access to the rich ecosystem of neuroimaging software libraries
-    (e.g. SPM, FSL, Freesurfer) through its \textit{Interfaces}. \weird{Should
-    users want to use their own custom tool, it is also possible to create an
-    interface for that}. To execute Python code in Nipype, the \textit{Function}
-    interface can be used \tristan{for arbitrary code?}. The framework is also
-    easy to use, and thus does not contribute to the steep learning curve new
-    researchers must face.
+    \begin{abstract} Text of abstract.  
+    \end{abstract} 
+    \tableofcontents
+    \chapter{Introduction}
+    In recent years, the volume of neuroimaging data acquired has exceeded
+    both the storage and computation capacity of a standard research 
+    lab workstation. With the advancement of data sharing technologies, this 
+    data has been made widely available, with the only limiting factor impeding
+    research on this data being access to infrastructure and software that 
+    enables efficient processing on such infrastructure. Many research labs do
+    have access to high performance computing (HPC) clusters, however, without
+    efficient software to process this data, processing can take an excessive 
+    amount of time. Ensuring efficient processing of data is complex and 
+    likely beyond the knowledge of an average neuroimaging researcher. As a 
+    result, frameworks for the efficient processing of neuroimaging data need
+    to be developed.
 
-        A pipeline in Nipype, consists of a data analysis. explicit
-        \textit{Workflow} of connected \textit{interfaces}. The workflow is a
-        directed acyclic graph that connects the outputs of one interface to be
-        the inputs of another. A workflow may also consist of connected
-        workflows.
+    Neuroimaging big data comes in two formats: 1) large images and 2) large 
+    datasets. Large images consist of a singular images ranging from 100s of
+    gigabytes to terabytes in size. A well-known example of such an image is the
+    BigBrain \note{cite here}, a histological image of a 69 year-old man. 
+    At its highest resolution, it is 1x1x20micrometers \note{fix this} or 1TB in
+    size. Other examples of large images can be found in electron microscopy 
+    (EM), polarized light images (PLI) and micro coherence tomography (microCT)
+    \note{need to do more research here}. Images at such high resolution are
+    important to researchers as they give insights into aspects not otherwise
+    detectible in images at lower resolutions. However, due to their size and 
+    lack of resources available to process such images, research on this data 
+    remains limited.
 
-                \tristan{The text below is about workflow execution while the
-                text above is about workflow description, this should be a new
-                paragraph.}  Workflows may be executed in parallel either
-                locally or an a cluster through the use of a plugin in the
-                Workflow's run function. For debugging, Nipype uses
-                \textit{graphviz} to generate a static graph representing the
-                Nodes and their relationships. \tristan{debugging might be a
-                third theme, in addition to description and execution.
-                Provenance might go in debugging.}
+    In contrast, large datasets are consists of many images, typically arising
+    from multiple different subjects, that are too big to fit in storage. 
+    The images alone being small enough for an average researcher to process. 
+    Example of these datasets include the UK Biobank, HCP Project, ADNI dataset,
+    OpenfMRI, amongst many others. These datasets are very large. For instance, 
+    the UK Biobank is expected to exceed \note{X} Petabytes in size. As subsets
+    of these datasets are manageable for the average researcher to process, 
+    much analysis has gone into them. Moreover, unlike for large images, 
+    toolkits have been developed for researchers to process this data. However,
+    the toolkits do not enable researchers to process these entire datasets as
+    a whole. Therefore, it is crucial to develop efficient frameworks for the 
+    processing of these datasets in order for researchers to harness the data
+    in its entirety.
 
-         Each \textit{interface} within the workflow must be contained within a
-         \textit{Node} or a \textit{MapNode} object, the MapNode class being a
-         subclass of the Node class. Wrapping an interface in such objects
-         ensures that interfaces are executed within a uniquely named
-         directories, which in turn enables provenance tracking. Node objects
-         also hash inputs, provide the ability to iterate over inputs and cache
-         results. Hash inputs are useful in the case of Workflow recomputation,
-         where Nipype will only recompute Nodes whose inputs have changed since
-         the previous run.
+        \section{Platforms for  neuroscience}
+        \section{Infrastructure}
 
-        Inspired by the MapReduce paradigm, the MapNode copies the interface to
-        each input and executes it independently. \tristan{How are inputs
-        defined/combined?} A reduce is subsequently performed by the MapNode to
-        return the output in the form of a list. This differs from a Node object
-        which can only execute the interface on a single input at a time.
-
-        To enable execution of different parameters on the input, Nodes have a
-        property known as \textit{Iterables}. Similarly to a MapNode, a copy of
-        the \textit{interface} will be made for each input. However, with
-        \textit{iterables}, a copy of each dependent node will also be made.
-
-        \subsection{Limitations} Nipype, although meeting the various needs of
-        researchers in Neuroinformatics \tristan{vague}, was not designed with
-        the processing of Big Data in mind. Moreover, its performance relative
-        to current Big Data frameworks remains unknown \tristan{You could refer
-        to your MS thesis here}.  \note{nipype performs a lot of disk I/O.
-        Recomputation of workflows vs RDD caching}
-
-    \section{PSOM} \subsection{summary} While Python is a popular programming
-    language amongst neuroinformaticians, it is not the only one that is
-    commonly used. To address the needs of the Matlab/Octave community, PSOM
-    (Pipeline System for Octave and Matlab) was developed. Similarly to Nipype,
-    PSOM is easy-to-use, can run in distributed environments and can interface
-    with tools written in other languages.
-
-        \subsection{pipeline description} A pipeline in PSOM, is a series of
-        connected jobs, which can be represented as a (static?) directed acyclic
-        graph. The pipeline itself is implemented using the Matlab/Octave data
-        type. Each job within the pipeline is assigned a unique name. There is
-        only one mandatory field in the job description (command). The remaining
-        four optional fields include the list input files (files\_in), the list
-        output files (files\_out), the list of files to be deleted after
-        execution (files\_clean) and any Octave or Matlab datatype containing
-        the remaining parameters necessary for job execution (opt).  Jobs in
-        PSOM run in a protected environment where they only have access to the
-        parameters that are passed.
-
-        In order to determine whether the opt field was correctly set, the
-        generic function \textit{psom\_struct\_defaults} may be applied. This
-        function will set the default values, ensure all mandatory fields were
-        supplied and issue warnings for all unidentified attributes.
-
-        To define modules in PSOM, a special Octave/Matlab function type, known
-        as "bricks" is used. Bricks employ the same parameters as jobs but lack
-        the file\_clean parameter. All bricks contain an opt.flag\_test boolean
-        flag, which, if set to true, allows the user to perform a dry-run of the
-        brick. A dry-run of a brick consists of updating the default parameters
-        and validating that the brick can successfully execute with the
-        parameters provided. Bricks also have an option to specify the output
-        folder (opt.output\_folder) for brick outputs to be stored.
-
-        The pipeline generator is an easy way to create complete pipelines with
-        minimal information provided. A pipeline generator may accept at most
-        the file\_in and opt arguments. It then builds the pipeline based on
-        this information and the default parameters specified in each brick
-        applied by the pipeline generator.
-
-        \subsection{pipeline execution}
-
-        To execute a pipeline, the user must provide the pipeline structure and
-        the pipeline configuration options. The latter consists of the log
-        directory, which must be provided, and the pipepline execution modes
-        (e.g., session, background, batch, qsub, msub), in which the default
-        mode is \textit{background}. Pipeline execution is achieved through the
-        execution of the \textit{psom\_run\_pipeline} command.
-
-        Pipeline execution is broken down into three distinct modules: 1)
-        initialization stage, 2) pipeline manager and 3) job manager. The
-        initialization stage performs some checks to ensure that the pipeline
-        can execute successfully (e.g., verify that pipeline is a DAG, check
-        that output files are generated only once, check that all input files
-        not generated by the pipeline exist on disk, delete any already existing
-        output files), as well as performance checks (e.g.  do not recompute
-        pipeline steps that have already been executed, create necessary output
-        folders to avoid creating them for each individual job). The
-        initialization manager can determine which jobs need to be executed by
-        comparing the current pipeline jobs with the details of those previously
-        executed stored in the logs folder. The jobs that have been executed can
-        have the following states: 1) none, 2) finished and 3) failed. The
-        initialization manager stage the jobs with a \textit{none} or
-        \textit{failed} status for recomputation. For jobs whose description has
-        changed since the previous \textit{finished} execution, the
-        initialization will also stage it for recomputation. Should the user
-        want to reexecution a previously executed \textit{finished} job, the
-        user may force reexecution by setting \textit{opt\_pipe.restart}
-        parameter to include all jobs to be reexecuted.
-
-        The pipeline manager is responsible for monitoring the state of the
-        pipeline and submitting jobs for execution. Its duties begin after the
-        initialization manager has completed execution. Jobs can be submitted by
-        the pipeline manager once all their dependencies have been satisfied.
-        The dependencies of a job are designated as met if there are none or
-        when its parent jobs within the dependency graph are tagged as
-        \textit{finished}. In addition, a job is considered to be successfully
-        completely when all the dependencies of its children jobs have been
-        satisfied. Depending on the pipeline configuration, there may be a limit
-        to how many jobs can run in parallel. Once a job has finished executing,
-        an empty tag file is created with a filename containing the final state
-        in the logs folder. The pipeline manager continuously parses through the
-        logs folder in search of these state files to delete and update the
-        history file with the job state. Both the pipeline manager and the jobs
-        can be executed within the current session, locally in a independent
-        session, or in an independent session on a cluster
-
-        Every time a job is submitted by the pipeline manager for execution, a
-        job manager is started. It is a Matlab function responsible for creating
-        the job profile and generating the job-specific logs and the job exit
-        tags. When the job terminates, the job manager will verify that all
-        output has been correctly generated and assigns a job tagged based on
-        this verification that will be reported back to the pipeline manager.
-
-        \subsection{debugging and provenance}
-
-        The information stored in the logs folder is complete with details
-        allowing for both debugging and provenance. The logs folder contains six
-        files: 1) PIPE\_history.txt, 2) PIPE\_jobs.mat, 3) PIPE\_status.mat, 4)
-        PIPE\_logs.mat, 5) PIPE\_profile.mat and 6) PIPE.mat. PIPE\_history.txt
-        includes the execution history of the pipeline manager. PIPE\_jobs.mat
-        and PIPE\_profile.mat are both Matlab/Octave files containing each
-        executed job as a Matlab variable. They differ in that PIPE\_jobs.mat
-        contains the latest executed jobs and can be used to reexecute the
-        pipeline at any moment in time and PIPE\_profile.mat contains the
-        execution time of each job \note{i think both can be used to reexecute
-        pipeline from scratch. Also unclear whether PIPE\_profile.mat contains
-        more than just latest jobs}. It is also possible to access any of the
-        previously used parameters through PIPE\_jobs.mat. PIPE\_status.mat
-        stores the status of each job as a string variable, whereas
-        PIPE\_log.mat stores each job's log as a string variable. Job logs, as
-        generated by the log manager, consist of a text copy of the execution,
-        in addition to the user name, date, time and operating system used to
-        execute the job. Finally, all configuration details are stored in
-        PIPE.mat.
-
-        \subsection{limitations of PSOM} \section{SPM and matlabbatch}
-        \note{This section may not make much sense. Have not really yet fully
-        understood matlabbatch} \subsection{summary} SPM is a very popular
-        neuroimaging toolkit, and as a result, neuroimaging pipeline engines
-        such as Nipype provide interfaces for it. SPM is also equipped with its
-        own neuroimaging pipeline engine known as \textit{matlabbatch}.
-        Similarly to PSOM and as its name suggest, it is written in Matlab.
-
-        \subsection{pipeline description} Pipelines in matlabbatch consist of a
-        series of connected data processing steps known as "modules" \note{DAG??
-        probably but not explicitly listed anywhere}. All SPM modules, helper
-        modules and their respective dependencies can be defined using SPM's GUI
-        \note{can probably also be done through command-line}.
-
-        For any matlabbatch job, it is possible to create a skeleton of it. A
-        skeleton is a Matlab script containing a commented list of necessary
-        inputs at the top and all code necessary to initialize and run the
-        script. Before running, the skeleton will need to be configured with the
-        input names and other necessary parameters.
-
-        All configuration options can be set/changed by modifying the
-        \textit{cfg\_$\langle type \rangle$ } files.
-
-        Defaults can be set within the configuration files, through a default
-        file, or in a \textit{.def} field for leaf nodes \note{terminal
-        jobs???}. The latter updates at the start of the new job execution
-        whereas the former are configured at the beginning of the
-        SPM/matlabbatch startup.
-
-        Virtual outputs represent the dependencies between jobs \note{unsure}.
-        They contain three paramaters of which only one is optional: sname,
-        src\_output, tgt\_spec. sname is the display name that will be stored in
-        the dependecies list, whereas src\_output is a subscript reference used
-        to refer to the output that is produced at runtime. The optional
-        parameter, tgt\_spec, is a description of the kind of inputs the outputs
-        produced can be a dependency for. The virtual output callbacks are
-        evaluated when cfg\_(ex)branch, cfg\_choice and cfg\_repeat have the
-        correct number of child nodes within all their in-tree nodes.
-
-        \subsection{pipeline execution} Pipeline execution can be achieved
-        either through the provided GUI or by the command-line interface. The
-        GUI, however, requires significant user interaction, as the user must
-        repeat the "Runs" command as many times as necessary to reexecute the
-        command on multiple data. The CLI, however, avoids this with the help of
-        its \textit{spm\_jobman} interface.
-
-        The first step to running a batch pipeline is to perform the
-        initialization. This may include configuring the
-        \textit{spm\_defaults.m} code to run in command line and executing the
-        \textit{spm(’defaults’, MODALITY)} (where MODALITY represents the
-        modality used in the pipeline) and \textit{spm\_jobman(’initcfg’)}.
-
-        After the intialization is complete, a job may be executed using the
-        \textit{spm\_jobman(’run’, job[, input1, input2 ...])} command. This
-        command requires the job names as well as its necessary inputs specified
-        in the correct order. The job can be specified as a variable, the name
-        of a script that will generate a job variable or a cell list containing
-        a mixture of scripts and job variables.
-
-        \subsection{debugging and provenance} All batch scripts can be saved as
-        a Matlab scripts. These scripts not only contain the code that was
-        executed, but also all input parameters. Should any script fail in the
-        process of executing, an error will be printed to the Matlab console. If
-        the user is using the GUI, the error output may also be saved to a
-        Matlab file.
-
-        \subsection{limitations of SPM}
-
-    \section{FastR} \subsection{summary} FastR is another Python-based
-    neuroimaging pipeline framework that provides added features such including
-    access to different versions of the same command-line tool \note{criticizes
-    nipype for not having such a feature, but nipype can access boutiques
-    tools...so...} and compatibility with various possible filesystems. In
-    addition to its added features, FastR also guarantees a simple-to-use
-    framework for building pipelines written in any languages, has built-in
-    provenance support, provides the seamless ability to parallelize pipelines
-    with the use of plugins and avoids the numerical instability issue with the
-    simple integration of containerized pipelines. Similary to many other
-    pipeline framework, it also employs a DAG to describe data flow.
-
-        \subsection{Pipeline description}
-
-        A pipeline in FastR is a \textit{Network} of connected \textit{Nodes} in
-        which the directed edges represent the flow of data.  Each \textit{node}
-        contains a JSON or XML template known as a \textit{Tool}.
-
-\textit{Tools} describe the details of the underlying application that is to be
-executed at the node. Tools consist of three components: general metadata,
-target and interface.  The general metadata provides details on the Tools such
-as name, version, authors, etc, whereas the target specifies details on the
-execution environment.  The interface, on the other hand, provides details on
-the inputs and outputs of the tool.
-
-A JSON schema is required to validate the python object generated by the Tool.
-Fastr currently provides two schemas for object validation: the \textit{Tool}
-schema and the \textit{FastrInterface} schema. It is, however, possible to add a
-custom schema to FastR and define an \textit{Interface} for it.
-
-The \textit{Interface} class is responsible for executing a call to the command
-line tool.  It dictates the tags that must be defined within the interface tag
-of the \textit{Tool}. For instance, the generic interface class requires details
-on the input and output's id, cardinality (number of input values) and DataType
-(a python object denoting the details of the Input or Output's data type).
-DataTypes can also be grouped, allowing an inputs containing various files types
-to be inputted to applications which can support the file types.  Verification
-is done by FastR to determine if the linked Input and Output data types are
-compatible.
-
-Input samples can consist of one or more values, in which the output is also a
-sample consisting of the same number of values. During the instances in which it
-is desirable to apply the same task to multiple inputs, a multidimensional array
-of samples, known as a collection, can be supplied to a \textit{Node} for
-processing. If there are multiple input samples, they can be passed to a node in
-pairwise behaviour (e.g. 1-1) or cross-product behaviour (e.g. nxn) \textit{need
-to better explain this}.
-
-Output samples may need to be expanded (i.e. a single sample of values becomes
-separate samples) or collapsed (i.e. multiple samples become a single sample of
-values).  FastR provides data link flow directives for such operations.
-
-In FastR, there are four types of \textit{Nodes}: normal Nodes, Sink Nodes,
-Source Nodes and Constant Nodes. Data enters the network through Source or
-Constant Nodes and exit through Sink Nodes. Constant nodes differ from Source
-Nodes in that their input data are defined by the network \note{Not sure if
-saying this correctly}. Normal nodes are responsible for processing the data.
-Nodes are connected to each other through links that highlight the dependencies
-and form the \textit{Network}. Once the \textit{Network} is defined, it can be
-executed.
-
-The data type of the Source Node and Sink Node must be defined in the
-\textit{Network} definition.  The data is define at runtime by evaluation of the
-uniform resource identifier (URI). Multiple URIs can be defined through the use
-of wildcards and searches. Plugins are used to read and write data to different
-filesystems specified in the URI scheme. Currently, FastR supports virtual file
-system (VSF), csv files and XNAT. Plugins to support other filesystems can
-easily be added to FastR. In addition, plugins that can only read, write, or
-perform searches can also be created.
-
-        \subsection{Pipeline execution} To commence pipeline execution, the
-        \textit{Network}'s execution function is called on the Submission host.
-        FastR then analyses the network and its dependencies and breaks down the
-        Network into chunks that can be further processed. From the chunks, the
-        execution order of the \textit{Nodes} is determined.
-
-During \textit{Node} execution, a job is created for each input/output
-combination. Jobs are then executed on the execution host (i.e. local, cluster,
-cloud), as determined by the execution engine plugin. FastR currently only
-offers support for local and cluster plugins, however can easily be extended to
-support other plugins.
-
-While the job is executing, input URIs are translated to their actual
-path/values, the execution environment is set up by the \textit{Tool} and the
-\textit{Tool} command is executed by the \textit{Interface}. When the
-\textit{Interface} returns the results, the output data is validated and
-translated to URI format.
-
-After job completion, the results are returned to the Submission Host which
-reads the output and updates the network accordingly. Should a chunk be
-completed, the next chunk will be processed with the updated information, until
-all chunks have been processed and network execution is complete.
-
-        \subsection{Debugging and Provenance} In FastR, a provenance document is
-        generated for each resulting file generated in the Prov-N format. Data
-        recorded in the provenance documents include \textit{Tool} versions,
-        input/output paths/values and their respective checksums, start and end
-        time of execution for each \textit{job}, stdout and stderr logs, the end
-        status and the complete environment description.
-
-    Similarly to Nipype, the FastR network can be visualized using graphviz.
-
-    \section{LONI} \subsection{Summary} The LONI pipeline processing environment
-    facilitates the development, sharing and execution of pipelines. Workflow
-    Activities, known as \textit{Modules} in LONI, are described using
-    Extensible Markup Language (XML), which can easily shared between user
-    groups. LONI runs on the Java platform, allowing it to be platform
-    independent and has support DRMAA clusters and symmetric multiprocessing
-    (SMP) systems. In order to capture provenance information, LONI is also
-    equipped with a provenance manager.
-
-        \subsection{Pipeline description} LONI uses the dataflow programming
-        model to represent the pipeline. Data flows between \textit{Modules}
-        through \textit{data pipes}. Sub-pipelines are known and
-        \textit{Pipelets}. Modules are defined by the input arguments/streams
-        and output arguments/stream. Arguments in LONI may be specific
-        filetypes, flags or values listed in the order in which the application
-        requires. The provided arguments may be required or optional and can
-        also be duplicated, if necessary. File inputs in the form of URLs are
-        accepted by LONI and will be downloaded at runtime. List of inputs may
-        also be provided to process multiple inputs with copies of the same
-        module in parallel (batch processing). Conversely, it is also possible
-        to group list of inputs so as to process them all by the single module
-        \note{Need to express this better}. Input/Output arguments may be added
-        to the table of global variables for easy access throughout the
-        pipeline. It is possible to apply variable modifiers to the variables
-        stored within the table.
-
-Intermediary outputs as well as the file output are treated as temporary by
-LONI. They are scheduled for deletion before a new pipeline is executed. It is,
-however, possible to preserve these files.
-
-        \subsection{Pipeline execution} As a module completes it execution, its
-        output gets transmitted through the dataflow and the dependent modules
-        are signaled that their input data has arrived.  Modules who have met
-        all their dependency requirements are ready for execution.  Should a
-        module have no dependecies, it can be executed immediately. Modules
-        ready to be executed are placed in a FIFO launch queue to avoid
-        straining the platform.
-
-       LONI offers a built in client/server strategy that uses Java's remote
-       method invocation (RMI) libraries to communicate between servers. This
-       enables modules, pipelets or pipelines to be executed on servers
-       different from the client. It also enables full pipelines to consist of
-       connected modules and pipelets located on different servers. It is
-       favorable to have such a configuration in LONI, as the tools required for
-       processing may not be available on client server, but located in another.
-       The client is responsible for dictating all file transfers and module
-       launches that are to occur. The client authenticates with the servers
-       through a one-way hash. Only files produced by the pipeline are
-       accessible to the client and only temporary files created by the server
-       may be written to the server.
-
-        \subsection{Debugging and provenance} For each dataset processed in a
-        LONI pipeline, there exists a corresponding provenance (.prov) XML file.
-        The \textit{Provenance Editor}, a standalone platform-independent
-        application, reads the metadata from the image headers, extracts
-        relevant provenance information from them and creates the provenance
-        files.  This data can be modified by the user in the to correct
-        inaccuracies or append additional details before being saved. During
-        pipeline execution, the provenance file will be edited by the LONI
-        pipeline to include details on each processing step, the environment in
-        which the data was processed and tool binaries used.
-
-    \section{Automatic Analysis} \subsection{Summary} Automatic analysis (aa) is
-    another Matlab-based modular workflow engine, having started of as a
-    workflow engine for SPM. User may define their pipeline through a Matlab
-    user script and specify the tasks in addition to their input and output
-    types through an XML module interface. Modules used in pipelines must be
-    integrated into aa which can be easily accomplished through XML interfaces.
-    aa workflows can be executed locally, or in parallel through qsub or condor.
-
-        \subsection{Pipeline description} aa pipelines have two main components:
-        1) the Matlab user script and 2) the XML tasklist. The user script is a
-        script that defines which data will be processed, any necessary
-        parameter, and how it will be processed. aa provides an example user
-        script, which can be modified to meet user's needs. The entire pipeline
-        in aa is defined in a single structure, known as aap, which makes the
-        task of record-keeping simple.
-
-        The XML tasklist is an ordered list of all the modules that will be
-        applied to the data, that is read by the user script. This list does not
-        specify any dependencies as those dependendecies can be inferred by the
-        pipeline through analysis of the module interfaces. Modules can be
-        separated into two sections: initialization and main. Modules listed in
-        the initialization section are to be executed for every task.
-        Conversely, those modules liste in the main section are only executed
-        once for every task, unless otherwise specified.
-
-        In order to ensure that the correct output data is being operated on,
-        each task is provided with its own unique folder. The aa engine will
-        transmit the necessary input data to the folder before task execution
-        and also acquiring the outputs after task completion. Output folders are
-        numbered such that if a given module is invoked multiple times, each of
-        those folders are assigned an incremental ordering corresponding to the
-        order in which they have been executed \note{sentence might not make
-        sense}. Should all data be required for each task in the tasklist, all
-        the data will be copied to each task folder.
-
-    Similarly to a pipeline, each module is comprised of an XML interface and a
-    Matlab source file. A module's XML file consist of specifications on
-    required input and output modalities, as well as the desired domain. One of
-    the most important properties include the domain at which the module
-    operates. While there exist mainy domains for modules and new ones can
-    easily be created and integrated by the user, the most prominently used ones
-    include: "study", "subject", "session". As their name suggests, a module
-    with the domain "study" will only be executed once for each time it appears
-    in the tasklist, whereas "subject" executes the module for each subject
-    within the task and "session" executes once for each session within a
-    subject.
-
-    In addition to module domains, another important module property is the
-    input and output data stream types. Data stream types denote what input and
-    output data type each module produces. Commonly used data stream types
-    include "epi", "structural" and "dicom\_header".
-
-    Optional fields specified in a module include qsub, for estimates of
-    requirements for parallel processing, and \textit{performanceoutput}, for
-    garbage collection.
-
-   Analysis parameters can easily be configured in aa. Any variable defined by
-   the \textit{aaprecipe} command may be modified. Parameters can also be
-   defined for an instance of a module within the XML tasklist. Alternatively,
-   it is also possible to create an XML file that inherits parameters from
-   another XML file and overrides some or all of the parameters.
-
-   To set defaults and site-specific configurations, it is possible to either
-   create a local config file or by creating another parameter default file
-   which inherits the original one.
-
-    aa also supports the creation of branched tasklist. This allows a section of
-    the pipeline steps to undergo different tasks or the same tasks in a
-    different order in one pipeline rather than having to create separate
-    pipelines.
-
-        \subsection{Pipeline execution} Once the definition of the user-script,
-        XML tasklist and modules is complete, the pipeline is ready for
-        execution. Dependencies in aa are established based on input and output
-        data stream types. The aa engine attempts to match the last task
-        executed with a corresponding data stream type to the current type. It
-        is, however, possible to specify which module provides the input to the
-        current module by using fully qualified stream references.
-
-        During execution, a map of all the modules that need to be executed and
-        their respective dependencies is generated. Modules that have completed
-        contain a flag file indicating their completion within their instance's
-        root directory. Such modules will not be re-executed by the engine.
-        Additionally, if it is required that a previous module is re-executed,
-        the done flag is removed from all dependent modules.
-
-        Should part of the processing be performed on a different server it is
-        possible to connect the pipelines together through an analysis script.
-        The aa engine will be reponsible for matching the data streams between
-        the servers in order to avoid the copying of data and modules.
-
-        \subsection{Debugging and provenance} In order to retrace the steps of
-        previous executions, aa keeps a detailed record of the events that
-        occurred. These detailes are stored in a Matlab object \note{similarly
-        to PSOM}, which permits reexcution of a previous run or extraction of
-        specific parameter values.
-
-        Additionally, to verify the correctness of the results produced, aa
-        implements tools for quality control and comes with a dedicated module
-        for low-level quality control.
-
-    \section{Pydiper} \subsection{Summary}
-
-        Pydiper is a neuroimaging registration workflow engine implemented in
-        Python. This differs from other workflow engines in that it is specific
-        to the process of image registration, but can, however, be extended for
-        use with any analysis process. Its main features include being easy to
-        use, particularly for non-developpers, undertaking a modular approach,
-        giving the ability to access output files at any stage, and eliminating
-        duplicate stages. Pydiper also provides the added advantage of being
-        equipped with four commonly used registration pipelines, such that
-        unexperienced researchers can use such pipelines out-of-the-box.
-
-        \subsection{Pipeline description} Pydiper workflows can be broken down
-        into three components: atoms, modules and applications. An application
-        is the command line application that encapsulates the pipeline which
-        consists of modules and atoms. Atoms are wrappers for individual
-        operations, typically frquently used MINC tools. Modules are reponsible
-        for linking atoms together.
-
-        Pydpiper provides three main classes to use to develop pipelines. These
-        include the \textit{PipelineStage}, \textit{CmdStage} and
-        \textit{Pipeline} classes. The \textit{PipelineStage} is the base class
-        wrapper for all pipeline tools.  Pipeline tools can be written in Python
-        or can be a command-line application.  For the tools that are
-        command-line applications, the \textit{CmdStage} class, a child class of
-        \textit{PipelineStage}, must be used. Input to \textit{CmdStage} is in a
-        adequately parsed array format containing the specific command-line tool
-        and its arguments. Reusable subclasses of the \textit{CmdStage} class
-        can be created, or \textit{CmdStage} can be called directly.
-
-        To combine pipelines, the \textit{AddPipeline} function of the
-        \textit{Pipeline} class may be used.
-
-        The base class common to all Pydpiper applications is
-        \textit{AbstractApplication}.  Classes that inherit from
-        \textit{AbstractApplication} are command-line tools, with their options
-        set by the \textit{AbstractApplication} class. The class is also
-        responsible for initializing or resuming the pipeline, setting up a
-        logger, and launching the pipeline daemon.
-
-        Pydpiper atoms inherit from the \textit{CmdStage} class in order to wrap
-        the MINC Tools.  They require as input a MINC file, and may additionally
-        also require a target MINC file as second argument. These arguments can
-        be in the form of a string filename or a file handler, however, the same
-        format must be used for the two. An atom may also accepts various
-        optional arguments. Using the inputs provided, the atom will put
-        together the command to be executed.
-
-        \subsection{Pipeline execution}
-
-        In Pydpiper, there are two main classes,\textit{Pipeline} and
-        \textit{PipelineExecutor}, that monitor and manage pipeline execution.
-        The \textit{Pipeline} class's main responsibility is to infer
-        dependencies between pipeline stages from their inputs and outputs. It
-        also manages a queue of runnable stages (stages in which all their
-        dependencies have been met), ensuring that any runnable stage is placed
-        in the queue. It also keeps track of the status of the runnable stages
-        (running, finished, or failed). Should an already existing stage be
-        added to the pipeline, it is automatically skipped by the
-        \textit{Pipeline} class.
-
-        As stages may be added to incomplete pipelines, Pydpiper provides the
-        functionality of resuming pipelines. This is achieved through a pickling
-        mechanism implemented by the \textit{Pipeline} class. The pickled
-        components include the directed acyclic graph, the array of pipeline
-        stages, the stage counter, the hashes of all stages and each of their
-        output files, and the status of each stage.
-
-        The \textit{PipelineExecutor} class, on the other hand, is responsible
-        for the efficient processing of the pipeline through parallelization.
-        Pydpiper's architecture mimics that of client/server, in which the
-        server is the pipeline daemon started by the command-line executable and
-        the executors undertaking the role of the clients. The pipeline's
-        executors each consist of a thread pool in which each thread is
-        responsible for running a single pipeline stage from the queue of
-        runnable stages. To obtain a stage for execution, the executor polls the
-        server to obtain the next available stage. Should the executor have
-        enough memory and processors to execute the stage, it will be executed
-        by the thread and release memory and processors and repoll the server
-        once stage has completed.  Otherwise, the executors will sleep for
-        certain duration and repoll the server again.  Should there be any
-        failed stages, the pipeline will be shut down after all possible stages
-        have finished executing. If additional executors are required by the
-        pipeline, they may be added at any time through the command-line.
-        Details related to the number of threads, number of threads per executor
-        and the amount of memory can be specified through the command-line
-        options.
-
-        \subsection{Debugging and provenance}
-
-        As mentioned previously, Pydpiper uses a pickling mechanism to gather
-        important pipeline details during execution in order to allow for the
-        pipeline to be resumed should any stages have failed.  In addition, to
-        keep track pipeline stage interdependencies in addition to file
-        versioning, Pydpiper utilizes the helper class
-        \textit{RegistrationPipeFH} and its parent \textit{RegistrationFHBase}
-        to create file handlers of the files.
-
-        \note{from github wiki} For debugging, a Pydpiper pipeline creats its
-        own log file and executors typically also each create their own log file
-        that are stored in a \textit{logs/} subdirectory. Verbosity of logging
-        information can be managed through the setting of the PYRO\_LOGLEVEL
-        environment variable. In addition, individual stage stdout/stderr is
-        also redirected to a log file.
-
-        To verify the status of an executing pipeline, the
-        check\_pipeline\_status.py may be executed.
-
-\chapter{Other domains and general scientific worflow engines} \section{Taverna}
-Taverna is an open-source bioinformatics workflow engine, designed to allow
-users to combine distributed web services, other service types (e.g. BioMart
-queries or R-scripts) and local scripts into complex analysis pipelines. The
-main goal of Taverna is to give a unified access to the plethora of services
-already existing on the web without requiring the user to download the services
-locally or to have knowledge on all of the platforms and programming languages
-the services are hosted under.
-
-        \subsection{Pipeline description} Workflows in Taverna are built by
-        combining web services, local scripts and other services together. Web
-        services compatible with Taverna must comply to the Web Service
-        Description Language (WSDL) using the Simple Object Access Protocol
-        (SOAP protocol).  These services can be added to the BioCatalogue or the
-        Biodiveristy catalogue for sharing and reuse of the web services.
-
-        Workflows can be built in the Taverna workbench. Taverna, through the
-        help of the BioCatalogue and the Biodiversity Catalogue is able to
-        provide the user with a variety of webservices to combine into workflows
-        in a drag-and-drop fashion. Workflows may also be composed of
-        sub-workflows. As service inputs and outputs may not match between
-        services, shims are provided.  Similarly to the services, the
-        user-generated workflows can be uploaded the myExperiment repository to
-        be shared with other user.
-
-        \subsection{Pipeline execution} The workflows can be executed remotely
-        utilizing the Taverna Server. The Taverna Server is a an application
-        that provides the full library of possible workflows through
-        webservices. It has an interface, known as the Taverna Player, that
-        enables workflow execution to occur through a web-interface or through
-        third-party clients. The Taverna Server, however, is limited in that it
-        can only execute workflows available on the Server. Taverna Lite is an
-        alternative to the Taverna Server which allows users to upload and
-        execute workflows from myExperiment or elsewhere.
-
-        \subsection{Debugging and provenance} Provenance capture is achieved
-        through the Taverna Provenance Suite. Taverna does not capture
-        provenance of the workflow definition, but generates a new UUID after
-        each definition change. Workflow run provenance captures information on
-        "individual processor iterations" and their respective inputs and
-        outputs and stores this information in a database. The results can then
-        be exported as a workflow trace or as a PROV-O RDF graph using the
-        Taverna-PROV Plugin. The model format used for exported provenance uses
-        both the W3C PROV model and Open Provenance Model format.
-
-    \section{Pegasus} Pegasus is a general-purpose scientific workflow engine
-    that specializes in mapping abstract descriptions of workflows to resources
-    for efficient processing.  Similarly to other workflows, a Pegasus workflow
-    can be represented by a DAG. This DAG is described by the user in DAX format
-    (Directed Acyclic graph in XML).  As the DAX abstract workflow is
-    resource-independent, Pegasus can convert the abstract workflow into an
-    execution workflow that is catered to the resources available.
-
-        \subsection{Pipeline description} Within the DAX, the nodes are the
-        tasks and the execution order is represented by the edges.  In the
-        instance of nested graphs, the nodes in the DAX may also represent a
-        DAX. Pipeline inputs and outputs, as well as pipeline executables, are
-        represented by logical identifiers.
-
-         Various APIs are provided by Pegasus in order for the user to generate
-         the DAX using their preferred language of choice. Available APIs
-         include Perl, Python and Java.  \subsection{Pipeline execution} With
-         the effecient execution of pipelines being Pegasus' main focus, there
-         exist various components that ensure optimal processing. The Mapper's
-         responsability is to map abstract workflow tasks to available
-         resources. In order for the Mapper to successfully convert an abstract
-         workflow into an execution workflow, it requires details on the
-         execution environment. The Mapper will also query various catalogs to
-         determine the location of executables and input and output files. These
-         catalogs can be either file-based or in database format. With all the
-         necessary information provided, the Mapper is equipped to convert
-         abstract tasks into executable tasks. Data nodes are also added for
-         input and output staging. Should the data required for a task already
-         exist in the catalog, the task and all its parents will be removed by
-         the Mapper from the workflow. This functionality is optional, and in
-         some instances, it may be found that recomputation is preferable. If
-         the executables required by the task are not pre-installed on the node,
-         the Mapper will query the \textit{Transformation Catalog} for the
-         stageable binaries.
-
-        The default, Pegasus relies on the HTCondor framework as both the
-        workflow engine and scheduler. The executable workflow, using HTCondor
-        DAGMan, consists of nodes representing jobs.  These jobs each have a
-        corresponding job submit file that details how they are meant to be
-        executed. A job that is ready for execution is released by the DAG and
-        placed into a Condor queue that is managed by the HTCondor Schedd
-        daemon. The default scheduling is done using a local HTCondor pool, but
-        can also be achieved through built-in interfaces with remote
-        job-management protocols, or through SSH to directly submit jobs to
-        remote resource managers.
-
-        Using HTCondor, Pegasus is able to achieve portability. Pegasus is
-        compatible with local, HPC, distributed and cloud execution
-        environments. As distributed environments may not employ a shared
-        filesystem, it is recommended to use HTCondor glideins to ensure that
-        tasks are mapped to nodes where the resources are located. Should a user
-        running their pipeline locally without HTCondor installed, it is
-        possible through \textit{submit host}. Using this method, however,
-        executes the pipeline in serial order. Should the user wish to exploit
-        the pipeline's parallelism, HTCondor is required. Pegasus pipelines are
-        not limited to a single execution environment, and can combine multiple
-        environments to select the best resource configuration to execute each
-        job.
-
-        In order to manage tasks and their dependencies on remote resources,
-        three different execution engines have been developed. The first and
-        simplest execution engine is the \textit{Pegasus Cluster}, which is
-        compatible with all the execution environments. It runs on a
-        single-thread, meaning that it can only executed a single task at a time
-        using an ordered-list of tasks.  The \textit{Pegasus Cluster} is
-        particularly useful for the efficient execution of short tasks and the
-        reduction of data transfers.
-
-        For execution environments lacking a shared filesystem, the
-        \textit{PegasusLite} execution engine can be used. In order for the
-        Pegasus to execute in such environments, the Mapper first adds data
-        transfer nodes to the pipeline to move the data to a data staging server
-        associated with the execution node. Once a job is assigned to a node by
-        the remote cluster scheduler, \textit{PegasusLite} is deployed.
-
-        The final execution engine provided by Pegasus is the \textit{Pegasus
-        MPI Cluster} (PMC). This execution engine is designed for petascale
-        systems for which Pegasus' other strategies leveraging Condor Glideins
-        are unsuitable. It employs the Master-Worker architecture, launching a
-        single Master and many Workers, and using MPI for communication between
-        the Master and Workers. The main limitation of this strategy is that a
-        task must use less resources than that available on a single node. Jobs
-        are not limited to a single execution engine, and thus, multiple can be
-        used in a job to optimize performance.
-
-        As executing a workflow on remote resources can be slow due to job
-        queuing times and data transfers, the Mapper will group short-running
-        task together to reduce to improve runtime.  When tasks located at the
-        same distance from the DAG root have similar runtimes, level-based
-        clustering can be used. When runtimes of same-level tasks may vary, it
-        may be preferable to use job runtime-based horizontal clustering, to
-        achieve load-balancing of clusters. Alternatively, the user may opt for
-        label-based clustering, in which the user can define the task clusters
-        themselves.
-
-        To avoid running out of disk space during data-intensive workflows, the
-        Mapper can add cleanup nodes that remove all unecessary data after a job
-        has completed.
-
-        \note{Dynamic workflows permitted}
-
-        \subsection{Debugging and provenance} Job monitoring, in Pegasus, is
-        achieved through \textit{pegasus-kickstart} (kickstart). Every job
-        within a pegasus workflow has a kickstart wrapper that collects runtime
-        provenance and performance. Kickstart relays this information to the
-        submit host after job completion. This data is ultimately collected and
-        stored in a relational database.
-
-        The Mapper Logs contain details on the abstract workflow, the jobs in
-        the executable workflow and the mapping of tasks to jobs. Using the
-        mapper logs, Pegasus can generate the provenance of the results. Should
-        it be required, Pegasus can also generate the provenance of the mapping
-        process.
-
-        In addition to the Mapper Logs, there also exists the DAGMan logs,
-        generated by HTCondor DAGMan and updated in near real-time. These logs
-        capture details on the status of the jobs and their associated pre- and
-        post- execute scripts.
-
-        All provenance and debugging data is accessibly through either the
-        \textit{pegasus-analyzer} tool or through the provided lightweight web
-        interface.
-
-           \section{Make} \subsection{Summary} As the creation of workflow
-           activities requires a certain level of programming expertise, Make
-           was proposed as a simpler alternative. Make is a tool used for the
-           development of directed acyclic dependency graphs. It is available in
-           all Linux and UNIX distributions, which relieves the users from
-           worflow engine installation. Additionally, Make is capable of
-           offering other features necessary to workflow engines, such as,
-           reproducibility, parallelization, fault-tolerance and quality
-           control. The motivation behind Make is essentially that the tools
-           available for simple and efficient pipeline creation already exist
-           and there is no need to 'reinvent the wheel'. In addition, the use of
-           Make is not limited to the neuroimaging community, and thus support
-           for it is widely available. Make has also been adapted for use in
-           other programming languages such as SnakeMake for python
-
-        \subsection{Pipeline description} Make pipelines are described in
-        makefiles, which consist of a set of rules which describe the flow of
-        data. Rules defined in make can be define by three components: targets,
-        dependecies and recipes. The target is the output (typically file)
-        produced by execution of the recipe, whereas the dependecy would be the
-        input file from which the target would be created from. The target is
-        only modified if its last modified date is early than that of a
-        depencies's.  A recipe, in make, is the set of shell commands used to
-        create the target from its dependencies.
-
-In order connect the rules together, it is possible to define "phony" targets.
-Phony targets do not correspond to any actual file and will be generated once
-all of their dependencies (targets of other rules) have been created. Chaining
-multiple rules together will automatically generate the directed acyclic graph
-that is the pipeline. This is contrary to other pipelines such as Nipype and
-LONI, in which the user must explicitly state all connections \note{as stated in
-paper}. Through the use of rules, it is easily possible to reexecute only the
-desired parts of the pipeline.
-
-Cleanup after rule execution can be defined using the phony target clean.
-
-To create different targets for the same dependency files organized in different
-ways (e.g. subject/timpepoint timepoint/session), it is possible to use symbolic
-links for the dependencies, however this may complexify the code. Otherwise, it
-may be simpler to copy subject and session makefiles.
-
-        \subsection{Pipeline execution}
-
-        Compared to the previous workflow engines, Make pipeline variables are
-        only evaluated at time of use (lazy evaluation). In addition, targets
-        are not executed in the order they are specified (similarly to other
-        pipeline engines).
-
-Make files are also well-equipped to handle workflows with conditionals. Should
-a conditional exist in a workflow, it will be evaluated during the creation of
-the DAG (i.e. not at runtime) and only the required components will be executed.
-
-        As parallelism can be inferred from the dependency graph, parallel
-        execution of a make pipeline can occur simply through the addition of
-        the -j flag for multicore machines, or by using qmake command to run the
-        pipeline on a cluster.
-
-        By default, make only rebuilds targets whose last modified preceedes
-        that of its dependencies. As such, when a task fails and is re-executed,
-        only the targets that need to be rebuilt will be rebuilt. It is possible
-        to specify reexcution in Make such that the task will be reexecuted in
-        case of failure.
-
-        \subsection{Debugging and provenance} As users of a preexisting Make
-        pipeline may not be aware of the proper usage of each target, it is
-        necessary to provide a `help` module that is adaptable to targets being
-        added removed or modified. It is possible to define in Make a macro for
-        such a purpose that will provide users with usage information when
-        requested (e.g. calling make help).
-
-    Debugging in Make can be rather difficult, particularly when executing code
-    in parallel. However, pipelines in Make are broken down into many targets
-    which can each be tested individually. In addition, Make provides built-in
-    flags to aid in debugging. These flags include -n, for printing out the
-    commands without any execution and -p to print out all the rules. In the
-    instance where pattern substistution is used, -p will also include the
-    results obtained from the pattern substitution. More recent versions of GNU
-    Make also include the --trace and --output-sync flags which display why each
-    recipe was executed and print the parallelized output in sequential order,
-    respectively.
-
-    Neuroimaging tools are typically equipped with quality assurance modules. As
-    a result, it is possible to utilize the quality assurance modules from these
-    tools to provide a complete report for the entire pipeline. The proposed
-    solution suggests to write all the quality assurance outputs from each tool
-    to a single HTML report with R Markdown or a script. The R Markdown or
-    script, with the metrics and images are listed as dependencies to the QA
-    report target in Make.
-
-    Only a basic form of data provenance has been implemented by the authors.
-    The implemented data provenance consisted of a makefile that calls a bash to
-    extract acquisition parameters from the raw input images. The makefile also
-    obtains tool versions from the installation directories and pipeline options
-    from their respective makefiles. All of these details are stored in a CSV
-    file.  The makefile will then call an R Markdown script which will then
-    generate an HTML provenance report from the corresponding CSV file.
-
-         \section{OpenMOLE} \subsection{summary} OpenMOLE (Open MOdeL
-         Experiment) is a domain agnostic pipelining framework that focuses on
-         simplifying workflow creation through its Domain Specific Language
-         (DSL) built on top of Scala. Through the use of its DSL, much of the
-         backend is abstracted from users. It also offers a GUI for users to
-         build workflows. Similarly to other existing pipelining frameworks,
-         OpenMOLE enables parallelization through plugins, meaning that the
-         workflow need not be changed between different distribution
-         environments.  Moreover, OpenMOLE supports the deployment of tasks or
-         groups of tasks on different computing environments.  Another key
-         feature of OpenMOLE is that it attempts to address the reproducibility
-         issue through the integration of the CARE archiver. CARE allows for the
-         execution of command-line applications, written in any language, and
-         attempts to simulate the environment in which the application was
-         designed on.  OpenMOLE also provides embedded methods for parameter
-         exploration.
-
-        \subsection{Pipeline description} An OpenMOLE workflow consists of tasks
-        connected by transitions. Each task contains details such as input,
-        output and optional parameters, in addition to embedding the application
-        that will be applied to the data. Input values are generally supplied by
-        the dataflow through an entry point known as \textit{Sources} and output
-        values are returned to the dataflow through \textit{Hooks}.
-
-      Variables in OpenMOLE can be native types (e.g. int, string), files,
-      directories, any Java or Scala types, and user-defined types (i.e. Java or
-      Scala classes). The output type of an input supplied to another task
-      should be the same as the input type, or be in a child class of the input
-      type.
-
-      There exist five types of transitions in OpenMOLE: flat, divergent,
-      convergent, explorative and aggregative. A flat transition is one in which
-      the output of one task is the input to a single other task. On the other
-      hand, a divergent transition is one in which the output of one task serves
-      as the input to several other tasks. The inverse of a divergent task is a
-      convergent task, in which several tasks supply input to a single task.
-      Explorative transitions allow for multiple excutions of the same task on
-      the sampling outputs produced by an exploration task. Conversly, an
-      aggregative transition gathers all the streams produced by an explorative
-      transition. It is also possible to add a stopping condition to terminate
-      all streams once a condition is satisfied.
-
-      Tasks are not directly connected to each other by transitions. They are
-      encapsulated within \textit{capsules} who are connected to each other by
-      transitions. This permits the same task to be used multiple times in the
-      workflow by encapsulating them in different capsules.
-
-    There also exists the notion of \textit{Input Slots} in OpenMOLE, which
-    enables cycles in the workflow. A task may have several input slots and will
-    execute each time that all the transitions belonging to the same input slots
-    has passed through.
-
-        \subsection{Pipeline execution}
-
-      In order for an OpenMOLE workflow to be executed, there must exist a
-      single encapsulated task. A task will execute once all its dependencies
-      have been met.
-
-    During execution of a task, the CARETask will decompress the CARE archive
-    containing the application and its dependencies and the application is
-    deployed within the confines of the decompressed archive.
-
-OpenMOLE relies on the exit code returned by the application in order to
-determine whether to reschedule the task. It is possible to prevent task
-rescheduling by setting the boolean flag \textit{errorOnReturnValue} to false.
-
-      OpenMOLE workflows can be executed on a variety of platforms such as
-      remote servers, clusters and compute grids. As these execution hosts may
-      not have all the required dependencies to execute a task, tasks are
-      archived using CARE. This ensures that all the required applications and
-      their dependencies can be shipped and executed at runtime on any Linux
-      host.
-
-        \subsection{Debugging and Provenance} A CARETasks standard output and
-        standard error are normally printed out to OpenMOLES's console. For
-        distributed executions, this might not be a practical solution, in which
-        case such information may be passed through the dataflow using OpenMOLE
-        variables.
-
-        \section{Nextflow} \subsection{summary} Nextflow is a POSIX-compatible
-        pipeline engine for scientific workflows.  It extends the Groovy
-        programming language to provide an  intuitive domain-specific language
-        (DSL) that enables users to easily describe their workflows. Unlike
-        pipeline engines such as Nipype and PSOM, Netflow does not employ DAG.
-        It instead adheres to the \textit{Dataflow} programming model which
-        decides how data will flow through the pipeline at runtime \note{fix
-        this sentence}. Nextflow prides itself on allowing the integration of
-        containerized tools to address the issue of numerical instability that
-        may arise from different environment setups.
-
-        \subsection{Pipeline description} A Nextflow pipeline is represented by
-        the flow of data between processes. These processes can be written in
-        any Linux-compatible scripting language. As in Nipype, these processes
-        are executed in an isolated environment. The only means of communication
-        between processes is through asynchronous FIFO queues known as
-        \textit{channels}. Each process can contain any number of input and
-        output channels. Through the use of channels, input and output can be
-        stored not only on disk but also in-memory \note{Nipype can also return
-        outputs that are not on disk, so this is nothing special}.
-
-        As with other pipeline engines (nipype, psom, etc.,), the order in which
-        the processes appear is irrelevant. Execution order is determined by the
-        dependencies.
-
-        \subsection{Pipeline execution} The \textit{executor} is responsible for
-        the execution of a Nextflow pipeline. By default, it assumes that the
-        pipeline is to be run locally. The local executor will use
-        multithreading in order to exploit available cores within the local
-        environment. To configure the executor to run in a remote distributed
-        environment, it is necessary to modify the configuration file to specify
-        the environment. By following this approach, Nextflow users can
-        seamlessly run their code on various processing platforms with a simple
-        modification to a configuration file. In addition to local, Nextflow
-        currently has executors for SGE, LSF, SLURM, PBS/Torque, NQSII,
-        HTCondor, DRMAA, Ignite, Kubernetes and AWS Batch.
-
-        \subsection{Debugging and Provenance} When executing a pipeline, it is
-        always possible to specify the generation of a execution report. The
-        report is written in HTML and consists of three sections: Summary,
-        Resources and Tasks. Summary provides details on the execution command,
-        execution status, overall execution time, in addtion to other metadata.
-        The Resources tab, on the other hand, contains several HighChart plots
-        representing CPU, memory, time and disk read/write rate of each pipeline
-        process. The final section, Tasks, contains many metrics including the
-        status and command script of each task.
-
-    Other reports that can be generated at runtime include the Trace report and
-    the Timeline report. The Trace report include details such as CPU usage,
-    start time, submission time, etc, of each process that was executed,
-    whereas, the Timeline report generates an HTML timeline of all processes
-    that have been executed.
-
-    \note{doesn't seem to be anything on provenance...}
-
-    \section{Galaxy}
-        Galaxy is another open-source genomics web-platformed designed to 
-        address the
-        reproducibility crisis in genomics in addition to facilitating workflow
-        creation process for non-programmers and ensuring the transparency of 
-        the workflows. It differs from workflow engines in that it is a full
-        platform designed for researchers to create and share their workflows
-        and data with the community. Building workflows in Galaxy is
-        entirely GUI-based enabling users with little to no programming
-        knowledge to build workflows. Metadata is extracted from the workflows
-        during executions and pipelines can be annoted and shared with other 
-        users.
-
-        \subsection{Pipeline description}
-        In order to make workflow creation accessible to researchers with little
-        -to-no programming skills, Galaxy provides a GUI. 
-        The GUI allows users to drag-and-drop CLI tools to the screen to create
-        workflows. In order for new tools to be added to Galaxy, the user must 
-        write a configuration file which contains details on input and output
-        specifications and how to execute the tool. During workflow creation, 
-        the workflow editor will determine if the inputs and outputs of 
-        connected tools are compatible.
-
-        Alternatively, users may also create workflows by editing workflows that
-        have been published to Galaxy or using their own analysis history to 
-        generate a workflow. Any workflow created by a user can be published to
-        the Galaxy server and made available to all other users.
-
-        \subsection{Pipeline execution}
-
-        \note{no details on workflow failure}
-        As Galaxy is meant to be easy-to-use for researchers, all that is 
-        required to execute pipelines is access to their webpage. Users have 
-        access to detailed records of the executed workflow's metadata and 
-        provenance indicating the outcome of the executed workflow.
-
-        Should users desired to host their own Galaxy server, it is possible to
-        install Galaxy locally, on HPC with SLURM or PBS installed or the cloud.
-
-        \subsection{Debugging and provenance} 
-        Provenance is a big part of Galaxy as it aims to improve accessibility, 
-        reproducibility and transparency of workflows. During workflow execution
-        , Galaxy automatically produces metadata for each of the workflow tasks.
-        This metadata includes details on the input and output datasets, tools 
-        user and parameter values.
-
-        In addition to Galaxy's metadata capture, it is also possible for the 
-        user to annotate and tag their workflow. This enables the user to 
-        communicate with others the intent of the analysis steps in addition to 
-        incorporating search capabilities to their workflows.
-
-        Galaxy also provides the Pages capability. Pages is a web-based document
-        that enables users to automatically generate detailed documentation of 
-        their analyses.
-
-    \section{Toil}
-        \note{obtained from paper's supplementary documentation}
-        Toil is another general workflow framework aimed at the processing of 
-        scientific big data, with its main focus being on genomics workflows. It 
-        prides itself in not only its ability to not only scale the processing 
-        of very large datasets, but also its portability, allowing it to be 
-        executed on any environment. Toil pipelines can be written in CWL, WDL 
-        (Workflow Description Language) and Python. The Toil workflow 
-        framework is also open-sourced, making it a community-driven effort. 
-        \note{as is the case with most}
-
-        \subsection{Pipeline description}
-           A Toil workflow consists of a DAG of Jobs connected to each other
-           through input and output dependencies. All custom user scripts must
-           inherit from the Job class. In order to absolve the user from having
-           to manually create these classes for each function, Toil provides
-           wrapper classes that convert Python functions to Toil jobs directly.
-           A Toil job need not be part of the static workflow, it can also be
-           added at runtime.
-
-           There exists three types of relationships between Toil jobs: parent, 
-           child and follow-on jobs. Child jobs begin execution immediately
-           after their Parent completes their execution. Follow-on jobs, however
-           , begin execution after the Parent, associated child job and their
-           successors complete execution. They are particularly useful as 
-           cleanup jobs. Both child jobs and follow-on jobs are executed in
-           parallel.
-
-           Variables may be passed at runtime to dynamically added child or
-           follow-on jobs, or they can be referenced statically using promises.
-           Promises enable the reference to the result to be passed prior to 
-           runtime, evaluating the result only at runtime.
-
-           Workflows can be executed or 
-           resumed through the Runner class, or alternatively, invoked by the
-           Toil class. The latter differing from the Runner class as it serves 
-           as the workflow's context manager. The Toil class performs
-           a preliminary configuration of the workflow (e.g. file staging).
-           All Toil workflows begin with a single job. The value of this root 
-           job is returned at the end of the workflow.
-
-            The FileStore class manages the permanent and temporary files
-            created by the workflow. This class ensures that all necessary files
-            will be cleaned up after job execution, but also that workflows
-            can resume from the last successful state. Both temporary 
-            directories and temporary files can be created at a user's desired
-            location (e.g. local file system, Amazon S3, etc.,). Should files
-            and directories need to persist during workflow execution and
-            be accessible to all jobs, the FileStore
-            can add these files to a global store. All files stored in the 
-            global store are immutable, preventing any updates to these files
-            from occuring other than deletion. There also does not exist any
-            kind of filesystem hierarchy in the global store.\note{allows use
-            of various different object stores/ enables caching of files as they
-            never change}
-
-            In order to add files to a workflow, it is necessary to stage them
-            prior to execution. This is achieved using the \textit{Toil} context
-            manager as it is running on the leader prior to execution. It is 
-            possible to stage the file as a shared file, if necessary. In the 
-            instance of a job failure, the staged file behaves like any other 
-            file in the job store. If the file is not cleaned up post execution,
-            the file will not need to be staged again during resumption.
-
-            Toil also provides the functionality to startup services, such as
-            databases or servers, during pipeline execution. Multiple services
-            can be define per job and services can have associated child
-            services. This functionality allows, for instance, Apache Spark 
-            clusters to be spawned during workflow execution. As services 
-            complicate workflow resumption post failure, Toil enables the
-            checkpointing of services such that the workflow will resume from a
-            checkpoint if any of the successors to the service fail to restart.
-
-        \subsection{Pipeline execution} 
-            \note{obtained from website documentation}
-            Toil is composed of three components: 1) the Job Store, 2) the Batch
-            System and 3) the Provisioner. The Job Store centralizes all the 
-            files necessary for the workflow. During a workflow re-execution,
-            the job store is used to resume the workflow from its last
-            successful task. Currently, there are two types of job stores 
-            supported: the file job store and the cloud job store.
-
-            The Batch System specifies which type of scheduler will be used.
-            A Toil workflow can either be executed locally, on HPC clusters or 
-            the cloud.
-
-            The Provisioner, on the other hand, is used to execute Toil
-            workflows on specific cloud platforms. Using the Cluster Utilities
-            command line tools, a user can provision nodes for their desired 
-            cloud platform.
-
-            Toil executes workflows using a leader-worker architecture. The
-            leader is a single-threaded read-only process responsible for 
-            determining which jobs are ready to be executed. Workers, on the 
-            other hand, are created by the batch-system and are responsible
-            for the execution of a job. To alleviate the single-threaded leader
-            from having to do an excessive amount of work, the worker maintains
-            track of its status (i.e. success or failure) and reports its status
-            back to the leader by writing it to the job-store. The worker may
-            also choose to immediately execute any child-jobs defined by the 
-            job it has executed.
-
-            In the instance when a job or a subset of jobs fail, the workflow
-            will continue to execute until it can proceed no more and will throw
-            an error. The user is given the option to resolve the bug and resume
-            the workflow or delete the job store and reexecute the pipeline 
-            entirely.
-        \subsection{Debugging and provenance}
-
-        By default, workflow standard input and output is hidden to the user.
-        It is however possible to make the log data visible to the user during
-        execution or to redirect this data to a logfile.
-
-        Toil also provide a fuctionality, stats, to capture statistics on the 
-        workflow in the job store. Examples of the statistics it captures
-        includes cpu, memory and job duration.
-
-    \section{Thunder}
-
-        Thunder is a Python-based modular set of tools for the analysis of
-        imaging and timeseries data belonging to various domains including
-        neuroscience. Although it is not specifically a pipelining framework, it
-        can run on top of the Apache Spark framework enabling the Thunder
-        pipeline to be extended and customized. By running on top of Spark,
-        Thunder leverages its features to improve performance on large scale
-        studies. Thunder does not need to be executed on top of a Spark cluster
-        and can easily be executed, with the same script, an a local system with
-        no Spark installation.
-
-        \note{The following was extracted from the documenation}
-        \subsection{Pipeline description} The API for Thunder, regardless of if
-        backed by an Apache Spark cluster or not, remains the same. If not using
-        an Spark Resilient Distributed Dataset (RDD), the data can be backed by
-        a \textit{numpy} array or a distributed \textit{bolt} array. Thunder is
-        broken down into several packages, in which the core package consists of
-        tools for reading and writing data, in addition to defining common data
-        structures. Users may choose to install other Thunder packages for added
-        functionality, or define their own using any Python library.
-
-        Thunder contains two data structures: \textit{images} and
-        \textit{series}. As implied by its name, the \textit{images} data type
-        may represent a collection of images or volumes.  The \textit{series}
-        data type, however, represents a one-dimensional collection that shares
-        a common index. Both \textit{images} and \textit{series} can be loaded
-        from a list, ndarray or binary file type, however \textit{series} can
-        also be loaded from a text file and \textit{images} can be loaded from
-        png and tiff files.
-
-        Methods that are common to both data types are, such as \textit{map} and
-        \textit{reduce}, are found in the \textit{data} class which is a
-        subclass of the \textit{base} class. In order to convert between the
-        \textit{images} and \textit{series} data types, the intermediary data
-        type \textit{blocks}, child of \textit{base} class,  is used.  Using the
-        \textit{blocks} class, it is also possible to convert the data type to
-        an ndarray for use with other Python modules.
-
-        \note{not really anything on execution, debugging and provenance in the
-        paper nor in the documentation} \subsection{Pipeline execution} In order
-        to use Spark within Thunder, a SparkContext must be started and passed
-        to the Thunder API. Should the user need to switch between local and
-        Spark frequently, it is possible to use the \textit{station} package as
-        a context manager for the various contexts.
-
-        As Thunder is an extension to Spark, numpy and \textit{bolt}, much of
-        the execution details are likely differed to the framework that it is
-        being executed on.
-
-        \subsection{Debugging and provenance} Thunder does not explicitly
-        provide any details on debugging and provenance tracking. Similarly to
-        execution, it is expected that much of the debugging and provenance
-        details are not obtained from Thunder itself, but rather, from the
-        frameworks it is being executed on. For instance, should Thunder be
-        exected on a Spark cluster, it is likely that much of the debugging and
-        provenance information can be obtained from the Spark logs and RDD
-        lineage.
-
-\chapter{MapReduce-based BigData frameworks and related components}
-\section{mapreduce} Summary of ~\cite{mapred}
-
-        Typically, simple computations need to be performed on increasingly
-        growing large datasets. Processing of such large datasets require
-        parallelization. As a result, these computations are complicated by
-        details of parallelization, fault tolerance, data distribution and load
-        balancing. The MapReduce library addresses this issue by allowing the
-        expression of simple computations while abstracting the other details.
-        The map and reduce paradigm was selected for this purpose as it was
-        already commonly used in functional programming languages, such as Lisp,
-        and it was observed that many computations could be expressed as such.
-
-        The MapReduce library operates by first dividing the data into
-        manageable chunks (16 to 64MB). It then creates multiple copies of the
-        program and distributes them across nodes. There exists a single master
-        node; the remainder are worker nodes. The master node is responsible for
-        delegating map and reduce tasks to the idle workers. There are two data
-        structures found in the master. The first stores the state
-        (\textit{idle}, \textit{in-progress} or \textit{completed}) of each
-        \textit{map} and \textit{reduce} task, and the second stores the
-        identity of the non-idle worker nodes. In contrast, the worker nodes are
-        responsible for the execution of \textit{map} or \textit{reduce} tasks.
-        A worker node assigned a \textit{map} task will parse out the key-value
-        pairs from the input file and pass it to the \textit{map} function. The
-        intermediate keys produced by the \textit{map} function are buffered in
-        memory and periodically written to disk.The locations of these
-        intermediate files are sent to the master who is responsible for
-        forwarding these locations to the reducer. A worker node assigned a
-        \textit{reduce} task uses remote procedure calls to obtain the
-        intermediate local files created by the map tasks and sorts the keys
-        such that identical keys are grouped together. Each unique key and its
-        list of values are then submitted to the \textit{reduce} task and the
-        output of the \textit{reduce} task is stored in a final output file.
-        Once all \textit{reduce} tasks have completed, the master resumes the
-        program and the output is returned to the user.
-
-        In order to ensure fault tolerance, the master regularly pings the
-        worker nodes. Should a worker node not respond after a certain delay,
-        the master labels the worker as failed. The \textit{map} tasks that were
-        completed by the worker are all returned to their original state
-        (\textit{idle}) as \textit{map} task data is stored locally. Both the
-        completed and in-progress worker's \textit{map} tasks are subsequently
-        reassigned to other workers. Only the failed worker's
-        \textit{in-progress} \textit{reduce} tasks need to be reassigned.
-
-        The master node undergoes periodic checkpoints, such that the master
-        node can be reinitialized from its last checkpoint in case of failure.
-        Master node failure is, however, unlikely.
-
-        As network bandwidth is a scarce resource, the master attempts to
-        schedule \textit{map} tasks closest to where the data is located. it
-        will first attempt to schedule a map task on the host that contains the
-        data; should that option not be available, the master will attempt to
-        schedule the \textit{map} task on a host nearest to the data (e.g. same
-        network switch)
-
-        Ideally, the number of \textit{map} and \textit{reduce} tasks should be
-        much larger than the number of available worker nodes.  Having it as
-        such improves both dynamic load balancing and recovery time. The number
-        of \textit{map} tasks used in practice typically corresponds to the size
-        of the input chunks, whereas the number of \textit{reduce} tasks are a
-        small multiple of the number of worker nodes used.
-
-        When nearing completion of the MapReduce program, it is possible for
-        ``stragglers" (\textit{in-progress} tasks running on nodes with below
-        average performance) to delay completion. As a measure to counteract
-        this, the master node assigns the same task to idle workers. The task is
-        then marked as completed as soon as either the primary or the backups
-        complete the task. This strategy has been found to significantly improve
-        performance when used with very large datasets.  \note{44\% speedup with
-        their sort program}.
-
-        A few extensions have also been added to the library in order to improve
-        user experience. An important extension is the introduction of a
-        \textit{combiner}, which enables the commutative and associative reduce
-        function to be applied to the same keys on the hosts where the
-        \textit{map} task was executed. Like the \textit{map} task, the
-        \textit{combiner} produces intermediate output which is sent to the
-        \textit{reducer}. This alleviates the amount of key-value pairs that
-        must be transferred over the network from the map host to the reduce
-        host, and as a result, improves performance. Other extensions include
-        the ability to create user-defined partitioning functions, a guaranteed
-        increasing key sort order, the ability for the user to specify input
-        types, an option to skip records that continuously result in execution
-        failure, the ability to produce auxiliary output files in \textit{map}
-        or \textit{reduce} task, live status updates stream to HTTP server, the
-        ability to run all code on a single local machine, and a
-        \textit{Counter} class.
-
-    \section{spark} 
-        \begin{itemize} 
-            \item based on map-reduce 
-            \item in-memory processing 
-            \item lazy evaluation 
-            \item general framework 
-        \end{itemize}
-
-    \subsection{Scientific workflows in Spark} 
-        \subsubsection{Introduction}
-            emphasis on data rather than on interface. transformation describe 
-            how interfaces interact with the data.  
-        \subsubsection{ariel's paper and thunders astronomy - } 
-        \subsubsection{Spark on hpc's or } \note{large dataset for valduriez} 
-            Prior to examining how Big Data frameworks could be improved on HPC 
-            clusters, it is important to have a baseline of how it performs 
-            natively on an HPC cluster.  Particularly, how it behave when
-            processing scientific workflows. In a recent study done by 
-            \cite{valduriez}, they examined the scalability of Spark on an HPC 
-            cluster using a black-box scientific workflow. Five tests were 
-            performed in their study: 1) Measuring execution time when task and 
-            resources available increase proportionally, 2) Measuring the 
-            execution time of very short and short tasks, 3) Measuring the
-            execution time of short (5s) to long tasks (120s), 4) Varying the 
-            number of tasks with fixed task durations, 5) Mixed task durations 
-            and varied number of tasks. Their results show Spark behaves as 
-            expected when tasks and resources are increased exponentially, and 
-            that Spark scales better than expected with tasks of longer 
-            durations. This is due to the fact that the scheduler is overloaded 
-            in the case of many short tasks in addition to their being a 
-            significant amount of I/O occurring between each task. Although I/O
-            is problematic in both data center and HPC infrastructures, HPC is 
-            optimized for data transfer over the network rather than disk I/O, 
-            and therefore, such frequent I/O negatively impacts the system. 
-            However, many scientific workflows are composed of longer tasks, 
-            and therefore, Spark scales very well with scientific workflows in 
-            an HPC environment.  
-    \section{Use of containers with Big Data frameworks} 
-    \section{Hadoop Distributed file system} 
-        \note{maybe should go in between mapreduce and spark}
-
-        The Hadoop Distributed File System (HDFS)\cite{hadoop} is the filesystem
-        component of Hadoop. The metadata in HDFS is stored in on a dedicated
-        server known as the NameNode, whereas application data is distributed
-        across many servers referred to as DataNode. To ensure fault tolerance,
-        HDFS replicates the data across multiple DataNodes (default replication
-        factor: 3). Not only does this ensure that the data is not lost in the
-        event of node failure, but it also increases data transfer bandwidth, as
-        the data can be accessed from multiple nodes, and thus, there are more
-        opportunities for computations to be performed nearest to where the data
-        is located.
-
-        The NameNode contains the namespace tree, a hierarchy of directories and
-        their files, as well as the mapping of the split file blocks to
-        DataNodes. The entire namespace is stored in memory. Information on the
-        directories and files, such as modification and access times, namespace
-        and disk quotes, are stored within \textit{inodes} on the NameNode.  The
-        name system's metadata, \textit{inode} and file block mapping, is known
-        as the \textit{image}. Persistent record of the image that is stored on
-        the local file system are known as \textit{checkpoint}. Locations of
-        block replicas are not stored in the checkpoint as they may change over
-        time. The \textit{journal} is a log of the modifications made to the
-        image. Both the checkpoint and the journal may be copied across servers
-        for increased durability. The journal is played back during the NameNode
-        restart in order to restore the cluster.
-
-        When a client requests to read data, it must first contact the NameNode.
-        The NameNode provides it with the replica locations, and the client
-        reads from the DataNode located closest to it. When the client requests
-        to write a file, it contacts the NameNode which selects the DataNodes
-        that will host the replicas. The client subsequently writes the data
-        directly to the DataNodes in a pipeline fashion (The data gets
-        propagated through each DataNode by being transferred from the nearest
-        DataNode to it).
-
-        There is only one NameNode assigned to each cluster, however, each
-        cluster can have multiple clients and execute multiple tasks
-        concurrently.
-
-        A block replica stored on a DataNode consists of two files stored on the
-        local file system: the metadata and the data.
-
-        The DataNode connects to the NameNode during startup and performs
-        handshake to for DataNode namespace ID and software version
-        verification. Should either namespace ID or software version not match
-        those of the NameNode, the DataNode shuts down.
-
-        To ensure integrity of the system, a \textit{namespace ID} is assigned
-        to all nodes during formatting of the namespace.  DataNodes with
-        different namespace IDs to the NameNode will not be permitted to join
-        the cluster. Should the DataNode have not been assigned a namespace ID,
-        it will be permitted to join the cluster. The cluster's namespace ID
-        will then be assigned to that DataNode.
-
-        Following the handshake, the DataNode register with the NameNode using
-        their storage ID. The storage ID is a unique ID generated after initial
-        registration to the NameNode and that is persistently stored on the
-        DataNode. It permits identification of the DataNode in the event of an
-        IP address or port change.
-
-        \section{Dask}
-            Dask contains three primary abstractions: the \textit{dask.array}, 
-            \textit{dask.bag} and \textit{dask.dataframe}. The 
-            \textit{dask.array} is an abstraction that enables parallel 
-            NumPy-like operations using task scheduling. It can additionally be 
-            used on data that does not fit in memory. The \textit{dask.array} 
-            used blocked array algorithms to break down large commonly-used 
-            linear algebra functions into smaller ones that can be partially 
-            parallelized.
-
-            Dask array functions all create an \textit{Array} object which 
-            contains the dask graph \textit{.dask}, details on the overall shape
-            and the shape of each subdivided portion (chunk) of the array 
-            (\textit{.chunks}), the \textit{.name} representing which keys map 
-            to the results, and the data type.
-
-            A limitation of the Dask array is that it must know the shape of the 
-            array at graph creation time.
-
-            The Dask bag is similar to that of a Python list, however, order 
-            cannot be guaranteed. As Dask bags typically operate on Python 
-            objects, performance is bound by the Global Interpreter Lock (GIL), 
-            therefore multiprocessing schedulers must be used instead of 
-            multithreading ones.
-
-            The Dask bag contains functions like \textit{map} and 
-            \textit{filter}, in addition to many other functions that can be 
-            found in the PyToolz API. It is particularly well suited for the 
-            handling of data dumps.
-
-            The Dask dataframe consists of a collection of Pandas DataFrames, 
-            and thus implements many of the functionalities found in an Pandas 
-            DataFrame. The dataframe is well suited for running on partioned 
-            datasets, where partitions are clearly defined by an index.
-
-            In order to execute the Dask graphs in an efficient manner, proper 
-            scheduling is required. Dask uses dynamic scheduling for all their 
-            schedulers. When a worker reports completion of a task, the runtime 
-            state is updated to reflect that the task has been completed, tasks 
-            that can now be executed are marked and the data which can be 
-            released is determined. To ensure that workers complete execution of
-            related tasks before commencing unrelated ones, a 
-            \textit{last in first out} policy is used. As Dask graphs are 
-            entirely decoupled from the scheduler that they use, Dask users are 
-            free to design their own scheduler that best suits their needs.
-
-\chapter{Optimizing Big Data frameworks for High-Performance Computing}
-
-    \note{precise the systems that will be used. both have distributed memory.
-    data centre is hpc cluster with slower network, because may span multiple
-    racks and domains} \note{variety of jobs on hpc whereas dedicated map-reduce
-    jobs in a data center} \note{talk about spark on cloud and the challenges
-    associated with} \note{talk about why this is important} \note{talk }
-    High-Performance Computing (HPC) clusters are extensively used by the
-    scientific community to process complex, computation intensive problems. Due
-    to their importance in the advancement of scientific research, it is
-    necessary that Big Data frameworks used to process scientific data are
-    compatible with them.
-
-    Big Data frameworks were optimized for a commodity cluster.  Such clusters
-    are made up of nodes with homogenous storage connected by lower-end network
-    cables (e.g. Ethernet). As a result, moving data from local storage to
-    memory was much less costly in these infrastructures than moving data across
-    the network. Thus, Big Data frameworks implemented data locality in an
-    attempt to reduce costly network (horizontal) transfers and favour more
-    performant local storage to memory (vertical) transfers. HPC clusters,
-    however, are not necessarily commodity clusters. Large HPC clusters are
-    commonly made up of supercomputing nodes connected by a high bandwidth,
-    low-latency network (e.g.  InfiniBand). Nodes may contain their own local
-    storage (e.g. HDD, SSD, tmpfs) and are all connected to a distributed file
-    system (e.g GPFS, Lustre).  Due to their employment of a high bandwidth low
-    latency network, network transfers on such an HPC cluster may not be as
-    costly as that of a typical data center, and it may in fact be more costly
-    to do vertical transfers. Furthermore, HPC systems may enforce a batch
-    submission system. This required users to specify number of nodes, cores and
-    total processing time; requirements which are unknown to MapReduce users.
-
-    This chapter will examine the various efforts by the community to improve
-    Big Data framework performance on HPC clusters.
-
-    \section{File system improvements}
-
-    In paper \note{Scaling Spark on HPC Systems} it was found that Spark on a
-    Lustre backend performed significantly (4x) slower than that of a
-    workstation using local fast SSDs for storage. This is due to the fact that
-    file system metadata latency (more specifically, file open) is the
-    determinant factor for performance on HPC, whereas network dominates
-    performance in a typical data centre configuration. It was found by the
-    authors that using either a local in-memory filesystem or a filesystem
-    mounted to a single Lustre file improved performance up to a point where it
-    matched that of a workstation. In addition, use of a cache was found to
-    reduce file system metadata operations, and thus improve overall
-    performance.  The issues related to using an in-memory filesystem is that it
-    is limited by the amount of physical memory available, and a job will fail
-    if physical memory runs out. Moreover, there is not persistence or
-    resilience of data as it is not saved to non-volatile storage (disk). Spark
-    fails with medium to large applications due to "lax garbage collection in
-    block and shuffle managers". To resolve these issues, the authors attempted
-    to make all executors share a descriptor pool. This allowed files that were
-    already opened by other executors to be accessed by the executor through the
-    descriptor rather than having to reopen it again. However, the node OS image
-    is subject to number of Inode constraints as well Lustre limits on number of
-    open files for a given job. Due to Inode constraints, a livelock may occur
-    when descriptor pool is at capacity and an executor is attempting to open a
-    file. The authors also created mounted a local filesystem backed by a Lustre
-    file (lustremount) to address the issues related to an in-memory file system
-    \note{requires admin privilege}. It was found that the use of an in-memory
-    file system improved performance by 7.7x and the lustremount improved
-    performance by 6.6x. It was also found that implementation of the filepool
-    improved performance.  Another issue observed by the authors is that of poor
-    block management as a result of memory constraints resulting in a
-    significant amount of vertical movement. When available memory is exhausted,
-    the least recently used block is evicted by the block-manager. When a call
-    is made to this removed block, it will need to be recomputed, and in the
-    process blocks required for recomputation may also be evicted and
-    necessitate recomputation of those blocks, thus leading to a chain of
-    eviction and recomputation. To resolve this, marking intermediate RDDs as
-    persistent, such that their results are saved to storage instead of
-    requiring recomputation, may be applied.  \note{they didn't look at
-    improving initial application I/O, though believe that it could be improved
-    with lustremount and filepool. } \note{for each partition, a shuffle file is
-    created and written to as many times as there are partitions. An index file
-    is also created which contains shuffle file locations and offsets. Metadata
-    access is therefore O(partitions2)}
-
-    \subsection{tachyon paper} I/O is a known bottleneck in Spark that is
-    exacerbated in HPC systems in which data is stored in a separate cluster
-    outside compute nodes. One possible solution used to mitigate the effects of
-    the I/O bottleneck is to store data in an in-memory filesystem rather than
-    on disk.  Alluxio and Triple-H are two known examples of such filesystems.
-
-    Alluxio, formerly known as Tachyon, is an in-memory file system designed to
-    improve read and write performance without impacting fault tolerance of the
-    system. As write performance is impeded by the need to replicate data,
-    Alluxio leverages the concept of lineage to recompute the output if data is
-    lost.  Tachyon exhibits 110x faster write throughput than in-memory HDFS and
-    4.4x faster end-to-end latency.
-
-    Alluxio has is made up of two layers: the lineage layer and the persistence
-    layer. The lineage layer is the layer that performs recomputations to obtain
-    the desired output. The persistence layer contains all the code used to
-    regenerate the data, in addition to any data that cannot be recovered
-    through lineage as it may be too large to fit in memory or has no lineage
-    information stored.
-
-    \subsection{triple-H}
-
-    %%%In other words, such frameworks would try to limit data movement, as
-    %network cables are expected to have low bandwidth and high latency. In
-    %addition, each node in a commodity cluster is expected to have access to
-    %its own local storage homogenous to that of the rest of the cluster, thus
-    %replication is necessary to ensure data availability. This is, however, not
-    %necessarily the case for computing grids. Computing grids may have access
-    %to Infinibands, which provide high bandwidth and low latency. Moreover,
-    %grids may use heterogenous storage in addition to a parallel file system
-    %such as Lustre. As a result, it may, in some cases, be faster to transfer
-    %data over the network than to access the nearest copy of the data. As well,
-    %data replication may burden a system with centralized storage as multiple
-    %copies of the data would be created, taking a significant amount of
-    %space.%%%
-
-    %%%Big Data frameworks were not designed with high-performance computing in
-    %mind. Big Data filesystems such as were intended to run in data centers
-    %with homogenous storage (ex. HDDs, SDDs. etc,.). In such infrastructures,
-    %data movement is expensive. To counteract this, strategies were employed to
-    %avoid moving data around (data locality). In contrast, HPC clusters use
-    %heterogenous storage (ex. a combination of HDD, SSD, Ram Disk and Parallel
-    %file systems) with compute nodes connected by high-performance
-    %interconnects. HPC nodes may therefore not all be equipped with the same
-    %type of storage devices, some of which may be faster than others. Unlike
-    %data centers, HPC compute nodes do not have local storage. Moreover, the
-    %high-performance interconnects ensure low latency and high throughput of
-    %the system. In some instances, it may be even more performant to move data
-    %around rather than ensuring data locality. When a single-node HPC cluster
-    %using the Luster parallel file system was compared to a workstation with
-    %local SSD, it was found that Spark performed about 4x slower on the HPC
-    %cluster \cite{Chaimov:2016}.  As HDFS is not conscientious of the
-    %underlying infrastructure and expects a data center-type infrastructure, it
-    %cannot achieve its full performance potential. %%%
-
-    %%%\section{Spark in the cloud}
-
-    %%%Using Spark against a typical HPC parallel file system such a Luster has
-    %been found to impede scalability. This was found to be due the file system
-    %metadata latency, which limits Spark scalability to
-    %\textit{O}(10\textsuperscript{2}) cores \cite{Chaimov:2016}.
-
-    %%%Compute Grids, interconnected networks of computing resources, are
-    %heavily used in scientific research for the processing of data. Due to
-    %their use and importance to the scientific community, it is necessary that
-    %big data frameworks used for processing neuroimaging data are compatible
-    %with them. Compute grids typically utilize heterogenous storage, such as a
-    %mixture of SSD, HDD, RAM Disk and parallel file systems (e.g. Luster).
-    %However, big data frameworks, such as Spark are designed to work
-    %efficiently with homogenous storage. Such differences result in reduced
-    %performance and inefficient storage use when big data frameworks are used
-    %on HPC clusters. This chapter will review the difference techniques applied
-    %to big data frameworks as well as HPC and grids to improve collocation of
-    %both services.%%%
-
-    %%%\subsection{Anatomy of Grid storage} %%Typical grids consist of a cluster
-    %of data nodes possibly connected to local storage (e.g. RAM Disk, HDD,
-    %SDD), but is also connected through a network to another cluster
-    %representing a parallel file system such as Lustre. RAM Disk is the fastest
-    %storage as all data is located in memory, whereas SSD provides slightly
-    %slower, persistent storage, but is a good alternative for big data. HDDs
-    %are the least performant form of local storage.%%% %%Lustre is a stateful,
-    %object-based parallel file system. It consists of three components: 1) the
-    %meta data server (MDS), 2) the Object Storage Server (OSS) and 3) the
-    %Luster Network (LNET), which enables clients to communicate with each
-    %other.%%%
-
-    In an attempt to improve performance of HDFS on HPC clusters, Triple-H was
-    created.  \section{Why is use of such systems important}
-    \section{Scheduling} \subsection{HPC schedulers: SLURM/QSUB} \subsection{Big
-    Data schedulers: YARN/MESOS} \subsection{Multi-level scheduling w/ BigData
-    schedulers} \chapter{Conclusion}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Body of Thesis goes here.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\addcontentsline{toc}{chapter}{Bibliography} \bibliography{bibliography}
+    \chapter{Workflow composition}
+    \chapter{Performance}
+    \chapter{Reproducibility}
+    \chapter{Portability}
+    \addcontentsline{toc}
+        {chapter}{Bibliography} 
+        \bibliography{bibliography}
         \bibliographystyle{ieeetr}
-%\bibliography{abbr,chalin,common,larch,tn}  %place your .bib files here
-%\bibliographystyle{alpha}                   %the bibliography style to use
-
 \end{document}

### 2018-09-25 paper.tex
minor text fix
some reformatting

@@
         lab workstation. With the advancement of data sharing technologies, this 
-    data has been made widely available, with the only limiting factor impeding
+        data has been made widely available, with the only factor impeding
         research on this data being access to infrastructure and software that 
-    enables efficient processing on such infrastructure. Many research labs do
-    have access to high performance computing (HPC) clusters, however, without
-    efficient software to process this data, processing can take an excessive 
-    amount of time. Ensuring efficient processing of data is complex and 
-    likely beyond the knowledge of an average neuroimaging researcher. As a 
-    result, frameworks for the efficient processing of neuroimaging data need
-    to be developed.
+        enables efficient processing on such infrastructure. Many research labs 
+        do have access to high performance computing (HPC) clusters, however, 
+        without efficient software to process this data, processing can take an 
+        excessive amount of time. Ensuring efficient processing of data is 
+        complex and likely beyond the knowledge of an average neuroimaging 
+        researcher. As a result, frameworks for the efficient processing of 
+        neuroimaging data need to be developed.
 
-    Neuroimaging big data comes in two formats: 1) large images and 2) large 
-    datasets. Large images consist of a singular images ranging from 100s of
-    gigabytes to terabytes in size. A well-known example of such an image is the
-    BigBrain \note{cite here}, a histological image of a 69 year-old man. 
-    At its highest resolution, it is 1x1x20micrometers \note{fix this} or 1TB in
-    size. Other examples of large images can be found in electron microscopy 
-    (EM), polarized light images (PLI) and micro coherence tomography (microCT)
-    \note{need to do more research here}. Images at such high resolution are
-    important to researchers as they give insights into aspects not otherwise
-    detectible in images at lower resolutions. However, due to their size and 
-    lack of resources available to process such images, research on this data 
-    remains limited.
+        Many neuroimaging workflows currently exist. Examples include
+        Nipype, Pipeline System for Octave and Matlab (PSOM), LONI, SPM, FastR,
+        Automated Analysis (AA) and Pydpiper\note{add citations here}. These 
+        workflow engines aim to satify four criteria: 1) straightforward 
+        workflow composition, 2) performance, 3) portability of the workflows 
+        and 4) reproducibility of the analysis. While these workflow engines
+        do tackle performance, it is mainly limited to minimizing computation
+        time and does not consider data transfer times -- which are typically
+        costly in Big Data settings. Therefore, these engines need to be adapted
+        for the processing of neuroimaging data.
 
-    In contrast, large datasets are consists of many images, typically arising
-    from multiple different subjects, that are too big to fit in storage. 
-    The images alone being small enough for an average researcher to process. 
-    Example of these datasets include the UK Biobank, HCP Project, ADNI dataset,
-    OpenfMRI, amongst many others. These datasets are very large. For instance, 
-    the UK Biobank is expected to exceed \note{X} Petabytes in size. As subsets
-    of these datasets are manageable for the average researcher to process, 
-    much analysis has gone into them. Moreover, unlike for large images, 
-    toolkits have been developed for researchers to process this data. However,
-    the toolkits do not enable researchers to process these entire datasets as
-    a whole. Therefore, it is crucial to develop efficient frameworks for the 
-    processing of these datasets in order for researchers to harness the data
-    in its entirety.
+        In sec
 
-        \section{Platforms for  neuroscience}
-        \section{Infrastructure}
+        \section{Big neuroimaging Datasets}\label{datasets}
+            Neuroimaging big data comes in two formats: 1) large images and 2) 
+            large datasets. Large images consist of a singular images ranging 
+            from 100s of gigabytes to terabytes in size. A well-known example of 
+            such an image is the BigBrain \note{cite here}, a histological image 
+            of the brain of a healthy 69 year-old man. At its highest resolution
+            , it is 1x1x20$\mu$m or 1TB in
+            size. Other examples of large images can be found in electron 
+            microscopy (EM), polarized light images (PLI) and micro coherence 
+            tomography (microCT) \note{need to do more research here}. Images at 
+            such high resolution are important to researchers as they provide 
+            insights into aspects not otherwise detectible in images at lower 
+            resolutions. However, due to their size and lack of resources 
+            available to process such images, research on this data remains 
+            limited.
 
-    \chapter{Workflow composition}
-    \chapter{Performance}
-    \chapter{Reproducibility}
-    \chapter{Portability}
+            In contrast, large datasets are consists of many images, typically 
+            arising from multiple different subjects, that are too big to fit in 
+            storage. The images alone being small enough for an average 
+            researcher to process. Example of these datasets include the 
+            UK Biobank, HCP Project, ADNI dataset,OpenfMRI, amongst many others. 
+            These datasets are very large. For instance, the UK Biobank is 
+            expected to exceed \note{X} Petabytes in size. As subsets
+            of these datasets are manageable for the average researcher to 
+            process, much analysis has gone into them. Moreover, unlike for 
+            large images, toolkits have been developed for researchers to 
+            process this data. However,the toolkits do not enable researchers to 
+            process these entire datasets as a whole. Therefore, it is crucial 
+            to develop efficient frameworks for the processing of these datasets 
+            in order for researchers to harness the data in its entirety.
+
+        \section{Platforms for  neuroscience}\label{platforms}
+        \section{Infrastructure}\label{infrastructure}
+
+    \chapter{Workflow composition}\label{workcomp}
+        \section{Programming language}
+        \section{Modularity}
+        \section{Workflow sharing}
+    \chapter{Performance}\label{performance}
+        \section{Filesystems}
+        \section{Scheduling}
+        \section{Performance enhancement strategies}
+    \chapter{Portability}\label{portability}
+        \section{Containers}
+    \chapter{Reproducibility}\label{reproducibility}
+        \section{Provenance Tracking}
+        \section{containerization}
+    \chapter{Discussion/Conclusion}
+
     \addcontentsline{toc}

### 2018-09-25 paper.tex
finished intro summary

@@
 
-        In sec
+        Big Data frameworks differ from those available in neuroimaging as their
+        focus is mainly on workflow composition and performance. Big Data 
+        frameworks such as Apache Spark and Dask both aim to be easy-to-use,
+        applicable to a wide variety of analyses and improve performance in the 
+        processing of Big Data. These types of systems were primarily designed 
+        to be efficient on commodity infrastructure in which transferring data
+        over a network would be extremely costly.
+
+        Currently, there exists no Big Data framework which is adapted for the
+        processing of neuroimaging data. However, tools such as the Thunder 
+        Project \note{add citation} allow for the processing of time-series
+        neuroimaging data using Spark as the backend framework. Tools such as 
+        these provide built-in functions to process the data. For users to 
+        extend functionality, it is necessary for them to understand the backend
+        framework. Other scientific workflow engines, such as Toil and Pegasus
+        \note{add citations}, have incorporated features that enable more 
+        performant workflow executions for scientific Big Data. They also enable
+        the integration of Big Data frameworks as subpipelines within their 
+        workflows. Despite integration with Big Data frameworks, it is not
+        possible to create Big Data workflows using their API exclusively.
+
+        A big limitation to adapting Big Data frameworks for neuroimaging 
+        workflows is that these frameworks are tightly-coupled with the 
+        infrastructure they were designed for -- commodity clusters. While
+        it is possible for some research labs to be using either a local cluster
+        or the cloud for their processing, many may only have access to HPC
+        clusters. As a result of this, most available scientific workflow 
+        frameworks have been designed for processing on HPC clusters. The 
+        set-up of an HPC cluster is quite different from that of a commodity 
+        cluster, and therefore, in order to effectively use Big Data frameworks
+        for scientific computing, it is necessary to adapt the infrastructure 
+        for processing on HPC.
+
+        In section \ref{datasets}, we will briefly discuss the two types of 
+        datasets that are giving rise to the neuroimaging big data. Section 
+        \ref{platforms} will provide a brief introduction to the data-sharing
+        and processing platforms available to researchers. We will conclude our 
+        introduction on the differences between commodity and HPC infrastructure
+        in section \ref{infrastructure}.
+
+        As we wish to evaluate and compare scientific workflow frameworks with
+        Big Data frameworks for the processing of neuroimaging big data, we will
+        break down our analyses by the four principal features necessary for
+        effective neuroimaging workflow engines. The workflow engines discussed
+        will include SPM, LONI, Nipype, PSOM, FastR, AA, Pydpiper 
+        for neuroimaging; Taverna, Pegasus, Make, OpenMOLE, Nextflow, Galaxy, 
+        Toil for other scientific workflows; and MapReduce, Spark and Dask for
+        Big Data Frameworks. Chapter \ref{workcomp} will evaluate these 
+        workflows on the basis of workflow composition. In other words, we will
+        look at the programming languages necessary to construct these workflows
+        , how easy is it to add new tools and will these workflows be understood
+        by the community. In Chapter \ref{performance}, we will proceed to 
+        investigate the strategies employed by these frameworks to improve the 
+        performance of their workflows. Chapter \ref{portability} will discuss
+        the portability of the workflows written using these frameworks. 
+        Workflow reproducibility will be then discussed in Chapter 
+        \ref{reproducibility}.
 

### 2018-09-25 paper.tex
rough draft of open-science platforms

@@
 
-        \section{Platforms for  neuroscience}\label{platforms}
+        \section{Open-science Platforms}\label{platforms}
+            Efficient processing of workflows are only but a small portion of 
+            the Big Data issue in neuroscience. Sharing and storing of datasets
+            and tools, secure access to and configuration of multiple computing 
+            sites and provenance capture of all activities must be managed by 
+            the researcher. This puts a heavy burden on researchers as they 
+            require some knowledge of all these components in order to manage it
+            correctly. These platforms also play a vital role in promoting 
+            research through the sharing of knowledge. Users can choose to share
+            their data and pipelines with others, reducing the need for others 
+            to collect their own data or to develop their own workflows. It 
+            also promotes robustness of results as the data and workflows will 
+            be shared and tested by multiple users.
+
+            Access to HPC resources may not always be straight-forward as 
+            configuration must be adapted to the execution site. This can 
+            range from scheduling policies, to available file systems, to 
+            environment and available libraries. Platforms relieve the user
+            from having to know site-specific details by providing seamless
+            execution on any cluster the Platform has knowledge of.
+
+            Examples of popular neuroinformatic open-science platforms include 
+            CBRAIN\note{cite paper}, OpenNeuro \note{cite poster} and Neurodata
+            \note{cite prepring}. OpenNeuro and NeuroData primarily focus on
+            cloud deployement whereas CBRAIN is configured for the deployment
+            of workflows on HPC resources. All three of these systems allow 
+            users to configure, manage, share and execute their workflows with 
+            the help of a web-based GUI. Integration with compute resources is
+            entirely abstracted by the platforms. These platforms additionally
+            provide tools to visualize their data.
+
+            As neuroimaging workflow engines are typically best suited for 
+            compute-intensive workflows, the workflows accessible in these 
+            platforms are not expected to be as suitable to process the 
+            volume of data made accessible by these platforms. Since these 
+            platforms make various infrastructures available to researchers, 
+            workflow engines must be adapted to efficiently execute on the 
+            infrastructure made available.
+
         \section{Infrastructure}\label{infrastructure}

### 2018-09-26 paper.tex
a very rough intro completed

@@
 \titlespacing*{\chapter}{0pt}{3.5ex plus 1ex minus .2ex}{2.3ex plus .2ex}
-\usepackage[left=2cm, right=2cm, top=2cm, bottom=2cm]{geometry}
+\usepackage[left=1in, right=1in, top=1in, bottom=1in]{geometry}
 \usepackage{hyperref} 
@@
             microscopy (EM), polarized light images (PLI) and micro coherence 
-            tomography (microCT) \note{need to do more research here}. Images at 
+            tomography (microCT) \note{need to do more research here.cite}. 
+            Images at 
             such high resolution are important to researchers as they provide 
@@
         \section{Infrastructure}\label{infrastructure}
+            HPC infrastructure was designed durign a time when compute resources
+            were scarce. As a result, resource managers for these systems employ
+            strategies to efficiently use the compute resources available for 
+            processing. However, with the advent of big data, there has been a 
+            shift from compute-intensive to data-intensive workloads. That is, 
+            with the accumulation of large datasets, data transfers have begun 
+            to monopolize processing time, and with it software has been 
+            designed to minimize data transfers.
+
+            We define HPC systems as those that employ shared parallel 
+            filesystem such as Lustre. This shared filesystem is the slowest of
+            all storage devices as it is shared amongst all users of the system.
+            Compute nodes in an HPC system may or may not contain any local 
+            storage and network interconnects between nodes are typically fast.
+            Resources managers used in HPC infrastructures typically employ 
+            batch scheduling and execute heterogenous workflows. 
+            Examples of schedulers used in HPC clusters 
+            include SLURM, PBS, SGE and HTCondor \note{cite here}.
+
+            Commodity infrastructure differs from HPC infrastructures in that
+            it uses less-costly hardware. It follows a shared-nothing paradigm
+            where the storage is not shared between nodes. That is all compute
+            nodes are responsible for storing the data as well. File systems
+            used in these clusters are distributed filesystems, such as the
+            Hadoop Distributed File System (HDFS) and Alluxio \note{cite}.
+            Fault-tolerance of the data is achieved through replication across
+            nodes. As transferring data across the network is typically costly 
+            in such infrastructures, frameworks designed to run on commodity 
+            infrastructure attempt to limit the transfer of data across nodes. 
+            Resource managers created specifically for these systems
+            typically expect dedicated infrastructure, as is the case with 
+            Yet Another Resource Negotiator (YARN), but can also execute 
+            heterogeneous workflows, as can be seen with Mesos \note{cite}.
+
+            Commodity clusters may also come in the form of cloud infrastructure
+            . While the hardware employed is typically the same as other
+            commodity clusters, the software used varies. Cloud infrastructures
+            have the added benefit of virtualization, allowing users to 
+            configure the infrastructure and software requirements to meet their
+            needs. Cloud infrastructure also enables researchers to request
+            dedicated resources without having to maintain these resources. One
+            of the biggest limiting factors of cloud resources is the cost 
+            associated with usage.
+
+            Researchers typically have access to HPC infrastructure, but may
+            also have access to commodity infrastructure in the format of a 
+            local cluster or cloud. Although both neuroimaging frameworks and
+            big data frameworks can be made to execute in either environments,
+            the performance improvements that they bring are very much 
+            dependent on infrastructure that they are executed on. For instance,
+            neuroimaging workflows typically read and write to a shared file
+            system at each workflow step. This is less of an issue in HPC 
+            infrastructure as network bandwidth is typically better than on 
+            commodity infrastructure. In contrasts, Big Data framework typically
+            rely on there being some form of local storage. Employing a Big Data
+            framework on HPC infrastructure may in turn result in less to no 
+            data locality -- a strategy employed by Big Data frameworks to 
+            minimize data movement across the network. Due to the variability in
+            infrastructure available to researchers, a workflow framework good
+            for neuroscience must be well adapted to handle these differences.
 

### 2018-09-26 paper.tex
fixed big datasets

@@
 
-        \section{Big neuroimaging Datasets}\label{datasets}
+        \section{Big Neuroimaging Datasets}\label{datasets}
+
+            As a result of the rise in datasharing in neuroscience, neuroimaging
+            big data has come to fruition. The field of neuroscience has now
+            reached a bottleneck where there exists sufficient data to make
+            impactful studies, but the software to process these datasets have 
+            not evolved with the size of the data.
+
             Neuroimaging big data comes in two formats: 1) large images and 2) 
@@
 
-            In contrast, large datasets are consists of many images, typically 
-            arising from multiple different subjects, that are too big to fit in 
-            storage. The images alone being small enough for an average 
-            researcher to process. Example of these datasets include the 
-            UK Biobank, HCP Project, ADNI dataset,OpenfMRI, amongst many others. 
-            These datasets are very large. For instance, the UK Biobank is 
-            expected to exceed \note{X} Petabytes in size. As subsets
-            of these datasets are manageable for the average researcher to 
-            process, much analysis has gone into them. Moreover, unlike for 
-            large images, toolkits have been developed for researchers to 
-            process this data. However,the toolkits do not enable researchers to 
-            process these entire datasets as a whole. Therefore, it is crucial 
-            to develop efficient frameworks for the processing of these datasets 
-            in order for researchers to harness the data in its entirety.
+            In contrast, large datasets are consist of many images acquired from
+            multiple subjects, which are too large to be processed in their
+            entirety. Notable examples of these datasets include the 
+            Alzheimer's Disease Neuroimaging Initiative (ADNI), OpenfMRI, 
+            the Autism Brain Imaging Data Exchange (ABIDE), the Consortium for
+            Reliability and Reproducibility (CoRR), the Human Connectome Project
+            (HCP) and the UK Biobank, with the later
+            two being flagship examples of large datasets in neuroimaging.
+
+            The ADNI intiative is a collection of images (structural and 
+            functional MRI, PET), urine,
+            serum, cerebrospinal fluid biomarkers, as assessments (clinical and
+            neuropsychometric) of 800 elderly subjects, which of whom 200 where
+            healthy, 400 exhibited mild cognitive impairment and the remaining
+            200 showed signs of Alzheimers. Imaging data is available both 
+            pre- and post-processed form.
+
+            OpenfMRI is a public repository originally conceived for the sharing
+            of task-based fMRI data. Over time, it has grown to include EEG, MEG
+            , resting-state fMRI (R-fMRI) and diffusion fMRI. Data made 
+            available through
+            OpenfMRI are obtained from both healthy and clinical individuals.
+            Raw data must be made available for all datasets published to 
+            OpenfMRI, however, preprocessed data may also accompany the raw data
+            . At present, OpenfMRI houses data of a total 3372 subjects
+            originating from 95 studies \note{obtained from website}.
+
+            To promote the advancement of knowledge of Autism Spectrum Disorders
+            (ASD), ABIDE was conceived. ABIDE is an open repository of 
+            preprocessed data obtained from 1112 R-fMRI datasets.
+            In addition to R-fMRI, datasets also contain the corresponding 
+            structural MRI and phenotypic data. As data made available in ABIDE
+            is preprocessed, standardized pipelines were developed for the 
+            preprocessing of the images. Popular neuroimaging processing tools
+            such as CIVET, ANTs, and FreeSurfer \note{cite} were used for the 
+            preprocessing of structural data, and several other pipelines were
+            used for the processing of functional data.
+
+            The CoRR was development as initiative to study and improve the 
+            reliability and reproducibility of functional connectomics. The 
+            data made available by the Consortium consists of test-retest data 
+            obtained from 33 datasets. The repository contains data from 1629 
+            subjects, in which anatomical, R-fMRI, dMRI, CBF and ASL scans are 
+            available \note{obtained from website}.
+
+            The HCP initiative aimed to characterize the human connectome of 
+            1200 healthy adults, ranging from 22 to 35 years of age, by studying
+            twins and their non-twin siblings. Data collected for this initiave
+            includes structural MRI, R-fMRI, task fMRI and dMRI. MRI data was
+            collected using a 3T scanner for all participants, with a 200 
+            participants additionally undergoing scans in a higher-resolution 7T
+            scanner. MEG and EEG data is available for a subset of the subjects. 
+            In addition to imaging data, behavioural and genetic data were also 
+            acquired. To faciliate analysis for research groups, HCP made 
+            available the raw data, minimally pre-processed data and processed 
+            data. Pipelines were developed by the HCP for pre- and post-
+            processing. As in ABIDE, these pipelines reused many commonly-used
+            neuroimaging tools such as FreeSurfer and FSL for structural data
+            and FSL and other tools for functional data. The total amount of
+            data made available by the HCP reaches up to nearly 1 Petabyte.
+
+            The UK Biobank is an initiative aimed at the detection of pre-
+            symptomatic markers of diseases occuring in middle-aged to elderly 
+            populations. The study will gather data from questionnaires, 
+            physical and cognative measures and biological samples of 500,000
+            individuals. 100,000 of these individuals will additionally be 
+            subjected to imaging studies. Imaging modalities used include
+            structural MRI, fMRI and dMRI of the brain heart and body; low-dose
+            X-ray bone and joint scans; and ultrasound of the carotid arteries.
+            As with other initiatives, the UK Biobank makes available both 
+            raw and processed data. This data is expected to exceed 0.2 
+            Petabytes. Since processed data is provided, the UK Biobank makes 
+            available it own pipelines based on tools in FSL, but also plans to
+            adapt HCP pipelines from cortical surface extraction in their images
+            .
+
+            All these initiave stress the importance of data-sharing and
+            large-scale studies for the advancement of science. Many researchers
+            are limited to the processing of subsets of the data available as 
+            they lack compute and storage resources to process the data in its
+            entirety. It is also important to note that many of the pipelines 
+            developed by these initiatives made use of popular neuroimaging 
+            toolboxes, such as FSL, FreeSurfer, ANTS, CIVET, rather that
+            creating their own. Therefore, there is a great need for bigdata
+            frameworks that can leverage existing tools for the efficient 
+            processing of large-scale datasets. With the availability of such 
+            tools, many more research labs will be able to perform
+            higher-impact studies \note{reword}.
 

### 2018-09-26 paper.tex
A few comments and corrections

@@
         lab workstation. With the advancement of data sharing technologies, this 
-        data has been made widely available, with the only factor impeding
+        data has been made widely available, \weird{with the only factor impeding
         research on this data being access to infrastructure and software that 
-        enables efficient processing on such infrastructure. Many research labs 
+        enables efficient processing on such infrastructure}. Many research labs 
         do have access to high performance computing (HPC) clusters, however, 
@@
         complex and likely beyond the knowledge of an average neuroimaging 
-        researcher. As a result, frameworks for the efficient processing of 
-        neuroimaging data need to be developed.
+        researcher \tristan{Be careful with 'knowledge', some neuroimaging researchers are very compute-savvy}. As a result, frameworks for the efficient processing of 
+        neuroimaging data need to be developed. \tristan{Add a sentence to transition from 'frameworks for efficient processing' to 'workflow engines'}
 
-        Many neuroimaging workflows currently exist. Examples include
+        Many neuroimaging workflow engines currently exist. Examples include
         Nipype, Pipeline System for Octave and Matlab (PSOM), LONI, SPM, FastR,
         Automated Analysis (AA) and Pydpiper\note{add citations here}. These 
-        workflow engines aim to satify four criteria: 1) straightforward 
+        workflow engines aim to satisfy four criteria: 1) straightforward \tristan{easy? facilitated?} 
         workflow composition, 2) performance, 3) portability of the workflows 
-        and 4) reproducibility of the analysis. While these workflow engines
+        and 4) reproducibility of the analyses. While these workflow engines
         do tackle performance, it is mainly limited to minimizing computation
@@
         focus is mainly on workflow composition and performance. Big Data 
-        frameworks such as Apache Spark and Dask both aim to be easy-to-use,
+        frameworks such as Apache Spark and Dask \tristan{Add references} both aim to be easy-to-use,
         applicable to a wide variety of analyses and improve performance in the 
@@
         Project \note{add citation} allow for the processing of time-series
-        neuroimaging data using Spark as the backend framework. Tools such as 
-        these provide built-in functions to process the data. For users to 
+        neuroimaging data using Spark as the backend framework. \weird{Tools such as 
+        these} provide built-in functions to process the data. For users to 
         extend functionality, it is necessary for them to understand the backend
@@
         performant workflow executions for scientific Big Data. They also enable
-        the integration of Big Data frameworks as subpipelines within their 
+        the integration of Big Data frameworks as sub-pipelines within their 
         workflows. Despite integration with Big Data frameworks, it is not
@@
 
-        In section \ref{datasets}, we will briefly discuss the two types of 
-        datasets that are giving rise to the neuroimaging big data. Section 
-        \ref{platforms} will provide a brief introduction to the data-sharing
-        and processing platforms available to researchers. We will conclude our 
+        \tristan{Add a sentence to present the goal of the report.}
+        In section~\ref{datasets}, we will briefly introduce the two 
+        types of datasets that are giving rise to the neuroimaging big 
+        data. Section~\ref{platforms} will provide a brief introduction 
+        to the data-sharing and processing platforms available to 
+        researchers. We will conclude our 
         introduction on the differences between commodity and HPC infrastructure
-        in section \ref{infrastructure}.
+        in section~\ref{infrastructure}.
 
@@
         Big Data Frameworks. Chapter \ref{workcomp} will evaluate these 
-        workflows on the basis of workflow composition. In other words, we will
+        workflows on the basis of workflow composition: we will % 'in other words' is on my list of expressions to avoid
         look at the programming languages necessary to construct these workflows
@@
 
-            As a result of the rise in datasharing in neuroscience, neuroimaging
-            big data has come to fruition. The field of neuroscience has now
+            As a result of the rise in data sharing in neuroscience, neuroimaging
+            Big Data has come to fruition. The field of neuroscience has now
             reached a bottleneck where there exists sufficient data to make
@@
             Neuroimaging big data comes in two formats: 1) large images and 2) 
-            large datasets. Large images consist of a singular images ranging 
+            large datasets. Large images consist of singular images ranging 
             from 100s of gigabytes to terabytes in size. A well-known example of 
@@
             size. Other examples of large images can be found in electron 
-            microscopy (EM), polarized light images (PLI) and micro coherence 
+            microscopy (EM), polarized light imaging (PLI) and micro computational
             tomography (microCT) \note{need to do more research here.cite}. 
@@
 
-            In contrast, large datasets are consist of many images acquired from
+            In contrast, large datasets consist of many images acquired from
             multiple subjects, which are too large to be processed in their
@@
             (HCP) and the UK Biobank, with the later
-            two being flagship examples of large datasets in neuroimaging.
+            two being flagship recent examples of large datasets in neuroimaging.
 
@@
 
-            To promote the advancement of knowledge of Autism Spectrum Disorders
-            (ASD), ABIDE was conceived. ABIDE is an open repository of 
+            ABIDE was conceived to promote the advancement of knowledge 
+            of Autism Spectrum Disorders (ASD). % inverted the sentence to keep symmetry with the other paragraphs
+            ABIDE is an open repository of 
             preprocessed data obtained from 1112 R-fMRI datasets.
@@
 
-            The CoRR was development as initiative to study and improve the 
+            The CoRR \tristan{expand acronym} was developed as an initiative to study and improve the 
             reliability and reproducibility of functional connectomics. The 
@@
             In addition to imaging data, behavioural and genetic data were also 
-            acquired. To faciliate analysis for research groups, HCP made 
+            acquired. To facilitate analysis for research groups, HCP made 
             available the raw data, minimally pre-processed data and processed 
@@
             The UK Biobank is an initiative aimed at the detection of pre-
-            symptomatic markers of diseases occuring in middle-aged to elderly 
+            symptomatic markers of diseases occurring in middle-aged to elderly 
             populations. The study will gather data from questionnaires, 
-            physical and cognative measures and biological samples of 500,000
+            physical and cognitive measures and biological samples of 500,000
             individuals. 100,000 of these individuals will additionally be 
@@
 
-            All these initiave stress the importance of data-sharing and
+            All these initiative stress the importance of data-sharing and
             large-scale studies for the advancement of science. Many researchers
@@
             toolboxes, such as FSL, FreeSurfer, ANTS, CIVET, rather that
-            creating their own. Therefore, there is a great need for bigdata
+            creating their own. Therefore, there is a great need for Big Data
             frameworks that can leverage existing tools for the efficient 
@@
             Efficient processing of workflows are only but a small portion of 
-            the Big Data issue in neuroscience. Sharing and storing of datasets
+            the Big Data issue in neuroscience. % nice transition, this section gives more context for your study
+            Sharing and storing of datasets
             and tools, secure access to and configuration of multiple computing 
@@
 
-            Examples of popular neuroinformatic open-science platforms include 
+            Examples of popular neuroinformatics open-science platforms include 
             CBRAIN\note{cite paper}, OpenNeuro \note{cite poster} and Neurodata
@@
         \section{Infrastructure}\label{infrastructure}
-            HPC infrastructure was designed durign a time when compute resources
-            were scarce. As a result, resource managers for these systems employ
+            HPC infrastructure was designed during a time when compute resources
+            were scarce. As a result, \tristan{I disagree with 'As a result' and I don't particularly agree with the previous sentence: even if
+            resources aren't scarce, they are good reasons to optimize resource managers: energy consumption, application makespan, fairness, etc. Also,
+            I'm not sure if resources are less scarce now than they used to be: with the rise of computational sciences, almost every researcher now
+            needs a cluster, and Compute Canada (but also other infrastructures) has been very busy.}
+            resource managers for these systems employ
             strategies to efficiently use the compute resources available for 
@@
             We define HPC systems as those that employ shared parallel 
-            filesystem such as Lustre. This shared filesystem is the slowest of
+            filesystem such as Lustre \tristan{and data-unaware schedulers?}.
+            \tristan{You should say that this definition is 'by contrast to big data platforms' to explain where it's coming from.}
+            This shared filesystem is the slowest of
             all storage devices as it is shared amongst all users of the system.
@@
 
-            Commodity infrastructure differs from HPC infrastructures in that
+            Commodity infrastructure \tristan{Call it 'Big Data infrastructure?'} differs from HPC infrastructures in that
             it uses less-costly hardware. It follows a shared-nothing paradigm
-            where the storage is not shared between nodes. That is all compute
+            where the storage is not shared between nodes. That is, all compute
             nodes are responsible for storing the data as well. File systems
@@
 
-            Commodity clusters may also come in the form of cloud infrastructure
-            . While the hardware employed is typically the same as other
+            Commodity clusters may also come in the form of cloud infrastructure.
+            While the hardware employed is typically the same as other
             commodity clusters, the software used varies. Cloud infrastructures

### 2018-09-27 paper.tex
start to chapter 2

--- 2018-09-26 paper.tex
@@
     \chapter{Workflow composition}\label{workcomp}
-        \section{Programming language}
-        \section{Modularity}
-        \section{Workflow sharing}
+
+        Workflows, also known as pipelines, describe the flow of data from 
+        activity to another. They are typically represented by a 
+        Directed Acyclic Graph (DAG). Workflow activites, also referred to as 
+        workflows stages, encapsulate a tool. Tools may either take the form of 
+        a command-line tool or a custom script. The engines determine which
+        workflow activity can be staged by determining if the stage dependencies
+        have been satisfied.
+
+        In order for a workflow engine to be successful, it must be simple to 
+        compose workflows using that framework. New researchers should be able 
+        to acquire the skills to compose worfklows in that framework without
+        investing a significant amount of time in understanding it. Reuse of
+        existing tools is also common occurrence in neuroinformatics and 
+        workflow engines need to accomodate this by undertaking a modular 
+        approach and being easily extensible. With the sharing of data also 
+        comes the sharing of workflows. For labs to share their workflows with
+        the community at large, these workflows must be easily understood by the
+        community.
+
+        In this Chapter, we will compare and contrast the workflow composition 
+        of seven neuroimaging frameworks (SPM, LONI, Nipype, PSOM, FastR, AA and
+        Pydpiper), seven scientific workflow engines stemming from other fields
+        (Taverna, Pegasus, Make, OpenMOLE, Nextflow, Galaxy, Toil), and three
+        BigData frameworks (MapReduce, Spark, Dask). We will use three axes to
+        evaluate workflow composition. Section~\ref{lang} will compare the 
+        framework on the basis of programming language used, whereas 
+        Section~\ref{mod} will look at how easy it is to integrate existing 
+        tools into the workflows. Section~\ref{sharing} will evaluate the ease
+        at which their corresponding workflows can be shared.
+
+        \section{Workflow language}\label{lang}
+            The choice of workflow language (e.g. GUI, scripting language,
+            programming language) greatly
+            affects how easily it will get picked up by the community. Complex
+            languages may mean greater efficiency of the the engines
+            , but also result in extensive time spent on learning and developing
+            the pipelines. Moreover, some researchers may not be well-versed in
+            programming and may require very easy-to-use engines.
+
+            To address this issue, some workflow engines have taken the route
+            of employing a graphical user interface (GUI). This makes workflow 
+            composition accessible to all researchers, including those with
+            little to no programming background. The caveat to employing a 
+            workflow engine with a GUI interface for workflow composition is
+            that users are greatly limited by the tasks they can perform. 
+            Workflow engines that employ a worfklow composition GUI include
+            SPM, Taverna, LONI and Galaxy. \note{double-check that none have a 
+            programming API.Loni does..i think SPM does too}
+
+            As more experienced programmers may feel limited by a GUI, workflow
+            engines have decided to provide users with the ability to describe
+            their workflows in more flexible formats. Even those with GUIs for 
+            workflow composition, such as LONI, enable more experienced users to
+            script their workflow. Six different programming
+            languages are used for describing scientific workflows: 1) Python, 
+            2) MATLAB and Octave, 3) Shell, 4) XML, 6) Common Workflow Language 
+            (CWL) and 6) Domain Specific Language (DSL). Workflows engines that 
+            employ usage of Python include: Nipype, FastR, Pydpiper, Pegasus
+            and Toil.
+            Python is an excellent choice as a programming language for 
+            scientific workflows, particularly neuroinformatics, due to its 
+            widespread use within the community. Other advantages of 
+            programming in Python including a wide array of available libraries,
+            including neuroimaging-specific ones such as Nilearn and Nibabel
+            \note{cite}, and an lower learning curve compared to other 
+            programming languages. Additionally, Python is platform independent
+            meaning it can be installed on any system. In terms of BigData 
+            frameworks that employ Python, Dask is exclusively written in Python
+            , whereas Hadoop MapReduce and Spark both have interfaces to Python
+            while being written in Java and Scala, respectively.
+
+            Another widely-used high-level programming language in the 
+            neuroinformatics community is MATLAB/Octave. Workflow engines which
+            use this language include: PSOM, SPM, and Automatic analysis. A 
+            downfall to imposing MATLAB as programming language is that the 
+            it is proprietary, thus imposing a fee on 
+            anyone wishes to program in that language. To offset this, PSOM 
+            ensure compatibility with the GNU Octave programming language, thus 
+            removing the requirement to purchase software for workflow 
+            composition. Similarly to Python, MATLAB and Octave are 
+            high-level platform-independent languages. MATLAB, in particular,
+            also provides a wide array of libraries, however, the software must
+            be purchased in order to access them.
+
+            Scientific workflows are typically composed in UNIX-based 
+            environments. These environments typically have Make installed by
+            default eliminating the need for users to install it manually.
+            Furthermore, all users who use Linux environments will typically 
+            need to write some form of Shell script. This is the basis to use 
+            workflow engines, such as Make, as scientific workflow engines. 
+            As Make is exclusive to UNIX environments, it is not 
+            platform-independent. However, Make has the flexibility of being
+            easily extensible and can interface directly with any command-line
+            tool. Shell scripting, nevertheless, is generally not as intuitive
+            as programming languages like Python and MATLAB.
+
+            Composing workflows in XML is another option provided by some 
+            workflow engines such as Pegasus and LONI. However, the use of a 
+            markup language removes some of the flexibility. Users wanting to 
+            include 
+            custom scripts as workflow activities must do so by creating a 
+            command-line interface for their custom script. Pegasus, however,
+            does provide many interfaces to describe the pipeline in including
+            Java, Python and Perl.
+
+            Another alternative is to compose workflows using Common Workflow
+            Language (CWL). This can be seen in
+
+        \section{Modularity}\label{mod}
+        \section{Workflow sharing}\label{sharing}
     \chapter{Performance}\label{performance}

### 2018-09-27 paper.tex
draft of workflow languages

@@
             Workflow engines that employ a worfklow composition GUI include
-            SPM, Taverna, LONI and Galaxy. \note{double-check that none have a 
-            programming API.Loni does..i think SPM does too}
+            SPM, Taverna, OpenMOLE, LONI and Galaxy. \note{double-check that 
+            none have a programming API.Loni does..i think SPM does too}
 
@@
             their workflows in more flexible formats. Even those with GUIs for 
-            workflow composition, such as LONI, enable more experienced users to
-            script their workflow. Six different programming
+            workflow composition, such as LONI and OpenMOLE, allow more 
+            experienced users to script their workflow. Six different 
+            programming
             languages are used for describing scientific workflows: 1) Python, 
@@
             Another alternative is to compose workflows using Common Workflow
-            Language (CWL). This can be seen in
+            Language (CWL). This can be seen in workflow engines like Toil. 
+            CWL is not a programming language, but a specification for workflows
+            . Files can either be written in JSON or YAML formats. CWL is 
+            based on workflow engines like Make, but differs in that tasks
+            are isolated and inputs and outputs must be explicit. Similarly to 
+            make, CWL is only available on UNIX platforms. \note{cite cwl 
+            website}
+
+            Domain Specific Languages (DSL) are also used to compose workflows.
+            DSLs are programming languages that have been adapted to specific
+            domains. Workflow engine examples include OpenMOLE, which was 
+            derived from Scala, and Netflow, which is derived from the Apache 
+            Groovy programming language. While both are written on top of the 
+            Java platform, Nextflow is only compatible with UNIX-based 
+            environments. These programming languages are not otherwise commonly
+            used in neuroimaging, meaning that users must learn specific 
+            languages that would not otherwise be used in their field.
+
+            With the exeption of LONI, neuroimaging workflows tend to stick to
+            commonly used programming languages such as Python and MATLAB/Octave
+            . This is likely due to the fact that the learning curve to adapt to
+            these languages is relatively low and users can harness the 
+            flexibility of these languages.
+
+            MapReduce, Spark and Dask all provide interfaces to Python, making 
+            them good candidates for neuroimaging workflow engines. In addition
+            Spark also interfaces with R, another programming language commonly
+            used in neuroimaging.
 

### 2018-09-27 paper.tex
start of modularity + some notes

@@
         \section{Modularity}\label{mod}
+            Often times, users will want to integrate existing tools, in 
+            addition to custom scripts, into their workflows. As workflow 
+            engines, with the exception of Make, are typically not written in 
+            Shell, it is necessary to create interfaces to these tools. These
+            interfaces may range from simple calls to the command-line, to more 
+            complex ones that track inputs, outputs and other parameters; ensure
+            that all parameters are specified in the correct order and that the
+            parameter and file types are all correct; capture all of the 
+            standard input/output; and create folders to encapsulate task inputs
+            and outputs. Workflow engines typically rely on more complex types 
+            of interfaces in order to ensure correct functioning of the pipeline
+            . As these interfaces can be rather complex, it is necessary for the 
+            workflow engine to provide an intuitive way for users to create 
+            interfaces for their custom tools.
+
+            Early neuroimaging workflow engines, such as SPM, are monolithic, 
+            thus preventing users from incorporating their custom workflows
+
         \section{Workflow sharing}\label{sharing}
@@
         \section{Filesystems}
+            scientific: shared FS such as Lustre, NFS, S3
+            big data: mapreduce - HDFS, spark and dask shared or distributed
         \section{Scheduling}
+            scientific: SLURM, PBS, SGE, HTCondor, Mesos (sometimes)
+            big data: mapreduce - local or YARN, Spark - standalone, YARN or 
+            Mesos, dask: any (easy to extend to any scheduler)
         \section{Performance enhancement strategies}
+            scientific: not much
+            big data: data locality, in memory computing, lazy evaluation
     \chapter{Portability}\label{portability}
+        java or python backend (most) typically platform independent (some 
+        exceptions)
+        all tools used in the pipeline must be installed on system
+        Taverna - exception as relies on web services
         \section{Containers}
+            use of containers increase portability. none of the scientific 
+            pipelines can be containerized and communicate with the scheduler 
+            outside of the container
+            big data frameworks can
     \chapter{Reproducibility}\label{reproducibility}
         \section{Provenance Tracking}
+        big data frameworks: limited to lineage
         \section{containerization}

### 2018-09-27 paper.tex
some more info on modularity

--- 2018-09-27 paper.tex
@@
             Early neuroimaging workflow engines, such as SPM, are monolithic, 
-            thus preventing users from incorporating their custom workflows
+            and therefore, do not provide a mechanism for users to incorporate
+            their custom tools. Workflow engines, such as AA, were thus 
+            conceived to enable engines such as SPM to be more modular. AA 
+            allows users to define their tasks through XML specification. AA's 
+            XML tasklist is simply an ordered list of tasks which also includes
+            details on analysis-specific parameters. Dependencies are
+            inferred by the program through analysis of the module interfaces.
+            Module interfaces consist of MATLAB user script and an XML 
+            descriptor. The descriptor contains details on the input and output
+            data stream types (e.g. epi, structural, dicom\_header) 
+            and the desired domain (e.g. study, subject, session). During 
+            execution, each task is given its own unique folder to operate on. 
+            The engine will copy data to and extract data from this folder.
+
+            LONI and FastR are other examples of workflow engines that use an
+            XML interface to describe CLIs. LONI enables generation of the 
+            XML-based module interface through its GUI. The module interface
+            consists of details on \note{from website} module metadata (e.g.
+            authors, package, version, name) and citation information, 
+            command-line syntax, parameter and file types, dependencies and 
+            output and error handling.
+
+            FastR allows for the definition of workflow modules in both XML and 
+            JSON. Module definitions are broken down into two components, 
+            \textit{Tool} and \textit{Interface} class. The \textit{Tool} is the
+            component that is written in either XML or JSON. It defines three
+            components: general metadata, target and interface. General metadata
+            consists of details such as name, version and authors. The target, 
+            on the other hand, consists of details on the execution environment
+            (e.g. operating system, interpreter and binary file). The last 
+            component of the \textit{Tool}, the interpreter, provides details on
+            the inputs and outputs of the tool (e.g. name, datatype, cardinality
+            , location). FastR provides to schemas to validate and convert the 
+            \textit{Tool} to a Python object
+            : the \textit{Tool} schema and \textit{FastRInterface}. Should these
+            schemas not meet the requirements of the user, the user may choose 
+            to define their own schema.
 

### 2018-09-27 paper.tex
some additions to workflow composition

@@
 
+            Additionally, BigData frameworks are better suited for the 
+            processing of large images. Scientific workflows are generally 
+            file-based with neuroimaging frameworks only providing parallelism
+            on the file front. Sparks main abstraction, the Resilient 
+            Distributed Dataset (RDD), permits users to process large datasets.
+            A library based on the RDD, \textit{linealg}, allows users to create
+            vectors and matrices and use common Python libraries, such as 
+            Numpy and Scikit-learn to manipulate the data. Dask similarly 
+            provides the Dask bag for file manipulation, and the Dask array for
+            the manipulation of large matrices. Dask provides its own API for
+            the processing of large matrices that is based on Numpy. 
+            Both frameworks also provide data structures to manipulate tabular
+            data: the Spark Dataframe and the Dask Dataframe. Unlike Spark, the
+            Dask Dataframe is based off of the Pandas Dataframe API.
+
         \section{Modularity}\label{mod}
@@
 
+            Some workflows do not require module interfaces, as is the case with
+            Make.
+
         \section{Workflow sharing}\label{sharing}
@@
         \section{Performance enhancement strategies}
-            scientific: not much
+            scientific: workflow checkpointing (looking at file hashes)
             big data: data locality, in memory computing, lazy evaluation

### 2018-09-27 paper.tex

--- 2018-09-27 paper.tex
@@
             Early neuroimaging workflow engines, such as SPM, are monolithic, 
-            thus preventing users from incorporating their custom workflows
+            and therefore, do not provide a mechanism for users to incorporate
+            their custom tools. Workflow engines, such as AA, were thus 
+            conceived to enable engines such as SPM to be more modular. AA 
+            allows users to define their tasks through XML specification. AA's 
+            XML tasklist is simply an ordered list of tasks which also includes
+            details on analysis-specific parameters. Dependencies are
+            inferred by the program through analysis of the module interfaces.
+            Module interfaces consist of MATLAB user script and an XML 
+            descriptor. The descriptor contains details on the input and output
+            data stream types (e.g. epi, structural, dicom\_header) 
+            and the desired domain (e.g. study, subject, session). During 
+            execution, each task is given its own unique folder to operate on. 
+            The engine will copy data to and extract data from this folder.
+
+            LONI and FastR are other examples of workflow engines that use an
+            XML interface to describe CLIs. LONI enables generation of the 
+            XML-based module interface through its GUI. The module interface
+            consists of details on \note{from website} module metadata (e.g.
+            authors, package, version, name) and citation information, 
+            command-line syntax, parameter and file types, dependencies and 
+            output and error handling.
+
+            FastR allows for the definition of workflow modules in both XML and 
+            JSON. Module definitions are broken down into two components, 
+            \textit{Tool} and \textit{Interface} class. The \textit{Tool} is the
+            component that is written in either XML or JSON. It defines three
+            components: general metadata, target and interface. General metadata
+            consists of details such as name, version and authors. The target, 
+            on the other hand, consists of details on the execution environment
+            (e.g. operating system, interpreter and binary file). The last 
+            component of the \textit{Tool}, the interpreter, provides details on
+            the inputs and outputs of the tool (e.g. name, datatype, cardinality
+            , location). FastR provides to schemas to validate and convert the 
+            \textit{Tool} to a Python object
+            : the \textit{Tool} schema and \textit{FastRInterface}. Should these
+            schemas not meet the requirements of the user, the user may choose 
+            to define their own schema.
+
+            Some workflows do not require module interfaces, as is the case with
+            Make.
 

### 2018-09-27 paper.tex
more of modularity

--- 2018-09-27 paper.tex
@@
 
-            LONI and FastR are other examples of workflow engines that use an
-            XML interface to describe CLIs. LONI enables generation of the 
-            XML-based module interface through its GUI. The module interface
+            LONI, FastR, Taverna and Pegasus are other examples of workflow 
+            engines that 
+            use an XML interface to describe CLIs. LONI enables generation of 
+            the XML-based module interface through its GUI. The module interface
             consists of details on \note{from website} module metadata (e.g.
@@
 
+            Tool addition in Taverna is primarily performed through the user 
+            interface, but can also achieved through the creation of an XML
+            descriptor. Taverna differs from other workflow engines in that it
+            execute generic-style services (i.e. WSDL and REST) in addition to
+            other types of Web Services, as well a external CLIs, R scripts and
+            Java services. To interface with a CLI, the XML file must contain 
+            the command to execute, file inputs and outputs (if required), if 
+            standard input, output and error streams must be conserved and the 
+            location where the command should execute\note{from website}.
+
+            Pegasus is another workflow engine that uses an XML format to 
+            define the CLI tools interface. The XML interfaces are typically 
+            incorporated directly into the workflow DAX definition. However, a 
+            separate multiline text-based 'Transformation Catalogue' file can
+            be provided. Information encoded within the interfaces include
+            tool identifier (typically namespace::version), the URL or filepath
+            to the executable, site identifier of the executable's host, 
+            executable type (e.g. installed or staged) and the profile. Pegasus
+            can also execute containerized tools. For this, a container 
+            identifier, image URL, site identifier of the image, and container
+            profiles must also be defined.
+
             Some workflows do not require module interfaces, as is the case with
-            Make.
+            Make. The pipeline is expressed in a makefile containing a set of 
+            rules. Rules consist of three components: targets, dependencies and
+            recipes. The target is the output (typically a file) of the 
+            execution of a recipe and the dependencies are the input required to 
+            produce the given target. Recipes are the shell commands that will
+            generate the target from its dependencies. As makefiles are written
+            entirely using shell commands, it is not necessarily to create 
+            interfaces to the shell commands of the tools.
 

### 2018-09-28 paper.tex
more on modularity

@@
 
-            LONI, FastR, Taverna and Pegasus are other examples of workflow 
-            engines that 
+            LONI, FastR, Taverna, Pegasus and Galaxy are other examples of 
+            workflow engines that 
             use an XML interface to describe CLIs. LONI enables generation of 
@@
             identifier, image URL, site identifier of the image, and container
-            profiles must also be defined.
+            profiles must also be defined \note{from website}.
 
-            Some workflows do not require module interfaces, as is the case with
-            Make. The pipeline is expressed in a makefile containing a set of 
-            rules. Rules consist of three components: targets, dependencies and
-            recipes. The target is the output (typically a file) of the 
+            While Galaxy is primarily a GUI-based application, an XML tool
+            configuration file is required to add the tool to the application.
+            The tool wrappers schema is very rich allowing for many tool 
+            features to be defined. Similarly to many other workflow engine tool
+            wrapper, the configuration file contains details on the tool name, 
+            type, id, versions, command and input and output specifications. It
+            also enables the execution of containerized tools. Furthermore,
+            Galaxy enables users to specific actions to be performed on the 
+            output files once created and define test cases.
+
+            While many worfklow engines use XML or JSON to describe their tools,
+            others wrap their tools in an interface defined by a
+            programming/scripting language. Nipype, Pydpiper and Toil all use
+            Python to create their wrappers; PSOM wrappers are written in
+            MATLAB/Octave; OpenMOLE uses Java/Scala; and Make interfaces are
+            written in Shell Script.
+
+            Nipype's wrapper, written in Python, can encapsulate CLIs, SPM, 
+            MATLAB and Python scripts. Attributes and methods that can be 
+            defined for all interfaces include InputSpec, OutputSpec and a 
+            Python dictionary of all expected outputs. InputSpec and OutputSpec
+            are two Nipype classes. The former defines the input parameters of 
+            a command and the second defines the outputs that are possibly 
+            generated by the command. For CLIs, it is necessary to define the
+            command. Additionally, Nipype will start X virtual framebuffer 
+            before executing a CLI, if specified. For SPM interfaces, the job 
+            type and name must be specified. Interfaces to MATLAB require 
+            definition of the run function, while Python interfaces require
+            the definition of the \textit{\_run\_interface} function. \note{from
+            website}.
+
+            Pydpiper's tools can either be custom Python code or CLIs. The 
+            \textit{PipelineStage} and \textit{CmdStage} are used to 
+            encapsulate the tools. The \textit{CmdStage} class is specifically 
+            used to interface with CLIs, and thus, accepts a parsed array of the
+            command and its parameters as input. Atoms, typically used to wrap
+            around MINC file-format related tools, inherit from the
+            \textit{CmdStage} class, and require the input MINC file (string or
+            file handler) to be provided. A output file can also be specifiec, 
+            but it must be of the same format as the MINC file (string or file 
+            handler). Additionally, it is possible to pass multiple optional 
+            arguments to the Atom.
+
+            Toil is another workflow engine that describes their tool interfaces
+            in Python. The wrappers, also known as \textit{Jobs} enable users to
+            interface with custom Python code or Dockerized \note{cite Docker} 
+            CLIs. Interfaces can be constructed either by creating a class that
+            inherits from the \textit{Job} class or through the use of a 
+            built-in function that converts the user-defined function (UDF) to a 
+            Toil \textit{Job}. To convert, all that is required is the UDF,
+            however, the user may also choose to include details on memory, 
+            number of cores and disk space required. For Dockerized tools, Toil
+            provides a built-in function that accepts the tool name, working dir
+            and parameters of the tool.
+
+            PSOM's tool interfaces are known as \textit{Bricks}. \textit{Bricks}
+            can interface with any MATLAB function or CLI. Bricks require the
+            definition of input and output files, in addition to a structure of
+            program options. A useful feature found in PSOM is the ability 
+            to perform a "dry-run" of the Brick without executing the command.
+
+            OpenMOLE has the ability to encapsulate UDF Scala tasks and CLI. 
+            In the case of CLI tasks, it is also possible to containerize or 
+            encapsulate with CARE \note{cite} to make it portable.
+
+            The Make pipeline is expressed in a makefile containing a set of 
+            rules (Tool interfaces). Rules consist of three components: targets, 
+            dependencies and recipes. The target is the output (typically a 
+            file) of the 
             execution of a recipe and the dependencies are the input required to 
             produce the given target. Recipes are the shell commands that will
-            generate the target from its dependencies. As makefiles are written
-            entirely using shell commands, it is not necessarily to create 
-            interfaces to the shell commands of the tools.
+            generate the target from its dependencies.
 

### 2018-09-28 paper.tex

--- 2018-09-27 paper.tex
@@
 
-            LONI and FastR are other examples of workflow engines that use an
-            XML interface to describe CLIs. LONI enables generation of the 
-            XML-based module interface through its GUI. The module interface
+            LONI, FastR, Taverna, Pegasus and Galaxy are other examples of 
+            workflow engines that 
+            use an XML interface to describe CLIs. LONI enables generation of 
+            the XML-based module interface through its GUI. The module interface
             consists of details on \note{from website} module metadata (e.g.
@@
 
-            Some workflows do not require module interfaces, as is the case with
-            Make.
+            Tool addition in Taverna is primarily performed through the user 
+            interface, but can also achieved through the creation of an XML
+            descriptor. Taverna differs from other workflow engines in that it
+            execute generic-style services (i.e. WSDL and REST) in addition to
+            other types of Web Services, as well a external CLIs, R scripts and
+            Java services. To interface with a CLI, the XML file must contain 
+            the command to execute, file inputs and outputs (if required), if 
+            standard input, output and error streams must be conserved and the 
+            location where the command should execute\note{from website}.
+
+            Pegasus is another workflow engine that uses an XML format to 
+            define the CLI tools interface. The XML interfaces are typically 
+            incorporated directly into the workflow DAX definition. However, a 
+            separate multiline text-based 'Transformation Catalogue' file can
+            be provided. Information encoded within the interfaces include
+            tool identifier (typically namespace::version), the URL or filepath
+            to the executable, site identifier of the executable's host, 
+            executable type (e.g. installed or staged) and the profile. Pegasus
+            can also execute containerized tools. For this, a container 
+            identifier, image URL, site identifier of the image, and container
+            profiles must also be defined \note{from website}.
+
+            While Galaxy is primarily a GUI-based application, an XML tool
+            configuration file is required to add the tool to the application.
+            The tool wrappers schema is very rich allowing for many tool 
+            features to be defined. Similarly to many other workflow engine tool
+            wrapper, the configuration file contains details on the tool name, 
+            type, id, versions, command and input and output specifications. It
+            also enables the execution of containerized tools. Furthermore,
+            Galaxy enables users to specific actions to be performed on the 
+            output files once created and define test cases.
+
+            While many worfklow engines use XML or JSON to describe their tools,
+            others wrap their tools in an interface defined by a
+            programming/scripting language. Nipype, Pydpiper and Toil all use
+            Python to create their wrappers; PSOM wrappers are written in
+            MATLAB/Octave; OpenMOLE uses Java/Scala; and Make interfaces are
+            written in Shell Script.
+
+            Nipype's wrapper, written in Python, can encapsulate CLIs, SPM, 
+            MATLAB and Python scripts. Attributes and methods that can be 
+            defined for all interfaces include InputSpec, OutputSpec and a 
+            Python dictionary of all expected outputs. InputSpec and OutputSpec
+            are two Nipype classes. The former defines the input parameters of 
+            a command and the second defines the outputs that are possibly 
+            generated by the command. For CLIs, it is necessary to define the
+            command. Additionally, Nipype will start X virtual framebuffer 
+            before executing a CLI, if specified. For SPM interfaces, the job 
+            type and name must be specified. Interfaces to MATLAB require 
+            definition of the run function, while Python interfaces require
+            the definition of the \textit{\_run\_interface} function. \note{from
+            website}.
+
+            Pydpiper's tools can either be custom Python code or CLIs. The 
+            \textit{PipelineStage} and \textit{CmdStage} are used to 
+            encapsulate the tools. The \textit{CmdStage} class is specifically 
+            used to interface with CLIs, and thus, accepts a parsed array of the
+            command and its parameters as input. Atoms, typically used to wrap
+            around MINC file-format related tools, inherit from the
+            \textit{CmdStage} class, and require the input MINC file (string or
+            file handler) to be provided. A output file can also be specifiec, 
+            but it must be of the same format as the MINC file (string or file 
+            handler). Additionally, it is possible to pass multiple optional 
+            arguments to the Atom.
+
+            Toil is another workflow engine that describes their tool interfaces
+            in Python. The wrappers, also known as \textit{Jobs} enable users to
+            interface with custom Python code or Dockerized \note{cite Docker} 
+            CLIs. Interfaces can be constructed either by creating a class that
+            inherits from the \textit{Job} class or through the use of a 
+            built-in function that converts the user-defined function (UDF) to a 
+            Toil \textit{Job}. To convert, all that is required is the UDF,
+            however, the user may also choose to include details on memory, 
+            number of cores and disk space required. For Dockerized tools, Toil
+            provides a built-in function that accepts the tool name, working dir
+            and parameters of the tool.
+
+            PSOM's tool interfaces are known as \textit{Bricks}. \textit{Bricks}
+            can interface with any MATLAB function or CLI. Bricks require the
+            definition of input and output files, in addition to a structure of
+            program options. A useful feature found in PSOM is the ability 
+            to perform a "dry-run" of the Brick without executing the command.
+
+            OpenMOLE has the ability to encapsulate UDF Scala tasks and CLI. 
+            In the case of CLI tasks, it is also possible to containerize or 
+            encapsulate with CARE \note{cite} to make it portable.
+
+            The Make pipeline is expressed in a makefile containing a set of 
+            rules (Tool interfaces). Rules consist of three components: targets, 
+            dependencies and recipes. The target is the output (typically a 
+            file) of the 
+            execution of a recipe and the dependencies are the input required to 
+            produce the given target. Recipes are the shell commands that will
+            generate the target from its dependencies.
 

### 2018-09-28 paper.tex

@@
         activity to another. They are typically represented by a 
-        Directed Acyclic Graph (DAG). Workflow activites, also referred to as 
-        workflows stages, encapsulate a tool. Tools may either take the form of 
-        a command-line tool or a custom script. The engines determine which
+        Directed Acyclic Graph (DAG) \tristan{Not necessarily a DAG, see Nipype for instance, it has loops.
+        I don't think you want to enter in a taxonomy of languages though, just say 'graph'}. Workflow activites, also referred to as 
+        workflows stages, encapsulate a tool \tristan{Mention 'interface' here as an intro to sub-sec 2}. Tools may either take the form of 
+        a command-line tool or a custom script \tristan{A script is a command-line tool. I think tools may also be web services (Taverna) or just 
+        software libraries (Dask)}. The engines determine which
         workflow activity can be staged by determining if the stage dependencies
@@
         the community at large, these workflows must be easily understood by the
-        community.
+        community. \tristan{Perhaps mention that this also relates to the sharing
+        of reproducible results}.
 
@@
             affects how easily it will get picked up by the community. Complex
-            languages may mean greater efficiency of the the engines
+            languages may mean greater efficiency of the engines
             , but also result in extensive time spent on learning and developing
@@
             none have a programming API.Loni does..i think SPM does too}
+            \tristan{My conclusion on GUIs is that they are suitable only
+            for very simple pipelines. Perhaps the intro could give an idea
+            of the types of pipelines encountered, showing a simple one and
+            a more complex one. No way fmriprep is described in a GUI.}
 
@@
             languages are used for describing scientific workflows: 1) Python, 
-            2) MATLAB and Octave, 3) Shell, 4) XML, 6) Common Workflow Language 
-            (CWL) and 6) Domain Specific Language (DSL). Workflows engines that 
+            2) MATLAB and Octave \tristan{I think the official name is GNU/Octave}, 3) Shell \tristan{scripting languages}, 4) XML, 6) Common Workflow Language 
+            (CWL) and 6) Domain Specific Language (DSL). \tristan{I don't see why CWL and XML don't qualify as DSL} Workflows engines that 
             employ usage of Python include: Nipype, FastR, Pydpiper, Pegasus
-            and Toil.
+            and Toil \tristan{Dask? Spark?}.
             Python is an excellent choice as a programming language for 
@@
             programming in Python including a wide array of available libraries,
-            including neuroimaging-specific ones such as Nilearn and Nibabel
+            including neuroimaging-specific ones such as Nilearn and Nibabel \tristan{mention scipy, sci-kit learn, etc}
             \note{cite}, and an lower learning curve compared to other 
@@
             it is proprietary, thus imposing a fee on 
-            anyone wishes to program in that language. To offset this, PSOM 
-            ensure compatibility with the GNU Octave programming language, thus 
+            anyone wishes to program in that language \tristan{You should also mention the
+            advantage of Matlab: easy to learn, integrated ecosystem, 
+            performance, see 
+            \url{https://www.mathworks.com/products/matlab/matlab-vs-python.html}}. 
+            To offset this, PSOM ensure compatibility with the GNU 
+            Octave programming language, thus 
             removing the requirement to purchase software for workflow 
@@
 
-            Scientific workflows are typically composed in UNIX-based 
+            \tristan{This paragraph should start with ``Shell scripting ...". The structuration of this section by language
+            is great, but the writing of this paragram doesn't put it forward.}Scientific workflows are typically composed in UNIX-based 
             environments. These environments typically have Make installed by
@@
             workflow engines such as Pegasus and LONI. However, the use of a 
-            markup language removes some of the flexibility. Users wanting to 
+            markup language removes some of the flexibility. \tristan{Before you explain the downside, 
+            you should say what are the advantages: mainly language-independent parsers and validation.} Users wanting to 
             include 
@@
             CWL is not a programming language, but a specification for workflows
-            . Files can either be written in JSON or YAML formats. CWL is 
-            based on workflow engines like Make, but differs in that tasks
-            are isolated and inputs and outputs must be explicit. Similarly to 
-            make, CWL is only available on UNIX platforms. \note{cite cwl 
+            \tristan{It is a YAML-based DSL for workflows. Just like Pegasus' DAX is an XML-based DSL}. Files can either be written in JSON or YAML formats. CWL is 
+            based on workflow engines like Make \tristan{? I don't think CWL relies on make, does it?}, but differs in that tasks
+            are isolated and inputs and outputs must be explicit \tristan{inputs and outputs are somewhat explicit in a Makefile}. Similarly to 
+            make, CWL is only available on UNIX platforms \tristan{strange, the CWL interpreter is in Python so I'd expect it to run anywhere}. \note{cite cwl 
             website}
@@
             domains. Workflow engine examples include OpenMOLE, which was 
-            derived from Scala, and Netflow, which is derived from the Apache 
+            derived from Scala, and Netflow \tristan{Nextflow?}, which is derived from the Apache 
             Groovy programming language. While both are written on top of the 
-            Java platform, Nextflow is only compatible with UNIX-based 
+            Java platform, Nextflow is only compatible with UNIX-based \tristan{How about using Linux everywhere instead of UNIX-based?} 
             environments. These programming languages are not otherwise commonly
@@
 
-            Additionally, BigData frameworks are better suited for the 
+            \tristan{So this is now about data structures, not just language.
+            This is really excellent, but this should perhaps go into a 
+            specific sub-section. I think you could mention that 
+            scientific workflow engines traditionally rely on files 
+            only, and simple types, but that big data engines have 
+            invested a lot in data structures, starting from key-value 
+            pair representations in MR to RDD and Dask arrays. Spark is 
+            actually built around RDDs, this is salient-enough to be 
+            discussed a bit more. It actually makes a lot of sense that 
+            data-focused engines are built around data structures 
+            instead of control constructs. The ACM Com Spark paper makes it clear
+            that unifying data structures is key to performance and interoperability, 
+            because it allows data to remain in-memory instead of having to go back
+            and forth to different environments. I would only mention linealg 
+            in the description of Dask arrays. So, how about a 
+            sub-section 'data structures' with 4 paragraphs: (1) 
+            scientific and neuroimaging engines, (2) Big-Data, MR, (3) 
+            Spark's RDDs and DataFrames, (4) Dask arrays, mentioning 
+            linealg. } Additionally, BigData frameworks are better 
+            suited for the 
             processing of large images. Scientific workflows are generally 
@@
         \section{Modularity}\label{mod}
+            \tristan{Since this section is about interfaces, I would
+            rather call it ``Interfaces" or ``Tool interfaces".}
             Often times, users will want to integrate existing tools, in 
@@
             and therefore, do not provide a mechanism for users to incorporate
-            their custom tools. Workflow engines, such as AA, were thus 
+            their custom tools \tristan{Users can still integrate tools as Matlab functions
+            doing system calls}. Workflow engines, such as AA, were thus 
             conceived to enable engines such as SPM to be more modular. AA 
@@
             executable type (e.g. installed or staged) and the profile. Pegasus
-            can also execute containerized tools. For this, a container 
+            can also execute containerized tools \tristan{how about the other ones?}. For this, a container 
             identifier, image URL, site identifier of the image, and container
@@
             command. Additionally, Nipype will start X virtual framebuffer 
-            before executing a CLI, if specified. For SPM interfaces, the job 
+            before executing a CLI, if specified \tristan{you need to explain why}. For SPM interfaces, the job 
             type and name must be specified. Interfaces to MATLAB require 
@@
             program options. A useful feature found in PSOM is the ability 
-            to perform a "dry-run" of the Brick without executing the command.
+            to perform a ``dry-run" of the Brick without executing the command.
 
@@
             generate the target from its dependencies.
+
+            \tristan{Good material in here, but it lacks a bit structure.
+            I like your structure of the previous sub-section, by language. I think
+            here you could have two groups: (1) DSL descriptions (XML, JSON, etc) and (2)
+            Python or Matlab descriptions. One advantage of DSLs is that they can 
+            be more easily reused in different engines, while programming languages
+            are more flexible, for instance Nipype allows outputs to be specified
+            at runtime (dynamic number of inputs with names only known after
+            the interface ran.)}
 

### 2018-09-29 paper.tex
intro on performance

@@
 
+            \subsection{Data structures}
+                \subsubsection{Scientific workflows}
+                \subsubsection{Big Data: MapReduce}
+
             \tristan{So this is now about data structures, not just language.
@@
 
-        \section{Modularity}\label{mod}
+        \section{Tool interfaces}\label{mod}
             \tristan{Since this section is about interfaces, I would
@@
 
-            OpenMOLE has the ability to encapsulate UDF Scala tasks and CLI. 
+            OpenMOLE has the ability to encapsulate UDF Scala scripts and CLI. 
             In the case of CLI tasks, it is also possible to containerize or 
-            encapsulate with CARE \note{cite} to make it portable.
+            encapsulate with CARE \note{cite} to make it portable. OpenMOLE
+            defines several classes to allow for the execution of different 
+            types of tools. Like other framworks, it is vital to describe the
+            inputs and outputs of the Tasks such that the execution of tasks 
+            can be performed.
 
@@
             generate the target from its dependencies.
+
+            Unlike scientific workflow engines, BigData frameworks do not 
+            provide default interfaces for the execution of tools written in 
+            languages different from their own. Users defining their workflows
+            in Dask or the Spark and MapReduce using the Python API integrate
+            Python UDFs in their workflow. While Dask users can reference their 
+            UDF of Python function directly in their workflow, Spark users
+            must define a lambda function within an action or transformation 
+            that executes their UDF or library on the data. Similarly to Spark,
+            functions referenced in MapReduce must be called from within a Map
+            or Reduce task. MapReduce can additionally reference Java 
+            code and Spark can call on code also written in Scala, Java and R.
+            \note{verify if it's possible to mix-and-match different libraries
+            ...don't think so}
+
+            For all workflow engines that do not, by default, interface with 
+            command-line tools or containers, there are libraries that can be
+            easily integrated to achieve that. Boutiques, for instance, is a 
+            Python library that can execute command-line or containerized tools
+            defined within a JSON schema. Furthermore, the library's richness
+            compete with Galaxy's XML. Python-based BigData frameworks can 
+            easily call on this library from within the workflow in order to
+            communicate with the tools.
+
+            \subsection{input and output definitions}
+                TBD
 
@@
         \section{Workflow sharing}\label{sharing}
+
     \chapter{Performance}\label{performance}
-        \section{Filesystems}
+        Neuroimaging frameworks are notoriously known to be compute-intensive.
+        The tools and their respective workflows needed to incorporate 
+        parallelism in order to save on computing time. Maintaining good 
+        performance has become especially critical now that BigData has been
+        introduced into the mix. The dichotomy remains between scientific 
+        workflows and BigData workflows. While scientific workflows have made
+        advances towards more performant engines, they still lack the 
+        strategies BigData frameworks employ to reduce data transfers.
+
+        Another notable aspect to consider is that both types of engines 
+        (scientific and BigData) are tightly coupled to the infrastructure
+        they were intended to be used on. That is, as scientific workflow 
+        try to remain efficient in their handling of compute resources, as such
+        resources were previously scarce. In contrast, BigData frameworks 
+        make efficient use of commodity clusters, where data transfers are 
+        burdensome, particularly when faced with BigData.
+
+        As some researchers are trying to shift towards the use of commodity 
+        infrastructure while others are trying to process BigData on HPC, the 
+        ideal workflow engine must run efficiently regardless of 
+        infrastructure available. Otherwise, researchers are either left with
+        workflows running inefficiently on resources or having to learn 
+        multiple frameworks and rewrite their workflows depending on the 
+        infrastructure selected.
+
+        In this Chapter, we will compare scientific
+        workflows to BigData frameworks on the basis of performance. 
+        Section~\ref{fs} will look at what filesystems are required by the 
+        engines. This is important because data movement of any kind is costly,
+        and may be worse depending on the filesystems available. 
+        Section~\ref{sched} will provide an overview of the schedulers 
+        available and their compatibility with the workflow engines. 
+        Section~\ref{other} will discuss other performance improvement 
+        strategies employed by the workflow engines.
+
+        \section{Filesystems}\label{fs}
+
+            Scientific workflows historically relied on I/O to transfer data
+            between nodes. As scientific workflows are usually executed on
+            HPC infrastructure, this meant writing to a filesystem such as 
+            Lustre
+
+            \subsection{Lustre}
+            \subsection{Hadoop Distributed File System}
+            \subsection{Amazon S3}
+
             scientific: shared FS such as Lustre, NFS, S3
             big data: mapreduce - HDFS, spark and dask shared or distributed
-        \section{Scheduling}
+        \section{Scheduling}\label{sched}
             scientific: SLURM, PBS, SGE, HTCondor, Mesos (sometimes)
@@
             Mesos, dask: any (easy to extend to any scheduler)
-        \section{Performance enhancement strategies}
+        \section{Performance enhancement strategies}\label{other}
             scientific: workflow checkpointing (looking at file hashes)

### 2018-09-30 paper.tex
some notes on performance

@@
 
-            Scientific workflows historically relied on I/O to transfer data
-            between nodes. As scientific workflows are usually executed on
-            HPC infrastructure, this meant writing to a filesystem such as 
-            Lustre
+            Unlike MapReduce and Spark-based workflows, Dask and scientific 
+            workflows are not data-centric. They focus on the efficient 
+            scheduling of tasks and rely on I/O and network transfers to
+            communicate the data between nodes. With BigData workflows, such
+            operations could incur significant performance penalties. 
+            Filesystems. To mitigate this and remain efficient in the face of
+            BigData, Dask implements lazy evaluation and in-memory processing.
+            BigData platforms such as MapReduce rely heavily on I/O, but the 
+            I/O performed occurs on the compute node rather than on 
+            shared storaged, making it generally more efficient than scientific
+            workflows. Additionally, MapReduce uses the concept of 
+            locality to ensure limit data transfers across the network. Spark
+            uses data locality, lazy evaluation and in-memory computing to 
+            ensure its optimal performance.
+            For the efficient processing of neuroimaging BigData, it will be
+            necessary for scientific worflow engines to incorporate these 
+            strategies.
 
-            \subsection{Lustre}
-            \subsection{Hadoop Distributed File System}
-            \subsection{Amazon S3}
+            \subsection{Strategies}
+                \subsubsection{in-memory computing}
+                    Random Access Memory (RAM) is the fastest memory available
+                    to workflow engines. BigData frameworks, such as Spark and 
+                    Dask, leverage this feature by maintaining data in memory 
+                    between tasks rather that saving them to local disk or 
+                    shared memory which may incur significant overhead.
 
-            scientific: shared FS such as Lustre, NFS, S3
-            big data: mapreduce - HDFS, spark and dask shared or distributed
-        \section{Scheduling}\label{sched}
-            scientific: SLURM, PBS, SGE, HTCondor, Mesos (sometimes)
+                    Spark and Dask can both read from filesystems available to
+                    all nodes. When an action (Spark) or a compute (Dask) is 
+                    called, the data is read from the storage and uploaded into
+                    memory. Tasks use the data located in memory to compute the
+                    results. Should data not fit in memory, the results may 
+                    spill to local disk, in the instance where nodes contain
+                    storage. Otherwise, the engines will spill data to shared 
+                    filesystem. The later is a common case in HPC systems as 
+                    storage and compute resources are typically separate, 
+                    however, these shared filesystems are typically slower than 
+                    local storage, and therefore, there will be additional 
+                    performance penalties.
+
+                    As data is not stored within a resilient distributed 
+                    filesystem such as Lustre and HDFS, it risks being 
+                    permanently lost if a node failure occurs. Spark and Dask
+                    have both addressed this through the notion of 
+                    \textit{lineage}. Lineage refers to the set of operations
+                    that the data has undergone. By recording the lineage, lost
+                    data will always be recoverable as the set of operations
+                    to reproduce the data are known. Should a node be lost to 
+                    the cluster, Spark and Dask reschedule the tasks required 
+                    to rebuild the dataset to other nodes.
+
+                    \note{mention caching}
+
+                    While Spark and Dask both rely on the concept of shared 
+                    memory, Spark avoids data transfers as often as possible by
+                    scheduling tasks to nodes where the memory is located. Dask
+                    , however, will schedule its tasks to the next available 
+                    node.
+
+                \subsubsection{Data locality}
+                    Data locality, with respect to BigData frameworks, is a
+                    scheduling
+                    strategy in which location of the data is known by the 
+                    scheduler and tasks are scheduled to the location where the
+                    necessary data is located. It was one of the first main 
+                    strategies implemented to make BigData processing more 
+                    efficient. The reasoning behind data locality is that 
+                    network transfers are expensive, particularly within 
+                    commodity infrastracture, and therefore transferring large 
+                    volumes of data across the network for every task would be 
+                    rather expensive.
+
+                    While MapReduce and Spark both partition data across the 
+                    network similarly, they achieve data locality slightly 
+                    differently. MapReduce can use the local filesystem in 
+                    standalone mode, but requires use of a distributed 
+                    filesystem executing on the compute nodes such as the 
+                    Hadoop Distributed File System (HDFS). The reason as to why
+                    parallel file systems such as Lustre do not work is that 
+                    the storage nodes on such file systems are separate from 
+                    the compute nodes. Separate storage and compute nodes 
+                    eliminate any possiblity of data locality as network 
+                    transfers will occur regardless of storage node. When the 
+                    next MapReduce task is available, the master node will 
+                    query the distributed file system to determine the 
+                    available nodes closest. The master will first attempt to
+                    schedule the task on a node containing the data, but if 
+                    that node is not available, it will attempt to schedule
+                    the task on the same rack. If all else fails, the task will
+                    be scheduled on a node found on a different rack.
+
+                    In contrast to MapReduce, Spark does not require data to 
+                    be stored on a HDFS-like filesystem. It can read data from
+                    any form of shared-filesystem, including Lustre. If data 
+                    is extracted from a non HDFS-like filesystem, the initial 
+                    loading of data will not use data locality. However, in all
+                    future tasks in which the data is already loaded into an 
+                    RDD, the scheduler will be able to assign task to the most
+                    favorable partition dictated by delay scheduling.
+
+                \subsubsection{lazy evaluation}
+
+                    Lazy evaluation is a strategy used throughout computer 
+                    science to improve performance of operations. Rather than
+                    evaluating a expression immediately, it will wait until it 
+                    is required in order to compute the result. This is 
+                    achieved in Spark through the calling of an \textit{action}
+                    on an RDD and in Dask through the calling of a 
+                    \textit{compute} function on one of its data structures.
+                    By using lazy evaluation, the engines are capable of 
+                    constructing graphs that result in the most optimal
+                    task order of execution.
+
+            \subsection{BigData Filesystem optimizations for HPC}
+
+                Writting BigData to Lustre is very costly compared to local 
+                node storage. BigData filesystems like HDFS may not necessarily
+                be the best solution for HPC as they will not only use up 
+                storage on compute nodes, but also replicate the data across 
+                the cluster taking up a significant amount of storage space.
+                Moreover, while BigData frameworks typically execute on a 
+                dedicated BigData cluster, they will be on a shared cluster in
+                HPC.
+
+                Alluxio, formerly known as tachyon, is an in-memory filesystem
+                that stores lineage data rather than the data itself. It is
+                assumed that recomputation of data will occur more rapidly than
+                data transfer over the network. However,
+
+                Triple-H leverages the heterogeneity of HPC clusters to 
+                optimize the data's placement across the cluster for most 
+                efficient use.
+
+            \note{scientific: shared FS such as Lustre, NFS, S3
+            big data: mapreduce - HDFS, spark and dask shared or distributed}
+        \section{BigData Scheduling}\label{sched}
+
+            Scientific workflows typically rely on HPC schedulers for 
+            distributed scheduling on clusters. These schedulers include SLURM,
+            PBS,TORQUE, SGE, HTCondor and Mesos. Additionally, they may provide
+            their own form of scheduling for computations on a single node. 
+            Nipype uses mutliproc... etc. BigData frameworks, such as Dask,
+            also provide interfaces to these HPC schedulers. However, other
+            frameworks such as MapReduce and Spark are limited to BigData 
+            clusters such as Yet Another Resource Negotiator (YARN and MESOS).
+            Spark additionally provides its own standalone scheduler. Similarly
+            to scientific workflows, all three BigData frameworks provide their
+            own schedulers for single-node scheduling.
+
+            Spark's standalone scheduler is a simple scheduler that can be used
+            in distributed workflows. It can execute in two modes: client and
+            cluster. In client mode, the  \note{cannot run in cluster mode in pyspark}
+
+            YARN needs hadoop cluster
+
+            Mesos can run on HPC
+
+            dask distributed
+            \note{scientific: SLURM, PBS, SGE, HTCondor, Mesos (sometimes)
             big data: mapreduce - local or YARN, Spark - standalone, YARN or 
-            Mesos, dask: any (easy to extend to any scheduler)
+            Mesos, dask: any (easy to extend to any scheduler)}
         \section{Performance enhancement strategies}\label{other}
@@
         \section{Containers}
+
+             Containeri
             use of containers increase portability. none of the scientific 

### 2018-09-30 paper.tex
some more text of FS

@@
                 dedicated BigData cluster, they will be on a shared cluster in
-                HPC.
+                HPC, with many processed trying to access the memory. The 
+                following are two optimizations made for
 
-                Alluxio, formerly known as tachyon, is an in-memory filesystem
-                that stores lineage data rather than the data itself. It is
-                assumed that recomputation of data will occur more rapidly than
-                data transfer over the network. However,
+                Alluxio, formerly known as Tachyon, is an in-memory filesystem
+                that stores lineage data to offset the cost of writing data to 
+                disk. The assumption that Alluxio makes is that it will be less
+                costly to recompute the data that to store the data to disk and
+                replicate across the network for fault tolerance. However,
+                unlike Spark applications, Alluxio is meant to run indefinitely
+                , making it unfavourable, in some cases, to recompute the data
+                from its starting point. For this, Alluxio evaluates the 
+                lineage data and selects
 

### 2018-10-01 paper.tex
some text on portability

@@
                 HPC, with many processed trying to access the memory. The 
-                following are two optimizations made for
+                following are two optimizations made for  \note{two or one}
 
@@
                 from its starting point. For this, Alluxio evaluates the 
-                lineage data and selects
+                lineage data and selects the
 
@@
                 optimize the data's placement across the cluster for most 
-                efficient use.
+                efficient use. It has two modes, default and Lustre integrated,
+                that allows it to execute on systems with and without access
+                to a distributed filesystem.
 
@@
     \chapter{Portability}\label{portability}
+        In order for a workflow engine to be utilized, it much be simple for 
+        users to install and execute the data. Platform-independence of both 
+        the frameworks and tools, as well as, simple installation of all 
+        necessary components and straightforward execution of workflows  are a 
+        requirement for frameworks and their respective pipelines to be 
+        portable. Without portability, users are limited to executing certain
+        frameworks only or dealing with the significant overhead of making 
+        these engines work on the system available to them. In this chapter, 
+        we evaluate the portability of workflows on the basis of the following
+        three components: 1) platform-independence, 2) tool installation and
+        3) execution setup. We also examine containerization, one of the 
+        strategies used to improve portability.
+
+        \section{Platform-independence}
+            Achieving platform-independence allows users to use the workflow 
+            engine in any available environment. This however does not imply 
+            that the tools, and as a consequence, the pipelines created will
+            also be platform-independent. Many workflow engines have opted to 
+            be platform independent. These workflow engines include: Nipype,
+            PSOM, SPM, FASTR, LONI, AA, Pydpiper, Taverna, MapReduce, Spark and
+            Dask. These workflow engines achieve platform independence through
+            running on the Java Virtual Machine (JVM) or being written in 
+            Python. Platform-independent engines that are Java-based include: 
+            PSOM, SPM, LONI, AA, Taverna, MapReduce and Spark. The remainder of
+            the pipelines are implemented in Python. The remainder of the 
+            workflow engines (Pegasus, Make, OpenMOLE, Nextflow, Galaxy, and 
+            Toil are not platform-independent, but can execute on UNIX-like 
+            systems.
+
+            Engines built ontop of a platform-independent language does not 
+            necessarily imply platform independence. For instance, OpenMOLE is 
+            built in Scala and Netflow runs on Groovy, yet neither are platform
+            independent. In the case of OpenMOLE, this is due to its CARETask
+            blackbox, which is only possible for Linux environments. However,
+            many tools and programs available for incorporation into workflows
+            are platform-dependent. For example, even though Nipype is 
+            platform-independent, common tools used in Nipype modules, such as 
+            FreeSurfer and FSL, are Linux-based. Moreover, HPC infrastructure 
+            is usually Linux-based and it is therefore crucial that all 
+            workflows are compatible with and tested on Linux systems.
+
+       \section{Tool installation}
+
+            Unless a user is building a workflow exclusively based on custom
+            functions or using a monolithic workflow engine such as SPM, they
+            will need to install the tools necessary for their workflow on
+            their computer or cluster. Depending on the workflow engine 
+            selected, this may be managed by the engine, or the user may be 
+            left to install the dependencies themselves. Workflow engines that
+            delegate tool installation to the user include: Nipype, PSOM,
+
+        Many of these platforms have selected to be written in 
+        platform-independent languages. However,  Due to the modularity of these
+        frameworks, it may be necessary
         java or python backend (most) typically platform independent (some 

### 2018-10-02 paper.tex
Comments on chap 2 and 3

@@
         performance has become especially critical now that BigData has been
-        introduced into the mix. The dichotomy remains between scientific 
+        introduced \weird{into the mix}. The dichotomy remains between scientific 
         workflows and BigData workflows. While scientific workflows have made
@@
         (scientific and BigData) are tightly coupled to the infrastructure
-        they were intended to be used on. That is, as scientific workflow 
+        they were intended to be used on. That is, as scientific workflows \tristan{use 'engine' and 'workflows' consistently (think nipype vs fmriprep).}
         try to remain efficient in their handling of compute resources, as such
@@
         workflows to BigData frameworks on the basis of performance. 
-        Section~\ref{fs} will look at what filesystems are required by the 
+        Section~\ref{fs} will look at what filesystems are \weird{required} \tristan{supported?} by the 
         engines. This is important because data movement of any kind is costly,
-        and may be worse depending on the filesystems available. 
+        and may be worse depending on the filesystems available. \tristan{Also, mention that neuroinformatics
+        pipelines manipulate files.}
         Section~\ref{sched} will provide an overview of the schedulers 
@@
             operations could incur significant performance penalties. 
-            Filesystems. To mitigate this and remain efficient in the face of
+            \weird{Filesystems}. To mitigate this and remain efficient in the face of
             BigData, Dask implements lazy evaluation and in-memory processing.
@@
                     memory, Spark avoids data transfers as often as possible by
-                    scheduling tasks to nodes where the memory is located. Dask
+                    scheduling tasks to nodes where the memory is located \tristan{You mean where RAM is available or where the data is located?}. Dask
                     , however, will schedule its tasks to the next available 
@@
                     RDD, the scheduler will be able to assign task to the most
-                    favorable partition dictated by delay scheduling.
-
-                \subsubsection{lazy evaluation}
+                    favorable partition dictated by delay scheduling \tristan{ignoring data locality?}.
+
+                \subsubsection{Lazy evaluation}
 
@@
                     constructing graphs that result in the most optimal
-                    task order of execution.
+                    task order of execution. \tristan{And further save on data transfers}
 
@@
             big data: data locality, in memory computing, lazy evaluation
+
+            \tristan{Mentioning fault tolerance would be wise too.}
+
     \chapter{Portability}\label{portability}
-        In order for a workflow engine to be utilized, it much be simple for 
-        users to install and execute the data. Platform-independence of both 
+        In order for a workflow engine to be utilized, \weird{it much be simple} for 
+        users to \weird{install and execute the data}. Platform-independence of both 
         the frameworks and tools, as well as, simple installation of all 
@@
             Python. Platform-independent engines that are Java-based include: 
-            PSOM, SPM, LONI, AA, Taverna, MapReduce and Spark. The remainder of
+            PSOM, SPM, LONI, AA, Taverna, MapReduce and Spark \tristan{this list contains Matlab
+            engines}. The remainder of
             the pipelines are implemented in Python. The remainder of the 
@@
             Toil are not platform-independent, but can execute on UNIX-like 
-            systems.
-
-            Engines built ontop of a platform-independent language does not 
-            necessarily imply platform independence. For instance, OpenMOLE is 
+            systems \tristan{macOS and Linux}.
+
+            Engines built on top of a platform-independent language does not 
+            necessarily imply platform independence \tristan{Grammar? Engines ... does not}. For instance, OpenMOLE is 
             built in Scala and Netflow runs on Groovy, yet neither are platform
@@
             will need to install the tools necessary for their workflow on
-            their computer or cluster. Depending on the workflow engine 
+            their computer or cluster \tristan{replace computer or cluster by infrastructure?}. Depending on the workflow engine 
             selected, this may be managed by the engine, or the user may be 

### 2018-10-02 paper.tex
More comments
some changes

--- 2018-10-01 paper.tex
@@
 
-            Unless a user is building a workflow exclusively based on custom
-            functions or using a monolithic workflow engine such as SPM, they
+            Unless building a workflow exclusively based on custom
+            functions or using a monolithic workflow engine such as SPM, users
             will need to install the tools necessary for their workflow on
             their computer or cluster. Depending on the workflow engine 
-            selected, this may be managed by the engine, or the user may be 
-            left to install the dependencies themselves. Workflow engines that
-            delegate tool installation to the user include: Nipype, PSOM,
+            selected, this may be managed by the engine or performed manually 
+            by the user. Most install the dependencies themselves. Workflow engines that
+            delegate tool installation to the user include: Nipype, PSOMi,
 

### 2018-10-02 paper.tex

--- 2018-10-02 paper.tex
@@
             \tristan{Mentioning fault tolerance would be wise too.}
+            \tristan{How about: burst buffers, pilot jobs, deployments of Spark on HPC?}
 
@@
 
-            Unless a user is building a workflow exclusively based on custom
-            functions or using a monolithic workflow engine such as SPM, they
+            Unless building a workflow exclusively based on custom
+            functions or using a monolithic workflow engine such as SPM, users
             will need to install the tools necessary for their workflow on
-            their computer or cluster \tristan{replace computer or cluster by infrastructure?}. Depending on the workflow engine 
-            selected, this may be managed by the engine, or the user may be 
-            left to install the dependencies themselves. Workflow engines that
+            their infrastructure. Depending on the workflow engine 
+            selected, this may be managed by the engine or performed manually 
+            by the user. Most install the dependencies themselves. Workflow engines that
             delegate tool installation to the user include: Nipype, PSOM,

### 2018-10-02 paper.tex
reproducibility intro

@@
         Neuroimaging frameworks are notoriously known to be compute-intensive.
-        The tools and their respective workflows needed to incorporate 
+        The workflows and their respective tools needed to incorporate 
         parallelism in order to save on computing time. Maintaining good 
-        performance has become especially critical now that BigData has been
-        introduced \weird{into the mix}. The dichotomy remains between scientific 
+        performance has become especially critical now that neuroimaging
+        BigData has emerged. The dichotomy remains between scientific 
         workflows and BigData workflows. While scientific workflows have made
@@
         (scientific and BigData) are tightly coupled to the infrastructure
-        they were intended to be used on. That is, as scientific workflows \tristan{use 'engine' and 'workflows' consistently (think nipype vs fmriprep).}
-        try to remain efficient in their handling of compute resources, as such
+        they were intended to be used on. Scientific workflow engines
+        try to remain efficient in their handling of compute resources since such
         resources were previously scarce. In contrast, BigData frameworks 
@@
                     results. Should data not fit in memory, the results may 
-                    spill to local disk, in the instance where nodes contain
-                    storage. Otherwise, the engines will spill data to shared 
+                    spill to local disk, if local storage is available. 
+                    Otherwise, the engines will spill data to a shared 
                     filesystem. The later is a common case in HPC systems as 
@@
                     memory, Spark avoids data transfers as often as possible by
-                    scheduling tasks to nodes where the memory is located \tristan{You mean where RAM is available or where the data is located?}. Dask
+                    scheduling tasks to nodes where the data is located. Dask
                     , however, will schedule its tasks to the next available 
@@
                     RDD, the scheduler will be able to assign task to the most
-                    favorable partition dictated by delay scheduling \tristan{ignoring data locality?}.
+                    favorable partition dictated by delay scheduling according
+                    to data locality.
 
@@
     \chapter{Reproducibility}\label{reproducibility}
-        \section{Provenance Tracking}
-        big data frameworks: limited to lineage
-        \section{containerization}
+        Reproducibility is a vital component to all scientific workflows. 
+        Without reproducibility, correctness of results could never be 
+        determined. Reproducibility is defined as the ability to obtain similar 
+        results by 
+        recreating or executing the original workflow on the same or different
+        infrastructure\note{cite}. It have been found that more than 70\% of
+        scientific experiments are not reproducible \note{cite Mbaker}
+
+        Current studies show that different tools 
+        performing the same function, in addition to the same versions of tools 
+        executing on top of different operating systems produce significantly 
+        dissimilar results. As a result, it is necessary that specific details 
+        on the workflow, such a data used, software versions and operating 
+        system used are provided. For this reason, many workflow engines have
+        incorporated provenance tracking into their design.
+
+        Getting specific versions of tools and their workflows is not always 
+        a straightforward process. To facilitate this, many workflow engines
+        have begun to use containers to ensure that users can reproduce the 
+        experiment within a contained environment identical to the one it that
+        was originally experimented on.
+
+        This chapter will examine how different workflow engines tackle the 
+        issue of reproducibility. Section~\ref{prov} will discuss how the 
+        engines assimilate and distribute the provenance data. 
+        Section~\ref{repcont} will examine how these engines use containers to
+        facilitate reproducibility of the experiments.
+        \section{Provenance Tracking}\label{prov}
+
+        \section{Containerization}\label{repcont}
     \chapter{Discussion/Conclusion}

### 2018-10-03 paper.tex
intro to reproducibility

@@
         \section{Provenance Tracking}\label{prov}
+            Provenance tracking is of vital importance to scientific reporting.
+            On its own, source code is insuffient to reproduce the workflow 
+            described in a research paper. This is because source code does 
+            not specify the specific dataset used nor the specific versions of
+            tools and the environments that they were executed on. Minor 
+            changes to any of these features can significantly affect results.
+
+            As explained by the W3 PROW specification, provenance details 
+            information on the data, the tools involved in its creation and the
+            responsible parties. Provenance tracking, in the context of 
+            workflows, is the the act of capturing provenance data during
+            workflow execution. With detailed provenance and access to source 
+            codeinfo, researchers should have enough information to be able 
+            to reproduce workflows.
+
+            To to the importance of provenance tracking for scientific 
+            workflows several workflow engines adapted strategies to capture 
+            comprehensive provenance data. These include Nipype, PSOM, FastR, 
+            LONI, Taverna, Pegasus, Nextflow, Galaxy. Other workflow engines,
+            such as SPM, AA, Pydpiper,
 

### 2018-10-03 paper.tex
Comments on reproducibility intro

--- 2018-10-02 paper.tex
@@
     \chapter{Reproducibility}\label{reproducibility}
-        Reproducibility is a vital component to all scientific workflows. 
+        Reproducibility is a vital \tristan{critical?} component to all scientific workflows. 
         Without reproducibility, correctness of results could never be 
@@
         recreating or executing the original workflow on the same or different
-        infrastructure\note{cite}. It have been found that more than 70\% of
+        infrastructure\note{cite Peng 2011}. It has been found that more than 70\% of
         scientific experiments are not reproducible \note{cite Mbaker}
@@
         performing the same function, in addition to the same versions of tools 
-        executing on top of different operating systems produce significantly 
-        dissimilar results. As a result, it is necessary that specific details 
+        executing on top of different operating systems, produce significantly 
+        dissimilar results \cite{bowring, gronenschild 2011, glatard 2015}. As a result, it is necessary that specific details 
         on the workflow, such a data used, software versions and operating 
@@
         \section{Provenance Tracking}\label{prov}
+            Provenance tracking is \weird{of vital importance} \tristan{twice} to scientific reporting.
+            On its own, source code is insufficient to reproduce the workflow 
+            described in a research paper. This is because source code does 
+            not specify the specific dataset used nor the specific versions of
+            tools and the environments that they were executed on. Minor 
+            changes to any of these features can significantly affect results.
+
+            As explained by the W3 PROV specification, provenance details 
+            information on the data, the tools involved in its creation, and the
+            responsible parties. Provenance tracking, in the context of 
+            workflows, is the act of capturing provenance data during
+            workflow execution. With detailed provenance and access to source 
+            code information, researchers should have enough information to be able 
+            to reproduce workflows.
+
+            To to the importance of provenance tracking for scientific 
+            workflows several workflow engines adapted strategies to capture 
+            comprehensive provenance data. These include Nipype, PSOM, FastR, 
+            LONI, Taverna, Pegasus, Nextflow, Galaxy. Other workflow engines,
+            such as SPM, AA, Pydpiper,
 

### 2018-10-03 paper.tex
some restructuring
completed reproducibility

@@
     \chapter{Reproducibility}\label{reproducibility}
-        Reproducibility is a vital \tristan{critical?} component to all scientific workflows. 
+        Reproducibility is a critical component to all scientific workflows. 
         Without reproducibility, correctness of results could never be 
@@
         This chapter will examine how different workflow engines tackle the 
-        issue of reproducibility. Section~\ref{prov} will discuss how the 
+        issue of reproducibility. Section~\ref{metadata} will investigate three 
+        different provenance metadata specifications available (W3C PROV, NIDM,
+        and BIDS Derivates).Section~\ref{prov} will discuss how the 
         engines assimilate and distribute the provenance data. 
@@
         facilitate reproducibility of the experiments.
+        \section{Metadata formats}\label{metadata}
+            The recent increase in data-sharing resulted in the rise of BigData 
+            in neuroscience. To continue the promotion of data sharing in 
+            neuroscience, the derived data's metadata must be well understood
+            by the community. Even if the provenance tracking is comprehensive,
+            it is useless if it is only understood by some. To address this 
+            issue, specifications for the standardization of metadata have been 
+            proposed. Here we look at three provenance metadata specifications.
+            The first is W3 PROV, a general-purpose provenance specification; 
+            NIDM-results
+            a comprehensive neuroimaging provenance specification; and BIDS 
+            Derivatives the recent provenance addition to the family of 
+            popular neuroimaging metadata specifications, BIDS.
+
+            \subsection{W3C PROV}
+            The W3C PROV specification came out from the need to explain the 
+            origin of data shared on the internet. 
+            The aim of the W3 PROV specification was gather information on
+            the activities, entities and involved parties in the creation of 
+            the data. This metadata would allow for assements of the quality, 
+            reliability and trustworthiness to be performed. Sharing of the 
+            provenance is faciliated through the use of widely-used formats 
+            such as RDF and XML.
+
+            The PROV model has three main objects involved: 1) entities, 2) 
+            activities and 3) agents. Any kind of thing is considered an entity
+            . Actions act upon entities to generate new ones. Activities have 
+            agents associated to them. These agents are responsible for the 
+            activity taking place. Agents can be individuals, companies and 
+            software, amongst others.
+
+            \subsection{NIDM-Results}
+                NIDM-results was designed to improve the limited XML-based 
+                Clinical and Experimental Data Exchange (XCEDE) neuroimaging
+                provenance metadata format. It is a domain-specific extension
+                to the W3 PROV model that aims to provide comprehensive and
+                machine-readable provenance from raw data to its final stage.
+                However, it is currently limited in the provenance 
+                information it can provide.
+
+            \subsection{BIDS Derivatives}
+                The BIDS-Derivatives metadata specification is the newest 
+                addition to the BIDS family of neuroimaging-related 
+                specifications. It is the sister
+                specification to the widely adopted BIDSRaw metadata format 
+                and it remains compatible with the BIDS Raw specification. 
+                Unlike NIDM-Results, BIDS derivatives, like BIDS Raw, do not
+                use the W3C PROV specification. Rather, the provenance can be
+                derived from its folder structure and filename. Any additional
+                metadata can be found in a JSON sidecar file.
+
         \section{Provenance Tracking}\label{prov}
-            Provenance tracking is \weird{of vital importance} \tristan{twice} to scientific reporting.
+            Provenance tracking is crucial to scientific reporting.
             On its own, source code is insufficient to reproduce the workflow 
@@
             As explained by the W3 PROV specification, provenance details 
-            information on the data, the tools involved in its creation, and the
-            responsible parties. Provenance tracking, in the context of 
+            information on the data, the tools involved in its creation, and 
+            the responsible parties. Provenance tracking, in the context of 
             workflows, is the act of capturing provenance data during
             workflow execution. With detailed provenance and access to source 
-            code information, researchers should have enough information to be able 
-            to reproduce workflows.
-
-            To to the importance of provenance tracking for scientific 
+            code information, researchers should have enough information to be 
+            able to reproduce workflows.
+
+            To address the importance of provenance tracking for scientific 
             workflows several workflow engines adapted strategies to capture 
-            comprehensive provenance data. These include Nipype, PSOM, FastR, 
-            LONI, Taverna, Pegasus, Nextflow, Galaxy. Other workflow engines,
-            such as SPM, AA, Pydpiper,
+            provenance metadata. Some leverage the W3C PROV specification (e.g.
+            Nipype, FastR, Taverna), while others may chose to use NIDM (e.g. 
+            SPM). The 
+            remainder use a custom format to share their metadata.
+
+            Although Nipype, FastR and Taverna all capture provenance using 
+            the W3C Prov specification, they differ slightly in how they
+            present provenance
+            information and what they capture.
+            Currently, Nipype's support for W3C PROV is experimental. However,
+            it is capable of outputting provenance using the W3C PROV data 
+            model for each workflow interface, as well as the workflow as a 
+            whole.
+
+            FastR performs provenance very similarly to Nipype. However,
+            FastR's agents can be Fastr, Networks, Tools, Nodes. Additionally,
+            activities, in
+            FastR map to Jobs and entities map to the data.
+
+            Taverna differs from Nipype and Taverna in that it provides 
+            tracking of
+            both workflow runs and of the derived data.
+
+            Likewise to the engines that capture provenance that follows the
+            W3C PROV specification, all MATLAB-based workflow engines capture
+            provenance data similarly.
+            By default, SPM saves an execution by writing the  workflow, with
+            all its input parameters as a MATLAB file. Saving the workflow as
+            a MATLAB file allows for distribution of the workflow and 
+            reexecution. To export to NIDM-Results, the option must be 
+            sslected.
+
+            As in SPM's MATLAB workflow execution files, PSOM and AA 
+            also save
+            their workflow to a MATLAB file for future reexecution. 
+            In addition, PSOM generates
+            also files containing additional provenance information such 
+            as username, date, time and operating system used.
+
+            While all workflow engines capture generally the same type of 
+            provenance information, they differ in how much provenance 
+            information is captured and how it's done. LONI, Pegasus, Pydpiper 
+            and Galaxy
+            capture an above minimum amount of provenance data, whereas engines 
+            like Nextflow, Make, Toil and MapReduce capture very limited  
+            provenance 
+            information. OpenMOLE handles provenance through encapsulation of
+            tasks using CARE to simulate runtime environments
+            and BigData frameworks like Spark and Dask mainly rely on 
+            lineage.
+
+            An above minimum amount of provenance details captured includes 
+            capturing details on the workflow stages as well as the inputs and
+            outputs. LONI stores this type of information in an XML file.
+
+            Pegasus uses an SQL database to store its provenace. Data collected 
+            includes runtime information,
+            in addition to, data locations, software used and parameters 
+            applied to the software.
+
+            Pydpiper has no formal form of provenance tracking, but will save 
+            essential pipeline information after each completed stage to a 
+            serialized Python object. Data captured includes: the DAG, the 
+            array of pipeline stages, the current stage counter, hash of stages 
+            and output files and an array of stage statuses.
+
+            As provenance is a major component of the Galaxy project, it 
+            provides comprehensive provenance tracking. This includes gathering
+            of metadata similarly to other workflow engines, but also allows
+            users to add annotations to make the intent of their analysis known
+            .
+
+            The workflows that capture limited provenance information typically
+            restrict themselves to capturing only runtime information.
+            Nextflow generates an
+            HTML report that contains details on execution commands, runtime, 
+            CPU usage and status.
+
+            Toil captures at most tool execution logs and provides runtime 
+            statistics.
+
+            Make is another workflow engine that does not provide built-in 
+            provenance tracking. However, Make can be adapted to capture 
+            provenance as in \note{cite make paper}.
+
+            Similarly, MapReduce also captures runtime information and relays 
+            it through a web UI.
+
+            OpenMOLE tackles provanence differently by wrapping workflow tools
+            into CARETasks. CARETasks simulate the runtime environment of the 
+            execution, such that workflows can be executed on any Linux 
+            distribution avoiding any silent errors that may arise from 
+            executing a workflow in a different environment. An issue with 
+            restricting provenance to CARETasks, however, is that provenance is
+            a bit more than just simulating an environment. Provenance 
+            information should also be accessible. Understanding of CARE should
+            not be a prerequisite to obtaining provenance information.
+
+            Dask and Spark rely on their lineage to relay details on which 
+            tasks modified the data. They also provide a web UI that provides
+            additional information on runtime statistics. Methods to improve 
+            provenance tracking in Spark, particularly interactively are 
+            currently being explored \note{cite bigdebug}. Unfortunately, both
+            of them lack the ability to provide provenance capture of external
+            tools at present.
 
         \section{Containerization}\label{repcont}
+
+            To handle issues with numerical instability of tools and relieve 
+            users from
+            having to configure their environment to be almost identical to 
+            that of the study they are trying to replicate, significant 
+            emphasis has been placed on containerization. Similarly to 
+            OpenMOLE's CARETasks, containerization provides an encapsulated
+            environment for the tools to run which mimics the environment of 
+            the original study. This ensures that users executing their 
+            workflows in the same environments will obtain the same results.
+            A major issue with containerization is that despite it controlling
+            environments and permitting experiments to be reproducible, it does
+            not solve the issue of numerical instability. Moreover, this 
+            mentality may lead researchers to limit themselves to older tool 
+            versions and further limit scientific advancement.
+
     \chapter{Discussion/Conclusion}

### 2018-10-04 paper.tex
Comments on reproducibility section

@@
         Without reproducibility, correctness of results could never be 
-        determined. Reproducibility is defined as the ability to obtain similar 
+        determined \tristan{Link between two first sentences could be smoothened}. Reproducibility is defined as the ability to obtain similar 
         results by 
@@
         system used are provided. For this reason, many workflow engines have
-        incorporated provenance tracking into their design.
+        incorporated provenance tracking into their design. \tristan{Provenance is also
+        a way to merely document what happened so that results can be traced.}
 
@@
         issue of reproducibility. Section~\ref{metadata} will investigate three 
-        different provenance metadata specifications available (W3C PROV, NIDM,
-        and BIDS Derivates).Section~\ref{prov} will discuss how the 
+        different provenance specifications used in neuroinformatics (W3C PROV, NIDM,
+        and BIDS Derivates) \tristan{I don't think you want to talk about 'metadata'. Just
+        mention 'provenance'.}. Section~\ref{prov} will discuss how the 
         engines assimilate and distribute the provenance data. 
@@
         facilitate reproducibility of the experiments.
-        \section{Metadata formats}\label{metadata}
+        \section{Provenance Formats}\label{metadata}
             The recent increase in data-sharing resulted in the rise of BigData 
-            in neuroscience. To continue the promotion of data sharing in 
+            in neuroscience \tristan{This feels like general intro to me}. To continue the promotion of data sharing in 
             neuroscience, the derived data's metadata must be well understood
@@
             Derivatives the recent provenance addition to the family of 
-            popular neuroimaging metadata specifications, BIDS.
+            popular neuroimaging metadata specifications, BIDS \tristan{Make sure that
+            the acronym is defined earlier.}.
 
@@
             The W3C PROV specification came out from the need to explain the 
-            origin of data shared on the internet. 
-            The aim of the W3 PROV specification was gather information on
+            origin of data shared on the Web. 
+            The aim of the W3 PROV specification was \tristan{to?} gather information on
             the activities, entities and involved parties in the creation of 
-            the data. This metadata would allow for assements of the quality, 
+            data. This metadata would allow for assessments of the quality, 
             reliability and trustworthiness to be performed. Sharing of the 
@@
             The PROV model has three main objects involved: 1) entities, 2) 
-            activities and 3) agents. Any kind of thing is considered an entity
-            . Actions act upon entities to generate new ones. Activities have 
+            activities and 3) agents. Any kind of thing is considered an entity.
+            Actions act upon entities to generate new ones. Activities have 
             agents associated to them. These agents are responsible for the 
@@
                 Clinical and Experimental Data Exchange (XCEDE) neuroimaging
-                provenance metadata format. It is a domain-specific extension
+                provenance metadata format \tristan{I guess you'll cite references for that}. It is a domain-specific extension
                 to the W3 PROV model that aims to provide comprehensive and
-                machine-readable provenance from raw data to its final stage.
+                machine-readable provenance from raw data to its final stage. \tristan{give example?}
                 However, it is currently limited in the provenance 
-                information it can provide.
+                information it can provide. \tristan{Be more specific, what are the limitations?}
 
@@
                 derived from its folder structure and filename. Any additional
-                metadata can be found in a JSON sidecar file.
+                metadata can be found in a JSON sidecar file. \tristan{Explain that
+                although BIDS doesn't rely on semantic web technologies, this is the
+                solution being adopted in the community, mostly for simplicity. Mention
+                that there is a community effort around it and that bidsraw has been
+                very successful.}
+
+            \tristan{You might want to add data lineage here.}
 
@@
             code information, researchers should have enough information to be 
-            able to reproduce workflows.
+            able to reproduce workflows. \tristan{These 2 paragraphs should be in the chapter intro.}
 
@@
             activities, in
-            FastR map to Jobs and entities map to the data.
+            FastR map to Jobs and entities map to the data \tristan{how is the mapping in nipype? If different,
+            it might be a good place to comment on the fact that PROV can be used
+            in many different ways}.
 
@@
             and BigData frameworks like Spark and Dask mainly rely on 
-            lineage.
-
+            lineage \tristan{which is?}.
+
+            \tristan{The rest of this section is very difficult to read because
+            it lacks structure. Perhaps you could classify provenance tracking in PROV-based, lineage-based,
+            and other custom formats?}
             An above minimum amount of provenance details captured includes 
@@
             versions and further limit scientific advancement.
+            \tristan{Nextflow should be mentioned here, as they make a big deal
+            of using containers for reproducibility. I would structure this section in 
+            2 paragraphs: (1) what engines support containers, (2) why containers help
+            reproducibility and why there aren't enough.}
 

### 2018-10-04 paper.tex
added tool installation

--- 2018-10-03 paper.tex
@@
             selected, this may be managed by the engine or performed manually 
-            by the user. Most install the dependencies themselves. Workflow engines that
-            delegate tool installation to the user include: Nipype, PSOM,
-
-        Many of these platforms have selected to be written in 
-        platform-independent languages. However,  Due to the modularity of these
-        frameworks, it may be necessary
-        java or python backend (most) typically platform independent (some 
-        exceptions)
-        all tools used in the pipeline must be installed on system
-        Taverna - exception as relies on web services
+            by the user. Although not a direct requirement of workflows engines, 
+            engines that simplify tool installation promote sharing of 
+            workflows. To simplify this process, many workflow engines rely on 
+            importing and running containerized workflow tools. These engines
+            include: Galaxy, Toil, Netflow, Pegasus and FastR. While most 
+            provide support for Docker and Singularity, Toil and Galaxy only 
+            support Docker containers. All that is required by these workflow 
+            engines is the URL to the image in DockerHub or SingularityHub and
+            the engine will abstract the process of acquiring the container and
+            running the tool.
+
+            Alternatively, engines such as OpenMOLE use a different tool to
+            sandbox the environment, known as CARE. CARE allows 
+            users to make a compressed file wrapping the tool in the 
+            environment it was meant to be executed on. It uses environment 
+            information to simulate the system calls of the environment when
+            interacting with the tool. CARE was suggested as an alternative to
+            Docker as Docker requires a Docker engine to be running on each 
+            node, which may not always be the case. Moreover, Docker engines 
+            cannot be launched on-the-fly as they require administrative 
+            privileges. However, should a user wish to run a Docker container,
+            OpenMOLE provides a mechanism to run DockerHub tools in 
+            unprivileged using udocker.
+
+            Other workflow engines do not use an form of containerization to
+            ensure that workflows can be shared easily amongst users. These 
+            include Taverna, Make, Pydpiper, AA, SPM, PSOM, Nipype, LONI, 
+            MapReduce, Spark and Dask.
+            Taverna is different from the other workflow engines, however, as 
+            it is webservice-based an thus it expects that a server is up, 
+            running the desired version of the tool.
+
+            As for the engines with no tool dependency setup, it is still 
+            possible for the user to manually reference a container within 
+            a UDF. Conversly, libraries that manage execution of containerized 
+            tools, such as Boutiques, can be intergrated into these engines.
+
         \section{Containers}

### 2018-10-05 paper.tex
more on portability

@@
 
-       \section{Tool installation}
+       \section{Containers}
+
+            Containers are lightweight solutions to isolating 
+            environments. Although similar, containers and virtual machines 
+            differ from each 
+            other in that VMs virtualize the
+            hardware by running an entirely different operating system, whereas
+            containers virtualize the software by running on top of the host's
+            kernel.
+
+            Containerization greatly increases the portability of workflows. 
+            They allows users to package their workflows and/or tools with all 
+            their dependencies in a package that users can download and 
+            execute. Moreover, these packages can be versioned permitting users 
+            to select the preferred version of the container. Although there 
+            exist many containerization technologies, the two most notable ones
+            in scientific computing are Docker and Singularity. Docker and 
+            Singularity are both platform independent meaning users can run
+            containerizaed versions of their workflows ontop of an otherwise
+            incompatible OS.
+
+            Singularity is the current preferred solution for running 
+            containers on HPC systems. It is an open-source project designed to
+            meet the security, portability and reproducibility requirements of 
+            the scientific community. The main difference between Docker and 
+            Singularity is that users gain root permissions in a Docker 
+            container, and thus obtain root access to the underlying host OS. 
+            This is a serious security concern in HPC systems where many users
+            share the same resources and therefore are given restricted access
+            only. Singularity addresses this concern by disabling the user from
+            acquiring elevated permissions within the container.
+
+            Containers are widely used to isolate tool executions and eliminate
+            the need to manually install dependencies in scientific workflows.
+            Pipeline tools can be containerized. In neuroinformatics, 
+            NeuroDocker, a command-line tool for creating Docker and 
+            Singularity containers with specific versions of neuroimaging tools
+            installed. Examples of 
+            containerized applications include all of the BIDS Apps. 
+            Additionally, entire workflows can also be containerized, as can be
+            observed in fMRIPrep, a Nipype-based workflow for fMRI 
+            Preprocessing.
+
+       \section{Tool sharing}\label{sharing}
 
@@
 
-        \section{Containers}
-
-             Containeri
-            use of containers increase portability. none of the scientific 
-            pipelines can be containerized and communicate with the scheduler 
-            outside of the container
-            big data frameworks can
     \chapter{Reproducibility}\label{reproducibility}

### 2018-10-05 paper.tex
more on scheduling

@@
             their own form of scheduling for computations on a single node. 
-            Nipype uses mutliproc... etc. BigData frameworks, such as Dask,
+            BigData frameworks, such as Dask,
             also provide interfaces to these HPC schedulers. However, other
@@
 
-            Spark's standalone scheduler is a simple scheduler that can be used
-            in distributed workflows. It can execute in two modes: client and
-            cluster. In client mode, the  \note{cannot run in cluster mode in pyspark}
-
-            YARN needs hadoop cluster
-
-            Mesos can run on HPC
-
-            dask distributed
-            \note{scientific: SLURM, PBS, SGE, HTCondor, Mesos (sometimes)
-            big data: mapreduce - local or YARN, Spark - standalone, YARN or 
-            Mesos, dask: any (easy to extend to any scheduler)}
+            Spark's standalone scheduler is a simple scheduler packaged with 
+            Spark that can be used in distributed workflows. It can be used
+            for single-node and cluster deployments. For cluster deployments,
+            Spark has two execution modes: client and cluster. For client 
+            execution, the application driver is intended to be executed on the
+            client node. In cluster mode, however, the driver is intended to be
+            running on a compute node. Cluster mode can ensure fault-tolerance 
+            of not only the workers and master, but also the driver. However,
+            cluster mode is not available to PySpark deployments using the 
+            standalone scheduler.
+
+            YARN is the main scheduler used for MapReduce applications. It was 
+            designed in order to decouple resource management from the 
+            MapReduce programming model while also addressing issues created 
+            by ad-hoc clusters, such as Hadoop on Demand (HoD), and helped 
+            users transition to shared Hadoop clusters. YARN is not restricted 
+            to MapReduce applications, and can also be used as a scheduler for 
+            Spark and Dask applications. Unlike Spark's standalone scheduler, 
+            YARN cluster deployment mode can be used with PySpark.
+
+            Mesos is another scheduler compatible with Hadoop and Spark 
+            applications. It differs with YARN in that it is designed for the
+            efficient resource sharing of many cluster computing frameworks,
+            not just those that belong to the Hadoop family. Cluster deployment
+            mode is also available to PySpark applications using Mesos.
+
+            Dask provides support for two built-in schedulers: single machine
+            and distributed. The single machine scheduler is Dask's default
+            scheduler. It allows users to parallelize their application on a
+            single machine at either the thread or process level. For any 
+            application that needs to scale to multiple nodes, the Dask 
+            distributed scheduler is recommended. Dask distributed can also be
+            run on a single machine as is useful to use in applications with 
+            significant data transfers as it employs more efficient 
+            locality-aware scheduling strategies.
+
+            Unlike Spark, Dask can interface with HPC schedulers. The 
+            dask-jobqueue interface is compatible with PBS, SLURM and SGE 
+            resource managers, while the dask-drmaa interface is compatible 
+            with any DRMAA compliant resource manager. It is recommended to use
+            these interfaces for interactive scheduling, but they can also be
+            used for batch processing.
+
+            Since Spark does not natively interface with HPC schedulers, 
+            the use of overlay Spark clusters, similar to HoD, has been 
+            investigated for Spark applications running on HPC. The reasons 
+            listed for HoD's failure include poor 
+            data locality when deploying a dynamic cluster on a shared HDFS 
+            instance, inability to resize cluster resulting in idle nodes and 
+            high cluster allocation latency. To address the issue of high 
+            cluster allocation, pilot-agent scheduling strategies can be used.
+            Pilot-agent scheduling strategies allow for clusters to grow 
+            dynamically and to add nodes when required. This relieves users, 
+            in the case of Spark, from having to wait until all nodes become 
+            available to begin execution. Both papers \note{jha and baer} use 
+            this strategy. Furthermore, pilot-agent scheduling could be adapted
+            to enable smart cluster resizing. Conversly, maintaining data 
+            locality using a shared HDFS instance will remain an issue.
+
         \section{Performance enhancement strategies}\label{other}
             scientific: workflow checkpointing (looking at file hashes)
-            big data: data locality, in memory computing, lazy evaluation
 

### 2018-10-07 paper.tex
Comment on scheduling
some more text

@@
         Nipype, Pipeline System for Octave and Matlab (PSOM), LONI, SPM, FastR,
-        Automated Analysis (AA) and Pydpiper\note{add citations here}. These 
+        Automatic Analysis (AA) and Pydpiper\note{add citations here}. These 
         workflow engines aim to satisfy four criteria: 1) straightforward \tristan{easy? facilitated?} 
@@
 
-        \section{Workflow sharing}\label{sharing}
-
     \chapter{Performance}\label{performance}
@@
                 Writting BigData to Lustre is very costly compared to local 
-                node storage. BigData filesystems like HDFS may not necessarily
-                be the best solution for HPC as they will not only use up 
-                storage on compute nodes, but also replicate the data across 
-                the cluster taking up a significant amount of storage space.
+                node storage. Distributed filesystems found on HPC 
+                infrastructure provide their own form of fault-tolerance, 
+                therefore, adding a BigData filesystem such as HDFS which 
+                replicates the data may burden the storage layer only to 
+                providea feature which already exist. This is especially 
+                problematic as storage on HPC compute nodes may be limited.
                 Moreover, while BigData frameworks typically execute on a 
@@
                 HPC, with many processed trying to access the memory. The 
-                following are two optimizations made for  \note{two or one}
+                following are two optimizations made for the storing of BigData
+                .
 
@@
                 from its starting point. For this, Alluxio evaluates the 
-                lineage data and selects the
+                lineage data and checkpoints the files which have been 
+                repeatedly accessed (i.e. hot files). This strategy also helps 
+                in avoiding the checkpointing of intermediary files.
 
@@
                 optimize the data's placement across the cluster for most 
-                efficient use. It has two modes, default and Lustre integrated,
-                that allows it to execute on systems with and without access
-                to a distributed filesystem.
-
-            \note{scientific: shared FS such as Lustre, NFS, S3
-            big data: mapreduce - HDFS, spark and dask shared or distributed}
+                efficient use. It aims to reduce the I/O bottleneck of writing 
+                through data buffering and caching. Triple-H has two modes of
+                operation, Default and Lustre integrated,
+                that allows it to execute on systems both with and without 
+                access to a distributed filesystem. To further reduce I/O in
+                HPC systems, Triple-H leverage's Lustre's fault tolerance 
+                instead of using HDFS's replication-based fault-tolerance. 
+                Triple-H improves 
+                performance of cluster I/O by writing the data to
+                the most optimal storage layer (RAMDisk, SSD, HDD, Lustre) 
+                dictacted by the selected placement policy (i.e. Greedy or 
+                Load-balanced).
+
+                Although both of these solutions have shown to improve I/O
+                performance
+
         \section{BigData Scheduling}\label{sched}

### 2018-10-07 paper.tex
added bibliography

@@
         Many neuroimaging workflow engines currently exist. Examples include
-        Nipype, Pipeline System for Octave and Matlab (PSOM), LONI, SPM, FastR,
-        Automatic Analysis (AA) and Pydpiper\note{add citations here}. These 
+        Nipype~\cite{nipype}, Pipeline System for Octave and Matlab (PSOM)~\cite{10.3389/fninf.2012.00007}, LONI~\cite{REX20031033}, SPM~\cite{spm}, FastR~\cite{10.3389/fict.2016.00015},
+        Automatic Analysis (AA)~\cite{10.3389/fninf.2014.00090} and Pydpiper~\cite{10.3389/fninf.2014.00067}. These 
         workflow engines aim to satisfy four criteria: 1) straightforward \tristan{easy? facilitated?} 
@@
         executing on top of different operating systems, produce significantly 
-        dissimilar results \cite{bowring, gronenschild 2011, glatard 2015}. As a result, it is necessary that specific details 
+        dissimilar results \cite{bowring, gronenschild2011, glatard2015}. As a result, it is necessary that specific details 
         on the workflow, such a data used, software versions and operating 
@@
         {chapter}{Bibliography} 
+        \bibliographystyle{ieeetr}
         \bibliography{bibliography}
-        \bibliographystyle{ieeetr}
 \end{document}

### 2018-10-08 paper.tex
some notes on checkpointing

@@
             locality using a shared HDFS instance will remain an issue. 
-
-        \section{Performance enhancement strategies}\label{other}
-            scientific: workflow checkpointing (looking at file hashes)
+            \tristan{The section is interesting, but it lacks a bit of context: one would like to 
+            understand why Spark and Hadoop developed their own scheduler instead of
+            reusing the ones available on HPC. And what are the main characteristics of these
+            schedulers: data locality should be mentioned here.}
+
+        \section{Checkpointing}\label{other}
+
+            Scientific workflow engines employ other strategies for improving   
+            performance. As scientific worflows are compute intensive, it is    
+            necessary for workflow engines to avoid repeating steps that have   
+            completed successfully. This is achieved by scientific workflow     
+            engines through the process of checkpointing. Workflow engines such 
+            as Nipype, PSOM, AA, Pydpiper, Nextflow, Taverna, Pegasus, Galaxy,  
+            Toil, Make and LONI all provide a mechanism to resume a pipeline    
+            from last successful state. This can be done either by detecting    
+            changes in the task definition (e.g. Nipype, PSOM, AA, Pydpiper,    
+            Nextflow), checking if modification of input                        
+            data is more recent than previously stored output data (e.g. Make), 
+            or requesting that the user manually restart execution from         
+            selected task (e.g. LONI).
+
+            Big Data frameworks such as MapReduce, Spark and Dask do not have   
+            any form of checkpointing functionality outside of a given          
+            execution. For instance, it is possible to checkpoint an RDD, but   
+            that RDD will only be accessible within the application's context   
+            and will not be shared with future executions. As failures do       
+            happen and individual neuroimaging tasks may take a non-negligable  
+            amount of time to execute, a mechanism to checkpoint workflows      
+            will need to be added to any BigData framework used for             
+            neuroinformatics.                                                   
+            \note{dont know how Taverna, Pegasus, Galaxy or Toil achieves this} 
+            \note{openMole says it checks to see if outputs exist. similar to   
+            make?}                                                              
+            \note{what about SPM, FastR??}
 
@@
             \tristan{How about: burst buffers, pilot jobs, deployments of Spark on HPC?}
+
+        \section{Evaluation of BigData Frameworks for Scientific Computing}     
+            The BigData frameworks have sparked significant interest within the 
+            scientific community, leading to research on the efficiency of      
+            these frameworks for scientific computing. ~\cite{valduriez} found  
+            that Spark scales better than expected with longer tasks. This is   
+            because                                                             
+            there is less I/O occurring and the scheduler is not as overloaded. 
+            Additionally, ~\cite{Boubela} found that Spark in conjunction with  
+            GPU acceleration greatly reduced the processing time of large       
+            neuroimaging datasets.                                              
+            As scientific, specifically neuroimaging workflows are typically    
+            comprised of long tasks, it                                         
+            is crucial that BigData frameworks scale well with tasks of longer  
+            durations.
+
+            To determine how different BigData frameworks compare in their 
+            efficiency for the processing of scientific workflows, 
+            ~\cite{ariel} compared Spark, Dask, SciDB, Myria and TensorFlow on
+            a neuroimaging and astronomy use case. It was found that Dask, 
+            Myria and Spark performed consistently better and had comparable 
+            performance. Dask had poorer performance when data did not fit in
+            memory. It was also found that Spark exhibited slower performance 
+            when the individual images are larger due to the lack of overlap 
+            between shuffle and map stages \note{revise this}. Overall it was
+            found that BigData Frameworks still required some adaptations to be
+            effectively used as scientific workflow engines.
 

### 2018-10-08 paper.tex
intro's intro fixed

@@
         both the storage and computation capacity of a standard research 
-        lab workstation. With the advancement of data sharing technologies, this 
-        data has been made widely available, \weird{with the only factor impeding
-        research on this data being access to infrastructure and software that 
-        enables efficient processing on such infrastructure}. Many research labs 
-        do have access to high performance computing (HPC) clusters, however, 
-        without efficient software to process this data, processing can take an 
-        excessive amount of time. Ensuring efficient processing of data is 
-        complex and likely beyond the knowledge of an average neuroimaging 
-        researcher \tristan{Be careful with 'knowledge', some neuroimaging researchers are very compute-savvy}. As a result, frameworks for the efficient processing of 
-        neuroimaging data need to be developed. \tristan{Add a sentence to transition from 'frameworks for efficient processing' to 'workflow engines'}
+        lab workstation. With the advancement of data sharing technologies, 
+        this data has been made widely available, however, accessibility to 
+        such data remains limited due to infrastructural demands and lack of 
+        performant software. Even with powerful infrastructure, such as High
+        Performance Computing (HPC) infrastructure, software configured to 
+        efficiently use resources is required to make execution time manageable
+        . Ensuring efficient processing of data is 
+        complex and limits accessibility to researchers with little programming 
+        knowledge. Existing neuroimaging workflow engines primarily rely on HPC
+        resource managers for performance and are therefore not well suited for
+        BigData processing. As a result, frameworks for the efficient 
+        processing of Big neuroimaging Data need to be developed\tristan{Add a sentence to transition from 'frameworks for efficient processing' to 'workflow engines'}
 
         Many neuroimaging workflow engines currently exist. Examples include
-        Nipype~\cite{nipype}, Pipeline System for Octave and Matlab (PSOM)~\cite{10.3389/fninf.2012.00007}, LONI~\cite{REX20031033}, SPM~\cite{spm}, FastR~\cite{10.3389/fict.2016.00015},
-        Automatic Analysis (AA)~\cite{10.3389/fninf.2014.00090} and Pydpiper~\cite{10.3389/fninf.2014.00067}. These 
-        workflow engines aim to satisfy four criteria: 1) straightforward \tristan{easy? facilitated?} 
+        Nipype~\cite{nipype}, Pipeline System for Octave and Matlab 
+        (PSOM)~\cite{10.3389/fninf.2012.00007}, LONI~\cite{REX20031033}, 
+        SPM~\cite{spm}, FastR~\cite{10.3389/fict.2016.00015},
+        Automatic Analysis (AA)~\cite{10.3389/fninf.2014.00090} and 
+        Pydpiper~\cite{10.3389/fninf.2014.00067}. These 
+        workflow engines aim to satisfy four criteria: 1) simplified
         workflow composition, 2) performance, 3) portability of the workflows 
@@
         focus is mainly on workflow composition and performance. Big Data 
-        frameworks such as Apache Spark and Dask \tristan{Add references} both aim to be easy-to-use,
+        frameworks such as Apache Spark~\cite{Zaharia:2016:ASU:3013530.2934664} 
+        and Dask~\cite{rocklin2015dask} both aim to be easy-to-use,
         applicable to a wide variety of analyses and improve performance in the 
@@
         to be efficient on commodity infrastructure in which transferring data
-        over a network would be extremely costly.
-
-        Currently, there exists no Big Data framework which is adapted for the
+        over a network is extremely costly.
+
+        Currently, there lacks a Big Data framework which is adapted for the
         processing of neuroimaging data. However, tools such as the Thunder 
-        Project \note{add citation} allow for the processing of time-series
-        neuroimaging data using Spark as the backend framework. \weird{Tools such as 
-        these} provide built-in functions to process the data. For users to 
-        extend functionality, it is necessary for them to understand the backend
-        framework. Other scientific workflow engines, such as Toil and Pegasus
-        \note{add citations}, have incorporated features that enable more 
+        Project~\cite{Freeman:2014aa} allow for the processing of time-series
+        neuroimaging data using Spark as the backend framework. Such tools
+        provide built-in functions to process the data. For users to 
+        extend functionality, it is necessary for them to understand the 
+        backend framework. Other scientific workflow engines, such as 
+        Toil~\cite{Vivian:2017aa} and Pegasus~\cite{DEELMAN201517}, have 
+        incorporated features that enable more 
         performant workflow executions for scientific Big Data. They also enable
@@
         \tristan{Add a sentence to present the goal of the report.}
+        This paper aims to evaluate the efficiency of current neuroimaging 
+        frameworks in comparison to BigData frameworks and to determine the 
+        extent to which BigData frameworks must be adapted to be effective
+        neuroimaging workflow engines. 
         In section~\ref{datasets}, we will briefly introduce the two 
@@
         workflows on the basis of workflow composition: we will % 'in other words' is on my list of expressions to avoid
-        look at the programming languages necessary to construct these workflows
+        examine the programming languages necessary to construct these workflows
         , how easy is it to add new tools and will these workflows be understood
@@
 
-            \subsection{input and output definitions}
-                TBD
-
             \tristan{Good material in here, but it lacks a bit structure.

### 2018-10-08 paper.tex
edited dataset paragraph

@@
         will include SPM, LONI, Nipype, PSOM, FastR, AA, Pydpiper 
-        for neuroimaging; Taverna, Pegasus, Make, OpenMOLE, Nextflow, Galaxy, 
-        Toil for other scientific workflows; and MapReduce, Spark and Dask for
+        for neuroimaging; Taverna~\cite{doi:10.1093/bioinformatics/bth361}, 
+        Pegasus, Make~\cite{10.3389/fninf.2016.00002}, 
+        OpenMOLE~\cite{passerat2014openmole}, Nextflow~\cite{Di-Tommaso:2017aa}
+        , Galaxy~\cite{Goecks2010}, 
+        Toil for other scientific workflows; and Hadoop 
+        MapReduce~\cite{Dean:2008:MSD:1327452.1327492}, Spark and Dask for
         Big Data Frameworks. Chapter \ref{workcomp} will evaluate these 
@@
         examine the programming languages necessary to construct these workflows
-        , how easy is it to add new tools and will these workflows be understood
-        by the community. In Chapter \ref{performance}, we will proceed to 
+        , as well as, how easy is it to add new tools. In Chapter 
+        \ref{performance}, we will proceed to 
         investigate the strategies employed by these frameworks to improve the 
@@
 
-            As a result of the rise in data sharing in neuroscience, neuroimaging
-            Big Data has come to fruition. The field of neuroscience has now
+            As a result of the rise in data sharing in neuroscience, 
+            neuroimaging
+            BigData has come to fruition. The field of neuroscience has now
             reached a bottleneck where there exists sufficient data to make
@@
             large datasets. Large images consist of singular images ranging 
-            from 100s of gigabytes to terabytes in size. A well-known example of 
-            such an image is the BigBrain \note{cite here}, a histological image 
+            from 100s of gigabytes to terabytes in size. A well-known example 
+            of 
+            such an image is the BigBrain~\cite{Amunts1472}, a histological 
+            image 
             of the brain of a healthy 69 year-old man. At its highest resolution
@@
             size. Other examples of large images can be found in electron 
-            microscopy (EM), polarized light imaging (PLI) and micro computational
-            tomography (microCT) \note{need to do more research here.cite}. 
+            microscopy (EM)~\cite{Hildebrand:2017aa}, polarized light imaging 
+            (PLI)~\cite{10.1007/978-3-319-12084-3_1} and micro computational
+            tomography (microCT)~\cite{10.1371/journal.pone.0035691}. 
             Images at 
@@
             entirety. Notable examples of these datasets include the 
-            Alzheimer's Disease Neuroimaging Initiative (ADNI), OpenfMRI, 
-            the Autism Brain Imaging Data Exchange (ABIDE), the Consortium for
-            Reliability and Reproducibility (CoRR), the Human Connectome Project
-            (HCP) and the UK Biobank, with the later
-            two being flagship recent examples of large datasets in neuroimaging.
+            Alzheimer's Disease Neuroimaging Initiative 
+            (ADNI)~\cite{doi:10.1002/jmri.21049}, 
+            OpenfMRI~\cite{POLDRACK2017259}, 
+            the Autism Brain Imaging Data Exchange 
+            (ABIDE)~\cite{Di-Martino:2013aa}, the Consortium for
+            Reliability and Reproducibility (CoRR)~\cite{Zuo:2014aa}, the Human 
+            Connectome Project (HCP)~\cite{VANESSEN201362} and the UK 
+            Biobank~\cite{Miller:2016aa}, with the later
+            two being flagship recent examples of large datasets in 
+            neuroimaging.
 
@@
             preprocessing of the images. Popular neuroimaging processing tools
-            such as CIVET, ANTs, and FreeSurfer \note{cite} were used for the 
+            such as CIVET~\cite{citation-0}, ANTs~\cite{avants2009advanced}, 
+            and FreeSurfer~\cite{FISCHL2012774} were used for the 
             preprocessing of structural data, and several other pipelines were
@@
 
-            The CoRR \tristan{expand acronym} was developed as an initiative to study and improve the 
+            The Consortium for Reliability and Reproducibility 
+            \tristan{expand acronym} was developed as an initiative to study 
+            and improve the 
             reliability and reproducibility of functional connectomics. The 
@@
             developed by these initiatives made use of popular neuroimaging 
-            toolboxes, such as FSL, FreeSurfer, ANTS, CIVET, rather that
+            toolboxes, such as FSL~\cite{SMITH2004S208}, FreeSurfer, ANTS, 
+            CIVET, rather than
             creating their own. Therefore, there is a great need for Big Data
             frameworks that can leverage existing tools for the efficient 
-            processing of large-scale datasets. With the availability of such 
-            tools, many more research labs will be able to perform
-            higher-impact studies \note{reword}.
+            processing of large-scale datasets. Widespread availability and 
+            accessibility of such frameworks will in turn lead to a greater
+            volume of higher-impact studies.
 

### 2018-10-08 paper.tex
edited intro

@@
         \section{Open-science Platforms}\label{platforms}
-            Efficient processing of workflows are only but a small portion of 
+            Efficient processing of workflows is only but a small portion of 
             the Big Data issue in neuroscience. % nice transition, this section gives more context for your study
@@
             Examples of popular neuroinformatics open-science platforms include 
-            CBRAIN\note{cite paper}, OpenNeuro \note{cite poster} and Neurodata
-            \note{cite prepring}. OpenNeuro and NeuroData primarily focus on
+            CBRAIN~\cite{10.3389/fninf.2014.00054}, 
+            OpenNeuro~\cite{gorgolewski2017openneuro} and 
+            Neurodata~\cite{burns2018community}. OpenNeuro and NeuroData primarily focus on
             cloud deployement whereas CBRAIN is configured for the deployment
@@
         \section{Infrastructure}\label{infrastructure}
-            HPC infrastructure was designed during a time when compute resources
-            were scarce. As a result, \tristan{I disagree with 'As a result' and I don't particularly agree with the previous sentence: even if
+            HPC infrastructure was initially designed for the processing of 
+            compute-intensive applications. As a result, \tristan{I disagree with 'As a result' and I don't particularly agree with the previous sentence: even if
             resources aren't scarce, they are good reasons to optimize resource managers: energy consumption, application makespan, fairness, etc. Also,
@@
             strategies to efficiently use the compute resources available for 
-            processing. However, with the advent of big data, there has been a 
-            shift from compute-intensive to data-intensive workloads. That is, 
-            with the accumulation of large datasets, data transfers have begun 
-            to monopolize processing time, and with it software has been 
-            designed to minimize data transfers.
-
-            We define HPC systems as those that employ shared parallel 
-            filesystem such as Lustre \tristan{and data-unaware schedulers?}.
+            processing. However, with the advent of big data, data-intensive 
+            workflows are now beginning to become more present on HPC systems.
+            Since the schedulers and workflow engines made for HPC systems are 
+            'data-unaware', they are not efficient with the processing of Big
+            Data
+
+            HPC and Big Data infrastructure differ in many ways. We define HPC 
+            systems as those that employ a shared parallel 
+            filesystem (distinct from any compute node), such as Lustre, and 
+            data-unaware schedulers. \tristan{and data-unaware schedulers?}.
             \tristan{You should say that this definition is 'by contrast to big data platforms' to explain where it's coming from.}
-            This shared filesystem is the slowest of
-            all storage devices as it is shared amongst all users of the system.
-            Compute nodes in an HPC system may or may not contain any local 
-            storage and network interconnects between nodes are typically fast.
+            The shared parallel filesystem is the slowest of
+            all storage devices found on HPC as it is shared amongst all users 
+            of the system.
+            Compute nodes in such infrastructure may not contain any local 
+            storage. However, network interconnects between nodes are typically 
+            fast.
             Resources managers used in HPC infrastructures typically employ 
@@
             Examples of schedulers used in HPC clusters 
-            include SLURM, PBS, SGE and HTCondor \note{cite here}.
-
-            Commodity infrastructure \tristan{Call it 'Big Data infrastructure?'} differs from HPC infrastructures in that
+            include SLURM~\cite{yoo2003slurm}, 
+            PBS~\cite{10.1007/3-540-60153-8_34}, SGE~\cite{SGE} and 
+            HTCondor~\cite{htcondor}.
+
+            Big Data infrastructure \tristan{Call it 'Big Data infrastructure?'} differs from HPC infrastructures in that
             it uses less-costly hardware. It follows a shared-nothing paradigm
@@
             nodes are responsible for storing the data as well. File systems
-            used in these clusters are distributed filesystems, such as the
-            Hadoop Distributed File System (HDFS) and Alluxio \note{cite}.
+            used within such infrastructure are distributed filesystems, such 
+            as the
+            Hadoop Distributed File System (HDFS)~\cite{5496972} and 
+            Alluxio~\cite{Li:2014:TRM:2670979.2670985}.
             Fault-tolerance of the data is achieved through replication across
             nodes. As transferring data across the network is typically costly 
-            in such infrastructures, frameworks designed to run on commodity 
+            in such infrastructures, frameworks designed to run on Big Data 
             infrastructure attempt to limit the transfer of data across nodes. 
@@
             typically expect dedicated infrastructure, as is the case with 
-            Yet Another Resource Negotiator (YARN), but can also execute 
-            heterogeneous workflows, as can be seen with Mesos \note{cite}.
-
-            Commodity clusters may also come in the form of cloud infrastructure.
+            Yet Another Resource Negotiator 
+            (YARN)~\cite{Vavilapalli:2013:AHY:2523616.2523633}, but can also 
+            execute heterogeneous workflows, as can be seen with 
+            Mesos~\cite{hindman2011mesos}.
+
+            Big Data infrastructure may also come in the form of cloud 
+            infrastructure.
             While the hardware employed is typically the same as other
-            commodity clusters, the software used varies. Cloud infrastructures
+            Big Data clusters, the software used varies. Cloud infrastructures
             have the added benefit of virtualization, allowing users to 
@@
             of the biggest limiting factors of cloud resources is the cost 
-            associated with usage.
+            associated with usage~\cite{10.3389/fninf.2017.00063}.
 
             Researchers typically have access to HPC infrastructure, but may
-            also have access to commodity infrastructure in the format of a 
+            also have access to Big Data infrastructure in the format of a 
             local cluster or cloud. Although both neuroimaging frameworks and
-            big data frameworks can be made to execute in either environments,
-            the performance improvements that they bring are very much 
+            Big Data frameworks can be configured to execute in either 
+            environments,
+            the performance enhancements that they bring are extremely
             dependent on infrastructure that they are executed on. For instance,

### 2018-10-08 paper.tex

@@
 
-        Workflows, also known as pipelines, describe the flow of data from 
-        activity to another. They are typically represented by a 
-        Directed Acyclic Graph (DAG) \tristan{Not necessarily a DAG, see Nipype for instance, it has loops.
-        I don't think you want to enter in a taxonomy of languages though, just say 'graph'}. Workflow activites, also referred to as 
-        workflows stages, encapsulate a tool \tristan{Mention 'interface' here as an intro to sub-sec 2}. Tools may either take the form of 
-        a command-line tool or a custom script \tristan{A script is a command-line tool. I think tools may also be web services (Taverna) or just 
-        software libraries (Dask)}. The engines determine which
-        workflow activity can be staged by determining if the stage dependencies
-        have been satisfied.
-
-        In order for a workflow engine to be successful, it must be simple to 
+        Workflows, also known as pipelines, describe the flow of data from one 
+        activity to another and are typically represented by a graph. \tristan{Not necessarily a DAG, see Nipype for instance, it has loops.
+        I don't think you want to enter in a taxonomy of languages though, just say 'graph'}. Workflow activities, also referred to as 
+        workflow interfaces or stages, encapsulate a tool through the use of an
+        interface\tristan{Mention 'interface' here as an intro to sub-sec 2}. 
+        Tools can be represented in many forms, including: command-line tools, 
+        custom scripts integrated directly into the workflow code, web 
+        services and software libraries.\tristan{A script is a command-line tool. I think tools may also be web services (Taverna) or just 
+        software libraries (Dask)}. A workflow activity with all its 
+        dependencies is considered ready to be staged.
+
+        For a workflow engine to be considered successful, it must be simple to 
         compose workflows using that framework. New researchers should be able 
-        to acquire the skills to compose worfklows in that framework without
+        to acquire the skills to compose worfklows without
         investing a significant amount of time in understanding it. Reuse of
-        existing tools is also common occurrence in neuroinformatics and 
+        existing tools is also common occurrence in neuroinformatics, and 
         workflow engines need to accomodate this by undertaking a modular 
@@
         the community at large, these workflows must be easily understood by the
-        community. \tristan{Perhaps mention that this also relates to the sharing
-        of reproducible results}.
+        community. This relates to the reproducibility of studies, as it is 
+        more than likely that easily understood workflows are reproducible. 
+        \tristan{Perhaps mention that this also relates to the sharing
+        of reproducible results -- VH: not sure if this is what you meant}.
 
@@
         (Taverna, Pegasus, Make, OpenMOLE, Nextflow, Galaxy, Toil), and three
-        BigData frameworks (MapReduce, Spark, Dask). We will use three axes to
+        BigData frameworks (MapReduce, Spark, Dask). We will use two axes to
         evaluate workflow composition. Section~\ref{lang} will compare the 
-        framework on the basis of programming language used, whereas 
+        framework on the basis of programming language used. Section~\ref{ds}
+        will then examine the differences in data structures between scientific
+        and BigData frameworks. 
         Section~\ref{mod} will look at how easy it is to integrate existing 
-        tools into the workflows. Section~\ref{sharing} will evaluate the ease
-        at which their corresponding workflows can be shared.
+        tools into the workflows.
 
@@
             affects how easily it will get picked up by the community. Complex
-            languages may mean greater efficiency of the engines
+            languages may mean greater efficiency of engines
             , but also result in extensive time spent on learning and developing
@@
 
-            To address this issue, some workflow engines have taken the route
+            To address this issue, some workflow engines have taken the role
             of employing a graphical user interface (GUI). This makes workflow 
@@
             workflow engine with a GUI interface for workflow composition is
-            that users are greatly limited by the tasks they can perform. 
-            Workflow engines that employ a worfklow composition GUI include
-            SPM, Taverna, OpenMOLE, LONI and Galaxy. \note{double-check that 
-            none have a programming API.Loni does..i think SPM does too}
+            that users are greatly limited by the tasks they can perform. For 
+            instance, workflows that perform any form of iterative operations
+            (e.g. machine-learning workflows) or workflows with conditionals
+            (e.g. given a certain input, do not execute this activity) cannot
+            be represented by a GUI. 
+            Workflow engines that employ a worfklow composition GUI include:
+            SPM, Taverna, OpenMOLE, LONI and Galaxy.
             \tristan{My conclusion on GUIs is that they are suitable only
@@
             workflow composition, such as LONI and OpenMOLE, allow more 
-            experienced users to script their workflow. Six different 
-            programming
+            experienced users to script their workflow. Four different 
             languages are used for describing scientific workflows: 1) Python, 
-            2) MATLAB and Octave \tristan{I think the official name is GNU/Octave}, 3) Shell \tristan{scripting languages}, 4) XML, 6) Common Workflow Language 
-            (CWL) and 6) Domain Specific Language (DSL). \tristan{I don't see why CWL and XML don't qualify as DSL} Workflows engines that 
-            employ usage of Python include: Nipype, FastR, Pydpiper, Pegasus
-            and Toil \tristan{Dask? Spark?}.
+            2) MATLAB and GNU/Octave \tristan{I think the official name is GNU/Octave}, 3) Shell Scripting languages \tristan{scripting languages}, 4) Domain Specific Language (DSL).
+\tristan{I don't see why CWL and XML don't qualify as DSL} Workflows engines that
+            employ usage of Python include: Nipype, FastR, Pydpiper, Pegasus, Toil,
+            MapReduce, Dask and Spark. \tristan{Dask? Spark?}.
             Python is an excellent choice as a programming language for 
@@
             widespread use within the community. Other advantages of 
-            programming in Python including a wide array of available libraries,
-            including neuroimaging-specific ones such as Nilearn and Nibabel \tristan{mention scipy, sci-kit learn, etc}
-            \note{cite}, and an lower learning curve compared to other 
+            programming in Python include: a wide array of available libraries,
+            particularly neuroimaging-specific ones such as Nilearn~\cite{nilearn} and Nibabel~\cite{matthew_brett_2018_1287921}
+            and more general, but heavily used in neuroimaging, libraries such
+            as SciPy~\cite{scipy} and Scikit Learn~\cite{pedregosa2011scikit} \tristan{mention scipy, sci-kit learn, etc}
+            , and an lower learning curve compared to other 
             programming languages. Additionally, Python is platform independent
@@
             neuroinformatics community is MATLAB/Octave. Workflow engines which
-            use this language include: PSOM, SPM, and Automatic analysis. A 
-            downfall to imposing MATLAB as programming language is that the 
-            it is proprietary, thus imposing a fee on 
-            anyone wishes to program in that language \tristan{You should also mention the
+            use this language include: PSOM, SPM, and Automatic analysis. Like 
+            Python, MATLAB and GNU/Octave are platform-independent and also easy to use. In addition, MATLAB has 
+            been found to exhibit better performance compared to Python for 
+            common technical tasks. It also provides an integrated environment
+            for programming. A 
+            downfall to imposing MATLAB as programming language is that it is proprietary, thus imposing a fee on 
+            anyone wishes to program in that language or have access to their diverse array of libraries \tristan{You should also mention the
             advantage of Matlab: easy to learn, integrated ecosystem, 
@@
             removing the requirement to purchase software for workflow 
-            composition. Similarly to Python, MATLAB and Octave are 
-            high-level platform-independent languages. MATLAB, in particular,
-            also provides a wide array of libraries, however, the software must
-            be purchased in order to access them.
+            composition.
 
             \tristan{This paragraph should start with ``Shell scripting ...". The structuration of this section by language
-            is great, but the writing of this paragram doesn't put it forward.}Scientific workflows are typically composed in UNIX-based 
+            is great, but the writing of this paragram doesn't put it forward.}
+            Many researchers have been exposed to Shell scripting languages as 
+            several scientific workflows are composed in UNIX-based 
             environments. These environments typically have Make installed by
             default eliminating the need for users to install it manually.
-            Furthermore, all users who use Linux environments will typically 
-            need to write some form of Shell script. This is the basis to use 
-            workflow engines, such as Make, as scientific workflow engines. 
+            These features make workflow engine Make desirable for scientific
+            computing.
             As Make is exclusive to UNIX environments, it is not 
@@
 
-            Composing workflows in XML is another option provided by some 
-            workflow engines such as Pegasus and LONI. However, the use of a 
+            Domain Specific Languages (DSL) are also used to compose workflows.
+            The can be made up of any language (e.g. XML, CWL, Scala and Groovy) that has been adapted for use in 
+            a specific domain.
+            XML is the language of choice in  
+            workflow engines such as Pegasus and LONI. Using a markup language
+            such as XML have the advantage of having access to 
+            language-independent parsers and validators. However, the use of a 
             markup language removes some of the flexibility. \tristan{Before you explain the downside, 
@@
 
-            Another alternative is to compose workflows using Common Workflow
-            Language (CWL). This can be seen in workflow engines like Toil. 
-            CWL is not a programming language, but a specification for workflows
+            Another example of a DSL is a Common Workflow
+            Language (CWL). CWL can be seen in workflow engines like Toil. 
+            It is similar to XML in that its main difference is that it is 
+            YAML/JSON-based.
             \tristan{It is a YAML-based DSL for workflows. Just like Pegasus' DAX is an XML-based DSL}. Files can either be written in JSON or YAML formats. CWL is 
             based on workflow engines like Make \tristan{? I don't think CWL relies on make, does it?}, but differs in that tasks
-            are isolated and inputs and outputs must be explicit \tristan{inputs and outputs are somewhat explicit in a Makefile}. Similarly to 
-            make, CWL is only available on UNIX platforms \tristan{strange, the CWL interpreter is in Python so I'd expect it to run anywhere}. \note{cite cwl 
+            are isolated \tristan{inputs and outputs are somewhat explicit in a Makefile}. Similarly to 
+            make, CWL is only available on Linux and OSX platforms \tristan{strange, the CWL interpreter is in Python so I'd expect it to run anywhere}. \note{cite cwl 
             website}
 
-            Domain Specific Languages (DSL) are also used to compose workflows.
-            DSLs are programming languages that have been adapted to specific
-            domains. Workflow engine examples include OpenMOLE, which was 
-            derived from Scala, and Netflow \tristan{Nextflow?}, which is derived from the Apache 
+            Programming language-based DSL workflow engines include OpenMOLE, which is  
+            derived from Scala, and Nextflow \tristan{Nextflow?}, which is derived from the Apache 
             Groovy programming language. While both are written on top of the 
-            Java platform, Nextflow is only compatible with UNIX-based \tristan{How about using Linux everywhere instead of UNIX-based?} 
+            Java platform, Nextflow is only compatible with Linux and OSX \tristan{How about using Linux everywhere instead of UNIX-based?} 
             environments. These programming languages are not otherwise commonly
@@
 
-            MapReduce, Spark and Dask all provide interfaces to Python, making 
-            them good candidates for neuroimaging workflow engines. In addition
+            Since MapReduce, Spark and Dask all provide interfaces to Python,  
+            they are good candidates for neuroimaging workflow engines. In addition
             Spark also interfaces with R, another programming language commonly
@@
 
-            \subsection{Data structures}
-                \subsubsection{Scientific workflows}
-                \subsubsection{Big Data: MapReduce}
+            \section{Data structures}
+
+                The data structures made available by workflow engines dictate
+                the maximum performance available to a workflow. Scientific
+                workflows are generally file-based, meaning that data, in the 
+                form of files, will be 
+                transferred across the network and I/O must be
+                performed at each workflow activity. It is possible for some 
+                data to remain in memory with scientific workflows, but these 
+                are typically basic types such as numbers, strings and lists.
+                More complex types, however, can be seen in workflow engines 
+                such as PSOM which accept nested structures as long as the
+                structures's terminal fields are either strings or cells of 
+                strings. Due to workflow engines being file-based, performance
+                is limited, particularly with BigData.
+
+                In order to improve performance, BigData frameworks introduced
+                data-centric data structures. MapReduce is key-value based, 
+                therefore, data with the same keys would be located at the
+                same nodes. As MapReduce still requires file-based I/O, the 
+                amount of I/O remains similar to that of scientific workflows. 
+                However, the key-value structures help to reduce the amount 
+                of data transfers by promoting data-locality.
+
+                The MapReduce key-value pair data structure is an effective 
+                data structure for describing workflows, but is limited to only
+                batch workflows and can be improved in terms of performance. 
+                The resilient distributed datasets 
+                (RDD)~\cite{Zaharia:2012:RDD:2228298.2228301}, Spark's main
+                abstractions addresses these 
+                issues by being a shared-memory structure. They 
+                are immutable and only permit coarse-grained edits to ensure 
+                fault-tolerance. The RDDs shared memory enables a reduction in
+                I/O operation as data can be maintained in memory. Spark will only read 
+                and write data to and from disk when data exceeds available 
+                memory or when data needs to be shuffled around, as is the case
+                with all Spark actions and certain transformations such as 
+                \textit{join} and \textit{reduceByKey}. Moreover, RDDs permit
+                workflows to be evaluated lazily. By being able to look at the
+                entire workflow before evaluation, Spark can select the most
+                performant methods for evaluating the data. 
+                The RDD additionally expands its use to additional applications
+                by enabling the expression of both iterative and interactive
+                workflows. Furthermore, Spark provides an additional data 
+                structure to enable the efficient processing of tabular data.
+                This data structure is know as the 
+                DataFrame~\cite{Armbrust:2015:SSR:2723372.2742797}.
+
+                Dask builds off of Spark by not only providing data 
+                structures similar to Spark's RDDs and Dataframes, Dask Bag and 
+                DataFrame, but by also providing an abstraction for 
+                larger-than-memory matrices, the Dask array. Furthermore, it is
+                possible to use common Python APIs (e.g. NumPy\note{cite} and 
+                Pandas DataFrame~\cite{mckinney2011pandas}) to process the 
+                data. Although Spark does not have a datastructure for large 
+                matrices, it has an RDD-based API, known as \textit{linealg}
+                for similar processing of large 
+                matrices~\cite{BosaghZadeh:2016:MCO:2939672.2939675}.
 
@@
             Spark's RDDs and DataFrames, (4) Dask arrays, mentioning 
-            linealg. } Additionally, BigData frameworks are better 
-            suited for the 
-            processing of large images. Scientific workflows are generally 
-            file-based with neuroimaging frameworks only providing parallelism
-            on the file front. Sparks main abstraction, the Resilient 
-            Distributed Dataset (RDD), permits users to process large datasets.
-            A library based on the RDD, \textit{linealg}, allows users to create
-            vectors and matrices and use common Python libraries, such as 
-            Numpy and Scikit-learn to manipulate the data. Dask similarly 
-            provides the Dask bag for file manipulation, and the Dask array for
-            the manipulation of large matrices. Dask provides its own API for
-            the processing of large matrices that is based on Numpy. 
-            Both frameworks also provide data structures to manipulate tabular
-            data: the Spark Dataframe and the Dask Dataframe. Unlike Spark, the
-            Dask Dataframe is based off of the Pandas Dataframe API.
+            linealg. }
 
@@
             rather call it ``Interfaces" or ``Tool interfaces".}
-            Often times, users will want to integrate existing tools, in 
-            addition to custom scripts, into their workflows. As workflow 
-            engines, with the exception of Make, are typically not written in 
-            Shell, it is necessary to create interfaces to these tools. These
-            interfaces may range from simple calls to the command-line, to more 
-            complex ones that track inputs, outputs and other parameters; ensure
-            that all parameters are specified in the correct order and that the
-            parameter and file types are all correct; capture all of the 
-            standard input/output; and create folders to encapsulate task inputs
+            Often times, users will want to integrate existing tools into their 
+            workflows. In order for workflows to communicate effectively with
+            the tools, it is required that an interface be created. Workflow 
+            interfaces may range from simple command-line calls, to more 
+            complex ones that track inputs, outputs and other parameters. 
+            Furthermore, they ensure
+            that all necessary parameters are correctly defined and in the 
+            correct order,  capture all of the 
+            standard input/output and create folders to encapsulate task inputs
             and outputs. Workflow engines typically rely on more complex types 
@@
             workflow engine to provide an intuitive way for users to create 
-            interfaces for their custom tools.
-
+            interfaces for their custom tools. Workflow interface can be 
+            divided into two groups: those that are described using a 
+            programming language and those that are described using DSL.
+
+            Workflow engines that utilize programming languages include: SPM,
+            One of the earliest neuroimaging workflow engines, 
             Early neuroimaging workflow engines, such as SPM, are monolithic, 
-            and therefore, do not provide a mechanism for users to incorporate
-            their custom tools \tristan{Users can still integrate tools as Matlab functions
-            doing system calls}. Workflow engines, such as AA, were thus 
-            conceived to enable engines such as SPM to be more modular. AA 
+            and therefore, do not provide a default interface by which users 
+            can integrate their tools\tristan{Users can still integrate tools as Matlab functions
+            doing system calls}. Workflow engines, such as AA, were 
+            created to enable engines such as SPM to be more modular. AA 
             allows users to define their tasks through XML specification. AA's 

### 2018-10-08 paper.tex

@@
             interfaces may range from simple command-line calls, to more 
-            complex ones that track inputs, outputs and other parameters. 
+            complex interfaces that track inputs, outputs and other parameters. 
             Furthermore, they ensure
@@
             standard input/output and create folders to encapsulate task inputs
-            and outputs. Workflow engines typically rely on more complex types 
+            and outputs. Early scientific workflow engines, such as SPM, are 
+            monolithic. As a consequence, these workflow engines do not provide
+            a complex class for interface definition and validation. Recent
+            Workflow engines take a modular approach and promote sharing of data
+            and pipelines. Thetypically rely on more complex types 
             of interfaces in order to ensure correct functioning of the pipeline
-            . As these interfaces can be rather complex, it is necessary for the 
+            . Since these interfaces can be rather complex, it is necessary for the 
             workflow engine to provide an intuitive way for users to create 
-            interfaces for their custom tools. Workflow interface can be 
+            interfaces for their custom tools. Workflow interfaces can be 
             divided into two groups: those that are described using a 
@@
 
-            Workflow engines that utilize programming languages include: SPM,
-            One of the earliest neuroimaging workflow engines, 
-            Early neuroimaging workflow engines, such as SPM, are monolithic, 
-            and therefore, do not provide a default interface by which users 
-            can integrate their tools\tristan{Users can still integrate tools as Matlab functions
-            doing system calls}. Workflow engines, such as AA, were 
-            created to enable engines such as SPM to be more modular. AA 
+            Making the choice between a programming or a markup language for 
+            interface definition is not obvious. One of the main reasons for 
+            selecting a programming language is that programming language is 
+            more versatile than are markup language, allowing any type of 
+            operation to be defined. For instance, a dynamic number of outputs
+            can be generated using a programming-language based interface 
+            definition. Workflow engines that utilize programming 
+            languages for the 
+            description of their workflow interfaces include: Nipype, PSOM
+            , Pydpiper, Toil, OpenMOLE and Make interfaces.
+
+            Nipype's wrapper, written in Python, can encapsulate CLIs, SPM, 
+            MATLAB and Python scripts. InputSpec and OutputSpec are two classes 
+            that can are used to describe the inputs and outputs of the 
+            interface. The former defines the input parameters of 
+            a command and the latter defines the outputs that are possibly 
+            generated by the command. For CLIs, it is necessary to describe the
+            command, such as order of parameters and required and optional parameters. 
+            Apart from Input and Output Spec, a Python dictionary containing all
+            expected outputs can be included.
+            Before executing a CLI, Nipype can start X virtual framebuffer if 
+            user interaction through a GUI is required by the tool if specified. 
+            For SPM interfaces, the job 
+            type and name must be specified. Interfaces to MATLAB require 
+            definition of the run function, while Python interfaces require
+            the definition of the \textit{\_run\_interface} function. \note{from
+            website}.
+
+            PSOM’s tool interfaces are known as \textit{Bricks}. They are a 
+            little less versatile than Nipype'sinterfaces since they can only
+            interface with MATLAB functions or CLIs. Bricks require the 
+            definition of input and output files, in addition to a structure of 
+            program options. A useful feature found in PSOM is the ability to 
+            perform a “dry-run” of the Brick without executing the command.
+
+            Pydpiper's tools can either be custom Python code or CLIs. The 
+            \textit{PipelineStage} and \textit{CmdStage} are used to 
+            encapsulate the tools. The \textit{CmdStage} class is specifically 
+            used to interface with CLIs, and thus, accepts a parsed array of the
+            command and its parameters as input. Atoms, typically used to wrap
+            around MINC file-format related tools, inherit from the
+            \textit{CmdStage} class, and require the input MINC file (string or
+            file handler) to be provided. A output file can also be specified, 
+            but it must be of the same format as the MINC file (string or file 
+            handler). Additionally, it is possible to pass multiple optional 
+            arguments to the Atom.
+
+            Toil is another workflow engine that describes their tool interfaces
+            in Python. The wrappers, also known as \textit{Jobs} enable users to
+            interface with custom Python code or Dockerized \cite{merkel2014docker} 
+            CLIs. Interfaces can be constructed either by creating a class that
+            inherits from the \textit{Job} class or through the use of a 
+            built-in function that converts the user-defined function (UDF) to a 
+            Toil \textit{Job}. To convert, all that is required is the UDF,
+            however, the user may also choose to include details on memory, 
+            number of cores and disk space required. For Dockerized tools, Toil
+            provides a built-in function that accepts the tool name, working dir
+            and parameters of the tool.
+
+            OpenMOLE has the ability to encapsulate UDF Scala scripts and CLI. 
+            In the case of CLI tasks, it is also possible to containerize or 
+            encapsulate with CARE \note{cite} to make it portable. OpenMOLE
+            defines several classes to allow for the execution of different 
+            types of tools. Like other framworks, it is vital to describe the
+            inputs and outputs of the Tasks such that the execution of tasks 
+            can be performed.
+
+            The Make pipeline is expressed in a makefile containing a set of 
+            rules (Tool interfaces). Rules consist of three components: targets, 
+            dependencies and recipes. The target is the output (typically a 
+            file) of the 
+            execution of a recipe and the dependencies are the input required to 
+            produce the given target. Recipes are the shell commands that will
+            generate the target from its dependencies.
+
+            \tristan{Users can still integrate tools as Matlab functions
+            doing system calls}.
+            Although DSL-based workflow interface schemas may be less versatile
+            than those written using programming languages, DSL based schemas
+            can be reused by any engine. This allows interface schemas to be 
+            portable. AA, LONI, FastR, Taverna, Pegasus and Galaxy all 
+            implement their iterface definition using a DSL.
+
+            AA 
             allows users to define their tasks through XML specification. AA's 
@@
 
-            LONI, FastR, Taverna, Pegasus and Galaxy are other examples of 
-            workflow engines that 
-            use an XML interface to describe CLIs. LONI enables generation of 
+            LONI enables generation of 
             the XML-based module interface through its GUI. The module interface
@@
             descriptor. Taverna differs from other workflow engines in that it
+            can 
             execute generic-style services (i.e. WSDL and REST) in addition to
-            other types of Web Services, as well a external CLIs, R scripts and
+            other types of Web Services, external CLIs, R scripts and
             Java services. To interface with a CLI, the XML file must contain 
@@
             to the executable, site identifier of the executable's host, 
-            executable type (e.g. installed or staged) and the profile. Pegasus
+            executable type (e.g. installed or staged) and the profile. In
+            addition to UDFs and CLIs, Pegasus
             can also execute containerized tools \tristan{how about the other ones?}. For this, a container 
@@
             configuration file is required to add the tool to the application.
-            The tool wrappers schema is very rich allowing for many tool 
+            The tool wrappers\ schema is very rich allowing for many tool 
             features to be defined. Similarly to many other workflow engine tool
-            wrapper, the configuration file contains details on the tool name, 
+            wrappers, the configuration file contains details on the tool name, 
             type, id, versions, command and input and output specifications. It
@@
 
-            While many worfklow engines use XML or JSON to describe their tools,
-            others wrap their tools in an interface defined by a
-            programming/scripting language. Nipype, Pydpiper and Toil all use
-            Python to create their wrappers; PSOM wrappers are written in
-            MATLAB/Octave; OpenMOLE uses Java/Scala; and Make interfaces are
-            written in Shell Script.
-
-            Nipype's wrapper, written in Python, can encapsulate CLIs, SPM, 
-            MATLAB and Python scripts. Attributes and methods that can be 
-            defined for all interfaces include InputSpec, OutputSpec and a 
-            Python dictionary of all expected outputs. InputSpec and OutputSpec
-            are two Nipype classes. The former defines the input parameters of 
-            a command and the second defines the outputs that are possibly 
-            generated by the command. For CLIs, it is necessary to define the
-            command. Additionally, Nipype will start X virtual framebuffer 
-            before executing a CLI, if specified \tristan{you need to explain why}. For SPM interfaces, the job 
-            type and name must be specified. Interfaces to MATLAB require 
-            definition of the run function, while Python interfaces require
-            the definition of the \textit{\_run\_interface} function. \note{from
-            website}.
-
-            Pydpiper's tools can either be custom Python code or CLIs. The 
-            \textit{PipelineStage} and \textit{CmdStage} are used to 
-            encapsulate the tools. The \textit{CmdStage} class is specifically 
-            used to interface with CLIs, and thus, accepts a parsed array of the
-            command and its parameters as input. Atoms, typically used to wrap
-            around MINC file-format related tools, inherit from the
-            \textit{CmdStage} class, and require the input MINC file (string or
-            file handler) to be provided. A output file can also be specifiec, 
-            but it must be of the same format as the MINC file (string or file 
-            handler). Additionally, it is possible to pass multiple optional 
-            arguments to the Atom.
-
-            Toil is another workflow engine that describes their tool interfaces
-            in Python. The wrappers, also known as \textit{Jobs} enable users to
-            interface with custom Python code or Dockerized \note{cite Docker} 
-            CLIs. Interfaces can be constructed either by creating a class that
-            inherits from the \textit{Job} class or through the use of a 
-            built-in function that converts the user-defined function (UDF) to a 
-            Toil \textit{Job}. To convert, all that is required is the UDF,
-            however, the user may also choose to include details on memory, 
-            number of cores and disk space required. For Dockerized tools, Toil
-            provides a built-in function that accepts the tool name, working dir
-            and parameters of the tool.
-
-            PSOM's tool interfaces are known as \textit{Bricks}. \textit{Bricks}
-            can interface with any MATLAB function or CLI. Bricks require the
-            definition of input and output files, in addition to a structure of
-            program options. A useful feature found in PSOM is the ability 
-            to perform a ``dry-run" of the Brick without executing the command.
-
-            OpenMOLE has the ability to encapsulate UDF Scala scripts and CLI. 
-            In the case of CLI tasks, it is also possible to containerize or 
-            encapsulate with CARE \note{cite} to make it portable. OpenMOLE
-            defines several classes to allow for the execution of different 
-            types of tools. Like other framworks, it is vital to describe the
-            inputs and outputs of the Tasks such that the execution of tasks 
-            can be performed.
-
-            The Make pipeline is expressed in a makefile containing a set of 
-            rules (Tool interfaces). Rules consist of three components: targets, 
-            dependencies and recipes. The target is the output (typically a 
-            file) of the 
-            execution of a recipe and the dependencies are the input required to 
-            produce the given target. Recipes are the shell commands that will
-            generate the target from its dependencies.
-
             Unlike scientific workflow engines, BigData frameworks do not 
-            provide default interfaces for the execution of tools written in 
-            languages different from their own. Users defining their workflows
-            in Dask or the Spark and MapReduce using the Python API integrate
-            Python UDFs in their workflow. While Dask users can reference their 
-            UDF of Python function directly in their workflow, Spark users
+            provide default interfaces for the execution of complex tools 
+            written in 
+            languages different from their own. While Dask users can reference
+            UDF Python functions directly in their workflow, Spark users
             must define a lambda function within an action or transformation 
@@
             code and Spark can call on code also written in Scala, Java and R.
-            \note{verify if it's possible to mix-and-match different libraries
-            ...don't think so}
 
@@
             command-line tools or containers, there are libraries that can be
-            easily integrated to achieve that. Boutiques, for instance, is a 
+            easily integrated to achieve that. 
+            Boutiques~\cite{doi:10.1093/gigascience/giy016}, for instance, is a 
             Python library that can execute command-line or containerized tools
@@
         workflows to BigData frameworks on the basis of performance. 
-        Section~\ref{fs} will look at what filesystems are \weird{required} \tristan{supported?} by the 
+        Section~\ref{fs} will look at what filesystems are supported \tristan{supported?} by the 
         engines. This is important because data movement of any kind is costly,
-        and may be worse depending on the filesystems available. \tristan{Also, mention that neuroinformatics
+        and may be worse depending on the filesystems available. Furthermore, 
+        typical neuroinformatics pipelines manipulate files, and therefore, 
+        reducing the amount of I/O is necessary.\tristan{Also, mention that neuroinformatics
         pipelines manipulate files.}
@@
         available and their compatibility with the workflow engines. 
-        Section~\ref{other} will discuss other performance improvement 
-        strategies employed by the workflow engines.
+        Section~\ref{other} will discuss the checkpointing of data in the context
+        of scientific workflows. This chapter will conclude with a discussion
+        on the efficacy of Big Data frameworks for scientific computing, in
+        Section~\ref{eval}.
 
@@
 
-            Unlike MapReduce and Spark-based workflows, Dask and scientific 
-            workflows are not data-centric. They focus on the efficient 
-            scheduling of tasks and rely on I/O and network transfers to
+            Although MapReduce, Spark and Dask are all data-intensive worflows,
+            Dask can be more similar to that of scientific workflow, 
+            particularly in a multiprocessing environment. It focused on the 
+            efficient 
+            scheduling of tasks and relies on I/O and network transfers to
             communicate the data between nodes. With BigData workflows, such
             operations could incur significant performance penalties. 
-            \weird{Filesystems}. To mitigate this and remain efficient in the face of
-            BigData, Dask implements lazy evaluation and in-memory processing.
-            BigData platforms such as MapReduce rely heavily on I/O, but the 
+            To mitigate this issue and remain efficient in the face of
+            BigData, Spark and Dask implement data locality, lazy evaluation 
+            and in-memory processing. Data locality is the scheduler's ability
+            to schedule tasks nearest to data location to reduce data transfers.
+            Lazy evaluation is the workflow's ability to optimize performance
+            based on the pipeline's graph structure. In-memory processing, on 
+            the other hand, tries to remove any I/O that occurs between tasks. 
+            Platforms such as MapReduce rely heavily 
+            on I/O, but the 
             I/O performed occurs on the compute node rather than on 
@@
             workflows. Additionally, MapReduce uses the concept of 
-            locality to ensure limit data transfers across the network. Spark
-            uses data locality, lazy evaluation and in-memory computing to 
-            ensure its optimal performance.
+            locality to ensure limit data transfers across the network.
             For the efficient processing of neuroimaging BigData, it will be
@@
 
-                    \note{mention caching}
-
-                    While Spark and Dask both rely on the concept of shared 
-                    memory, Spark avoids data transfers as often as possible by
-                    scheduling tasks to nodes where the data is located. Dask
-                    , however, will schedule its tasks to the next available 
-                    node.
+                    As subsets of the lineage can be reused amongst different
+                    structures in the pipeline, it is possible to cache the 
+                    data in order to avoid any form of recomputation, which 
+                    could otherwise incur performance penalties.
 
@@
                     network transfers are expensive, particularly within 
-                    commodity infrastracture, and therefore transferring large 
+                    Big Data infrastracture, and therefore transferring large 
                     volumes of data across the network for every task would be 
@@
 
-                    In contrast to MapReduce, Spark does not require data to 
-                    be stored on a HDFS-like filesystem. It can read data from
+                    In contrast to MapReduce, Spark and Dask do not require 
+                    data to 
+                    be stored on a HDFS-like filesystem. They can read data 
+                    from
                     any form of shared-filesystem, including Lustre. If data 
@@
                     constructing graphs that result in the most optimal
-                    task order of execution. \tristan{And further save on data transfers}
+                    task order of execution and further save on data transfers. 
+                    \tristan{And further save on data transfers}
 
@@
 
-                Alluxio, formerly known as Tachyon, is an in-memory filesystem
+                Alluxio~\cite{Li:2014:TRM:2670979.2670985}, 
+                formerly known as Tachyon, is an in-memory filesystem
                 that stores lineage data to offset the cost of writing data to 
@@
                 unlike Spark applications, Alluxio is meant to run indefinitely
-                , making it unfavourable, in some cases, to recompute the data
+                , making it unfavourable in some cases, to recompute the data
                 from its starting point. For this, Alluxio evaluates the 
@@
 
-                Triple-H leverages the heterogeneity of HPC clusters to 
+                Triple-H~\cite{7152476} leverages the heterogeneity of HPC 
+                clusters to 
                 optimize the data's placement across the cluster for most 
@@
                 dictacted by the selected placement policy (i.e. Greedy or 
-                Load-balanced).
+                Load-balanced). This allows alternative storage to be used as 
+                a form of burst buffer to Lustre as data is written into a more
+                optimal storage layer before being transferred to a slower 
+                storage such as Lustre.
 
                 Although both of these solutions have shown to improve I/O
-                performance
+                performance, their benefits are limited to certain 
+                configurations. Alluxio needs to be installed throughout the 
+                cluster occupying available memory. Data locality may also be
+                and issue as the user may only be scheduled to a subset of
+                nodes that do not store the data. Moreover, neuroimaging 
+                operations are compute-intensive, with some tasks being 
+                extremely lengthy. Recomputation may not be affordable and long
+                tasks can be frequent, meaning Alluxio will need to store more 
+                data that desirable.
+
+                Triple-H's main limitation is that it assumes the compute nodes
+                have local storage. This is certainly not always the case and
+                any performance improvements in systems without local storage 
+                may be negligable at best.
 
@@
             Scientific workflows typically rely on HPC schedulers for 
-            distributed scheduling on clusters. These schedulers include SLURM,
-            PBS,TORQUE, SGE, HTCondor and Mesos. Additionally, they may provide
+            distributed scheduling on clusters. These schedulers include 
+            SLURM~\cite{yoo2003slurm},
+            PBS~\cite{10.1007/3-540-60153-8_34}, 
+            TORQUE~\cite{computing2015torque}, 
+            SGE~\cite{SGE}, HTCondor~\cite{htcondor} and 
+            Mesos~\cite{hindman2011mesos}. Additionally, they may provide
             their own form of scheduling for computations on a single node. 
@@
             also provide interfaces to these HPC schedulers. However, other
-            frameworks such as MapReduce and Spark are limited to BigData 
-            clusters such as Yet Another Resource Negotiator (YARN and MESOS).
-            Spark additionally provides its own standalone scheduler. Similarly
+            frameworks such as MapReduce and Spark are limited to Big Data 
+            schedulers such as Yet Another Resource Negotiator (YARN) and MESOS.
+            The reason MapReduce and Spark do not leverage HPC resource managers
+            is that HPC resources managers are unaware of the locality of the 
+            data and cannot perform any form of lazy evaluation.
+            In addition to YARN and MEsos, Spark provides its own standalone 
+            scheduler. The standalone scheduler is simple providing only a 
+            basic scheduler. It allows users to use Spark out-of-the-box. Similarly
             to scientific workflows, all three BigData frameworks provide their
@@
             Since Spark does not natively interface with HPC schedulers, 
-            the use of overlay Spark clusters, similar to HoD, has been 
+            the use of overlay Spark clusters, similar to Hadoop on Demand 
+            (HoD), has been 
             investigated for Spark applications running on HPC. The reasons 
@@
             in the case of Spark, from having to wait until all nodes become 
-            available to begin execution. Both papers \note{jha and baer} use 
+            available to begin execution. 
+            Papers~\cite{7530058}~\cite{Baer:2015:IAS:2792745.2792779} both use 
             this strategy. Furthermore, pilot-agent scheduling could be adapted
@@
 
-        \section{Evaluation of BigData Frameworks for Scientific Computing}     
+        \section{Evaluation of BigData Frameworks for Scientific Computing}\label{eval}     
             The BigData frameworks have sparked significant interest within the 
             scientific community, leading to research on the efficiency of      
-            these frameworks for scientific computing. ~\cite{valduriez} found  
+            these frameworks for scientific computing. ~\cite{souza2017spark} found  
             that Spark scales better than expected with longer tasks. This is   
-            because                                                             
-            there is less I/O occurring and the scheduler is not as overloaded. 
-            Additionally, ~\cite{Boubela} found that Spark in conjunction with  
+            a consequence of there being less I/O occurring and the scheduler 
+            not being as overloaded. 
+            Additionally, ~\cite{boubela2016big} found that Spark in conjunction with  
             GPU acceleration greatly reduced the processing time of large       
@@
             efficiency for the processing of scientific workflows, 
-            ~\cite{ariel} compared Spark, Dask, SciDB, Myria and TensorFlow on
+            ~\cite{mehta2017comparative} compared Spark, Dask and other Big Data 
+            frameworks on
             a neuroimaging and astronomy use case. It was found that Dask, 
-            Myria and Spark performed consistently better and had comparable 
+            and Spark performed consistently better than other frameworks and 
+            had comparable 
             performance. Dask had poorer performance when data did not fit in
             memory. It was also found that Spark exhibited slower performance 
-            when the individual images are larger due to the lack of overlap 
-            between shuffle and map stages \note{revise this}. Overall it was
+            when the individual images are larger due to the amount of transfers
+            occuring between map and shuffle phases. Overall it was
             found that BigData Frameworks still required some adaptations to be
@@
     \chapter{Portability}\label{portability}
-        In order for a workflow engine to be utilized, \weird{it much be simple} for 
-        users to \weird{install and execute the data}. Platform-independence of both 
+        Platform-independence of both 
         the frameworks and tools, as well as, simple installation of all 
@@
             running on the Java Virtual Machine (JVM) or being written in 
-            Python. Platform-independent engines that are Java-based include: 
-            PSOM, SPM, LONI, AA, Taverna, MapReduce and Spark \tristan{this list contains Matlab
-            engines}. The remainder of
-            the pipelines are implemented in Python. The remainder of the 
+            Python. Platform-independent engines that run on the JVM include: 
+            MATLAB engine frameworks, such as PSOM, SPM and AA; LONI; Taverna; 
+            MapReduce; and Spark \tristan{this list contains Matlab
+            engines}. Nipype, FastR, Pydpiper are all python-based platform 
+            independent workflow engines. The remainder of the 
             workflow engines (Pegasus, Make, OpenMOLE, Nextflow, Galaxy, and 
-            Toil are not platform-independent, but can execute on UNIX-like 
-            systems \tristan{macOS and Linux}.
-
-            Engines built on top of a platform-independent language does not 
+            Toil are not platform-independent, but can execute on macOS and Linux.
+
+            Engines built on top of a platform-independent language do not 
             necessarily imply platform independence \tristan{Grammar? Engines ... does not}. For instance, OpenMOLE is 
@@
             exist many containerization technologies, the two most notable ones
-            in scientific computing are Docker and Singularity. Docker and 
+            in scientific computing are Docker~\cite{merkel2014docker} and 
+            Singularity~\cite{10.1371/journal.pone.0177459}. Docker and 
             Singularity are both platform independent meaning users can run
@@
             Pipeline tools can be containerized. In neuroinformatics, 
-            NeuroDocker, a command-line tool for creating Docker and 
+            NeuroDocker~\cite{neurodoc}, a command-line tool for creating Docker and 
             Singularity containers with specific versions of neuroimaging tools
             installed. Examples of 
-            containerized applications include all of the BIDS Apps. 
+            containerized applications include all of the BIDS Apps~\cite{10.1371/journal.pcbi.1005209}. 
             Additionally, entire workflows can also be containerized, as can be
-            observed in fMRIPrep, a Nipype-based workflow for fMRI 
-            Preprocessing.
+            observed in fMRIPrep~\cite{Esteban306951}, a Nipype-based workflow for fMRI 
+            preprocessing.
 
@@
             importing and running containerized workflow tools. These engines
-            include: Galaxy, Toil, Netflow, Pegasus and FastR. While most 
+            include: Galaxy, Toil, Nextflow, Pegasus and FastR. While most 
             provide support for Docker and Singularity, Toil and Galaxy only 
@@
         Reproducibility is a critical component to all scientific workflows.
+        It ensures that results reported are valid through repeated testing. 
         Without reproducibility, correctness of results could never be 
@@
         recreating or executing the original workflow on the same or different
-        infrastructure\note{cite Peng 2011}. It has been found that more than 70\% of
-        scientific experiments are not reproducible \note{cite Mbaker}
+        infrastructure~\cite{Peng1226}. It has been found that more than 70\% of
+        scientific experiments are not reproducible~\cite{baker2016there}.
 
@@
         executing on top of different operating systems, produce significantly 
-        dissimilar results \cite{bowring, gronenschild2011, glatard2015}. As a result, it is necessary that specific details 
+        dissimilar results \cite{bowring:inserm-01760535, 10.1371/journal.pone.0038234, 10.3389/fninf.2015.00012}. 
+        As a result, it is necessary that specific details 
         on the workflow, such a data used, software versions and operating 
-        system used are provided. For this reason, many workflow engines have
+        system used are provided. Additionally, having records of such information
+        enable the tracing of results. For this reason, many workflow engines have
         incorporated provenance tracking into their design. \tristan{Provenance is also
         a way to merely document what happened so that results can be traced.}
+
+            Provenance tracking is crucial to scientific reporting.
+            On its own, source code is insufficient to reproduce the workflow 
+            described in a research paper. This is because source code does 
+            not specify the specific dataset used nor the specific versions of
+            tools and the environments that they were executed on. Minor 
+            changes to any of these features can significantly affect results.
+
+            As explained by the W3C PROV specification~\cite{missier2013w3c}, provenance details 
+            information on the data, the tools involved in its creation, and 
+            the responsible parties. Provenance tracking, in the context of 
+            workflows, is the act of capturing provenance data during
+            workflow execution. With detailed provenance and access to source 
+            code information, researchers should have enough information to be 
+            able to reproduce workflows. \tristan{These 2 paragraphs should be in the chapter intro.}
 
@@
         \section{Provenance Formats}\label{metadata}
-            The recent increase in data-sharing resulted in the rise of BigData 
-            in neuroscience \tristan{This feels like general intro to me}. To continue the promotion of data sharing in 
-            neuroscience, the derived data's metadata must be well understood
+            In order to improve data sharing in 
+            neuroscience, the derived data's provenance must be well understood
             by the community. Even if the provenance tracking is comprehensive,
@@
             issue, specifications for the standardization of metadata have been 
-            proposed. Here we look at three provenance metadata specifications.
-            The first is W3 PROV, a general-purpose provenance specification; 
-            NIDM-results
-            a comprehensive neuroimaging provenance specification; and BIDS 
-            Derivatives the recent provenance addition to the family of 
-            popular neuroimaging metadata specifications, BIDS \tristan{Make sure that
+            proposed. Here we look at four provenance metadata specifications.
+            The first is W3C PROV~\cite{missier2013w3c}, a general-purpose provenance specification; 
+            NIDM-results~\cite{Maumet:2016aa}
+            a comprehensive neuroimaging provenance specification; BIDS 
+            Derivatives~\cite{bidsderiv} the recent provenance addition to the family of 
+            popular neuroimaging metadata specifications, Brain Imaging Data Structure (BIDS); 
+            and data lineage, the Big Data solution to provenance tracking \tristan{Make sure that
             the acronym is defined earlier.}.
@@
             origin of data shared on the Web. 
-            The aim of the W3 PROV specification was \tristan{to?} gather information on
+            The aim of the W3 PROV specification was to \tristan{to?} gather information on
             the activities, entities and involved parties in the creation of 
@@
                 Clinical and Experimental Data Exchange (XCEDE) neuroimaging
-                provenance metadata format \tristan{I guess you'll cite references for that}. It is a domain-specific extension
-                to the W3 PROV model that aims to provide comprehensive and
-                machine-readable provenance from raw data to its final stage. \tristan{give example?}
+                provenance metadata format~\cite{Maumet:2016aa} \tristan{I guess you'll cite references for that}. It is a domain-specific extension
+                to the W3C PROV model that aims to provide comprehensive and
+                machine-readable provenance from raw data to its final stage. \tristan{give example? -- vhs: like a figure?}
                 However, it is currently limited in the provenance 
-                information it can provide. \tristan{Be more specific, what are the limitations?}
+                information it can provide, such as details on computational 
+                environments. It is also limited by the fact that consensus lacks
+                within the community on paradigms. \tristan{Be more specific, what are the limitations?}
 
@@
                 specifications. It is the sister
-                specification to the widely adopted BIDSRaw metadata format 
-                and it remains compatible with the BIDS Raw specification. 
+                specification to the widely adopted BIDS Raw metadata format 
+                and it remains compatible with BIDS Raw. 
                 Unlike NIDM-Results, BIDS derivatives, like BIDS Raw, do not
@@
                 derived from its folder structure and filename. Any additional
-                metadata can be found in a JSON sidecar file. \tristan{Explain that
+                metadata can be found in a JSON sidecar file. Due to its simplicity,
+                BIDS Derivatives will be the solution adopted by the community 
+                despite not using semantic web technologies such as W3C PROV and NIDM-Results.
+                This is primarily a result of its simplicity, as seen in its 
+                sister-project BIDS Raw, a widely-adopted community-driven effort. \tristan{Explain that
                 although BIDS doesn't rely on semantic web technologies, this is the
@@
 
+            \subsection{Data Lineage}
+                Data lineage is primarily used by Big Data Frameworks such as 
+                Spark and Dask to avoid storing large amounts data to disk for
+                fault tolerance, however, it is also an effective way to capture
+                provenance. Data lineage provides a significantly detailed 
+                provenance representation, by recording which tasks interact with
+                which data to produce results, that allows data to be recomputed 
+                in case of loss.
+
             \tristan{You might want to add data lineage here.}
@@
         \section{Provenance Tracking}\label{prov}
-            Provenance tracking is crucial to scientific reporting.
-            On its own, source code is insufficient to reproduce the workflow 
-            described in a research paper. This is because source code does 
-            not specify the specific dataset used nor the specific versions of
-            tools and the environments that they were executed on. Minor 
-            changes to any of these features can significantly affect results.
-
-            As explained by the W3 PROV specification, provenance details 
-            information on the data, the tools involved in its creation, and 
-            the responsible parties. Provenance tracking, in the context of 
-            workflows, is the act of capturing provenance data during
-            workflow execution. With detailed provenance and access to source 
-            code information, researchers should have enough information to be 
-            able to reproduce workflows. \tristan{These 2 paragraphs should be in the chapter intro.}
 
@@
             Nipype, FastR, Taverna), while others may chose to use NIDM (e.g. 
-            SPM). The 
-            remainder use a custom format to share their metadata.
+            SPM). Big Data frameworks, such as Dask and Spark utilize data lineage
+            to record their provenance. Other scienific workflow engines prefer 
+            to record provenance in a custom format.
 
@@
             activities, in
-            FastR map to Jobs and entities map to the data \tristan{how is the mapping in nipype? If different,
+            FastR map to Jobs and entities map to the data. This representation
+            differs from Nipype's representation. For example, a Nipype entity
+            may not only be the data, but also the definition of a UDF. Despite
+            W3C Prov being well-documented, definitions can still vary based on 
+            interpretation. For this reason, domain-specific metadata representations
+            are still required. \tristan{how is the mapping in nipype? If different,
             it might be a good place to comment on the fact that PROV can be used
@@
             Taverna differs from Nipype and Taverna in that it provides 
-            tracking of
-            both workflow runs and of the derived data.
-
-            Likewise to the engines that capture provenance that follows the
+            tracking of both workflow runs and of the derived data.
+
+            Similarly to the engines that capture provenance adhering the
             W3C PROV specification, all MATLAB-based workflow engines capture
@@
             and Galaxy
-            capture an above minimum amount of provenance data, whereas engines 
+            capture some provenance data, whereas engines 
             like Nextflow, Make, Toil and MapReduce capture very limited  
@@
             and BigData frameworks like Spark and Dask mainly rely on 
-            lineage \tristan{which is?}.
+            lineage \tristan{which is? -- described above}.
 
@@
             and other custom formats?}
-            An above minimum amount of provenance details captured includes 
+            At minumum, provenance tracking should include 
             capturing details on the workflow stages as well as the inputs and
-            outputs. LONI stores this type of information in an XML file.
-
-            Pegasus uses an SQL database to store its provenace. Data collected 
-            includes runtime information,
-            in addition to, data locations, software used and parameters 
-            applied to the software.
-
-            Pydpiper has no formal form of provenance tracking, but will save 
-            essential pipeline information after each completed stage to a 
-            serialized Python object. Data captured includes: the DAG, the 
+            outputs. LONI, Pegasus, Pydpiper and Galaxy all satisfy this 
+            requirement, but store the data differently. LONI stores this 
+            information in an XML file, whereas Pegasus stores the data in an 
+            SQL database and Pydpiper saves it a serialized Python object.
+
+            Additionally, the details captured may differ slightly between 
+            these frameworks. In addition to the basic amount of  provenance 
+            captured, Pegasus records runtime information, data locations, 
+            software used and parameters applied. Pydpiper captures the DAG, the 
             array of pipeline stages, the current stage counter, hash of stages 
-            and output files and an array of stage statuses.
-
-            As provenance is a major component of the Galaxy project, it 
+            and output files and an array of stage statuses.As provenance is a 
+            major component of the Galaxy project, it 
             provides comprehensive provenance tracking. This includes gathering
@@
 
-            The workflows that capture limited provenance information typically
-            restrict themselves to capturing only runtime information.
+            Some workflows engines do not really capture provenance details. At 
+            most, the provenance captured is limited to runtime information.
+            Nextflow, Make, Toil and MapReduce are all examples of these types
+            of systems.
+
             Nextflow generates an
             HTML report that contains details on execution commands, runtime, 
-            CPU usage and status.
-
-            Toil captures at most tool execution logs and provides runtime 
-            statistics.
-
-            Make is another workflow engine that does not provide built-in 
-            provenance tracking. However, Make can be adapted to capture 
-            provenance as in \note{cite make paper}.
-
-            Similarly, MapReduce also captures runtime information and relays 
-            it through a web UI.
+            CPU usage and status. In comparison, Toil captures at most tool 
+            execution logs and provides runtime 
+            statistics. Make does not even capture runtime details, but could be
+            easily configured to capture even more comprehensive provenance as
+            shown in ~\cite{10.3389/fninf.2016.00002}. MapReduce captures at 
+            most runtime details, which it makes accessible through a web UI.
 
@@
             Dask and Spark rely on their lineage to relay details on which 
-            tasks modified the data. They also provide a web UI that provides
-            additional information on runtime statistics. Methods to improve 
-            provenance tracking in Spark, particularly interactively are 
-            currently being explored \note{cite bigdebug}. Unfortunately, both
+            tasks modified the data. While Dask uses common Python structures to 
+                detail its lineage, Spark does not provide an intuitive 
+                representation of its lineage. Tools such as 
+                Titian~\cite{interlandi2018adding} improve the readability of 
+                Spark's lineage capture and  debugging of errors in 
+                Spark's interactive mode. They also provide a web UI that provides
+            additional information on runtime statistics. Unfortunately, both
             of them lack the ability to provide provenance capture of external
@@
             environment for the tools to run which mimics the environment of 
-            the original study. This ensures that users executing their 
-            workflows in the same environments will obtain the same results.
+            the original study. Workflow engines that support the encapsulation of
+            tools include: Galaxy, Toil, Nextflow, Pegasus and FastR.
+
+            All workflows, despite the framework, can theoretically be 
+            containerized. This is an important feature as custom UDFs which 
+            are not integrated as a CLI would not be containerized. Such 
+            functions also risk contributing to numerical instability of 
+            workflows. One limiting factor to this technique may be the complexity
+            associated with parallelizing the execution of a containerized 
+            workflow, particularly those using HPC schedulers, across cluster 
+            nodes.
+
+            Containerization ensures that users executing their 
+            workflows in the ``same" (containerized) environments will obtain the same results.
             A major issue with containerization is that despite it controlling
@@
 
-    \chapter{Discussion/Conclusion}
+    \chapter{Discussion}
+        While mainy neuroimaging workflow engines exist, none of them are well
+        suited for the processing of neuroimaging Big Data. Even more general
+        scientific workflows are not equipped for the handling of these datasets.
+        Currently existing workflow engines cannot be adapted for Big Data without
+        necessitating a major structural changes, as they would need to be converted
+        from compute-centric to data-centric. Furthermore, it may not be desirable
+        to reuse these frameworks even if adapting them was simple as Big Data 
+        frameworks are supported by a much larger community, contributing to 
+        more robust workflows.
+
+        Research on the efficiency of Big Data frameworks have been performed 
+        and results indicate that Big Data frameworks are effective at 
+        improving the performance of scientific workflows. This was even shown
+        to be the case when workflows were compute-intensive, as is the case 
+        with many neuroimaging frameworks. Performance of Spark and Dask is 
+        comparable, however, Dask has been found to be lose efficiency during 
+        the processing of larger datasets, whereas Spark exhibits reduced 
+        performance when significant amounts of shuffling occur. However, both
+        of these frameworks need to be adapted to better support the processing
+        of scientific data.
+
+        Effective neuroimaging workflows need to meet certain criteria: 
+        ease-of-use, performance, portability and reproducibility. In order for
+        a workflow language to be easy to use, it is recommended that it is 
+        based on simple programming languages commonly used in the field, such 
+        as Python or MATLAB (GNU/Octave). Using other programming languages 
+        would force researchers to learn a particular language to simply use the
+        framework. GUIs are ineffective solutions as they limit the flexibility
+        of the pipeline. Moreover, languages such as Python and MATLAB have a 
+        diverse set of libraries that can help with processing the data. Big
+        Data frameworks such as MapReduce, Spark and Dask all have a Python 
+        interface, making them all good candidates for ease-of-use.
+
+        Another feature relating to ease-of-use is the simplicity to integrate
+        custom tools into the workflow. Interfaces can be defined using programming
+        languages or DSLs. While programming languages provide more flexibility,
+        DSL written using markup languages can be easily shared and used by 
+        many workflow languages. Currently, MapReduce, Spark and Dask do not
+        provide any such functionality, however, libraries, such as Boutiques,
+        can manage such operations on behalf of the frameworks.
+
+        Performance is a major component to the effective processing of 
+        neuroimaging Big Data. Big Data frameworks employ strategies such as 
+        in-memory processing, data locality and lazy evaluation to improve
+        performance of processing large volumes of data. However, they are not
+        well adapted to the processing of data on HPC, particularly as HPC 
+        shared filesystems are separate from compute nodes and schedulers are 
+        locality unaware. Filesystem improvements are effective at improving 
+        performance, but make assumptions on the hardware configuration. More
+        research will need to be investigated into making Big Data frameworks
+        more performant on HPC infrastructure.
+
+        Although scientific workflow engines lack important features found in 
+        Big Data frameworks, the employ their own strategies, such as 
+        checkpointing to improve performance. Checkpointing is of particular 
+        importance to neuroimaging workflows as tasks are compute-intensive and
+        may take significant time to complete. Therefore, recomputing data after
+        loss, as Big Data frameworks do, may not be an effective solutions. Big
+        Data frameworks do not employ any checkpointing capabilities outside of
+        a given execution. They will need to be adapted to support this feature.
+
+        For workflows to be portable, their workflow engine should be platform
+        independent and accept containerized tools which can be manually 
+        fetched by the framework. This ensures that programmers can develop
+        and execute their pipelines easily on any system. Many scientific 
+        workflow engines decided to forego platform independence due to 
+        likelihood of using non-Linux computers. However, the main neuroimaging
+        platform engines, such as Nipype and PSOM maintain their 
+        platform-independence. All three Big Data frameworks are also
+        platform-independent.
+
+        Through the use of containerized tools users executing shared workflows
+        need not be concerned by all the dependency installations. Moreover, 
+        if workflow engines can share interfaces between each other, pipeline
+        developpers may not even need to create interfaces for their pipelines.
+        As scientific workflows are expected to be executed on HPC environments,
+        compatibility with Singularity containers is requires. Big Data frameworks
+        cannot natively interface command-line tools. DSL-based libraries that
+        enable interfacing with such tools can be used or a custom API for
+        the definition of interfaces for a particular Framework can be generated.
+
+        Reproducibility of workflows is currently a major concern within the 
+        scientific community. Issues resulting in significant errors may stem
+        from differences in methodologies imposed by tools performing the same
+        action or operating system differences. To ensure the validity of 
+        results, it is necessary to collect and share provenance tracking data,
+        which details each steps of the workflow such that users know which data,
+        tools, workflow engines, and operating system was used, in addition to
+        the order of operations. Of the three provenance metadata formats used 
+        in neuroimaging, BIDS derivatives is currently the format of choice due
+        to it simplicity, community-driven effort and large adoption. Big Data
+        frameworks collect provenance through data lineage. Although a good 
+        amount of provenance data is captured through lineage, it is not always
+        readable, as is the case with Spark, and lacks details on any of the
+        custom functions defined. Big Data frameworks should be adapted to use
+        a neuroimaging-related provenance representation format, such as BIDS 
+        Derivative in order to ensure it is well understood by the community.
+
+        Another important aspect to ensuring the reproducibility of workflows
+        is allowing for the incorporation of containerized tools and workflows.
+        Big Data frameworks, by default, do not provide a mechanism to interface
+        directly with containerized tools and will need to be adapted to do so,
+        or employ the use of a library to manage tool interactions, such as 
+        Boutiques. As with all workflow engines, Big Data frameworks can also 
+        be containerized.
+
+        Big Data frameworks, particularly Dask and Spark, appear well suited for
+        the processing of Big neuroimaging Data. However, these frameworks still
+        require some work in order to be effectively used for neuroimaging data.
+        Furthermore, it is not yet clear if one is more performant than the other
+        for the processing of both large images and large datasets. Additional
+        comparative evaluations between Dask and Spark will need to be performed.
 

### 2018-10-08 paper.tex
fixed erroneous text

@@
             be platform independent. These workflow engines include: Nipype,
-            PSOM, SPM, FASTR, LONI, AA, Pydpiper, Taverna, MapReduce, Spark and
+            PSOM, SPM, FASTR, LONI, AA, Pydpiper, Taverna, Toil, MapReduce, Spark and
             Dask. These workflow engines achieve platform independence through
-            running on the Java Virtual Machine (JVM) or being written in 
-            Python. Platform-independent engines that run on the JVM include: 
+            running on the Java Virtual Machine (JVM) being written in 
+            Python or CWL. Platform-independent engines that run on the JVM include: 
             MATLAB engine frameworks, such as PSOM, SPM and AA; LONI; Taverna; 
@@
             engines}. Nipype, FastR, Pydpiper are all python-based platform 
-            independent workflow engines. The remainder of the 
-            workflow engines (Pegasus, Make, OpenMOLE, Nextflow, Galaxy, and 
-            Toil are not platform-independent, but can execute on macOS and Linux.
+            independent workflow engines and Toil is CWL-based platform independent. 
+            The remainder of the 
+            workflow engines (Pegasus, Make, OpenMOLE, Nextflow and Galaxy are 
+            not platform-independent, but can execute on macOS and Linux.
 

### 2018-10-08 paper.tex
more fixes

@@
             engines}. Nipype, FastR, Pydpiper are all python-based platform 
-            independent workflow engines and Toil is CWL-based platform independent. 
+            independent workflow engines and Toil and Galaxy are CWL-based 
+            platform independent engines. 
             The remainder of the 
-            workflow engines (Pegasus, Make, OpenMOLE, Nextflow and Galaxy are 
+            workflow engines (Pegasus, Make, OpenMOLE and Nextflow are 
             not platform-independent, but can execute on macOS and Linux.

### 2018-10-09 paper.tex
Minor fixes

@@
 \usepackage{ulem}                     
+\usepackage{graphicx} 
+\usepackage{rotating}
 
@@
 
-    \begin{abstract} Text of abstract.  
+    \begin{abstract}
+
     \end{abstract} 
@@
 
-            \section{Data structures}
+            \section{Data structures}\label{ds}
 
@@
 
-        \section{Filesystems}\label{fs}
+        \section{Data-centric optimizations}\label{fs}
 
@@
             \subsection{Strategies}
-                \subsubsection{in-memory computing}
+                \subsubsection{In-memory computing}
                     Random Access Memory (RAM) is the fastest memory available
@@
             provenance metadata. Some leverage the W3C PROV specification (e.g.
-            Nipype, FastR, Taverna), while others may chose to use NIDM (e.g. 
+            Nipype, FastR, Taverna), while others may chose to use NIDM-results (e.g. 
             SPM). Big Data frameworks, such as Dask and Spark utilize data lineage
@@
 
+    \begin{sidewaysfigure}[!p]
+        \centering
+        \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{Figures/summary.pdf}
+        \caption{Summary table comparing the 17 workflow engines on four axes:
+        workflow composition, performance, portability and reproducibility.}
+    \end{sidewaysfigure} 
     \addcontentsline{toc}

### 2018-10-09 paper.tex
fixed comment on matlab engines

@@
             Dask. These workflow engines achieve platform independence through
-            running on the Java Virtual Machine (JVM) being written in 
+            running on the Java Virtual Machine (JVM), running on the MATLAB
+            engine, being written in 
             Python or CWL. Platform-independent engines that run on the JVM include: 
-            MATLAB engine frameworks, such as PSOM, SPM and AA; LONI; Taverna; 
-            MapReduce; and Spark \tristan{this list contains Matlab
-            engines}. Nipype, FastR, Pydpiper are all python-based platform 
+            LONI, Taverna, MapReduce and Spark \tristan{this list contains Matlab
+            engines}. MATLAB engine frameworks include PSOM, SPM and AA. 
+            Nipype, FastR, Pydpiper are all python-based platform 
             independent workflow engines and Toil and Galaxy are CWL-based 
@@
             array of pipeline stages, the current stage counter, hash of stages 
-            and output files and an array of stage statuses.As provenance is a 
+            and output files and an array of stage statuses. As provenance is a 
             major component of the Galaxy project, it 
@@
 
-            Some workflows engines do not really capture provenance details. At 
+            Some workflows engines capture limited provenance details. At 
             most, the provenance captured is limited to runtime information.

### 2018-10-09 paper.tex
Some edits and comments

--- 2018-10-08 paper.tex
@@
 \usepackage{ulem}                     
+\usepackage{pdfpages} 
+\usepackage{rotating}
 
@@
 \newcommand{\weird}[1]{\uwave{#1}}
+
+% Fix link colors
+\hypersetup{
+    colorlinks = true,
+    linkcolor=red,
+    citecolor=red,
+    urlcolor=blue,
+    linktocpage % so that page numbers are clickable in toc
+}
+\setcounter{tocdepth}{1}
 
@@
         BigData processing. As a result, frameworks for the efficient 
-        processing of Big neuroimaging Data need to be developed\tristan{Add a sentence to transition from 'frameworks for efficient processing' to 'workflow engines'}
+        processing of Big neuroimaging Data need to be developed.
 
@@
 
-        \tristan{Add a sentence to present the goal of the report.}
         This paper aims to evaluate the efficiency of current neuroimaging 
@@
         workflows on the basis of workflow composition: we will % 'in other words' is on my list of expressions to avoid
-        examine the programming languages necessary to construct these workflows
-        , as well as, how easy is it to add new tools. In Chapter 
+        examine the programming languages necessary to construct these workflows,
+        as well as, how easy is it to add new tools. In Chapter 
         \ref{performance}, we will proceed to 
@@
             image 
-            of the brain of a healthy 69 year-old man. At its highest resolution
-            , it is 1x1x20$\mu$m or 1TB in
+            of the brain of a healthy 69 year-old man. At its highest resolution,
+            it is 1x1x20$\mu$m or 1TB in
             size. Other examples of large images can be found in electron 
@@
 
+            \tristan{It could be useful to refer to this special issue on workflows
+            in neuroimaging in intro:
+            https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2844796/}
+            \tristan{This special issue on data sharing is also worth citing:
+            https://www.sciencedirect.com/journal/neuroimage/vol/124/part/PB}
             In contrast, large datasets consist of many images acquired from
@@
             OpenfMRI is a public repository originally conceived for the sharing
-            of task-based fMRI data. Over time, it has grown to include EEG, MEG
-            , resting-state fMRI (R-fMRI) and diffusion fMRI. Data made 
+            of task-based fMRI data. Over time, it has grown to include EEG, MEG,
+            resting-state fMRI (R-fMRI) and diffusion fMRI. Data made 
             available through
@@
             developed by these initiatives made use of popular neuroimaging 
-            toolboxes, such as FSL~\cite{SMITH2004S208}, FreeSurfer, ANTS, 
+            toolboxes, such as FSL~\cite{SMITH2004S208} \tristan{You should cite it the first time you mention it}, FreeSurfer, ANTS, 
             CIVET, rather than
@@
             HPC infrastructure was initially designed for the processing of 
-            compute-intensive applications. As a result, \tristan{I disagree with 'As a result' and I don't particularly agree with the previous sentence: even if
-            resources aren't scarce, they are good reasons to optimize resource managers: energy consumption, application makespan, fairness, etc. Also,
-            I'm not sure if resources are less scarce now than they used to be: with the rise of computational sciences, almost every researcher now
-            needs a cluster, and Compute Canada (but also other infrastructures) has been very busy.}
+            compute-intensive applications. As a result,
             resource managers for these systems employ
@@
             filesystem (distinct from any compute node), such as Lustre, and 
-            data-unaware schedulers. \tristan{and data-unaware schedulers?}.
-            \tristan{You should say that this definition is 'by contrast to big data platforms' to explain where it's coming from.}
+            data-unaware schedulers.
             The shared parallel filesystem is the slowest of
@@
 
-            Big Data infrastructure \tristan{Call it 'Big Data infrastructure?'} differs from HPC infrastructures in that
+            Big Data infrastructure differs from HPC infrastructures in that
             it uses less-costly hardware. It follows a shared-nothing paradigm
@@
         Workflows, also known as pipelines, describe the flow of data from one 
-        activity to another and are typically represented by a graph. \tristan{Not necessarily a DAG, see Nipype for instance, it has loops.
-        I don't think you want to enter in a taxonomy of languages though, just say 'graph'}. Workflow activities, also referred to as 
-        workflow interfaces or stages, encapsulate a tool through the use of an
-        interface\tristan{Mention 'interface' here as an intro to sub-sec 2}. 
-        Tools can be represented in many forms, including: command-line tools, 
-        custom scripts integrated directly into the workflow code, web 
-        services and software libraries.\tristan{A script is a command-line tool. I think tools may also be web services (Taverna) or just 
-        software libraries (Dask)}. A workflow activity with all its 
+        activity to another and are typically represented by a graph. Workflow activities, also referred to as 
+        workflow interfaces or stages, encapsulate a tool through the 
+        use of an interface. Tools can be represented in many forms, 
+        including: command-line tools, custom scripts integrated 
+        directly into the workflow code, web services and software 
+        libraries. A workflow activity with all its 
         dependencies is considered ready to be staged.
@@
         \tristan{Perhaps mention that this also relates to the sharing
-        of reproducible results -- VH: not sure if this is what you meant}.
+        of reproducible results -- VH: not sure if this is what you meant --
+        yeah, we might want to explicit this further in the paper but
+        good enough for now}.
 
@@
             affects how easily it will get picked up by the community. Complex
-            languages may mean greater efficiency of engines
-            , but also result in extensive time spent on learning and developing
+            languages may mean greater efficiency of engines,
+            but also result in extensive time spent on learning and developing
             the pipelines. Moreover, some researchers may not be well-versed in
@@
             (e.g. machine-learning workflows) or workflows with conditionals
-            (e.g. given a certain input, do not execute this activity) cannot
+            (e.g. given a certain input, do not execute this activity) can hardly
             be represented by a GUI. 
@@
             SPM, Taverna, OpenMOLE, LONI and Galaxy.
-            \tristan{My conclusion on GUIs is that they are suitable only
-            for very simple pipelines. Perhaps the intro could give an idea
-            of the types of pipelines encountered, showing a simple one and
-            a more complex one. No way fmriprep is described in a GUI.}
 
@@
             languages are used for describing scientific workflows: 1) Python, 
-            2) MATLAB and GNU/Octave \tristan{I think the official name is GNU/Octave}, 3) Shell Scripting languages \tristan{scripting languages}, 4) Domain Specific Language (DSL).
-\tristan{I don't see why CWL and XML don't qualify as DSL} Workflows engines that
+            2) MATLAB and GNU/Octave, 3) Shell Scripting languages, 4) Domain Specific Language (DSL). Workflows engines that 
             employ usage of Python include: Nipype, FastR, Pydpiper, Pegasus, Toil,
-            MapReduce, Dask and Spark. \tristan{Dask? Spark?}.
+            MapReduce, Dask and Spark.
             Python is an excellent choice as a programming language for 
@@
             and more general, but heavily used in neuroimaging, libraries such
-            as SciPy~\cite{scipy} and Scikit Learn~\cite{pedregosa2011scikit} \tristan{mention scipy, sci-kit learn, etc}
-            , and an lower learning curve compared to other 
+            as SciPy~\cite{scipy} and Scikit Learn~\cite{pedregosa2011scikit},
+             and an lower learning curve compared to other 
             programming languages. Additionally, Python is platform independent
             meaning it can be installed on any system. In terms of BigData 
-            frameworks that employ Python, Dask is exclusively written in Python
-            , whereas Hadoop MapReduce and Spark both have interfaces to Python
+            frameworks that employ Python, Dask is exclusively written in Python,
+            whereas Hadoop MapReduce and Spark both have interfaces to Python
             while being written in Java and Scala, respectively.
@@
             downfall to imposing MATLAB as programming language is that it is proprietary, thus imposing a fee on 
-            anyone wishes to program in that language or have access to their diverse array of libraries \tristan{You should also mention the
-            advantage of Matlab: easy to learn, integrated ecosystem, 
-            performance, see 
-            \url{https://www.mathworks.com/products/matlab/matlab-vs-python.html}}. 
+            anyone wishes to program in that language or have access to their diverse array of libraries. 
             To offset this, PSOM ensure compatibility with the GNU 
@@
 
-            \tristan{This paragraph should start with ``Shell scripting ...". The structuration of this section by language
-            is great, but the writing of this paragram doesn't put it forward.}
             Many researchers have been exposed to Shell scripting languages as 
@@
             Domain Specific Languages (DSL) are also used to compose workflows.
-            The can be made up of any language (e.g. XML, CWL, Scala and Groovy) that has been adapted for use in 
+            They can be made up of any language (e.g. XML, Scala and Groovy) that has been adapted for use in 
             a specific domain.
@@
             workflow engines such as Pegasus and LONI. Using a markup language
-            such as XML have the advantage of having access to 
+            such as XML has the advantage of having access to 
             language-independent parsers and validators. However, the use of a 
-            markup language removes some of the flexibility. \tristan{Before you explain the downside, 
-            you should say what are the advantages: mainly language-independent parsers and validation.} Users wanting to 
+            markup language removes some of the flexibility. Users wanting to 
             include 
@@
 
-            Another example of a DSL is a Common Workflow
+            Another example of a DSL is the Common Workflow
             Language (CWL). CWL can be seen in workflow engines like Toil. 
             It is similar to XML in that its main difference is that it is 
-            YAML/JSON-based.
+            YAML/JSON-based. \tristan{Review the previous sentence, it doesn't make sense. And the end of the paragraph is till fuzzy too}
             \tristan{It is a YAML-based DSL for workflows. Just like Pegasus' DAX is an XML-based DSL}. Files can either be written in JSON or YAML formats. CWL is 
@@
             Programming language-based DSL workflow engines include OpenMOLE, which is  
-            derived from Scala, and Nextflow \tristan{Nextflow?}, which is derived from the Apache 
+            derived from Scala, and Nextflow, which is derived from the Apache 
             Groovy programming language. While both are written on top of the 
-            Java platform, Nextflow is only compatible with Linux and OSX \tristan{How about using Linux everywhere instead of UNIX-based?} 
+            Java platform, Nextflow is only compatible with Linux and OSX \tristan{How about using Linux everywhere instead of UNIX-based? You still have UNIX here and there.} 
             environments. These programming languages are not otherwise commonly
@@
             With the exeption of LONI, neuroimaging workflows tend to stick to
-            commonly used programming languages such as Python and MATLAB/Octave
-            . This is likely due to the fact that the learning curve to adapt to
+            commonly used programming languages such as Python and MATLAB/Octave.
+            This is likely due to the fact that the learning curve to adapt to
             these languages is relatively low and users can harness the 
@@
                 The data structures made available by workflow engines dictate
-                the maximum performance available to a workflow. Scientific
+                the maximum performance available to a workflow. \tristan{I don't understand the previous sentence, 
+                in particular the 'maximum' idea. You should review that.} Scientific
                 workflows are generally file-based, meaning that data, in the 
@@
                 issues by being a shared-memory structure. They 
-                are immutable and only permit coarse-grained edits to ensure 
+                are immutable and only permit coarse-grained edits, to ensure 
                 fault-tolerance. The RDDs shared memory enables a reduction in
@@
                 Pandas DataFrame~\cite{mckinney2011pandas}) to process the 
-                data. Although Spark does not have a datastructure for large 
+                data \tristan{I think Dask provides clones of these libraries, but you can't use
+            them directly}. Although Spark does not have a datastructure for large 
                 matrices, it has an RDD-based API, known as \textit{linealg}
@@
 
-            \tristan{So this is now about data structures, not just language.
-            This is really excellent, but this should perhaps go into a 
-            specific sub-section. I think you could mention that 
-            scientific workflow engines traditionally rely on files 
-            only, and simple types, but that big data engines have 
-            invested a lot in data structures, starting from key-value 
-            pair representations in MR to RDD and Dask arrays. Spark is 
-            actually built around RDDs, this is salient-enough to be 
-            discussed a bit more. It actually makes a lot of sense that 
-            data-focused engines are built around data structures 
-            instead of control constructs. The ACM Com Spark paper makes it clear
-            that unifying data structures is key to performance and interoperability, 
-            because it allows data to remain in-memory instead of having to go back
-            and forth to different environments. I would only mention linealg 
-            in the description of Dask arrays. So, how about a 
-            sub-section 'data structures' with 4 paragraphs: (1) 
-            scientific and neuroimaging engines, (2) Big-Data, MR, (3) 
-            Spark's RDDs and DataFrames, (4) Dask arrays, mentioning 
-            linealg. }
-
         \section{Tool interfaces}\label{mod}
-            \tristan{Since this section is about interfaces, I would
-            rather call it ``Interfaces" or ``Tool interfaces".}
             Often times, users will want to integrate existing tools into their 
@@
             a complex class for interface definition and validation. Recent
-            Workflow engines take a modular approach and promote sharing of data
-            and pipelines. Thetypically rely on more complex types 
-            of interfaces in order to ensure correct functioning of the pipeline
-            . Since these interfaces can be rather complex, it is necessary for the 
+            workflow engines take a modular approach and promote sharing of data
+            and pipelines. They typically rely on more complex types 
+            of interfaces in order to ensure correct functioning of the pipeline.
+            Since these interfaces can be rather complex, it is necessary for the 
             workflow engine to provide an intuitive way for users to create 
@@
             languages for the 
-            description of their workflow interfaces include: Nipype, PSOM
-            , Pydpiper, Toil, OpenMOLE and Make interfaces.
+            description of their workflow interfaces include: Nipype, PSOM,
+            Pydpiper, Toil, OpenMOLE and Make interfaces.
 
@@
             website}.
+            \tristan{Expand UDF, CLI, CARE.}
 
@@
 
-            The Make pipeline is expressed in a makefile containing a set of 
+            The Make pipeline is expressed in a Makefile containing a set of 
             rules (Tool interfaces). Rules consist of three components: targets, 
@@
 
-            \tristan{Users can still integrate tools as Matlab functions
-            doing system calls}.
             Although DSL-based workflow interface schemas may be less versatile
-            than those written using programming languages, DSL based schemas
+            than those written using programming languages, they
             can be reused by any engine. This allows interface schemas to be 
             portable. AA, LONI, FastR, Taverna, Pegasus and Galaxy all 
-            implement their iterface definition using a DSL.
+            implement their interface definition using a DSL.
 
@@
             addition to UDFs and CLIs, Pegasus
-            can also execute containerized tools \tristan{how about the other ones?}. For this, a container 
+            can also execute containerized tools. For this, a container 
             identifier, image URL, site identifier of the image, and container
@@
 
-            \tristan{Good material in here, but it lacks a bit structure.
-            I like your structure of the previous sub-section, by language. I think
-            here you could have two groups: (1) DSL descriptions (XML, JSON, etc) and (2)
-            Python or Matlab descriptions. One advantage of DSLs is that they can 
-            be more easily reused in different engines, while programming languages
-            are more flexible, for instance Nipype allows outputs to be specified
-            at runtime (dynamic number of inputs with names only known after
-            the interface ran.)}
-
     \chapter{Performance}\label{performance}
@@
         workflows to BigData frameworks on the basis of performance. 
-        Section~\ref{fs} will look at what filesystems are supported \tristan{supported?} by the 
+        Section~\ref{fs} will look at what filesystems are supported by the 
         engines. This is important because data movement of any kind is costly,
@@
         typical neuroinformatics pipelines manipulate files, and therefore, 
-        reducing the amount of I/O is necessary.\tristan{Also, mention that neuroinformatics
-        pipelines manipulate files.}
+        reducing the amount of I/O is necessary.
         Section~\ref{sched} will provide an overview of the schedulers 
@@
             Although MapReduce, Spark and Dask are all data-intensive worflows,
-            Dask can be more similar to that of scientific workflow, 
-            particularly in a multiprocessing environment. It focused on the 
+            Dask can be more similar \weird{to that of} scientific workflow, 
+            particularly in a multiprocessing environment. It focuses on the 
             efficient 
@@
 
-                    As data is not stored within a resilient distributed 
+                    As data is not always stored within a resilient distributed 
                     filesystem such as Lustre and HDFS, it risks being 
@@
 
-                    Lazy evaluation is a strategy used throughout computer 
-                    science to improve performance of operations. Rather than
+                    Lazy evaluation is a strategy used throughout programming languages
+                    to improve performance of operations. Rather than
                     evaluating a expression immediately, it will wait until it 
@@
                     task order of execution and further save on data transfers.
-                    \tristan{And further save on data transfers}
 
@@
                 replicates the data may burden the storage layer only to 
-                providea feature which already exist. This is especially 
+                provide a feature which already exists. This is especially 
                 problematic as storage on HPC compute nodes may be limited.
@@
                 replicate across the network for fault tolerance. However,
-                unlike Spark applications, Alluxio is meant to run indefinitely
-                , making it unfavourable in some cases, to recompute the data
+                unlike Spark applications, Alluxio is meant to run indefinitely,
+                making it unfavourable in some cases, to recompute the data
                 from its starting point. For this, Alluxio evaluates the 
@@
             frameworks such as MapReduce and Spark are limited to Big Data 
-            schedulers such as Yet Another Resource Negotiator (YARN) and MESOS.
+            schedulers such as Yet Another Resource Negotiator (YARN) and Mesos.
             The reason MapReduce and Spark do not leverage HPC resource managers
@@
             data and cannot perform any form of lazy evaluation.
-            In addition to YARN and MEsos, Spark provides its own standalone 
+            In addition to YARN and Mesos, Spark provides its own standalone 
             scheduler. The standalone scheduler is simple providing only a 
@@
             locality using a shared HDFS instance will remain an issue.
-            \tristan{The section is interesting, but it lacks a bit of context: one would like to 
-            understand why Spark and Hadoop developed their own scheduler instead of
-            reusing the ones available on HPC. And what are the main characteristics of these
-            schedulers: data locality should be mentioned here.}
 
@@
             \note{what about SPM, FastR??}
-
-            \tristan{Mentioning fault tolerance would be wise too.}             
-            \tristan{How about: burst buffers, pilot jobs, deployments of Spark on HPC?}
 
@@
             MATLAB engine frameworks, such as PSOM, SPM and AA; LONI; Taverna; 
-            MapReduce; and Spark \tristan{this list contains Matlab
-            engines}. Nipype, FastR, Pydpiper are all python-based platform 
+            MapReduce; and Spark \tristan{you don't have to mention that Matlab has a JVM}. Nipype, FastR, Pydpiper are all Python-based platform 
             independent workflow engines and Toil and Galaxy are CWL-based 
@@
             Engines built on top of a platform-independent language do not 
-            necessarily imply platform independence \tristan{Grammar? Engines ... does not}. For instance, OpenMOLE is 
+            necessarily imply platform independence. For instance, OpenMOLE is 
             built in Scala and Netflow runs on Groovy, yet neither are platform
@@
 
-            Other workflow engines do not use an form of containerization to
+            Other workflow engines do not use any form of containerization to
             ensure that workflows can be shared easily amongst users. These 
@@
         Without reproducibility, correctness of results could never be 
-        determined \tristan{Link between two first sentences could be smoothened}. Reproducibility is defined as the ability to obtain similar 
+        determined. Reproducibility is defined as the ability to obtain similar 
         results by 
@@
         enable the tracing of results. For this reason, many workflow engines have
-        incorporated provenance tracking into their design. \tristan{Provenance is also
-        a way to merely document what happened so that results can be traced.}
+        incorporated provenance tracking into their design.
 
@@
             code information, researchers should have enough information to be 
-            able to reproduce workflows. \tristan{These 2 paragraphs should be in the chapter intro.}
+            able to reproduce workflows.
 
@@
         different provenance specifications used in neuroinformatics (W3C PROV, NIDM,
-        and BIDS Derivates) \tristan{I don't think you want to talk about 'metadata'. Just
-        mention 'provenance'.}. Section~\ref{prov} will discuss how the 
+        and BIDS Derivates). Section~\ref{prov} will discuss how the 
         engines assimilate and distribute the provenance data. 
@@
             popular neuroimaging metadata specifications, Brain Imaging Data Structure (BIDS); 
-            and data lineage, the Big Data solution to provenance tracking \tristan{Make sure that
-            the acronym is defined earlier.}.
+            and data lineage, the Big Data solution to provenance tracking.
 
@@
             origin of data shared on the Web. 
-            The aim of the W3 PROV specification was to \tristan{to?} gather information on
+            The aim of the W3 PROV specification was to gather information on
             the activities, entities and involved parties in the creation of 
@@
                 Clinical and Experimental Data Exchange (XCEDE) neuroimaging
-                provenance metadata format~\cite{Maumet:2016aa} \tristan{I guess you'll cite references for that}. It is a domain-specific extension
+                provenance metadata format~\cite{Maumet:2016aa}. It is a domain-specific extension
                 to the W3C PROV model that aims to provide comprehensive and
-                machine-readable provenance from raw data to its final stage. \tristan{give example? -- vhs: like a figure?}
+                machine-readable provenance from raw data to its final stage. \tristan{give example? -- vhs: like a figure? No, like what can be described which couldn't in BIDS. 
+                Forget about it for the comp exam, keep this note for the paper.}
                 However, it is currently limited in the provenance 
@@
                 environments. It is also limited by the fact that consensus lacks
-                within the community on paradigms. \tristan{Be more specific, what are the limitations?}
+                within the community on paradigms. \tristan{I disagree with the last sentence
+                but this is fine. The main limitation imo is due to the use of RDF and an ontology: it is
+                very powerful but difficult to use: performance sucks, queries are hell to write, 
+                and format is not human-readable.}
 
@@
                 This is primarily a result of its simplicity, as seen in its 
-                sister-project BIDS Raw, a widely-adopted community-driven effort. \tristan{Explain that
-                although BIDS doesn't rely on semantic web technologies, this is the
-                solution being adopted in the community, mostly for simplicity. Mention
-                that there is a community effort around it and that bidsraw has been
-                very successful.}
+                sister-project BIDS Raw, a widely-adopted community-driven effort.
 
@@
 
-            \tristan{You might want to add data lineage here.}
-
         \section{Provenance Tracking}\label{prov}
@@
             interpretation. For this reason, domain-specific metadata representations
-            are still required. \tristan{how is the mapping in nipype? If different,
-            it might be a good place to comment on the fact that PROV can be used
-            in many different ways}.
+            are still required.
 
@@
             and BigData frameworks like Spark and Dask mainly rely on 
-            lineage \tristan{which is? -- described above}.
+            lineage.
 
@@
             it lacks structure. Perhaps you could classify provenance tracking in PROV-based, lineage-based,
-            and other custom formats?}
+            and other custom formats? -- It is still a bit 'boring' :) No worries for the comp report,
+            but could be discussed and fixed in the paper.}
             At minumum, provenance tracking should include 
@@
             versions and further limit scientific advancement.
-            \tristan{Nextflow should be mentioned here, as they make a big deal
-            of using containers for reproducibility. I would structure this section in 
-            2 paragraphs: (1) what engines support containers, (2) why containers help
-            reproducibility and why there aren't enough.}
 
     \chapter{Discussion}
+    \tristan{You should refer to figure 1 in the text. You could start your discussion by mentioning that
+    a summary of your review is available in Figure 1.}
         While mainy neuroimaging workflow engines exist, none of them are well
-        suited for the processing of neuroimaging Big Data. Even more general
+        suited for the processing of neuroimaging Big Data \tristan{due to? be more specific}. Even more general
         scientific workflows are not equipped for the handling of these datasets.
@@
         with many neuroimaging frameworks. Performance of Spark and Dask is 
-        comparable, however, Dask has been found to be lose efficiency during 
+        comparable, however, Dask has been found \weird{to be lose efficiency} during 
         the processing of larger datasets, whereas Spark exhibits reduced 
@@
         if workflow engines can share interfaces between each other, pipeline
-        developpers may not even need to create interfaces for their pipelines.
+        developers may not even need to create interfaces for their pipelines.
         As scientific workflows are expected to be executed on HPC environments,
-        compatibility with Singularity containers is requires. Big Data frameworks
+        compatibility with Singularity containers is required. Big Data frameworks
         cannot natively interface command-line tools. DSL-based libraries that
         enable interfacing with such tools can be used or a custom API for
-        the definition of interfaces for a particular Framework can be generated.
+        the definition of interfaces for a particular framework can be generated.
 
@@
 
+    \clearpage
+    \begin{sidewaysfigure}[!p]
+        \caption{Summary table comparing the 17 workflow engines on four axes:
+        workflow composition, performance, portability and reproducibility.}
+        \includepdf[landscape=true]{Figures/summary.pdf}
+    \end{sidewaysfigure} 
     \addcontentsline{toc}

### 2018-10-09 paper.tex

@@
 \usepackage{ulem}                     
-\usepackage{pdfpages} 
+\usepackage{graphicx} 
 \usepackage{rotating}
@@
             Dask. These workflow engines achieve platform independence through
-            running on the Java Virtual Machine (JVM) being written in 
+            running on the Java Virtual Machine (JVM), running on the MATLAB
+            engine, being written in 
             Python or CWL. Platform-independent engines that run on the JVM include: 
-            MATLAB engine frameworks, such as PSOM, SPM and AA; LONI; Taverna; 
-            MapReduce; and Spark \tristan{you don't have to mention that Matlab has a JVM}. Nipype, FastR, Pydpiper are all Python-based platform 
+            LONI, Taverna, MapReduce and Spark. MATLAB engine frameworks include PSOM, SPM and AA. 
+            Nipype, FastR, Pydpiper are all Python-based platform 
             independent workflow engines and Toil and Galaxy are CWL-based 
@@
 
-    \clearpage
     \begin{sidewaysfigure}[!p]
+        \centering
+        \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{Figures/summary.pdf}
         \caption{Summary table comparing the 17 workflow engines on four axes:
         workflow composition, performance, portability and reproducibility.}
-        \includepdf[landscape=true]{Figures/summary.pdf}
     \end{sidewaysfigure} 

### 2018-10-09 paper.tex
abstract

--- 2018-10-09 paper.tex
@@
     \begin{abstract}
+        The rate of data aquisition coupled with the volume of data obtained 
+        from datasharing initiatives has lead to unprecedented amount of 
+        neuroimaging data. This surge of open neuroimaging data is of great 
+        importance to scientific advancement, however, it is currently 
+        inaccessible to many research labs due to limited infrastructure and 
+        lack of software available to faciliate processing. Despite the existance
+        of many neuroimaging workflow engines, none of them are well adapted for
+        the processing of neuroimaging Big Data. Generalized Big Data frameworks
+        exist, however they may lack crucial aspects necessary to process neuroimaging
+        data. Here we evaluate 7 neuroimaging workflow engines (Nipype, PSOM, SPM, FastR,
+        LONI, Automatic Analysis and Pydpiper with 7 other scientific workflow
+        engines (Taverna, Pegasus, Make, OpenMOLE, Nextflow, Galaxy and Toil) and
+        3 popular Big Data frameworks (Hadoop MapReduce, Apache Spark and Dask) 
+        against 4 different criteria: Workflow composition, Performance, Portability
+        and Reproducibility. Scientific workflow engines are found to lack the 
+        inherent qualities required to be successful scientific Big Data workflow 
+        engines. However, Big Data frameworks are found exhibit features crucial
+        to the processing of neuroimaging Big Data, although may lack some features
+        necessary for the processing of neuroimaging data, such as data checkpointing and 
+        standardized provenance tracking formats. Adaptations required to make Big
+        Data frameworks compatible with the processing of neuroimaging Big Data 
+        appear less significant than adaptation required to make current neuroimaging 
+        workflows efficient in the processing of Big Data, thus making Big Data frameworks
+        such as Dask and Spark good contenders for the processing of neuroimaging Big Data.
 

### 2018-10-09 paper.tex
added space in big data and fixed title page

@@
             and open processing of Big neuroimaging Data} 
-    \author{Valerie Hayot-Sasson}
+    \author{Valerie Hayot-Sasson \\
+    PhD Comprehensive Examination\\
+    Departement of Computer Science and Software Engineering \\
+    Concordia University \\
+    Montreal, Quebec
+    }
+    \date{\today}
     \maketitle
@@
         Performance Computing (HPC) infrastructure, software configured to 
-        efficiently use resources is required to make execution time manageable
-        . Ensuring efficient processing of data is 
+        efficiently use resources is required to make execution time manageable. 
+        Ensuring efficient processing of data is 
         complex and limits accessibility to researchers with little programming 
@@
         resource managers for performance and are therefore not well suited for
-        BigData processing. As a result, frameworks for the efficient 
+        Big Data processing. As a result, frameworks for the efficient 
         processing of Big neuroimaging Data need to be developed.
@@
         This paper aims to evaluate the efficiency of current neuroimaging 
-        frameworks in comparison to BigData frameworks and to determine the 
-        extent to which BigData frameworks must be adapted to be effective
+        frameworks in comparison to Big Data frameworks and to determine the 
+        extent to which Big Data frameworks must be adapted to be effective
         neuroimaging workflow engines. 
@@
             neuroimaging
-            BigData has come to fruition. The field of neuroscience has now
+            Big Data has come to fruition. The field of neuroscience has now
             reached a bottleneck where there exists sufficient data to make
@@
         (Taverna, Pegasus, Make, OpenMOLE, Nextflow, Galaxy, Toil), and three
-        BigData frameworks (MapReduce, Spark, Dask). We will use two axes to
+        Big Data frameworks (MapReduce, Spark, Dask). We will use two axes to
         evaluate workflow composition. Section~\ref{lang} will compare the 
@@
         will then examine the differences in data structures between scientific
-        and BigData frameworks. 
+        and Big Data frameworks. 
         Section~\ref{mod} will look at how easy it is to integrate existing 
@@
             programming languages. Additionally, Python is platform independent
-            meaning it can be installed on any system. In terms of BigData 
+            meaning it can be installed on any system. In terms of Big Data 
             frameworks that employ Python, Dask is exclusively written in Python,
@@
                 strings. Due to workflow engines being file-based, performance
-                is limited, particularly with BigData.
-
-                In order to improve performance, BigData frameworks introduced
+                is limited, particularly with Big Data.
+
+                In order to improve performance, Big Data frameworks introduced
                 data-centric data structures. MapReduce is key-value based, 
@@
 
-            Unlike scientific workflow engines, BigData frameworks do not 
+            Unlike scientific workflow engines, Big Data frameworks do not 
             provide default interfaces for the execution of complex tools 
@@
             defined within a JSON schema. Furthermore, the library's richness
-            compete with Galaxy's XML. Python-based BigData frameworks can 
+            compete with Galaxy's XML. Python-based Big Data frameworks can 
             easily call on this library from within the workflow in order to
@@
         performance has become especially critical now that neuroimaging
-        BigData has emerged. The dichotomy remains between scientific 
-        workflows and BigData workflows. While scientific workflows have made
+        Big Data has emerged. The dichotomy remains between scientific 
+        workflows and Big Data workflows. While scientific workflows have made
         advances towards more performant engines, they still lack the 
-        strategies BigData frameworks employ to reduce data transfers.
+        strategies Big Data frameworks employ to reduce data transfers.
 
         Another notable aspect to consider is that both types of engines 
-        (scientific and BigData) are tightly coupled to the infrastructure
+        (scientific and Big Data) are tightly coupled to the infrastructure
         they were intended to be used on. Scientific workflow engines
         try to remain efficient in their handling of compute resources since such
-        resources were previously scarce. In contrast, BigData frameworks 
+        resources were previously scarce. In contrast, Big Data frameworks 
         make efficient use of commodity clusters, where data transfers are 
-        burdensome, particularly when faced with BigData.
+        burdensome, particularly when faced with Big Data.
 
         As some researchers are trying to shift towards the use of commodity 
-        infrastructure while others are trying to process BigData on HPC, the 
+        infrastructure while others are trying to process Big Data on HPC, the 
         ideal workflow engine must run efficiently regardless of 
@@
         In this Chapter, we will compare scientific
-        workflows to BigData frameworks on the basis of performance. 
+        workflows to Big Data frameworks on the basis of performance. 
         Section~\ref{fs} will look at what filesystems are supported by the 
@@
             scheduling of tasks and relies on I/O and network transfers to
-            communicate the data between nodes. With BigData workflows, such
+            communicate the data between nodes. With Big Data workflows, such
             operations could incur significant performance penalties. 
             To mitigate this issue and remain efficient in the face of
-            BigData, Spark and Dask implement data locality, lazy evaluation 
+            Big Data, Spark and Dask implement data locality, lazy evaluation 
             and in-memory processing. Data locality is the scheduler's ability
@@
             locality to ensure limit data transfers across the network.
-            For the efficient processing of neuroimaging BigData, it will be
+            For the efficient processing of neuroimaging Big Data, it will be
             necessary for scientific worflow engines to incorporate these 
@@
                     Random Access Memory (RAM) is the fastest memory available
-                    to workflow engines. BigData frameworks, such as Spark and 
+                    to workflow engines. Big Data frameworks, such as Spark and 
                     Dask, leverage this feature by maintaining data in memory 
@@
                 \subsubsection{Data locality}
-                    Data locality, with respect to BigData frameworks, is a
+                    Data locality, with respect to Big Data frameworks, is a
                     scheduling
@@
                     necessary data is located. It was one of the first main 
-                    strategies implemented to make BigData processing more 
+                    strategies implemented to make Big Data processing more 
                     efficient. The reasoning behind data locality is that 
@@
 
-            \subsection{BigData Filesystem optimizations for HPC}
-
-                Writting BigData to Lustre is very costly compared to local 
+            \subsection{Big Data Filesystem optimizations for HPC}
+
+                Writting Big Data to Lustre is very costly compared to local 
                 node storage. Distributed filesystems found on HPC 
                 infrastructure provide their own form of fault-tolerance, 
-                therefore, adding a BigData filesystem such as HDFS which 
+                therefore, adding a Big Data filesystem such as HDFS which 
                 replicates the data may burden the storage layer only to 
@@
                 problematic as storage on HPC compute nodes may be limited.
-                Moreover, while BigData frameworks typically execute on a 
-                dedicated BigData cluster, they will be on a shared cluster in
+                Moreover, while Big Data frameworks typically execute on a 
+                dedicated Big Data cluster, they will be on a shared cluster in
                 HPC, with many processed trying to access the memory. The 
-                following are two optimizations made for the storing of BigData
+                following are two optimizations made for the storing of Big Data
                 .
@@
 
-        \section{BigData Scheduling}\label{sched}
+        \section{Big Data Scheduling}\label{sched}
 
@@
             their own form of scheduling for computations on a single node. 
-            BigData frameworks, such as Dask,
+            Big Data frameworks, such as Dask,
             also provide interfaces to these HPC schedulers. However, other
@@
             basic scheduler. It allows users to use Spark out-of-the-box. Similarly
-            to scientific workflows, all three BigData frameworks provide their
+            to scientific workflows, all three Big Data frameworks provide their
             own schedulers for single-node scheduling.
@@
             amount of time to execute, a mechanism to checkpoint workflows      
-            will need to be added to any BigData framework used for             
+            will need to be added to any Big Data framework used for             
             neuroinformatics.                                                   
@@
 
-        \section{Evaluation of BigData Frameworks for Scientific Computing}\label{eval}     
-            The BigData frameworks have sparked significant interest within the 
+        \section{Evaluation of Big Data Frameworks for Scientific Computing}\label{eval}     
+            The Big Data frameworks have sparked significant interest within the 
             scientific community, leading to research on the efficiency of      
@@
             comprised of long tasks, it                                         
-            is crucial that BigData frameworks scale well with tasks of longer  
+            is crucial that Big Data frameworks scale well with tasks of longer  
             durations.
 
-            To determine how different BigData frameworks compare in their 
+            To determine how different Big Data frameworks compare in their 
             efficiency for the processing of scientific workflows, 
@@
             occuring between map and shuffle phases. Overall it was
-            found that BigData Frameworks still required some adaptations to be
+            found that Big Data Frameworks still required some adaptations to be
             effectively used as scientific workflow engines.
@@
             tasks using CARE to simulate runtime environments
-            and BigData frameworks like Spark and Dask mainly rely on 
+            and Big Data frameworks like Spark and Dask mainly rely on 
             lineage.

### 2018-10-09 paper.tex
final modifications

@@
 
-            \tristan{It could be useful to refer to this special issue on workflows
-            in neuroimaging in intro:
-            https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2844796/}
-            \tristan{This special issue on data sharing is also worth citing:
-            https://www.sciencedirect.com/journal/neuroimage/vol/124/part/PB}
+            %\tristan{It could be useful to refer to this special issue on workflows
+            %in neuroimaging in intro:
+            %https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2844796/}
+            %\tristan{This special issue on data sharing is also worth citing:
+            %https://www.sciencedirect.com/journal/neuroimage/vol/124/part/PB}
             In contrast, large datasets consist of many images acquired from
@@
             . At present, OpenfMRI houses data of a total 3372 subjects
-            originating from 95 studies \note{obtained from website}.
+            originating from 95 studies.
 
@@
 
-            The Consortium for Reliability and Reproducibility 
-            \tristan{expand acronym} was developed as an initiative to study 
+            The Consortium for Reliability and Reproducibility (CoRR) was developed as an initiative to study 
             and improve the 
@@
             subjects, in which anatomical, R-fMRI, dMRI, CBF and ASL scans are 
-            available \note{obtained from website}.
+            available.
 
@@
             developed by these initiatives made use of popular neuroimaging 
-            toolboxes, such as FSL~\cite{SMITH2004S208} \tristan{You should cite it the first time you mention it}, FreeSurfer, ANTS, 
+            toolboxes, such as FSL~\cite{SMITH2004S208}, FreeSurfer, ANTS, 
             CIVET, rather than
@@
         more than likely that easily understood workflows are reproducible. 
-        \tristan{Perhaps mention that this also relates to the sharing
-        of reproducible results -- VH: not sure if this is what you meant --
-        yeah, we might want to explicit this further in the paper but
-        good enough for now}.
+        %\tristan{Perhaps mention that this also relates to the sharing
+        %of reproducible results -- VH: not sure if this is what you meant --
+        %yeah, we might want to explicit this further in the paper but
+        %good enough for now}}.
 
@@
             Many researchers have been exposed to Shell scripting languages as 
-            several scientific workflows are composed in UNIX-based 
+            several scientific workflows are composed in Linux and OSX-based 
             environments. These environments typically have Make installed by
@@
             computing.
-            As Make is exclusive to UNIX environments, it is not 
+            As Make is exclusive to Linux environments, it is not 
             platform-independent. However, Make has the flexibility of being
@@
 
-            Another example of a DSL is the Common Workflow
-            Language (CWL). CWL can be seen in workflow engines like Toil. 
-            It is similar to XML in that its main difference is that it is 
-            YAML/JSON-based. \tristan{Review the previous sentence, it doesn't make sense. And the end of the paragraph is till fuzzy too}
-            \tristan{It is a YAML-based DSL for workflows. Just like Pegasus' DAX is an XML-based DSL}. Files can either be written in JSON or YAML formats. CWL is 
-            based on workflow engines like Make \tristan{? I don't think CWL relies on make, does it?}, but differs in that tasks
-            are isolated \tristan{inputs and outputs are somewhat explicit in a Makefile}. Similarly to 
-            make, CWL is only available on Linux and OSX platforms \tristan{strange, the CWL interpreter is in Python so I'd expect it to run anywhere}. \note{cite cwl 
-            website}
+            Another example of a DSL is the YAML/JSON-based Common Workflow
+            Language (CWL). CWL can be seen in workflow engines like Toil and Galaxy.
+            CWL is similar to Make, however, it differs from Make in the tasks are 
+            isolated. CWL-based workflow engines are platform-independent.
 
@@
             Groovy programming language. While both are written on top of the 
-            Java platform, Nextflow is only compatible with Linux and OSX \tristan{How about using Linux everywhere instead of UNIX-based? You still have UNIX here and there.} 
+            Java platform, Nextflow is only compatible with Linux and OSX
             environments. These programming languages are not otherwise commonly
@@
 
-                The data structures made available by workflow engines dictate
-                the maximum performance available to a workflow. \tristan{I don't understand the previous sentence, 
-                in particular the 'maximum' idea. You should review that.} Scientific
-                workflows are generally file-based, meaning that data, in the 
-                form of files, will be 
+                Performance of workflow engines can be linked to the data structures
+                employed by the workflow. Scientific
+                workflows are generally file-based, meaning that data in the 
+                form of files will be 
                 transferred across the network and I/O must be
@@
                 larger-than-memory matrices, the Dask array. Furthermore, it is
-                possible to use common Python APIs (e.g. NumPy\note{cite} and 
+                possible to use Dask clones of common Python APIs (e.g. NumPy~\cite{Oliphant:2015:GN:2886196} and 
                 Pandas DataFrame~\cite{mckinney2011pandas}) to process the 
-                data \tristan{I think Dask provides clones of these libraries, but you can't use
-            them directly}. Although Spark does not have a datastructure for large 
+                data. Although Spark does not have a datastructure for large 
                 matrices, it has an RDD-based API, known as \textit{linealg}
@@
 
-            Nipype's wrapper, written in Python, can encapsulate CLIs, SPM, 
+            Nipype's wrapper, written in Python, can encapsulate command-line interfaces (CLIs), SPM, 
             MATLAB and Python scripts. InputSpec and OutputSpec are two classes 
@@
             definition of the run function, while Python interfaces require
-            the definition of the \textit{\_run\_interface} function. \note{from
-            website}.
-            \tristan{Expand UDF, CLI, CARE.}
+            the definition of the \textit{\_run\_interface} function.
 
@@
             In the case of CLI tasks, it is also possible to containerize or 
-            encapsulate with CARE \note{cite} to make it portable. OpenMOLE
+            encapsulate with the Comprehensive Archiver for Reproducible Execution (CARE) \cite{Janin:2014:CCA:2618137.2618138} to make it portable. OpenMOLE
             defines several classes to allow for the execution of different 
@@
             the XML-based module interface through its GUI. The module interface
-            consists of details on \note{from website} module metadata (e.g.
+            consists of details on module metadata (e.g.
             authors, package, version, name) and citation information, 
@@
             standard input, output and error streams must be conserved and the 
-            location where the command should execute\note{from website}.
+            location where the command should execute.
 
@@
             identifier, image URL, site identifier of the image, and container
-            profiles must also be defined \note{from website}.
+            profiles must also be defined.
 
@@
             neuroinformatics.                                                   
-            \note{dont know how Taverna, Pegasus, Galaxy or Toil achieves this} 
-            \note{openMole says it checks to see if outputs exist. similar to   
-            make?}                                                              
-            \note{what about SPM, FastR??}
+            %\note{dont know how Taverna, Pegasus, Galaxy or Toil achieves this} 
+            %\note{openMole says it checks to see if outputs exist. similar to   
+            %make?}                                                              
+            %\note{what about SPM, FastR??}
 
@@
                 to the W3C PROV model that aims to provide comprehensive and
-                machine-readable provenance from raw data to its final stage. \tristan{give example? -- vhs: like a figure? No, like what can be described which couldn't in BIDS. 
-                Forget about it for the comp exam, keep this note for the paper.}
+                machine-readable provenance from raw data to its final stage. %\tristan{give example? -- vhs: like a figure? No, like what can be described which couldn't in BIDS. 
+                %Forget about it for the comp exam, keep this note for the paper.}
                 However, it is currently limited in the provenance 
                 information it can provide, such as details on computational 
-                environments. It is also limited by the fact that consensus lacks
-                within the community on paradigms. \tristan{I disagree with the last sentence
-                but this is fine. The main limitation imo is due to the use of RDF and an ontology: it is
-                very powerful but difficult to use: performance sucks, queries are hell to write, 
-                and format is not human-readable.}
+                environments. It is also limited by the fact that it is difficult to use and that consensus lacks
+                within the community on paradigms. %\tristan{I disagree with the last sentence
+                %but this is fine. The main limitation imo is due to the use of RDF and an ontology: it is
+                %very powerful but difficult to use: performance sucks, queries are hell to write, 
+                %and format is not human-readable.}
 
@@
 
-            \tristan{The rest of this section is very difficult to read because
-            it lacks structure. Perhaps you could classify provenance tracking in PROV-based, lineage-based,
-            and other custom formats? -- It is still a bit 'boring' :) No worries for the comp report,
-            but could be discussed and fixed in the paper.}
+            %\tristan{The rest of this section is very difficult to read because
+            %it lacks structure. Perhaps you could classify provenance tracking in PROV-based, lineage-based,
+            %and other custom formats? -- It is still a bit 'boring' :) No worries for the comp report,
+            %but could be discussed and fixed in the paper.}
             At minumum, provenance tracking should include 
@@
     \chapter{Discussion}
-    \tristan{You should refer to figure 1 in the text. You could start your discussion by mentioning that
-    a summary of your review is available in Figure 1.}
+
         While mainy neuroimaging workflow engines exist, none of them are well
-        suited for the processing of neuroimaging Big Data \tristan{due to? be more specific}. Even more general
+        suited for the processing of neuroimaging Big Data due to lack of data-centric optimizations. Even more general
         scientific workflows are not equipped for the handling of these datasets.
@@
         with many neuroimaging frameworks. Performance of Spark and Dask is 
-        comparable, however, Dask has been found \weird{to be lose efficiency} during 
+        comparable, however, Dask has been found to become less efficient during 
         the processing of larger datasets, whereas Spark exhibits reduced 
@@
         Effective neuroimaging workflows need to meet certain criteria: 
-        ease-of-use, performance, portability and reproducibility. In order for
+        ease-of-use, performance, portability and reproducibility. A summary of
+        the workflow engines and how they meet this criteria can be found in Figure~\ref{fig:summary}. In order for
         a workflow language to be easy to use, it is recommended that it is 
@@
         workflow composition, performance, portability and reproducibility.}
+        \label{fig:summary}
     \end{sidewaysfigure} 
